---
title: "Initial QC `r plate`"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: html_document
---

## Experiment Details {.tabset}

***

+ Plate 1: 96 samples(60/36), on MEGA kit v2, seq 15 sub-lib X 3 runs
+ Plate 2: 96 samples (88/8), on MEGA kit v2, seq 16 sub-lib X 3 runs
+ ~30 Billion reads per plate
+ Parse workflow has a low [doublet rate](https://support.parsebiosciences.com/hc/en-us/articles/360053107311-What-is-the-expected-doublet-rate#:~:text=Doublet%20rates%20are%20low%2C%20less,through%20the%20Whole%20Transcriptome%20workflow.): ~3% per 100K cells

***

- I had to run QC on objects separately due to combined object size
- FC samples included
- 2 filtering steps
- Stage 1
  - Remove cells with < 300 genes OR < 500 reads
  - Remove samples with fewer than 500 cells
- Stage 2
  - UMI - 3 Median absolute deviations
  - Genes - 3 Median absolute deviations
  - Mito - 5%
  - Ribo - 5%
  - Filtering MT genes and MALAT1 removed

### QC {.tabset}

#### Pre-filter counts {.tabset}

##### Boxplots

```{r tscp_gene_boxplt , results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_boxplt 
```

***

##### Median counts

```{r median_cnt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_tbl
```

***

##### Low count cells

```{r samples_low_cnts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
samples_low_cnts_tbl
```

***

##### Metadata

```{r seurat_metadata, results='asis', fig.dim = c(10,10), echo=FALSE}
metadata_lst[['seurat_meta']] |>
  flextable() |> 
  set_caption(caption = "Seurat object metadata") |>
  theme_vanilla() |>
  flextable_to_rmd()
```

#### Filter 1 {.tabset}

##### Counts post-filter 1

```{r cell_cnts_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
cell_cnts_post_filt_tbl
```

***

##### Median counts post-filter 1

```{r median_cnt_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_post_filt_tbl
```

***

##### Boxplots post-filter 1

```{r tscp_gene_post_filt_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_post_filt_boxplt
```

***

#### Filter 2 {.tabset}

##### Counts post-filter 2

```{r cell_cnts_final_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
cell_cnts_final_tbl
```

***

##### Median counts post-filter 2

```{r median_cnt_final_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_final_tbl
```

***

##### Boxplots post-filter 2

```{r tscp_gene_final_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_final_boxplt
```

***

##### All filters


```{r filt_all_counts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
filt_all_counts_tbl
```

***
