---
title: "Sample swap check"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(ggplot2)
library(DT)
library(viridis)
```

***

## Sample swap check process {.tabset}

1. For each sublib, copy BAM file and cell whitelist from neurocluster, subset BAM by chr
2. `cellsnp-lite` is run for each cell in the chr-specific BAM to generate per-cell allele 
counts (pile-ups) at informative SNP positions. 
    - SNP ref: TOPMed freeze 5, hg38, filtered for informative SNPs (MAF ≥ 0.1, depth ≥ 20) - 
    to save running time / mem
    - Parse cell whitelist is used to track cell / sample ids
    - Need to vary the SNP filtering based on chromosome, running times etc. chr22 has to few SNPs 
3. `vireo` is run to infer donor (per cell) by probabilistically matching each cell’s SNP profile 
(from GeX data) to the reference donor genotypes.
4. Finally, cross referenced vireo output with plate specific sample list, to see if vireo 
cell / sample predictions match our sample assignments

***

```{r, helper_function, message=FALSE, echo=FALSE}
process_vireo <- function(sublibrary, plate, chr,
                          base_dir = "~/Desktop/eQTL_study_2025/") {
  # Paths
  vireo_path <- file.path(
    base_dir,
    "results/00CHECK_SMPL_CONTAMINATION/vireo",
    paste0(sublibrary, "_plate", plate, "_chr", chr, "_vireo/donor_ids.tsv")
  )
  whitelist_path <- file.path(
    base_dir,
    "workflow/reports/00CHECK_SMPL_CONTAMINATION/whitelists",
    paste0(sublibrary, "_plate", plate, "_cell_metadata.csv")
  )
  
  # Load files
  vireo_tbl <- read_tsv(vireo_path, show_col_types = FALSE) 
  whitelist_tbl <- read_delim(whitelist_path,
                              col_names = c("donor_id", "raw_sample_number"),
                              show_col_types = FALSE) %>%
      filter(!str_detect(raw_sample_number, "Thal|Hipp|WGE")) |>
      filter(donor_id != 'bc_wells') |>
      dplyr::select(cell = donor_id, donor_id_wl = raw_sample_number) |>
      mutate(donor_id_wl = str_remove(donor_id_wl, '^sample_'))

  # Summary
  summary_tbl <- vireo_tbl %>%
    summarise(
      Assigned = sum(donor_id != "unassigned" & donor_id != "doublet"),
      Unassigned = sum(donor_id == "unassigned"),
      Doublet = sum(donor_id == "doublet"),
      Total = n()
    ) %>%
    mutate(sublibrary = sublibrary, plate = plate, chr = chr,
           Assigned_Pct = (Assigned / Total) * 100,
           Unassigned_Pct = (Unassigned / Total) * 100,
           Doublet_Pct = (Doublet / Total) * 100)
  
  # Assigned cells
  assigned_cells <- vireo_tbl %>%
    filter(donor_id != "unassigned" & donor_id != "doublet") %>%
    mutate(donor_id = if_else( # Deal with double IDs
      str_detect(donor_id, "_"),
      if_else(
        word(donor_id, 1, sep = "_") == word(donor_id, -1, sep = "_"),
        word(donor_id, 1, sep = "_"),
        donor_id
      ),
      donor_id
    )) |>
    dplyr::rename(donor_id_geno = donor_id)
  
  # Join with expected
  matched_tbl <- assigned_cells |>
    left_join(whitelist_tbl, by = "cell") |>
    mutate(match = donor_id_geno == donor_id_wl) |>
    dplyr::select(cell, donor_id_geno, donor_id_wl, match, prob_max, n_vars) |>
    mutate(sublibrary = sublibrary, plate = plate, chr = chr)
    
  # # Quantify NAs
  # na_tbl <- matched_tbl %>% filter(!complete.cases(.)) |> print(n=3)
  # common_cells <- na_tbl$cell %in% whitelist_tbl$cell
  # 
  # # Summarize results
  # if (any(common_cells)) {
  #   cat("Cells found in both na_tbl and matched_tbl:\n")
  #   print(na_tbl$cell[common_cells])
  # } else {
  #   cat("No cells from na_tbl are found in matched_tbl.\n")
  # }
  
  matched_tbl <- matched_tbl %>% filter(complete.cases(.))
    
  list(summary_tbl = summary_tbl, matched_tbl = matched_tbl)
}
```

```{r load data, message=FALSE, echo=FALSE, warning=FALSE}
# Define combinations of sublibrary, plate, and chr
inputs <- expand_grid(
  sublibrary = c("3", "6", "10", "12", "15"),
  plate = as.character(seq(1,3)),
  chr = as.character(seq(1,22))
)

# Process all datasets
all_results <- pmap(inputs, process_vireo)

rm_smpl_tbl <- read_tsv("~/Desktop/eQTL_study_2025/workflow/reports/00SAMPLE-SURVEY/removed_samples.tsv")

# Combine outputs
all_summaries <- bind_rows(map(all_results, "summary_tbl"))
all_matched   <- bind_rows(map(all_results, "matched_tbl"))
```

***

### Summarise cell assignments


```{r sum_tbl, echo=FALSE}
datatable(
  all_summaries,
  caption = 'Table 1. Cell Assignment Summary per sublibrary, plate, and chromosome'
)
```

***

### Predictions

```{r, predictions, echo=FALSE, eval =F}
datatable(
  all_matched,
  caption = 'Table 2. Matched cells across all datasets',
  options = list(pageLength = 25, autoWidth = TRUE)
)
```

### Plot (by donor)

***

```{r, counts, echo=FALSE, fig.dim=c(20,30)}
matched_summary <- all_matched %>%
  group_by(sublibrary, plate, chr, donor_id_wl) %>%
  summarise(cell_count = n(),
            match_rate = mean(match, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(cell_count))

ggplot(matched_summary,
       aes(x = reorder(donor_id_wl, cell_count),  # reorder for horizontal bars
           y = cell_count,
           fill = match_rate)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = cell_count), 
            hjust = -0.1, size = 3.5) +  # counts on bars
  scale_fill_viridis(name = "Match rate", option = "C", direction = -1) +
  labs(
    title = "Cell counts per expected donor ID",
    x = "Donor (annotation in sample list)",
    y = "Cell count"
  ) +
  coord_flip() +  # horizontal bars
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  facet_grid(sublibrary + plate ~ chr, scales = "free_y", space = "free_y")  # free_y for horizontal bars

```

***

### Plot

***

```{r plot, echo=FALSE, fig.dim=c(10,20), warning=FALSE}
global_summary <- all_matched %>%
  group_by(donor_id_wl) %>%
  summarise(cell_count = n(),
            match_rate = mean(match, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(cell_count)

samples <- unique(unlist(strsplit(paste(rm_smpl_tbl$samples, collapse = ", "), ", ")))
num_donors <- nrow(global_summary)
y_axis_colors <- ifelse(global_summary$donor_id_wl %in% samples, "red", "black")

ggplot(global_summary,
       aes(x = reorder(donor_id_wl, cell_count),
           y = cell_count,
           fill = match_rate)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = cell_count), 
            hjust = -0.1, size = 4) +  # Show counts to the right of bars
  annotate("text", 
           x = length(global_summary$donor_id_wl)/2, 
           y = max(global_summary$cell_count) * 1.05, 
           label = paste0("Donors: ", num_donors), 
           hjust = 2, vjust = 0, size = 5, fontface = "bold") +
  scale_fill_viridis(name = "Match rate", option = "C", direction = -1) +
  labs(
    title = "Combined Cell Counts per Expected Donor ID (all plates/chr)",
    x = "Donor (expected annotation in cell whitelist)",
    y = "Cell count"
  ) +
  coord_flip() +  # Horizontal bars
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 12),
    axis.text.y = element_text(color = y_axis_colors),  # Color y-axis text
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

```