---
title: "Sample swap check"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
params:
  in_dir: null
  bmark_dir: null
  report_dir: null
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(ggplot2)
library(DT)
library(viridis)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
report_dir <- params$report_dir
```

***

## Sample swap check process {.tabset}

***

1. `copy_bam`: For each sublibrary and plate combination, copy BAM file from neurocluster to scratch:
    - BAM located here: `<src_dir>/plate3/1_plate3/process/barcode_headAligned_anno.bam`
2. `make_whitelist`: Generated cell whitelist files from cell metadata for each sublibrary and plate.
    - Whitelist located here: `<src_dir>/plate3/1_plate3/all-sample/DGE_filtered/cell_metadata.csv`
3. `subset_vcf`: Subset TOPMed dbSNP VCF to chr-specific files
    - SNP ref: TOPMed freeze 5, hg38, filtered for informative SNPs (MAF ≥ 0.1, depth ≥ 20) - 
    to save running time / mem (same ref used for pre-imputation genotyping harmonisation)
4. `sort_bam`: Sorted BAM files by coordinates for each sublibrary.
5. `index_bam`: Indexed sorted BAM files
6. `extract_and_recode_chr`: Extracted chr-specific reads from sorted BAMs
    - Recoded headers to standardise chr names (e.g., hg38_chrX to chrX)
7. `cellsnp-lite` generate per-cell allele counts (pile-ups) at informative SNP positions
    - Quantifies SNP genotypes for each cell using chr-specific BAMs and filtered ref VCF
    - [PMID:33963851](https://academic.oup.com/bioinformatics/article/37/23/4569/6272512?login=false)
    - Cell / sample ids are tracked using Parse cell whitelist for each sublibrary, plate, chr combination
    - Need to vary the SNP filtering based on chr, running times etc. chr22 has too few SNPs 
8. `vireo`: infer donor identities by matching cell SNP profiles to actual donor genotypes
    - Assigns cells to donors based on probabilistic SNP profile matching
    - [PMID:31836005](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1865-2)
9. `process_vireo`: Compare predicted donor IDs with expected sample assignments from whitelist.
    - Evaluates accuracy of donor assignments by cross-referencing with known cell metadata
    - Run in this markdown file
10. `smpl_swap_report`: Assess donor assignment accuracy and potential sample swaps.

***

```{r, helper_function, message=FALSE, echo=FALSE}
process_vireo <- function(sublibrary, plate, chr,
                          base_dir = in_dir) {
  # Paths
  vireo_path <- file.path(
    base_dir,
    "vireo",
    paste0(sublibrary, "_plate", plate, "_chr", chr, "_vireo/donor_ids.tsv")
  )
  whitelist_path <- file.path(
    base_dir,
    "refs",
    paste0(sublibrary, "_plate", plate, "_cell_metadata.csv")
  )
  
  # Load files
  vireo_tbl <- read_tsv(vireo_path, show_col_types = FALSE) 
  whitelist_tbl <- read_delim(whitelist_path,
                              col_names = c("donor_id", "raw_sample_number"),
                              show_col_types = FALSE) %>%
      filter(!str_detect(raw_sample_number, "Thal|Hipp|WGE")) |>
      filter(donor_id != 'bc_wells') |>
      dplyr::select(cell = donor_id, donor_id_wl = raw_sample_number) |>
      mutate(donor_id_wl = str_remove(donor_id_wl, '^sample_'))

  # Summary
  summary_tbl <- vireo_tbl %>%
    summarise(
      Assigned = sum(donor_id != "unassigned" & donor_id != "doublet"),
      Unassigned = sum(donor_id == "unassigned"),
      Doublet = sum(donor_id == "doublet"),
      Total = n()
    ) %>%
    mutate(sublibrary = sublibrary, plate = plate, chr = chr,
           Assigned_Pct = (Assigned / Total) * 100,
           Unassigned_Pct = (Unassigned / Total) * 100,
           Doublet_Pct = (Doublet / Total) * 100)
  
  # Assigned cells
  assigned_cells <- vireo_tbl %>%
    filter(donor_id != "unassigned" & donor_id != "doublet") %>%
    mutate(donor_id = if_else( # Deal with double IDs
      str_detect(donor_id, "_"),
      if_else(
        word(donor_id, 1, sep = "_") == word(donor_id, -1, sep = "_"),
        word(donor_id, 1, sep = "_"),
        donor_id
      ),
      donor_id
    )) |>
    dplyr::rename(donor_id_geno = donor_id)
  
  # Join with expected
  matched_tbl <- assigned_cells |>
    left_join(whitelist_tbl, by = "cell") |>
    mutate(match = donor_id_geno == donor_id_wl) |>
    dplyr::select(cell, donor_id_geno, donor_id_wl, match, prob_max, n_vars) |>
    mutate(sublibrary = sublibrary, plate = plate, chr = chr)
    
  # # Quantify NAs
  # na_tbl <- matched_tbl %>% filter(!complete.cases(.)) |> print(n=3)
  # common_cells <- na_tbl$cell %in% whitelist_tbl$cell
  # 
  # # Summarize results
  # if (any(common_cells)) {
  #   cat("Cells found in both na_tbl and matched_tbl:\n")
  #   print(na_tbl$cell[common_cells])
  # } else {
  #   cat("No cells from na_tbl are found in matched_tbl.\n")
  # }
  
  matched_tbl <- matched_tbl %>% filter(complete.cases(.))
    
  list(summary_tbl = summary_tbl, matched_tbl = matched_tbl)
}
```

```{r load data, message=FALSE, echo=FALSE, warning=FALSE}
# Dynamically discover available inputs from directories
vireo_dir <- file.path(in_dir, "vireo")
dirs <- dir(vireo_dir, pattern = "_vireo$")

inputs <- tibble(dir = dirs) %>%
  mutate(name = str_remove(dir, "_vireo$")) %>%
  separate(name, into = c("sublibrary", "plate_str", "chr_str"), sep = "_") %>%
  mutate(plate = str_remove(plate_str, "plate"),
         chr = str_remove(chr_str, "chr")) %>%
  select(sublibrary, plate, chr)

# Process all datasets
all_results <- pmap(inputs, process_vireo)

rm_smpl_tbl <- read_tsv(paste0(report_dir, "00SAMPLE-SURVEY/removed_samples.tsv"))

# Combine outputs
all_summaries <- bind_rows(map(all_results, "summary_tbl"))
all_matched   <- bind_rows(map(all_results, "matched_tbl"))
```

***

### Summarise cell assignments

***

```{r sum_tbl, echo=FALSE}
datatable(
  all_summaries,
  caption = 'Table 1. Cell Assignment Summary per sublibrary, plate, and chromosome'
)
```

***

### Predictions

***

```{r, predictions, echo=FALSE, eval =F}
datatable(
  all_matched,
  caption = 'Table 2. Matched cells across all datasets',
  options = list(pageLength = 25, autoWidth = TRUE)
)
```

***

### Plot (by donor)

***

```{r, counts, echo=FALSE, fig.dim=c(20,30)}
matched_summary <- all_matched %>%
  group_by(sublibrary, plate, chr, donor_id_wl) %>%
  summarise(cell_count = n(),
            match_rate = mean(match, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(cell_count))

ggplot(matched_summary,
       aes(x = reorder(donor_id_wl, cell_count),  # reorder for horizontal bars
           y = cell_count,
           fill = match_rate)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = cell_count), 
            hjust = -0.1, size = 3.5) +  # counts on bars
  scale_fill_viridis(name = "Match rate", option = "C", direction = -1) +
  labs(
    title = "Cell counts per expected donor ID",
    x = "Donor (annotation in sample list)",
    y = "Cell count"
  ) +
  coord_flip() +  # horizontal bars
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  facet_grid(sublibrary + plate ~ chr, scales = "free_y", space = "free_y")  # free_y for horizontal bars

```

***

### Plot

***

```{r plot, echo=FALSE, fig.dim=c(10,20), warning=FALSE}
global_summary <- all_matched %>%
  group_by(donor_id_wl) %>%
  summarise(cell_count = n(),
            match_rate = mean(match, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(cell_count) |>
  write_tsv(paste0(report_dir, "00CHECK_SMPL_CONTAMINATION/sample_swap_summary.tsv"))

samples <- unique(unlist(strsplit(paste(rm_smpl_tbl$samples, collapse = ", "), ", ")))
num_donors <- nrow(global_summary)
y_axis_colors <- ifelse(global_summary$donor_id_wl %in% samples, "red", "black")

ggplot(global_summary,
       aes(x = reorder(donor_id_wl, cell_count),
           y = cell_count,
           fill = match_rate)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = cell_count), 
            hjust = -0.1, size = 4) +  # Show counts to the right of bars
  annotate("text", 
           x = length(global_summary$donor_id_wl)/2, 
           y = max(global_summary$cell_count) * 1.05, 
           label = paste0("Donors: ", num_donors), 
           hjust = 2, vjust = 0, size = 5, fontface = "bold") +
  scale_fill_viridis(name = "Match rate", option = "C", direction = -1) +
  labs(
    title = "Combined Cell Counts per Expected Donor ID (all plates/chr)",
    x = "Donor (expected annotation in cell whitelist)",
    y = "Cell count"
  ) +
  coord_flip() +  # Horizontal bars
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 12),
    axis.text.y = element_text(color = y_axis_colors),  # Color y-axis text
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )


```

***

### Plot (only low counts)

***

```{r plot_log10, echo=FALSE, fig.dim=c(10,20), warning=FALSE}
global_summary <- all_matched %>%
  group_by(donor_id_wl) %>%
  summarise(cell_count = n(),
            match_rate = mean(match, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(cell_count) |>
  filter(cell_count <= 20)

samples <- unique(unlist(strsplit(paste(rm_smpl_tbl$samples, collapse = ", "), ", ")))
num_donors <- nrow(global_summary)
y_axis_colors <- ifelse(global_summary$donor_id_wl %in% samples, "red", "black")

ggplot(global_summary,
       aes(x = reorder(donor_id_wl, cell_count),
           y = cell_count,
           fill = match_rate)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = cell_count), 
            hjust = -0.1, size = 4) +  # Show counts to the right of bars
  annotate("text", 
           x = length(global_summary$donor_id_wl)/2, 
           y = max(global_summary$cell_count) * 1.05, 
           label = paste0("Donors: ", num_donors), 
           hjust = 2, vjust = 0, size = 5, fontface = "bold") +
  scale_fill_viridis(name = "Match rate", option = "C", direction = -1) +
  labs(
    title = "Combined Cell Counts per Expected Donor ID (all plates/chr)",
    x = "Donor (expected annotation in cell whitelist)",
    y = "Cell count"
  ) +
  coord_flip() +  # Horizontal bars
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 12),
    axis.text.y = element_text(color = y_axis_colors),  # Color y-axis text
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "check_sample_contamination.*.benchmark\\.txt", full.names = TRUE)

# Read and combine benchmark files
benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  # Extract rule and optional chromosome from filename
  parts <- strsplit(basename(f), "\\.")[[1]]
  rule_name <- parts[1]  # e.g., "check_sample_contamination_cellsnp_lite"
  
  # Initialize base_rule and chr_part
  base_rule <- rule_name
  chr_part <- "single"
  
  # Check for chromosome in additional parts (e.g., "chr21")
  for (i in 2:(length(parts) - 2)) {
    p <- parts[i]
    if (grepl("^chr[0-9XY]+$", p)) {
      chr_part <- gsub("chr", "", p)
      break  # Assume only one chr part
    }
  }
  
  # Also check if rule_name ends with "_chr[0-9XY]+"
  chr_match <- str_match(rule_name, "^(.+)_chr([0-9XY]+)$")
  if (!is.na(chr_match[1, 1])) {
    base_rule <- chr_match[1, 2]
    chr_part <- chr_match[1, 3]
  }
  
  tibble(
    file = f,
    rule = base_rule,
    chromosome = chr_part,     # Chromosome (e.g., "21", "X") or "single" for non-chromosome-specific
    runtime_sec = df$s,        # Runtime in seconds
    runtime_hms = df$h.m.s.,   # hh:mm:ss string
    max_rss = df$max_rss       # Peak RAM in MB
  )
})

# Summarize across rules, collapsing variations (e.g., chromosome-specific, plates, sublibs)
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  # Replace NA in sd columns with 0 for single runs
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  # Round numeric columns
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  # Sort by rule for clarity
  arrange(rule) |>
  mutate(rule = str_remove(rule, 'check_sample_contamination_'))

# Create a DT table
datatable(
  summary,
  caption = 'Table 2. Rule Benchmarking. Sd is 0 when only one instance of a job is run.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)
```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule (± SD)") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```

***