---
title: "cTWAS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
  lookup_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(gt)
library(tidyverse)
library(ctwas)
library(EnsDb.Hsapiens.v86)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
lookup_dir <- params$lookup_dir 

cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")
gwas <- c("scz", "bpd", "mdd", "adhd", "ocd", "ptsd")
gene_lookup <- read_tsv(paste0(lookup_dir, "gene_lookup_hg38.tsv")) |>
  dplyr::select(chr = chromosome_name, start = start_position, end = end_position, 
         gene_biotype, ensembl_id = ensembl_gene_id, symbol = external_gene_name)
gwas_n <- 126849 # scz approx - need to pull from ctwas_run

ens_db <- EnsDb.Hsapiens.v86
```

***

## cTWAS pipeline steps {.tabset}

***

We used [cTWAS](https://xinhe-lab.github.io/multigroup_ctwas/index.html) to integrate our sc-eQTL data with neuropsychiatric GWAS data to identify putative candidate genes associated with these conditions.   [Wen et al 2024]() to boost our eQTL discovery power. 
The citation for cTWAS is [Zhao et al. (2025), DOI: 10.1038/s41588-023-01648-9](https://www.nature.com/articles/s41588-023-01648-9).

***

1. `create_ld_matrices`: Create (1000G hg38)LD matrices and variant info files for causal TWAS dataset
2. `copy_fusion_weights`: Copy genuine FUSION weight files (cTWAS chokes if any empty files are in weight dir)
3. `run_ctwas`: Run cTWAS on sc-eQTL and GWAS data
4. `ctwas_report`: Generate cTWAS report

TODO: 

1. Rpt configured for 1 cell type and scz atm, need to rejig to handle all
2. Need option to alter gene / region for plot(s) in smk call

***

### Z-scores {.tabset}

```{r, cTWAS_z_score_tbl, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
# === TWO-TIER LOOP: GWAS (####) → Cell types (#####) ===
for (g in gwas) {
  
  cat("\n#### ", toupper(g), " {.tabset .tabset-fade}\n\n")
  
  for (ct in cell_types) {
    
    # Construct filename
    file_path <- paste0(in_dir, "ctwas_", ct, "_", g, "_ctwas.rds")
    
    # Read data
    ctw_res <- read_rds(file_path)
    
    if (!"z_gene" %in% names(ctw_res) || nrow(ctw_res$z_gene) == 0) {
      cat("\n##### ", ct, "\n")
      cat("No gene results available.\n\n")
      next
    }
    
    # Process z_gene table
    z_gene <- as_tibble(ctw_res$z_gene) |>
    separate(id, into = c('gene', 'id'), sep = "\\|") |>
    separate(id, into = c('cell_type', 'id'), sep = "_") |>
      left_join(gene_lookup, by = join_by(gene == ensembl_id)) |>
      mutate(
        pval = 2 * pnorm(-abs(z)),
        fdr  = p.adjust(pval, method = "BH")
      ) |>
      dplyr::filter(fdr < 0.05) |>
      arrange(fdr) |>
      distinct() |>
      dplyr::select(cell_type, symbol, gene, z, pval, fdr, gene_biotype) 
    
    # Sub-tab header
    cat("\n#### ", ct, "\n\n")
    
    # If no significant genes
    if (nrow(z_gene) == 0) {
      cat("**No significant genes (FDR < 0.05)**\n\n")
      next
    }
    
    # Create and print gt table
    z_gene %>%
      dplyr::select(symbol, gene, z, pval, fdr, gene_biotype, chr) %>%
      gt() %>%
      tab_header(
        title = md(paste0("**", toupper(g), " – ", ct, "**")),
        subtitle = "cTWAS significant genes (FDR < 0.05)"
      ) %>%
      fmt_scientific(columns = c("pval", "fdr"), decimals = 2) %>%
      cols_label(
        cell_type    = "Cell Type",
        symbol       = "Gene Symbol",
        gene         = "ENSG ID",
        z            = "Z-score",
        pval         = "P-value",
        fdr          = "FDR",
        gene_biotype = "Gene Type",
        chr          = "Chr"
      ) %>%
      tab_options(table.font.size = "90%") %>%
      print()
    
    cat("\n\n")  # spacing between tables
  }
}

# ctwas_res <- read_rds(paste0(in_dir, "ctwas_ExN-UL_scz_ctwas.rds"))
# z_gene <- as_tibble(ctwas_res$z_gene) |>
#   separate(id, into = c('gene', 'id'), sep = "\\|") |>
#   separate(id, into = c('cell_type', 'id'), sep = "_") |>
#   left_join(gene_lookup, by = join_by(gene == ensembl_id )) |>
#   mutate(pval = 2 * pnorm(-abs(z)),
#          fdr = p.adjust(pval, method = "BH")) |>
#   dplyr::filter(fdr < 0.05) |>
#   arrange(fdr) |>
#   dplyr::select(cell_type, gene, symbol, z, pval, fdr, gene_biotype, chr, start, end) |>
#   distinct() |>
#   print(n = Inf)
# 
# # Table
# z_gene %>%
#   dplyr::select(cell_type, symbol, gene, z, pval, fdr, gene_biotype) %>%
#   gt() %>%
#   tab_header("Table 1. cTWAS significant genes (FDR < 0.05)") %>%
#   fmt_scientific(columns = c(pval, fdr), decimals = 2) %>%
#   cols_label(
#     cell_type = "Cell Type",
#     symbol = "Gene",
#     z = "Z-score",
#     pval = "P-value",
#     fdr = "FDR",
#     gene_biotype = "Type"
#   )
```

***

### Fine-mapping

***

```{r ctwas_fine_mapping_tbl, echo = FALSE, message = FALSE, warning = FALSE}
finemap_res <- as_tibble(ctwas_res$finemap_res) |>
  dplyr::filter(group != 'SNP' & susie_pip > 0.8 & !is.na(cs)) |>
  separate(id, into = c('gene', 'id'), sep = "\\|") |>
  separate(id, into = c('cell_type', 'id'), sep = "_") |>
  left_join(gene_lookup, by = join_by(gene == ensembl_id )) |>
  arrange(desc(susie_pip))
susie_alpha_res <- ctwas_res$susie_alpha_res
region_data <- ctwas_res$region_data
screen_res <- ctwas_res$screen_res

finemap_res %>%
  dplyr::filter(susie_pip > 0.8) %>%
  dplyr::select(symbol, cell_type, susie_pip, cs) %>%
  gt() %>%
  tab_header("Table 2. Fine-mapped causal genes (SuSiE PIP > 0.8)")
```

***

### Plot

***

```{r, ctwas_plot, echo = FALSE, message = FALSE, warning = FALSE}
# make_locusplot(finemap_res,
#                region_id = "22_40149793_42294812",
#                ens_db = ens_db,
#                weights = weights,
#                highlight_pip = 0.8,
#                filter_protein_coding_genes = TRUE,
#                filter_cs = TRUE,
#                color_pval_by = "cs",
#                color_pip_by = "cs")
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "12ctwas.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
datatable(
  summary,
  caption = 'Table 3. Rule benchmarking. Means and SD across replicates.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```