---
title: "cTWAS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
  lookup_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(gt)
library(tidyverse)
library(ctwas)
#library(EnsDb.Hsapiens.v86)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
lookup_dir <- params$lookup_dir 

# in_dir <- '~/Desktop/test/ctwas/'
# lookup_dir <- '~/Desktop/eQTL_study_2025/resources/sheets/'
# bmark_dir <- ''

gwas <- c("scz", "bpd", "mdd", "adhd", "ocd", "ptsd")
gene_lookup <- read_tsv(paste0(lookup_dir, "gene_lookup_hg38.tsv")) |>
  dplyr::select(chr = chromosome_name, start = start_position, end = end_position, 
         gene_biotype, ensembl_id = ensembl_gene_id, symbol = external_gene_name)
#gwas_n <- 126849 # scz approx - need to pull from ctwas_run

#ens_db <- EnsDb.Hsapiens.v86
```

***

## cTWAS pipeline steps {.tabset}

***

We used [cTWAS](https://xinhe-lab.github.io/multigroup_ctwas/index.html) to integrate our sc-eQTL data with neuropsychiatric GWAS data to identify putative candidate genes associated with these conditions. The citation for cTWAS is [Zhao et al. (2025), DOI: 10.1038/s41588-023-01648-9](https://www.nature.com/articles/s41588-023-01648-9).

***

1. `create_ld_matrices`: Create (1000G hg38)LD matrices and variant info files for causal TWAS dataset
2. `copy_fusion_weights`: Copy genuine FUSION weight files (cTWAS chokes if any empty files are in weight dir)
3. `run_ctwas`: Run cTWAS on sc-eQTL and GWAS data
4. `ctwas_report`: Generate cTWAS report

TODO: 

1. Rpt configured for 1 cell type and scz atm, need to rejig to handle all
2. Need option to alter gene / region for plot(s) in smk call

***

### Correction

```{r, get_gene_lists_for correction, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
perm_files <- tibble(
  cell_type = cell_types,
  file = glue::glue(
    "../results/05TENSORQTL/tensorqtl_perm/{cell_types}_quantile_genPC_4_expPC_40/{cell_types}_quantile_perm.cis_qtl.txt.gz"
  )
)

# Read, filter, and annotate
perm_tbl <- tibble(file = perm_files) %>%
  mutate(
    cell_type = str_extract(file, "(?<=tensorqtl_perm/)[^_/]+")
  ) %>%
  mutate(
    data = map(
      file,
      ~ read_tsv(.x, show_col_types = FALSE) %>%
          filter(pval_nominal < 0.01) %>%
          select(phenotype_id)
    )
  ) %>%
  select(cell_type, data) %>%
  unnest(data)

perm_tbl <- perm_files %>%
  mutate(
    data = map(
      file,
      ~ read_tsv(.x, show_col_types = FALSE) %>%
          filter(pval_nominal < 0.01) %>%
          select(phenotype_id)
    )
  ) %>%
  select(cell_type, data) %>%
  unnest(data)

per_celltype_counts <- perm_tbl %>%
  distinct(cell_type, phenotype_id) %>%
  count(cell_type, name = "n_sig_genes") %>%
  arrange(desc(n_sig_genes))

unique_gene_count <- perm_tbl %>%
  distinct(phenotype_id) %>%
  summarise(
    cell_type = "Unique across all cell types",
    n_sig_genes = n()
  )

final_table <- bind_rows(
  per_celltype_counts,
  unique_gene_count
) 

bf_denominator <- final_table %>%
  filter(cell_type == "Unique across all cell types") %>%
  pull(n_sig_genes)

final_table %>%
  gt() %>%
  cols_label(
    cell_type = "Cell type",
    n_sig_genes = "Genes with cis-eQTL (P < 0.01)"
  ) %>%
  fmt_number(
    columns = n_sig_genes,
    decimals = 0,
    use_seps = TRUE
  ) %>%
  tab_spanner(
    label = "TensorQTL cis-eQTL discovery",
    columns = n_sig_genes
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = cell_type == "Unique across all cell types"
    )
  ) %>%
  tab_style(
    style = list(
      cell_borders(
        sides = "top",
        weight = px(2)
      )
    ),
    locations = cells_body(
      rows = cell_type == "Unique across all cell types"
    )
  ) %>%
  tab_source_note(
    source_note = md(
      "Genes were defined as cis-regulated if at least one variant showed nominal association with expression (TensorQTL nominal *P* < 0.01). The final row reports the union of genes across all cell types and defines the Bonferroni correction denominator for causal TWAS analyses."
    )
  )
```

### Multi

***

#### Z-scores {.tabset}

```{r, cTWAS_z_score_tbl, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
cell_types <- c("ExN-UL", "ExN-DL", "RG", "InN", "Endo-Peri", "MG", "OPC")
for (g in gwas) {
  cat("##### ", toupper(g), " {.tabset}\n\n")
  
  file <- paste0(in_dir, "ctwas_multi_", g, "_ctwas.rds")
  ctw_res  <- read_rds(file)
  # Process data
  z_gene <- as_tibble(ctw_res$z_gene) |>
    separate(id, into = c('gene', 'id'), sep = "\\|") |>
    separate(id, into = c('cell_type', 'id'), sep = "_") |>
    left_join(gene_lookup, by = join_by(gene == ensembl_id)) |>
    mutate(pval = 2 * pnorm(-abs(z)),
           fdr  = p.adjust(pval, method = "BH"),
           pval_bf = pval * bf_denominator) |>
    mutate(pval_bf = pmin(pval_bf, 1)) |>
    arrange(pval_bf) |>
    filter(pval_bf < 0.05) |>
   # arrange(fdr) |>
    distinct(gene, .keep_all = TRUE) |>
    dplyr::select(cell_type, symbol, gene, z, pval, pval_bf, fdr, gene_biotype, chr)
    
  if (nrow(z_gene) == 0) {
    cat("No significant genes (FDR < 0.05)\n\n")
  } else {
    z_gene |>
      gt() |>
      tab_header(title = md(paste0("**", toupper(g), " – ", ct, "**")),
                 subtitle = md(paste0("cTWAS significant genes (FDR < 0.05): ", nrow(z_gene)))) |>
      fmt_scientific(columns = c(pval, fdr), decimals = 2) |>
      cols_label(cell_type = "Cell Type",
                 symbol = "Gene Symbol",
                 gene = "ENSG ID",
                 z = "Z-score",
                 pval = "P-value",
                 fdr = "FDR",
                 gene_biotype = "Gene Type",
                 chr = "Chr") |>
  # MAGIC SETTINGS BELOW
  tab_options(
      table.font.size = "85%",                 # smaller font
      data_row.padding = px(4),                # less vertical space per row
      column_labels.font.weight = "bold",
      table.width = pct(100),                  # use full width
      table.border.top.style = "none",
      table.border.bottom.style = "solid",
      table.border.bottom.width = px(2),
      table.border.bottom.color = "gray70"
  ) |>
  cols_width(gene ~ px(150),           
             symbol ~ px(100),         
             cell_type ~ px(80),
             z ~ px(70),
             pval ~ px(90),
             fdr ~ px(90),
             chr ~ px(50),
             gene_biotype ~ px(110)
  ) |>
  opt_table_font(font = "Arial") |>
  tab_style(
      style = cell_text(size = px(12)),
      locations = cells_body()
  ) |>
  opt_css(
      css = "
    .gt_row { white-space: nowrap; }       /* prevent wrapping in cells */
    .gt_table { overflow-x: auto; }        /* horizontal scroll if needed */
  "
  ) |>
  print()
    cat("\n\n")
  }
  
}
```

***

#### Fine-mapping {.tabset}

***

```{r ctwas_fine_mapping_tbl, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}

for (g in gwas) {
  cat("##### ", toupper(g), " {.tabset}\n\n")
  
  file <- paste0(in_dir, "ctwas_multi_", g, "_ctwas.rds")
  ctw_res  <- read_rds(file)
  
  if (is.null(ctw_res$finemap_res) || nrow(ctw_res$finemap_res) == 0) {

    cat("**No finemapped genes (no credible sets identified)**\n\n")
    next
  }
  
  finemap_res <- as_tibble(ctw_res$finemap_res) |>
    dplyr::filter(group != 'SNP' & susie_pip > 0.8 & !is.na(cs)) |>
    separate(id, into = c('gene', 'id'), sep = "\\|") |>
    separate(id, into = c('cell_type', 'id'), sep = "_") |>
    left_join(gene_lookup, by = join_by(gene == ensembl_id )) |>
    arrange(desc(susie_pip)) |>
    mutate(pval = 2 * pnorm(-abs(z))) |>
    dplyr::select(cell_type, symbol, gene, z, pval, susie_pip, cs, gene_biotype, chr)
  
  if (nrow(finemap_res) == 0) {
    cat("**No genes with SuSiE PIP > 0.8 in credible sets**\n\n")
} else {
  finemap_res |>
  gt() |>
  tab_header(
    title = md(paste0("**", toupper(g), " – ", ct, "**")),
    subtitle = md(paste0("Finemapped genes (SuSiE PIP > 0.8, in credible set): **", nrow(finemap_res), "**"))
  ) |>
  fmt_scientific(columns = c(pval, z), decimals = 2) |>
  fmt_number(columns = susie_pip, decimals = 3) |>
  cols_label(
    cell_type = "Cell", symbol = "Symbol", gene = "ENSG", z = "Z",
    pval = "P", susie_pip = "PIP", cs = "CS",
    gene_biotype = "Type", chr = "Chr"
  ) |>
  cols_width(
    gene ~ px(150),           
    symbol ~ px(100),         
    cell_type ~ px(80),
    z ~ px(70),
    pval ~ px(90),
    susie_pip ~ px(70),
    cs ~ px(50),
    chr ~ px(50),
    gene_biotype ~ px(110)
  ) |>
  tab_options(
    table.font.size = "84%",
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  ) |>
  print()
    cat("\n\n")
    
    }
    
  
}
```

***

### Plot

***

```{r, ctwas_plot, echo = FALSE, message = FALSE, warning = FALSE}
# make_locusplot(finemap_res,
#                region_id = "22_40149793_42294812",
#                ens_db = ens_db,
#                weights = weights,
#                highlight_pip = 0.8,
#                filter_protein_coding_genes = TRUE,
#                filter_cs = TRUE,
#                color_pval_by = "cs",
#                color_pip_by = "cs")
```

***

### Benchmarking {.tabset}

***

##### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "12ctwas.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
summary %>%
  gt() %>%
  tab_header(
    title = "Table X. Benchmarking",
    subtitle = "Means and SD across replicates"
  ) %>%
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) %>%
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```