---
title: "Developmental Specificity of Fetal Brain eQTL"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
params:
  eqtl_universe: null
  classification: null
  ocr_enrich: null
  eqtl_effects: null
  beta_correlation: null
  perm_dir: null
  gene_lookup: null
  bmark_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(gt)
  library(DT)
  library(data.table)
  library(tidyverse)
})

# --- Params, not snakemake@input: this Rmd is rendered via
# rmarkdown::render(params = list(...)), not Snakemake's script: directive ---
eqtl_universe_file    <- params$eqtl_universe
classification_file   <- params$classification
ocr_enrich_file        <- params$ocr_enrich
eqtl_effects_file      <- params$eqtl_effects
beta_correlation_file  <- params$beta_correlation
perm_dir               <- params$perm_dir
bmark_dir              <- params$bmark_dir

gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)

# Smaller-font DT wrapper for wide tables (Opposite effects, Subcluster-
# specific gene table) so all columns fit on screen without horizontal scroll
small_dt <- function(df, page_length = 20) {
  datatable(
    df, filter = "top", rownames = FALSE, class = "compact stripe hover",
    options = list(
      pageLength = page_length, scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '75%'});",
        "}"
      )
    )
  )
}
```

```{r load-data}
eqtl_universe    <- read_rds(eqtl_universe_file)
classification   <- read_rds(classification_file)
ocr_enrich       <- read_rds(ocr_enrich_file)
eqtl_effects     <- read_rds(eqtl_effects_file)
beta_correlation <- read_rds(beta_correlation_file)

# Unpack classification
any_table    <- classification$any_table
any_summary  <- classification$any_summary
cs_table     <- classification$cell_specific_table
cs_summary   <- classification$cell_specific_summary

# Unpack subcluster/supercluster effect-size object
subcluster_tbl   <- eqtl_effects$subcluster_specific
supercluster_tbl <- eqtl_effects$supercluster_specific
de_markers       <- eqtl_effects$de_markers_raw

# Full L1+L2+adult pairwise slope matrix (reinstated from the prior report
# version). ASSUMED still present under these two names in eqtl_effects --
# please confirm eqtl_effect_sizes.R still writes effect_size_matrix /
# effect_size_long alongside the subcluster/supercluster tables.
effect_matrix <- eqtl_effects$effect_size_matrix
effect_long   <- eqtl_effects$effect_size_long

# Unpack beta correlation object (real structure from devspec_beta_correlation.R)
paired_betas     <- beta_correlation$paired_betas
opposite_effects <- beta_correlation$opposite_effects
celltype_map     <- beta_correlation$celltype_map
af_tolerance     <- beta_correlation$af_tolerance

# Add gene symbols (paired_betas/opposite_effects only carry ENSG in `gene`)
paired_betas <- paired_betas |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

opposite_effects <- opposite_effects |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

# Three-way point classification for the scatter grid: grey (not significant in
# both datasets), blue (significant in both, concordant), red (significant in
# both, discordant). Plotted on every overlapping pair regardless of category --
# only the colour/highlight is restricted to the both-significant subset.
paired_betas <- paired_betas |>
  mutate(
    significant_both = significant_fugita == "Yes",
    concordant = sign(beta_my) == sign(beta_fugita),
    point_category = case_when(
      significant_both & concordant  ~ "concordant_both_sig",
      significant_both & !concordant ~ "discordant_both_sig",
      TRUE                            ~ "other"
    )
  )

# Per-comparison Pearson r (no cor_summary is written by the beta_correlation
# rule, so compute it here)
beta_cor_summary <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n = n(),
    r = cor(beta_my, beta_fugita, use = "complete.obs"),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type))

comparison_levels <- paste(celltype_map$my_cell, "vs", celltype_map$fugita_cell)

# Significance-count summary table (one row per comparison): significant in
# fetal & present in adult; also significant in Fujita; of those, concordant;
# of those, discordant.
significance_counts <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n_fetal_sig_present_adult = n(),
    n_also_sig_fujita = sum(significant_both, na.rm = TRUE),
    n_both_sig_concordant = sum(significant_both & concordant, na.rm = TRUE),
    n_both_sig_discordant = sum(significant_both & !concordant, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type)) |>
  arrange(factor(comparison, levels = comparison_levels))

# exp_pc_map for all 19 cell types (L1 + L2), hardcoded per instruction
exp_pc_map <- c(
  "Glu-UL" = 50, "Glu-DL" = 40, "NPC" = 30, "GABA" = 30,
  "Endo-Peri" = 30, "OPC" = 30, "MG" = 30,
  "Glu-UL-0" = 30, "Glu-UL-1" = 50, "Glu-UL-2" = 40,
  "Glu-DL-0" = 40, "Glu-DL-1" = 40, "Glu-DL-2" = 30,
  "GABA-0" = 30, "GABA-1" = 40, "GABA-2" = 40,
  "NPC-0" = 50, "NPC-1" = 30, "NPC-2" = 20
)
geno_pc     <- 4
norm_method <- "quantile"

# GO gene universe = every phenotype_id tested in that cell type's own perm
# file (unfiltered), matching the same path pattern used in
# devspec_beta_correlation.R
read_perm_universe <- function(cell_type, perm_dir) {
  expPC <- exp_pc_map[[cell_type]]
  f <- paste0(perm_dir, cell_type, "_", norm_method, "_genPC_", geno_pc,
              "_expPC_", expPC, "/", cell_type, "_", norm_method, "_perm.cis_qtl.txt.gz")
  if (!file.exists(f)) {
    warning(paste("perm file not found for", cell_type, ":", f))
    return(character(0))
  }
  fread(f, select = "phenotype_id") |> pull(phenotype_id) |> unique()
}
```

## Developmental specificity pipeline steps {.tabset}

***

This report addresses reviewer comments requesting formal quantification of
developmental-stage specificity in fetal brain eQTL, using Fujita et al.
(2024) adult cortical snRNA-seq eQTL data as the adult reference.

***

1. `extract_eqtl_universe`: Extract unique fetal eGene-eSNP universe (L1+L2)
2. `classify_eqtl_specificity`: Classify pairs as testable/untestable and shared/fetal-specific vs Fujita
3. `ocr_enrich`: Test fetal-specific eQTL enrichment in fetal (Ziffra) and adult (Li et al.) OCRs
4. `eqtl_effect_sizes`: Extract subcluster/supercluster-specific eQTL and DE marker crossref
5. `beta_correlation`: Fetal-adult eQTL beta correlations and opposite-effect allele checks
6. `dev_specificity_report`: Summarise developmental specificity results

***

### Classification of fetal eQTL {.tabset}

Unique eGene-eSNP pairs classified against Fujita et al.
adult eGene-eSNP pairs under two matching tiers: 

- "Any": tested against all 7 Fujita cell types, shared if sig. in at least Fugita cell type
- "Cell-specific": Eeach fetal cell type tested only against its a priori
label-matched Fujita cell type; NPC has no adult equivalent

***

#### "Any" tier {.tabset}

##### Summary table

```{r any-tier-summary}
any_summary |>
  mutate(pct = percent(prop, accuracy = 0.1)) |>
  select(Category = final_category, Count = n, Proportion = pct) |>
  gt() |>
  tab_header(title = '"Any" tier classification')
```

##### Barplot

```{r any-tier-barplot, fig.height=5, fig.width=7}
any_summary |>
  mutate(
    final_category = fct_reorder(final_category, n),
    fill_group = case_when(
      str_detect(final_category, "shared")   ~ "Shared",
      str_detect(final_category, "specific") ~ "Fetal-specific",
      TRUE                                   ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = final_category, y = n, fill = fill_group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70")) +
  labs(x = NULL, y = "Number of unique (eGene, eSNP) pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

##### Testable-pairs pie

```{r any-tier-testable-pie, fig.height=4, fig.width=5}
any_table |>
  filter(str_detect(final_category, "^testable")) |>
  count(final_category) |>
  ggplot(aes(x = "", y = n, fill = final_category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("testable_shared" = "#4DAF4A", "testable_specific" = "#E41A1C")) +
  labs(fill = "Classification") +
  theme_void(base_size = 13)
```

***

#### Cell-specific tier {.tabset}

##### Per-cell-type table

```{r cs-tier-summary}
cs_summary |>
  filter(!is.na(ref_cell_type)) |>
  select(Cell_Type = cell_type, Fujita_Match = ref_cell_type,
         Category = final_category, Count = n, Proportion = prop) |>
  datatable(filter = "top", rownames = FALSE,
            options = list(pageLength = 20, scrollX = TRUE))
```

##### Barplot

```{r cs-tier-barplot, fig.height=8, fig.width=10}
cs_summary |>
  filter(!is.na(ref_cell_type)) |>
  mutate(
    fill_group = case_when(
      str_detect(final_category, "shared")    ~ "Shared",
      str_detect(final_category, "specific")  ~ "Fetal-specific",
      final_category == "no_adult_equivalent" ~ "No adult equivalent",
      TRUE                                    ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = cell_type, y = n, fill = fill_group)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70", "No adult equivalent" = "grey40")) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Proportion of eGene-eSNP pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

### Effect-size concordance {.tabset}

Pearson correlation of eQTL betas between each fetal cell type and its
matched Fujita adult cell type. Points are highlighted only where the pair 
is sig. in both datasets  (significant_by_2step_FDR == “Yes”):

- Blue: sig. in both and concordant sign
- Red: sig. in both and and discordant sign (also shown in oppsite effects tab) 
- Grey: sig. in fetal, not sig. in Fugita

***

#### Significance counts

```{r beta-significance-counts}
significance_counts |>
  select(Comparison = comparison,
         `Fetal-sig, present in adult` = n_fetal_sig_present_adult,
         `Also Fujita-sig` = n_also_sig_fujita,
         `Both-sig, concordant` = n_both_sig_concordant,
         `Both-sig, discordant` = n_both_sig_discordant) |>
  gt() |>
  tab_header(title = "Significance counts per fetal-adult comparison")
```

***

#### Scatter grid

```{r beta-scatter-grid, fig.height=8, fig.width=14}
plot_df <- paired_betas |>
  mutate(
    comparison = factor(paste(my_cell_type, "vs", fugita_cell_type), levels = comparison_levels)
  ) |>
  # Plot "other" (grey) points first so blue/red highlighted points draw on top
  arrange(point_category == "other")

r_labels <- beta_cor_summary |>
  mutate(comparison = factor(comparison, levels = comparison_levels),
         label = paste0("r = ", round(r, 3), "\nn = ", n))

ggplot(plot_df, aes(x = beta_my, y = beta_fugita)) +
  geom_point(aes(colour = point_category, alpha = point_category != "other", size = point_category != "other")) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#377EB8", linewidth = 0.6) +
  geom_text(data = r_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  scale_colour_manual(values = c(concordant_both_sig = "#377EB8",
                                 discordant_both_sig = "#E41A1C",
                                 other = "grey70"),
                       labels = c(concordant_both_sig = "Concordant (sig. both)",
                                  discordant_both_sig = "Discordant (sig. both)",
                                  other = "Not sig. in both")) +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.25), guide = "none") +
  scale_size_manual(values = c(`TRUE` = 1.1, `FALSE` = 0.6), guide = "none") +
  facet_wrap(~ comparison, ncol = 4, scales = "free") +
  labs(x = "Fetal slope", y = "Adult (Fujita) slope", colour = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

***

#### Correlation heatmap {.tabset}

Pairwise Pearson correlations of eQTL effect sizes (slope) across all fetal
L1 cell types, fetal L2 subclusters, and Fujita adult cell types, computed
on the subset of (eGene, eSNP) pairs with non-missing slopes in both cell
types of each comparison.

##### Clustered (original)

```{r effect-size-heatmap, fig.height=8, fig.width=9}
slope_cols <- grep("^slope_", names(effect_matrix), value = TRUE)
slope_mat <- as.matrix(effect_matrix[, slope_cols])
colnames(slope_mat) <- str_remove(colnames(slope_mat), "^slope_")

fetal_L1_cols <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
fetal_L2_cols <- sort(setdiff(
  colnames(slope_mat)[!str_starts(colnames(slope_mat), "Fujita_")],
  fetal_L1_cols
))
fujita_cols <- sort(colnames(slope_mat)[str_starts(colnames(slope_mat), "Fujita_")])

col_order <- c(fetal_L1_cols, fetal_L2_cols, fujita_cols)
col_order <- col_order[col_order %in% colnames(slope_mat)]
slope_mat <- slope_mat[, col_order]

cor_mat <- cor(slope_mat, use = "pairwise.complete.obs")

stage_anno <- ifelse(str_starts(colnames(cor_mat), "Fujita_"), "Adult", "Fetal")
ha <- HeatmapAnnotation(
  Stage = stage_anno,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

Heatmap(
  cor_mat,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Cell-type ordered

Dendrogram removed, rows/columns fixed so each L1 cell type sits next to its
own L2 subclusters

```{r effect-size-heatmap-grouped, fig.height=8, fig.width=9}
grouped_order <- c(
  "GABA", "GABA-0", "GABA-1", "GABA-2",
  "Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
  "Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2",
  "NPC", "NPC-0", "NPC-1", "NPC-2",
  "OPC", "MG", "Endo-Peri",
  "Fujita_Exc", "Fujita_Inh", "Fujita_Mic", "Fujita_OPC", "Fujita_End",
  "Fujita_Ast", "Fujita_Oli"
)
grouped_order <- grouped_order[grouped_order %in% colnames(cor_mat)]

cor_mat_grouped <- cor_mat[grouped_order, grouped_order]
stage_anno_grouped <- ifelse(str_starts(colnames(cor_mat_grouped), "Fujita_"), "Adult", "Fetal")
ha_grouped <- HeatmapAnnotation(
  Stage = stage_anno_grouped,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

Heatmap(
  cor_mat_grouped,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha_grouped,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_grouped[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Fetal only

Adult (Fujita) cell types removed entirely; fetal L1/L2 comparisons only,
re-clustered on this fetal-only subset.

```{r effect-size-heatmap-fetal-only, fig.height=8, fig.width=8}
fetal_only_cols <- colnames(cor_mat)[!str_starts(colnames(cor_mat), "Fujita_")]
cor_mat_fetal <- cor_mat[fetal_only_cols, fetal_only_cols]

Heatmap(
  cor_mat_fetal,
  name = "Pearson r",
  col = col_fun,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_fetal[i, j]), x, y, gp = gpar(fontsize = 7))
  },
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9)
)
```

***

#### Pairwise n

```{r pairwise-n, fig.height=8, fig.width=9}
n_mat <- matrix(NA, ncol(slope_mat), ncol(slope_mat),
                dimnames = list(colnames(slope_mat), colnames(slope_mat)))
for (i in seq_len(ncol(slope_mat))) {
  for (j in seq_len(ncol(slope_mat))) {
    n_mat[i, j] <- sum(complete.cases(slope_mat[, c(i, j)]))
  }
}

col_fun_n <- colorRamp2(c(0, max(n_mat, na.rm = TRUE)), c("white", "#08519C"))

Heatmap(
  n_mat,
  name = "n pairs",
  col = col_fun_n,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = ha,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(n_mat[i, j], x, y, gp = gpar(fontsize = 5))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

***

#### Full slope table

Betas for every (eGene, eSNP) pair across all 19 fetal cell types (L1+L2)
and all 7 Fujita adult cell types

```{r full-slope-table}
# ASSUMED id column names in effect_matrix (phenotype_id, variant_id); adjust
# the two names below if eqtl_effect_sizes.R actually names them differently.
slope_full_tbl <- effect_matrix |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(external_gene_name, phenotype_id, variant_id,
         all_of(paste0("slope_", col_order))) |>
  rename(Symbol = external_gene_name, ENSG_ID = phenotype_id, SNP = variant_id) |>
  rename_with(~ str_remove(., "^slope_"), starts_with("slope_"))

small_dt(slope_full_tbl)
```

***

#### Opposite effects

- Sig. eQTL Pairs in both fetal and Fujita datasets with discordant effect
sign (the red points in the scatter grid). 
- Fetal/Fujita/1000G columns show `REF/ALT; allele frequency` for each dataset 
- AF concordance tolerance: `r af_tolerance`.

```{r opposite-effects-table}
opposite_effects |>
  mutate(
    Fetal  = paste0(ref_my, "/", alt_my, "; ", round(af_my, 4)),
    Fujita = paste0(ref_fugita, "/", alt_fugita, "; ", round(af_fugita, 4)),
    `1000G` = ifelse(is.na(a1_1000G), NA_character_,
                      paste0(a1_1000G, "/", a2_1000G, "; ", round(maf_1000G, 4)))
  ) |>
  select(Gene = external_gene_name, ENSG = gene, eSNP = snp,
         Fetal_Cell = my_cell_type, Fujita_Cell = fugita_cell_type,
         Beta_Fetal = beta_my, Beta_Fujita = beta_fugita,
         Fetal, Fujita, `1000G`,
         Allele_Status = overall_allele_status) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### OCR enrichment {.tabset}

Fetal-specific and shared eQTL SNPs tested for enrichment in fetal (Ziffra
et al. 2021) and adult (Li et al. 2023) open chromatin regions via block
bootstrap permutation.

***

#### Fold enrichment

- Fetal-specific SNP enrichment in fetal and adult OCRs (shared SNPs omitted, but still intable) here as a contrast set not needed
- NPC (no adult progenitor equivalent) is excluded from this pairing.

```{r ocr-enrichment-barplot, fig.height=7, fig.width=9}
# Fixed pair order: alternating fetal / matched-adult, grouped by comparison
ocr_pair_order <- c(
  "ulEN", "Glu-UL",
  "dlEN", "Glu-DL",
  "GABA", "IN_MGE",
  "OPC", "AstroOligo",
  "MG", "Microglia",
  "EndoMural", "Endo-Peri",
  "union", "cortical_union"
)

ocr_enrich |>
  filter(snp_set == "fetal_specific") |>
  mutate(
    peak_set = str_remove(peak_set, "^(fetal_|adult_)"),
    peak_set = str_remove(peak_set, "_MACSpeaks$"),
    peak_set = factor(peak_set, levels = rev(ocr_pair_order)),
    sig_label = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns")
  ) |>
  filter(!is.na(peak_set)) |>
  ggplot(aes(x = peak_set, y = fold_enrichment, fill = stage)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_text(aes(label = sig_label), hjust = -0.3, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c(fetal = "#984EA3", adult = "#FF7F00")) +
  labs(x = NULL, y = "Fold enrichment (fetal-specific SNPs)", fill = "Peak stage") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

#### Paired test (fetal vs adult)

- Paired Wilcoxon signed-rank test on fold enrichment (fetal-specific SNPs
only) across the 7 matched fetal/adult comparisons
- testing whether fetal fold enrichment exceeds matched adult fold enrichment.
- NPC excluded, no adult equivalent 

```{r ocr-paired-test}
ocr_pairs <- tribble(
  ~comparison,   ~fetal_peak,           ~adult_peak,
  "Glu-UL",      "fetal_ulEN_MACSpeaks",       "adult_Glu-UL",
  "Glu-DL",      "fetal_dlEN_MACSpeaks",       "adult_Glu-DL",
  "GABA",        "fetal_IN_MGE_MACSpeaks",     "adult_GABA",
  "OPC",         "fetal_AstroOligo_MACSpeaks", "adult_OPC",
  "MG",          "fetal_Microglia_MACSpeaks",  "adult_MG",
  "Endo-Peri",   "fetal_EndoMural_MACSpeaks",  "adult_Endo-Peri",
  "union",       "fetal_union",                "adult_cortical_union"
)

fold_lookup <- ocr_enrich |>
  filter(snp_set == "fetal_specific") |>
  select(peak_set, fold_enrichment)

paired_fold <- ocr_pairs |>
  left_join(fold_lookup, by = c("fetal_peak" = "peak_set")) |>
  rename(fetal_fold = fold_enrichment) |>
  left_join(fold_lookup, by = c("adult_peak" = "peak_set")) |>
  rename(adult_fold = fold_enrichment) |>
  mutate(diff = fetal_fold - adult_fold)

paired_fold |>
  select(comparison, fetal_fold, adult_fold, diff) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  gt() |>
  tab_header(title = "Paired fold enrichment (fetal-specific SNPs)")

wilcox_result <- wilcox.test(paired_fold$fetal_fold, paired_fold$adult_fold,
                              paired = TRUE, alternative = "greater", exact = TRUE)
print(wilcox_result)
```

***

#### Results table

```{r ocr-enrichment-table}
ocr_enrich |>
  mutate(p_value = signif(p_value, 3),
         fold_enrichment = round(fold_enrichment, 2),
         mean_permuted_overlap = round(mean_permuted_overlap, 1)) |>
  select(Peak_Set = peak_set, Stage = stage, SNP_Set = snp_set,
         N_SNPs = n_snps, N_Peaks = n_peaks,
         Observed = observed_overlap, Mean_Permuted = mean_permuted_overlap,
         Fold_Enrichment = fold_enrichment, P_Value = p_value) |>
  gt() |>
  tab_header(title = "OCR enrichment results") |>
  tab_options(table.font.size = px(11))
```

***

### Subcluster-specific eQTL {.tabset}

eGenes significant at FDR < 0.05 in an L2 subcluster but not its parent L1
("subcluster-specific"), and vice versa ("supercluster-specific"). GO
universe is every gene tested in that cell type's own TensorQTL perm file
(unfiltered).

***

#### Counts

```{r subcluster-counts}
subcluster_tbl |>
  count(cell_type_L2, parent_L1, name = "n_specific_eGenes") |>
  gt()
```

***

#### Parent nominal statistics

```{r subcluster-parent-stats, fig.height=6, fig.width=9}
p1 <- subcluster_tbl |>
  filter(!is.na(pval_nom_parent)) |>
  ggplot(aes(x = pval_nom_parent)) +
  geom_histogram(bins = 50, fill = "#377EB8", colour = "white") +
  facet_wrap(~ cell_type_L2, scales = "free_y") +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "red") +
  labs(x = "Nominal p-value in parent L1", y = "Count") +
  theme_classic(base_size = 11)

p2 <- subcluster_tbl |>
  filter(!is.na(slope_L2), !is.na(slope_parent)) |>
  ggplot(aes(x = slope_L2, y = slope_parent)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linetype = "dashed") +
  facet_wrap(~ cell_type_L2, scales = "free") +
  labs(x = "Slope in L2 subcluster", y = "Slope in parent L1") +
  theme_classic(base_size = 11)

p1 / p2
```

***

#### Summary per L2 type

```{r subcluster-summary-table}
subcluster_tbl |>
  group_by(cell_type_L2, parent_L1) |>
  summarise(
    n_eGenes = n_distinct(phenotype_id),
    median_pval_parent = round(median(pval_nom_parent, na.rm = TRUE), 4),
    prop_same_direction = round(mean(sign(slope_L2) == sign(slope_parent), na.rm = TRUE), 3),
    prop_nom_sig_parent = round(mean(pval_nom_parent < 0.05, na.rm = TRUE), 3),
    n_de_markers = sum(is_de_marker, na.rm = TRUE),
    .groups = "drop"
  ) |>
  gt() |>
  tab_options(table.font.size = px(11))
```

***

#### Supercluster-specific counts

```{r supercluster-counts}
supercluster_tbl |>
  count(parent_L1, name = "n_L1_only_eGenes") |>
  gt()
```

***

#### DE marker overlap

```{r de-overlap, fig.height=5, fig.width=7}
subcluster_tbl |>
  count(cell_type_L2, is_de_marker) |>
  ggplot(aes(x = cell_type_L2, y = n, fill = is_de_marker)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "grey70"),
                     labels = c("TRUE" = "DE marker", "FALSE" = "Not DE")) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Proportion of subcluster-specific eGenes", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

#### GO enrichment (subcluster-specific)

```{r go-subcluster, fig.height=10, fig.width=10, results='asis'}
go_results_sub <- list()

for (ct in unique(subcluster_tbl$cell_type_L2)) {
  genes <- subcluster_tbl |> filter(cell_type_L2 == ct) |> pull(phenotype_id) |> unique()
  if (length(genes) < 5) next

  universe_genes <- read_perm_universe(ct, perm_dir)

  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)
  universe_entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = universe_genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)

  if (is.null(entrez) || sum(!is.na(entrez)) < 5) {
    cat("GO universe/entrez mapping failed or insufficient for", ct, "\n\n")
    next
  }

  ego <- tryCatch(
    enrichGO(gene = na.omit(entrez), universe = na.omit(universe_entrez),
             OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH",
             pvalueCutoff = 0.05, readable = TRUE),
    error = function(e) { cat("enrichGO failed for", ct, ":", conditionMessage(e), "\n\n"); NULL })

  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) go_results_sub[[ct]] <- ego
}

for (ct in names(go_results_sub)) {
  cat("\n\n#####", ct, "\n\n")
  print(dotplot(go_results_sub[[ct]], showCategory = 15, title = ct))
}
if (length(go_results_sub) == 0) cat("No cell types produced GO output.\n")
```

***

#### GO enrichment (supercluster-specific)

```{r go-supercluster, fig.height=10, fig.width=10, results='asis'}
go_results_super <- list()

for (ct in unique(supercluster_tbl$parent_L1)) {
  genes <- supercluster_tbl |> filter(parent_L1 == ct) |> pull(phenotype_id) |> unique()
  if (length(genes) < 5) next

  universe_genes <- read_perm_universe(ct, perm_dir)

  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)
  universe_entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = universe_genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)

  if (is.null(entrez) || sum(!is.na(entrez)) < 5) {
    cat("GO universe/entrez mapping failed or insufficient for", ct, "\n\n")
    next
  }

  ego <- tryCatch(
    enrichGO(gene = na.omit(entrez), universe = na.omit(universe_entrez),
             OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH",
             pvalueCutoff = 0.05, readable = TRUE),
    error = function(e) { cat("enrichGO failed for", ct, ":", conditionMessage(e), "\n\n"); NULL })

  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) go_results_super[[ct]] <- ego
}

for (ct in names(go_results_super)) {
  cat("\n\n#####", ct, "\n\n")
  print(dotplot(go_results_super[[ct]], showCategory = 15, title = ct))
}
if (length(go_results_super) == 0) cat("No cell types produced GO output.\n")
```

***

#### Subcluster-specific gene table

```{r subcluster-gene-table}
subcluster_tbl |>
  select(Gene = gene_symbol, ENSG = phenotype_id, eSNP = variant_id,
         L2_Type = cell_type_L2, Parent_L1 = parent_L1,
         Slope_L2 = slope_L2, QVal_L2 = qval_L2,
         Slope_Parent = slope_parent, Pval_Parent = pval_nom_parent,
         DE_Marker = is_de_marker) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
files <- list.files(bmark_dir, pattern = "dev_specificity\\..*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]
  # "dev_specificity.extract_universe.txt" -> parts[2] = "extract_universe"
  rule <- parts[2]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

summary <- benchmarks |>
  group_by(rule) |>
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) |>
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  arrange(rule)

summary |>
  gt() |>
  tab_header(
    title = "Benchmarking",
    subtitle = "Means and SD across replicates"
  ) |>
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) |>
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) |>
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

```{r runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```

```{r session-info}
sessionInfo()
```

