---
title: "Developmental Specificity of Fetal Brain eQTL"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
params:
  eqtl_universe: null
  classification: null
  heterogeneity_test: null
  ocr_enrich: null
  eqtl_effects: null
  beta_correlation: null
  perm_dir: null
  gene_lookup: null
  bmark_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(gt)
  library(DT)
  library(data.table)
  library(tidyverse)
})

# --- Params, not snakemake@input: this Rmd is rendered via
# rmarkdown::render(params = list(...)), not Snakemake's script: directive ---
eqtl_universe_file    <- params$eqtl_universe
classification_file   <- params$classification
heterogeneity_file    <- params$heterogeneity_test
ocr_enrich_file        <- params$ocr_enrich
eqtl_effects_file      <- params$eqtl_effects
beta_correlation_file  <- params$beta_correlation
bmark_dir              <- params$bmark_dir

gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)

# Smaller-font DT wrapper for wide tables (Opposite effects, Heterogeneity
# results, Subcluster-specific gene table) so all columns fit on screen
# without horizontal scroll
small_dt <- function(df, page_length = 20) {
  datatable(
    df, filter = "top", rownames = FALSE, class = "compact stripe hover",
    options = list(
      pageLength = page_length, scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '75%'});",
        "}"
      )
    )
  )
}
```

```{r load-data}
eqtl_universe    <- read_rds(eqtl_universe_file)
classification   <- read_rds(classification_file)
het_test         <- read_rds(heterogeneity_file)
ocr_enrich       <- read_rds(ocr_enrich_file)
eqtl_effects     <- read_rds(eqtl_effects_file)
beta_correlation <- read_rds(beta_correlation_file)

# Unpack classification -- "any" tier only; cell_specific tier retired
# (superseded by the heterogeneity test, below)
any_table    <- classification$any_table
any_summary  <- classification$any_summary

# Collapse the two untestable-not-expressed categories for display only --
# B's underlying categories stay granular in its own TSV output.
any_summary <- any_summary |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable_not_expressed",
    untestable_not_genotyped_not_expressed = "untestable_not_expressed"
  )) |>
  group_by(final_category) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(prop = round(n / sum(n), 3))

any_table <- any_table |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable_not_expressed",
    untestable_not_genotyped_not_expressed = "untestable_not_expressed"
  ))

# Unpack heterogeneity test (Cochran's Q on fetal-specific pairs)
het_usable <- het_test$usable
het_funnel <- het_test$funnel

# Unpack subcluster/supercluster effect-size object
subcluster_tbl   <- eqtl_effects$subcluster_specific
supercluster_tbl <- eqtl_effects$supercluster_specific

# Full L1+L2+adult pairwise slope matrix
effect_matrix <- eqtl_effects$effect_size_matrix
effect_long   <- eqtl_effects$effect_size_long

# Unpack beta correlation object (real structure from devspec_beta_correlation.R)
paired_betas     <- beta_correlation$paired_betas
opposite_effects <- beta_correlation$opposite_effects
celltype_map     <- beta_correlation$celltype_map
af_tolerance     <- beta_correlation$af_tolerance

# Add gene symbols (paired_betas/opposite_effects only carry ENSG in `gene`)
paired_betas <- paired_betas |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

opposite_effects <- opposite_effects |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

# Three-way point classification for the scatter grid: grey (not significant in
# both datasets), blue (significant in both, concordant), red (significant in
# both, discordant). Plotted on every overlapping pair regardless of category --
# only the colour/highlight is restricted to the both-significant subset.
paired_betas <- paired_betas |>
  mutate(
    significant_both = significant_fugita == "Yes",
    concordant = sign(beta_my) == sign(beta_fugita),
    point_category = case_when(
      significant_both & concordant  ~ "concordant_both_sig",
      significant_both & !concordant ~ "discordant_both_sig",
      TRUE                            ~ "other"
    )
  )

# Per-comparison Pearson r (no cor_summary is written by the beta_correlation
# rule, so compute it here)
beta_cor_summary <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n = n(),
    r = cor(beta_my, beta_fugita, use = "complete.obs"),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type))

comparison_levels <- paste(celltype_map$my_cell, "vs", celltype_map$fugita_cell)

# Significance-count summary table (one row per comparison): significant in
# fetal & present in adult; also significant in Fujita; of those, concordant;
# of those, discordant.
significance_counts <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n_fetal_sig_present_adult = n(),
    n_also_sig_fujita = sum(significant_both, na.rm = TRUE),
    n_both_sig_concordant = sum(significant_both & concordant, na.rm = TRUE),
    n_both_sig_discordant = sum(significant_both & !concordant, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type)) |>
  arrange(factor(comparison, levels = comparison_levels))
```

## Developmental specificity pipeline steps {.tabset}

***

1. `extract_eqtl_universe`: Extract unique fetal eGene-eSNP universe (L1+L2)
2. `classify_eqtl_specificity`: Classify pairs as testable/untestable and shared/fetal-specific vs Fujita ("any" tier)
3. `heterogeneity_test`: Cochran's Q test on fetal-specific pairs vs matched adult Fujita cell type
4. `prep_ld_clump_input` / `run_ld_clump` / `parse_ld_clump`: LD-independence pruning of the eGene-eSNP universe
5. `ocr_enrich`: Test fetal-specific/shared eQTL enrichment in adult (Li et al.) cortical union OCRs
6. `eqtl_effect_sizes`: Extract subcluster/supercluster-specific eQTL
7. `beta_correlation`: Fetal-adult eQTL beta correlations and opposite-effect allele checks
8. `dev_specificity_report`: Summarise developmental specificity results

***

### Classification of fetal eQTL {.tabset}

Unique eGene-eSNP pairs sig. in fetal data classified against adult eGene-eSNP pairs:

- Testable shared: if sig. in >= 1 Fugita cell type
  - Feeds: LD-clumping, OCR enrichment, Heterogeneity test*
- Testable specific: present in Fugita, but not sig. in any Fugita cell type

***

#### Summary table

```{r any-tier-summary}
any_summary |>
  mutate(pct = percent(prop, accuracy = 0.1)) |>
  select(Category = final_category, Count = n, Proportion = pct) |>
  gt() |>
  tab_header(title = '"Any" tier classification')
```

#### Barplot

```{r any-tier-barplot, fig.height=5, fig.width=7}
any_summary |>
  mutate(
    final_category = fct_reorder(final_category, n),
    fill_group = case_when(
      str_detect(final_category, "shared")   ~ "Shared",
      str_detect(final_category, "specific") ~ "Fetal-specific",
      TRUE                                   ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = final_category, y = n, fill = fill_group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70")) +
  labs(x = NULL, y = "Number of unique (eGene, eSNP) pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

#### Testable-pairs pie

```{r any-tier-testable-pie, fig.height=4, fig.width=5}
any_table |>
  filter(str_detect(final_category, "^testable")) |>
  count(final_category) |>
  ggplot(aes(x = "", y = n, fill = final_category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("testable_shared" = "#4DAF4A", "testable_specific" = "#E41A1C")) +
  labs(fill = "Classification") +
  theme_void(base_size = 13)
```

***

### Heterogeneity test (fetal-specific eQTL) {.tabset}

- Cochran's Q: Tests whether the fetal and matched-adult effect  sizes differ 
by more than combined sampling error predicts
- FDR (BH) applied once, pooled across all usable rows
- Notes:
  - Q test is not directional. It tells you if the two betas differ but not which way
  - Many SNP-gene pairs in input annotation (testable specific) dropped. Mainly due to
    pair lacking overlap in relevant adult cell type. For example pair is sig. in 
    GABA, but only expressed (but not sig.) in adult Mic. Het tests can only be run in
    matching fetal and adult cell populations.

***

#### Summary

```{r het-test-funnel}
het_funnel |>
  gt() |>
  tab_header(title = "Population funnel")
```

```{r het-test-summary-stats, results = "asis"}
n_het   <- sum(het_usable$heterogeneous)
n_total <- nrow(het_usable)

cat(sprintf(
  "%d of %d usable rows (%s) were formally heterogeneous (FDR < 0.05).",
  n_het, n_total, percent(n_het / n_total, accuracy = 0.1)
))
```

***

#### Results table

```{r het-test-results-table}
het_usable |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(
    Cell_Type = cell_type,
    ENSG = phenotype_id,
    Symbol = external_gene_name,
    eSNP = variant_id,
    Fetal_Beta = beta_my, Fetal_SE = se_my, Fetal_AF = af_my,
    Adult_Cell_Type = fugita_label_match,
    Adult_Beta = beta_fugita_harmonized, Adult_SE = se_fugita, Adult_AF = af_fugita,
    Delta_Beta = delta_beta,
    Q, P = p, Q_value = q, Heterogeneous = heterogeneous
  ) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Effect-size concordance {.tabset}

Pearson correlation of eQTL betas between each fetal cell type and its
matched Fujita adult cell type. Points are highlighted only where the pair 
is sig. in both datasets  (significant_by_2step_FDR == "Yes"):

- Blue: sig. in both and concordant sign
- Red: sig. in both and and discordant sign (also shown in oppsite effects tab) 
- Grey: sig. in fetal, not sig. in Fugita

***

#### Significance counts

```{r beta-significance-counts}
significance_counts |>
  select(Comparison = comparison,
         `Fetal-sig, present in adult` = n_fetal_sig_present_adult,
         `Also Fujita-sig` = n_also_sig_fujita,
         `Both-sig, concordant` = n_both_sig_concordant,
         `Both-sig, discordant` = n_both_sig_discordant) |>
  gt() |>
  tab_header(title = "Significance counts per fetal-adult comparison")
```

***

#### Scatter grid

```{r beta-scatter-grid, fig.height=8, fig.width=14}
plot_df <- paired_betas |>
  mutate(
    comparison = factor(paste(my_cell_type, "vs", fugita_cell_type), levels = comparison_levels)
  ) |>
  # Plot "other" (grey) points first so blue/red highlighted points draw on top
  arrange(point_category == "other")

r_labels <- beta_cor_summary |>
  mutate(comparison = factor(comparison, levels = comparison_levels),
         label = paste0("r = ", round(r, 3), "\nn = ", n))

ggplot(plot_df, aes(x = beta_my, y = beta_fugita)) +
  geom_point(aes(colour = point_category, alpha = point_category != "other", size = point_category != "other")) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#377EB8", linewidth = 0.6) +
  geom_text(data = r_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  scale_colour_manual(values = c(concordant_both_sig = "#377EB8",
                                 discordant_both_sig = "#E41A1C",
                                 other = "grey70"),
                       labels = c(concordant_both_sig = "Concordant (sig. both)",
                                  discordant_both_sig = "Discordant (sig. both)",
                                  other = "Not sig. in both")) +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.25), guide = "none") +
  scale_size_manual(values = c(`TRUE` = 1.1, `FALSE` = 0.6), guide = "none") +
  facet_wrap(~ comparison, ncol = 4, scales = "free") +
  labs(x = "Fetal slope", y = "Adult (Fujita) slope", colour = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

***

#### Correlation heatmap {.tabset}

Pairwise Pearson correlations of eQTL effect sizes (slope) across all fetal
L1 cell types, fetal L2 subclusters, and Fujita adult cell types, computed
on the subset of (eGene, eSNP) pairs with non-missing slopes in both cell
types of each comparison.

##### Clustered (original)

```{r effect-size-heatmap, fig.height=8, fig.width=9}
slope_cols <- grep("^slope_", names(effect_matrix), value = TRUE)
slope_mat <- as.matrix(effect_matrix[, slope_cols])
colnames(slope_mat) <- str_remove(colnames(slope_mat), "^slope_")

fetal_L1_cols <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
fetal_L2_cols <- sort(setdiff(
  colnames(slope_mat)[!str_starts(colnames(slope_mat), "Fujita_")],
  fetal_L1_cols
))
fujita_cols <- sort(colnames(slope_mat)[str_starts(colnames(slope_mat), "Fujita_")])

col_order <- c(fetal_L1_cols, fetal_L2_cols, fujita_cols)
col_order <- col_order[col_order %in% colnames(slope_mat)]
slope_mat <- slope_mat[, col_order]

cor_mat <- cor(slope_mat, use = "pairwise.complete.obs")

stage_anno <- ifelse(str_starts(colnames(cor_mat), "Fujita_"), "Adult", "Fetal")
ha <- HeatmapAnnotation(
  Stage = stage_anno,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

Heatmap(
  cor_mat,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Cell-type ordered

Dendrogram removed, rows/columns fixed so each L1 cell type sits next to its
own L2 subclusters

```{r effect-size-heatmap-grouped, fig.height=8, fig.width=9}
grouped_order <- c(
  "GABA", "GABA-0", "GABA-1", "GABA-2",
  "Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
  "Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2",
  "NPC", "NPC-0", "NPC-1", "NPC-2",
  "OPC", "MG", "Endo-Peri",
  "Fujita_Exc", "Fujita_Inh", "Fujita_Mic", "Fujita_OPC", "Fujita_End",
  "Fujita_Ast", "Fujita_Oli"
)
grouped_order <- grouped_order[grouped_order %in% colnames(cor_mat)]

cor_mat_grouped <- cor_mat[grouped_order, grouped_order]
stage_anno_grouped <- ifelse(str_starts(colnames(cor_mat_grouped), "Fujita_"), "Adult", "Fetal")
ha_grouped <- HeatmapAnnotation(
  Stage = stage_anno_grouped,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

Heatmap(
  cor_mat_grouped,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha_grouped,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_grouped[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Fetal only

Adult (Fujita) cell types removed entirely; fetal L1/L2 comparisons only,
re-clustered on this fetal-only subset.

```{r effect-size-heatmap-fetal-only, fig.height=8, fig.width=8}
fetal_only_cols <- colnames(cor_mat)[!str_starts(colnames(cor_mat), "Fujita_")]
cor_mat_fetal <- cor_mat[fetal_only_cols, fetal_only_cols]

Heatmap(
  cor_mat_fetal,
  name = "Pearson r",
  col = col_fun,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_fetal[i, j]), x, y, gp = gpar(fontsize = 7))
  },
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9)
)
```

***

#### Pairwise n

```{r pairwise-n, fig.height=8, fig.width=9}
n_mat <- matrix(NA, ncol(slope_mat), ncol(slope_mat),
                dimnames = list(colnames(slope_mat), colnames(slope_mat)))
for (i in seq_len(ncol(slope_mat))) {
  for (j in seq_len(ncol(slope_mat))) {
    n_mat[i, j] <- sum(complete.cases(slope_mat[, c(i, j)]))
  }
}

col_fun_n <- colorRamp2(c(0, max(n_mat, na.rm = TRUE)), c("white", "#08519C"))

Heatmap(
  n_mat,
  name = "n pairs",
  col = col_fun_n,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = ha,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(n_mat[i, j], x, y, gp = gpar(fontsize = 5))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

***

#### Full slope table

Betas for every (eGene, eSNP) pair across all 19 fetal cell types (L1+L2)
and all 7 Fujita adult cell types

```{r full-slope-table}
slope_full_tbl <- effect_matrix |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(external_gene_name, phenotype_id, variant_id,
         all_of(paste0("slope_", col_order))) |>
  rename(Symbol = external_gene_name, ENSG_ID = phenotype_id, SNP = variant_id) |>
  rename_with(~ str_remove(., "^slope_"), starts_with("slope_"))

small_dt(slope_full_tbl)
```

***

#### Opposite effects

- Sig. eQTL Pairs in both fetal and Fujita datasets with discordant effect
sign (the red points in the scatter grid). 
- Fetal/Fujita/1000G columns show `REF/ALT; allele frequency` for each dataset 
- AF concordance tolerance: `r af_tolerance`.

```{r opposite-effects-table}
opposite_effects |>
  mutate(
    Fetal  = paste0(ref_my, "/", alt_my, "; ", round(af_my, 4)),
    Fujita = paste0(ref_fugita, "/", alt_fugita, "; ", round(af_fugita, 4)),
    `1000G` = ifelse(is.na(a1_1000G), NA_character_,
                      paste0(a1_1000G, "/", a2_1000G, "; ", round(maf_1000G, 4)))
  ) |>
  select(Gene = external_gene_name, ENSG = gene, eSNP = snp,
         Fetal_Cell = my_cell_type, Fujita_Cell = fugita_cell_type,
         Beta_Fetal = beta_my, Beta_Fujita = beta_fugita,
         Fetal, Fujita, `1000G`,
         Allele_Status = overall_allele_status) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### OCR enrichment {.tabset}

- LD-pruned fetal-specific and shared eQTL SNPs tested
for enrichment in adult cortical OCRs (Li et al. 2023)
- Cortical adult union peak set manually curated 

***

#### Fold enrichment

```{r ocr-enrichment-barplot, fig.height=4, fig.width=6}
ocr_enrich |>
  mutate(
    sig_label = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns"),
    snp_set = recode(snp_set, fetal_specific = "Fetal-specific", shared = "Shared")
  ) |>
  ggplot(aes(x = snp_set, y = fold_enrichment, fill = snp_set)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_text(aes(label = sig_label), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Fetal-specific" = "#E41A1C", "Shared" = "#4DAF4A")) +
  labs(x = NULL, y = "Fold enrichment (adult cortical union)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

***

#### Results table

```{r ocr-enrichment-table}
ocr_enrich |>
  mutate(p_value = signif(p_value, 3),
         fold_enrichment = round(fold_enrichment, 2),
         mean_permuted_overlap = round(mean_permuted_overlap, 1)) |>
  select(Peak_Set = peak_set, SNP_Set = snp_set,
         N_SNPs = n_snps, N_Peaks = n_peaks,
         Observed = observed_overlap, Mean_Permuted = mean_permuted_overlap,
         Fold_Enrichment = fold_enrichment, P_Value = p_value) |>
  gt() |>
  tab_header(title = "OCR enrichment results") |>
  tab_options(table.font.size = px(11))
```

***

### Subcluster-specific eQTL {.tabset}

- FDR sig. eGenes in an L2 subcluster but not its parent L1
("subcluster-specific"), and vice versa ("supercluster-specific").

***

#### Counts

```{r subcluster-counts}
subcluster_tbl |>
  count(cell_type_L2, parent_L1, name = "n_specific_eGenes") |>
  gt()
```

***

#### Supercluster-specific counts

```{r supercluster-counts}
supercluster_tbl |>
  count(parent_L1, name = "n_L1_only_eGenes") |>
  gt()
```

***

#### Subcluster-specific gene table

```{r subcluster-gene-table}
subcluster_tbl |>
  select(Gene = gene_symbol, ENSG = phenotype_id, eSNP = variant_id,
         cell_type_L1 = parent_L1, Slope_L1 = slope_parent, SE_L1 = se_parent, Pval_L1 = pval_nom_parent,
         cell_type_L2 = cell_type_L2, Slope_L2 = slope_L2, SE_L2 = se_L2, Pval_L2 = pval_nom_L2) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
files <- list.files(bmark_dir, pattern = "dev_specificity\\..*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]
  # "dev_specificity.extract_universe.txt" -> parts[2] = "extract_universe"
  rule <- parts[2]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

summary <- benchmarks |>
  group_by(rule) |>
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) |>
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  arrange(rule)

summary |>
  gt() |>
  tab_header(
    title = "Benchmarking",
    subtitle = "Means and SD across replicates"
  ) |>
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) |>
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) |>
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

```{r runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```

```{r session-info}
sessionInfo()
```
