---
title: "Developmental Specificity of Fetal Brain eQTL"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
params:
  eqtl_universe: null
  classification: null
  heterogeneity_test: null
  ocr_enrich: null
  eqtl_effects: null
  beta_correlation: null
  pi1_dir: null
  perm_dir: null
  gene_lookup: null
  bmark_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(gt)
  library(DT)
  library(data.table)
  library(tidyverse)
})

# --- Params, not snakemake@input: this Rmd is rendered via
# rmarkdown::render(params = list(...)), not Snakemake's script: directive ---
eqtl_universe_file    <- params$eqtl_universe
classification_file   <- params$classification
heterogeneity_file    <- params$heterogeneity_test
ocr_enrich_file        <- params$ocr_enrich
eqtl_effects_file      <- params$eqtl_effects
beta_correlation_file  <- params$beta_correlation
pi1_dir                <- params$pi1_dir
bmark_dir              <- params$bmark_dir

gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)

# Smaller-font DT wrapper for wide tables (Opposite effects, Heterogeneity
# results, Subcluster-specific gene table) so all columns fit on screen
# without horizontal scroll
small_dt <- function(df, page_length = 20) {
  datatable(
    df, filter = "top", rownames = FALSE, class = "compact stripe hover",
    options = list(
      pageLength = page_length, scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '75%'});",
        "}"
      )
    )
  )
}
```

```{r load-data}
eqtl_universe    <- read_rds(eqtl_universe_file)
classification   <- read_rds(classification_file)
het_test         <- read_rds(heterogeneity_file)
ocr_enrich       <- read_rds(ocr_enrich_file)
eqtl_effects     <- read_rds(eqtl_effects_file)
beta_correlation <- read_rds(beta_correlation_file)

# Unpack classification -- "any" tier only; cell_specific tier retired
# (superseded by the heterogeneity test, below)
any_table    <- classification$any_table
any_summary  <- classification$any_summary

# Collapse categories that aren't actionable for the reader into a single
# "untestable" display bucket -- display only, any_table's real categories
# (and therefore what OCR/heterogeneity actually consume) stay granular in
# B's own TSV output:
#   - untestable_genotyped_not_expressed / untestable_not_genotyped_not_expressed
#   - testable_pair_untested (gene expressed somewhere, SNP genotyped somewhere,
#     but the two were never tested TOGETHER in any single Jang cell type --
#     a genuine edge case, present in the underlying logic even before the
#     Fujita->Jang migration, just 0 pairs under Fujita vs 4 under Jang)
any_summary <- any_summary |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable",
    untestable_not_genotyped_not_expressed = "untestable",
    testable_pair_untested                 = "untestable"
  )) |>
  group_by(final_category) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(prop = round(n / sum(n), 3))

any_table <- any_table |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable",
    untestable_not_genotyped_not_expressed = "untestable",
    testable_pair_untested                 = "untestable"
  ))

# Unpack heterogeneity test (Cochran's Q on fetal-specific pairs). het_usable
# is the detail level (one row per fetal-adult comparison); het_heterogeneous
# is the aggregated, stringent pair-level call (heterogeneous in ALL tested
# adult comparisons).
het_usable        <- het_test$usable
het_heterogeneous <- het_test$heterogeneous_pairs
het_funnel        <- het_test$funnel

# Unpack subcluster/supercluster effect-size object
subcluster_tbl   <- eqtl_effects$subcluster_specific
supercluster_tbl <- eqtl_effects$supercluster_specific

# Full L1+L2+adult pairwise slope matrix
effect_matrix <- eqtl_effects$effect_size_matrix
effect_long   <- eqtl_effects$effect_size_long

# Unpack beta correlation object (Jang et al. 2026 -- real structure from
# devspec_beta_correlation.R)
paired_betas     <- beta_correlation$paired_betas
opposite_effects <- beta_correlation$opposite_effects
celltype_map     <- beta_correlation$celltype_map

# Add gene symbols (paired_betas/opposite_effects only carry ENSG in `gene`)
paired_betas <- paired_betas |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

opposite_effects <- opposite_effects |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

# Three-way point classification for the scatter grid: grey (not significant in
# both datasets), blue (significant in both, concordant), red (significant in
# both, discordant). Plotted on every overlapping pair regardless of category --
# only the colour/highlight is restricted to the both-significant subset.
# significant_jang is already boolean (two-check design: is the gene's Jang
# lead SNP AND qval<0.05) -- not a "Yes"/"No" string like Fujita's flag was.
paired_betas <- paired_betas |>
  mutate(
    concordant = sign(beta_my) == sign(beta_jang),
    point_category = case_when(
      significant_jang & concordant  ~ "concordant_both_sig",
      significant_jang & !concordant ~ "discordant_both_sig",
      TRUE                            ~ "other"
    )
  )

# Per-comparison Pearson r (no cor_summary is written by the beta_correlation
# rule, so compute it here)
beta_cor_summary <- paired_betas |>
  group_by(my_cell_type, jang_cell_type) |>
  summarise(
    n = n(),
    r = cor(beta_my, beta_jang, use = "complete.obs"),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", jang_cell_type))

comparison_levels <- paste(celltype_map$my_cell, "vs", celltype_map$jang_cell)

# Significance-count summary table (one row per comparison): significant in
# fetal & present in adult; also significant in Jang; of those, concordant;
# of those, discordant.
significance_counts <- paired_betas |>
  group_by(my_cell_type, jang_cell_type) |>
  summarise(
    n_fetal_sig_present_adult = n(),
    n_also_sig_jang = sum(significant_jang, na.rm = TRUE),
    n_both_sig_concordant = sum(significant_jang & concordant, na.rm = TRUE),
    n_both_sig_discordant = sum(significant_jang & !concordant, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", jang_cell_type)) |>
  arrange(factor(comparison, levels = comparison_levels))
```

## Developmental specificity pipeline steps {.tabset}

***

1. `extract_eqtl_universe`: Extract unique fetal eGene-eSNP universe (L1+L2)
2. `classify_eqtl_specificity`: Classify pairs as testable/untestable and shared/fetal-specific vs Jang et al. 2026 ("any" tier, all L1+L2 Jang types pooled)
3. `heterogeneity_test`: Cochran's Q test on fetal-specific pairs vs all tested adult Jang cell types
4. `prep_ld_clump_input` / `run_ld_clump` / `parse_ld_clump`: LD-independence pruning of the eGene-eSNP universe
5. `ocr_enrich`: Test fetal-specific/shared/specific-heterogeneous eQTL enrichment in adult (Li et al.) cortical union OCRs
6. `eqtl_effect_sizes`: Extract subcluster/supercluster-specific eQTL (fetal L1 vs Jang L1 effect-size correlation)
7. `beta_correlation`: Fetal-adult eQTL beta correlations and opposite-effect allele checks vs Jang et al. 2026
8. `pi1_enrich_jang`: pi1 replication test between fetal L1 sn-eQTL and Jang et al. 2026 L1 eQTL
9. `dev_specificity_report`: Summarise developmental specificity results

***

### Classification of fetal eQTL {.tabset}

Classify fetal eQTL against Jang et al. 2026 (SingleBrain) adult eGene-eSNP pairs, 
pooled across all L1+L2 Jang cell types (35 types; MiGA3 excluded -- 
a separate microglia-only meta-analysis, not part of the core L1+L2 hierarchy).

- `testable_shared`: fetal eQTL is present in nom file, and top assoc file, AND `qval < 0.05` in latter (in >= 1 Jang cell type)
  - Feeds: OCR enrichment (as the "shared" SNP set)
- `testable_specific`: present in Jang nom file, but not sig. in any Jang cell type
  - Feeds: Heterogeneity test (next section), OCR enrichment (as the "fetal-specific" SNP set)

**Significance note**: Jang has no `significant_by_2step_FDR` column, only a qval
column in the top assoc. file (i.e. the perm file). 

This makes `testable_specific` more permissive here than under Fujita.

Both categories are LD-pruned (see LD-clumping steps) before being used in OCR enrichment.

*Note*: qval in top assoc. is derived from Random_P, beta and se in Jang derived from
fixed_P and fixed_SD.

***

#### Summary table

```{r any-tier-summary}
any_summary |>
  mutate(pct = percent(prop, accuracy = 0.1)) |>
  select(Category = final_category, Count = n, Proportion = pct) |>
  gt() |>
  tab_header(title = '"Any" tier classification')
```

#### Barplot

```{r any-tier-barplot, fig.height=5, fig.width=7}
any_summary |>
  mutate(
    final_category = fct_reorder(final_category, n),
    fill_group = case_when(
      str_detect(final_category, "shared")   ~ "Shared",
      str_detect(final_category, "specific") ~ "Fetal-specific",
      TRUE                                   ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = final_category, y = n, fill = fill_group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70")) +
  labs(x = NULL, y = "Number of unique (eGene, eSNP) pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

#### Testable-pairs pie

```{r any-tier-testable-pie, fig.height=4, fig.width=5}
any_table |>
  filter(str_detect(final_category, "^testable")) |>
  count(final_category) |>
  ggplot(aes(x = "", y = n, fill = final_category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("testable_shared" = "#4DAF4A", "testable_specific" = "#E41A1C")) +
  labs(fill = "Classification") +
  theme_void(base_size = 13)
```

***

### Heterogeneity test (fetal-specific eQTL) {.tabset}

- Cochran's Q: Tests whether the fetal and adult effect sizes differ by more
than combined sampling error predicts
- Each `testable_specific` pair is collapsed to a single fetal reference row
(whichever fetal cell type it's most significant in), then tested against
EVERY adult Jang cell type that pair was actually tested in 
- FDR (BH) applied once, pooled across all usable rows
- STRINGENT final call: a pair only counts as fetal-specific-and-heterogeneous
if heterogeneous in ALL of its tested adult comparisons, not just one
- **Allele harmonization**: No AF column in Jang, so
palindromic (A/T, C/G) SNPs cannot be disambiguated via frequency the way
they could against Fujita. Every palindromic pair is therefore uniformly `palindromic_unresolved` and excluded from the Q-test.
- Notes:
  - Q is not directional -- it tells you the betas differ, not which way
  - Pairs tested against more adult cell types face a harder bar under the
    "all" rule -- see the stringency table below

***

#### Summary

```{r het-test-funnel}
het_funnel |>
  gt() |>
  tab_header(title = "Population funnel")
```

```{r het-test-summary-stats, results = "asis"}
n_row_het    <- sum(het_usable$heterogeneous)
n_row_total  <- nrow(het_usable)
n_pair_het   <- sum(het_heterogeneous$heterogeneous_all)
n_pair_total <- nrow(het_heterogeneous)

cat(sprintf(
  paste0(
    "At the individual fetal-adult comparison level, %d of %d tested comparisons (%s) ",
    "were formally heterogeneous (FDR < 0.05). Applying the stringent rule -- heterogeneous ",
    "in **all** tested adult comparisons -- %d of %d testable_specific pairs (%s) pass."
  ),
  n_row_het, n_row_total, percent(n_row_het / n_row_total, accuracy = 0.1),
  n_pair_het, n_pair_total, percent(n_pair_het / n_pair_total, accuracy = 0.1)
))
```

```{r het-test-stringency-effect}
het_heterogeneous |>
  mutate(n_bucket = if_else(n_comparisons_tested >= 4, "4+", as.character(n_comparisons_tested))) |>
  group_by(n_bucket) |>
  summarise(N_Pairs = n(), N_Passed_All = sum(heterogeneous_all), .groups = "drop") |>
  mutate(Pass_Rate = percent(N_Passed_All / N_Pairs, accuracy = 0.1)) |>
  arrange(n_bucket) |>
  gt() |>
  tab_header(title = "Pass rate by number of adult comparisons tested",
             subtitle = "Pairs tested against more adult cell types face a harder bar under the 'all' rule")
```

***

#### Final pairs (stringent)

```{r het-test-final-pairs-table}
het_heterogeneous |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(
    ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
    Fetal_Cell_Type = fetal_cell_type,
    N_Adult_Comparisons_Tested = n_comparisons_tested,
    N_Heterogeneous = n_comparisons_heterogeneous,
    Heterogeneous_All = heterogeneous_all,
    Matched_Type_Tested = matched_type_tested,
    Confirmed_By_Matched_Type = confirmed_by_matched_type
  ) |>
  arrange(desc(Heterogeneous_All), desc(N_Adult_Comparisons_Tested)) |>
  small_dt()
```

***

#### Magnitude-consistent pairs (|fetal| > |adult|)

Restricts the `r n_pair_het` stringent (heterogeneous-in-all) pairs further: retains
only those where |fetal beta| > |adult beta| in EVERY tested comparison for
that pair -- not just heterogeneous, but the fetal effect strictly larger in
magnitude, regardless of sign. This sidesteps delta_beta's ambiguity when
fetal and adult betas have opposite signs (delta_beta can look "fetal-bigger"
even when adult's magnitude is actually larger). NOT yet fed to OCR
enrichment -- report-only for now.

```{r het-test-magnitude-prep}
stringent_magnitude <- het_usable |>
  semi_join(het_heterogeneous |> filter(heterogeneous_all),
            by = c("phenotype_id", "variant_id")) |>
  mutate(
    magnitude_diff = abs(beta_my) - abs(beta_jang_harmonized),
    fetal_bigger = magnitude_diff > 0
  )

pair_magnitude <- stringent_magnitude |>
  group_by(phenotype_id, variant_id) |>
  summarise(
    fetal_cell_type = first(cell_type),
    n_comparisons = n(),
    n_fetal_bigger = sum(fetal_bigger),
    magnitude_consistent = all(fetal_bigger),
    min_magnitude_diff = min(magnitude_diff),
    max_magnitude_diff = max(magnitude_diff),
    .groups = "drop"
  )
```

```{r het-test-magnitude-summary-stats, results = "asis"}
n_pass <- sum(pair_magnitude$magnitude_consistent)
n_total <- nrow(pair_magnitude)

cat(sprintf(
  "%d of %d stringent pairs (%s) also have |fetal beta| > |adult beta| in EVERY tested comparison.",
  n_pass, n_total, percent(n_pass / n_total, accuracy = 0.1)
))
```

```{r het-test-magnitude-summary-table}
pair_magnitude |>
  count(magnitude_consistent, name = "N_Pairs") |>
  mutate(Proportion = percent(N_Pairs / sum(N_Pairs), accuracy = 0.1)) |>
  gt() |>
  tab_header(title = "Magnitude consistency (|fetal beta| > |adult beta|) across a pair's comparisons")
```

```{r het-test-magnitude-histogram, fig.height=4, fig.width=7}
ggplot(stringent_magnitude, aes(x = magnitude_diff)) +
  geom_histogram(bins = 30, fill = "#4DAF4A", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "|fetal beta| - |adult beta| (per comparison)", y = "N comparisons") +
  theme_classic(base_size = 12)
```

```{r het-test-magnitude-final-table}
pair_magnitude |>
  filter(magnitude_consistent) |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
         Fetal_Cell_Type = fetal_cell_type, N_Comparisons = n_comparisons,
         Min_Magnitude_Diff = min_magnitude_diff, Max_Magnitude_Diff = max_magnitude_diff) |>
  arrange(Min_Magnitude_Diff) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  small_dt()
```

***

#### Adult-larger comparisons

```{r het-test-adult-bigger-prep}
adult_bigger_rows <- stringent_magnitude |>
  filter(magnitude_diff < 0) |>
  arrange(magnitude_diff)

target_pairs <- adult_bigger_rows |>
  distinct(phenotype_id, variant_id, cell_type)

target_rows <- stringent_magnitude |>
  semi_join(target_pairs, by = c("phenotype_id", "variant_id", "cell_type"))

side_by_side <- target_rows |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(phenotype_id, external_gene_name, variant_id, cell_type,
         beta_my, jang_cell_type, beta_jang_harmonized, magnitude_diff) |>
  group_by(phenotype_id, variant_id, cell_type) |>
  mutate(Fetal_Beta = first(beta_my)) |>
  ungroup() |>
  select(-beta_my) |>
  pivot_wider(
    id_cols = c(phenotype_id, external_gene_name, variant_id, cell_type, Fetal_Beta),
    names_from = jang_cell_type,
    values_from = c(beta_jang_harmonized, magnitude_diff),
    names_glue = "{jang_cell_type}_{.value}"
  ) |>
  rename(Gene = external_gene_name, ENSG = phenotype_id, eSNP = variant_id,
         Fetal_Cell_Type = cell_type)

mixed_flag <- target_rows |>
  distinct(phenotype_id, variant_id) |>
  left_join(
    stringent_magnitude |>
      group_by(phenotype_id, variant_id) |>
      summarise(all_adult_bigger = all(!fetal_bigger), .groups = "drop"),
    by = c("phenotype_id", "variant_id")
  ) |>
  mutate(pair_type = if_else(all_adult_bigger, "adult_bigger_everywhere",
                              "mixed -- fetal bigger in >=1 other comparison"))
```

```{r het-test-adult-bigger-summary-stats, results = "asis"}
cat(sprintf(
  "%d individual comparisons (across %d unique gene/SNP/fetal-cell-type triples) show |adult beta| > |fetal beta|.",
  nrow(adult_bigger_rows), nrow(target_pairs)
))
```

```{r het-test-adult-bigger-breakdown}
mixed_flag |>
  count(pair_type, name = "N_Triples") |>
  mutate(Proportion = percent(N_Triples / sum(N_Triples), accuracy = 0.1)) |>
  gt() |>
  tab_header(title = "Breakdown: adult-bigger-everywhere vs mixed")
```

```{r het-test-adult-bigger-table}
side_by_side |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Per-comparison detail

Every fetal-adult comparison underlying the table above -- for checking why
a specific pair passed or failed the "all" rule.

```{r het-test-detail-table}
het_usable |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(
    Fetal_Cell_Type = cell_type,
    ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
    Fetal_Beta = beta_my, Fetal_SE = se_my, Fetal_AF = af_my,
    Adult_Cell_Type = jang_cell_type, Biologically_Matched = biologically_matched,
    Adult_Beta = beta_jang_harmonized, Adult_SE = se_jang,
    Delta_Beta = delta_beta,
    Q, P = p, Q_value = q, Heterogeneous = heterogeneous
  ) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Effect-size concordance {.tabset}

Pearson correlation of eQTL betas between each fetal cell type and its
matched Jang et al. 2026 adult cell type. Points are highlighted only where
the pair is sig. in both datasets:

- Blue: sig. in both and concordant sign
- Red: sig. in both and discordant sign (also shown in opposite effects tab)
- Grey: sig. in fetal, not sig. in Jang

***

#### Significance counts

```{r beta-significance-counts}
significance_counts |>
  select(Comparison = comparison,
         `Fetal-sig, present in adult` = n_fetal_sig_present_adult,
         `Also Jang-sig` = n_also_sig_jang,
         `Both-sig, concordant` = n_both_sig_concordant,
         `Both-sig, discordant` = n_both_sig_discordant) |>
  gt() |>
  tab_header(title = "Significance counts per fetal-adult comparison")
```

***

#### Scatter grid

```{r beta-scatter-grid, fig.height=8, fig.width=14}
plot_df <- paired_betas |>
  mutate(
    comparison = factor(paste(my_cell_type, "vs", jang_cell_type), levels = comparison_levels)
  ) |>
  # Plot "other" (grey) points first so blue/red highlighted points draw on top
  arrange(point_category == "other")

r_labels <- beta_cor_summary |>
  mutate(comparison = factor(comparison, levels = comparison_levels),
         label = paste0("r = ", round(r, 3), "\nn = ", n))

ggplot(plot_df, aes(x = beta_my, y = beta_jang)) +
  geom_point(aes(colour = point_category, alpha = point_category != "other", size = point_category != "other")) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#377EB8", linewidth = 0.6) +
  geom_text(data = r_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  scale_colour_manual(values = c(concordant_both_sig = "#377EB8",
                                 discordant_both_sig = "#E41A1C",
                                 other = "grey70"),
                       labels = c(concordant_both_sig = "Concordant (sig. both)",
                                  discordant_both_sig = "Discordant (sig. both)",
                                  other = "Not sig. in both")) +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.25), guide = "none") +
  scale_size_manual(values = c(`TRUE` = 1.1, `FALSE` = 0.6), guide = "none") +
  facet_wrap(~ comparison, ncol = 4, scales = "free") +
  labs(x = "Fetal slope", y = "Adult (Jang) slope", colour = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

***

#### Correlation heatmap {.tabset}

```{r effect-size-heatmap-setup}
slope_cols <- grep("^slope_", names(effect_matrix), value = TRUE)
slope_mat <- as.matrix(effect_matrix[, slope_cols])
colnames(slope_mat) <- str_remove(colnames(slope_mat), "^slope_")

fetal_l1_order <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
jang_l1_order  <- c("Jang_Ext", "Jang_IN", "Jang_MG", "Jang_OPC", "Jang_End", "Jang_Ast", "Jang_OD")

full_order <- c(fetal_l1_order, jang_l1_order)
full_order <- full_order[full_order %in% colnames(slope_mat)]
slope_mat <- slope_mat[, full_order]

cor_mat <- cor(slope_mat, use = "pairwise.complete.obs")
col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))
```

##### Fetal vs Fetal

```{r effect-size-heatmap-fetal, fig.height=9, fig.width=9}
fetal_family_order <- c(
  "Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
  "Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2",
  "GABA", "GABA-0", "GABA-1", "GABA-2",
  "NPC", "NPC-0", "NPC-1", "NPC-2",
  "MG", "OPC", "Endo-Peri"
)

fetal_long <- effect_long |>
  filter(cell_type %in% fetal_family_order) |>
  select(phenotype_id, variant_id, cell_type, slope) |>
  distinct(phenotype_id, variant_id, cell_type, .keep_all = TRUE)

fetal_wide <- fetal_long |>
  pivot_wider(names_from = cell_type, values_from = slope)

fetal_slope_mat <- as.matrix(fetal_wide[, setdiff(names(fetal_wide), c("phenotype_id", "variant_id"))])
fetal_family_order_present <- fetal_family_order[fetal_family_order %in% colnames(fetal_slope_mat)]
fetal_slope_mat <- fetal_slope_mat[, fetal_family_order_present]

cor_mat_fetal <- cor(fetal_slope_mat, use = "pairwise.complete.obs")

Heatmap(
  cor_mat_fetal,
  name = "Pearson r",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_fetal[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Fetal vs Jang

```{r effect-size-heatmap-fetal-jang, fig.height=6, fig.width=7}
fetal_cols <- fetal_l1_order[fetal_l1_order %in% colnames(cor_mat)]
jang_cols <- jang_l1_order[jang_l1_order %in% colnames(cor_mat)]
cor_mat_cross <- cor_mat[fetal_cols, jang_cols]
colnames(cor_mat_cross) <- str_remove(colnames(cor_mat_cross), "^Jang_")

Heatmap(
  cor_mat_cross,
  name = "Pearson r",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_cross[i, j]), x, y, gp = gpar(fontsize = 9))
  },
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10)
)
```

***

#### Pairwise n

```{r pairwise-n, fig.height=7, fig.width=8}
n_mat <- matrix(NA, ncol(slope_mat), ncol(slope_mat),
                dimnames = list(colnames(slope_mat), colnames(slope_mat)))
for (i in seq_len(ncol(slope_mat))) {
  for (j in seq_len(ncol(slope_mat))) {
    n_mat[i, j] <- sum(complete.cases(slope_mat[, c(i, j)]))
  }
}

stage_anno <- ifelse(str_starts(colnames(n_mat), "Jang_"), "Adult", "Fetal")
ha <- HeatmapAnnotation(
  Stage = stage_anno,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

col_fun_n <- colorRamp2(c(0, max(n_mat, na.rm = TRUE)), c("white", "#08519C"))

Heatmap(
  n_mat,
  name = "n pairs",
  col = col_fun_n,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = ha,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(n_mat[i, j], x, y, gp = gpar(fontsize = 7))
  },
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9)
)
```

***

#### Full slope table

Betas for every (eGene, eSNP) pair across the 7 fetal L1 cell types and 7
Jang adult L1-pooled cell types.

```{r full-slope-table}
slope_full_tbl <- effect_matrix |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(external_gene_name, phenotype_id, variant_id,
         all_of(paste0("slope_", full_order))) |>
  rename(Symbol = external_gene_name, ENSG_ID = phenotype_id, SNP = variant_id) |>
  rename_with(~ str_remove(., "^slope_"), starts_with("slope_"))

small_dt(slope_full_tbl)
```

***

#### Opposite effects

- Sig. eQTL pairs in both fetal and Jang datasets with discordant effect
sign (the red points in the scatter grid).
- Fetal/Jang columns show `REF/ALT` for each dataset. No 
AF column in Jang, so  `Allele_Status` is REF/ALT string comparison only. 
Palindromic pairs (A/T, C/G) cannot be resolved so were excluded upstream (see beta_correlation.R).

```{r opposite-effects-table}
opposite_effects |>
  mutate(
    Fetal = paste0(ref_my, "/", alt_my),
    Jang  = paste0(ref_jang, "/", alt_jang)
  ) |>
  select(Gene = external_gene_name, ENSG = gene, eSNP = snp,
         Fetal_Cell = my_cell_type, Jang_Cell = jang_cell_type,
         Beta_Fetal = beta_my, Beta_Jang = beta_jang,
         Fetal, Jang,
         Allele_Status = allele_status) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Jang pi1 enrichment {.tabset}

pi1 replication test (Storey & Tibshirani): what proportion of the tested
eQTL set shows non-null evidence in the comparison dataset, estimated from
the p-value distribution rather than a fixed significance threshold.

- Forward: pi1 of fetal-sig eQTL in Jang's nominal (full) data
- Reverse: pi1 of Jang-sig eQTL in fetal nominal data
- Forward and reverse are not symmetric: Jang (a 4-cohort meta-analysis) is
much better powered than any single fetal cell type, so forward pi1 tends
to run higher than reverse -- an asymmetric pair here reflects a power
difference between datasets, not necessarily a biological asymmetry

***

```{r pi1-jang-setup}
# Shared setup for both the Table and Heatmap subtabs below -- defined here,
# before either, since tabs render in document order regardless of heading
# nesting (same issue as the effect-size correlation heatmap's col_order).
pi1_fetal_l1 <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
pi1_jang_l1  <- c("Ext", "IN", "MG", "OPC", "End", "Ast", "OD")

pi1_combinations <- expand.grid(cell_type = pi1_fetal_l1, jang_cell_type = pi1_jang_l1,
                                 stringsAsFactors = FALSE)
```

#### Table

```{r pi1-jang-enrich-table-prep}
read_pi1_jang_enrich <- function(ct, jang_ct, pi1_dir) {
  file_path <- paste0(pi1_dir, ct, "_vs_", jang_ct, "_pi1_enrich_tbl.rds")
  if (file.exists(file_path)) {
    read_rds(file_path)
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

pi1_jang_enrich_tbl <- map2_dfr(pi1_combinations$cell_type, pi1_combinations$jang_cell_type,
                                 ~ read_pi1_jang_enrich(.x, .y, pi1_dir))
```

```{r pi1-jang-enrich-table}
pi1_jang_enrich_tbl |>
  mutate(
    pi1 = ifelse(is.na(pi1), NA, round(as.numeric(pi1), 2)),
    prop_replicating = ifelse(is.na(prop_replicating), NA, round(as.numeric(prop_replicating), 2)),
    prop_same_direction = ifelse(is.na(prop_same_direction), NA, round(as.numeric(prop_same_direction), 2)),
    pi1_cell_ref = ifelse(is.na(pi1_cell_ref), NA, round(as.numeric(pi1_cell_ref), 2)),
    prop_replicating_cell_ref = ifelse(is.na(prop_replicating_cell_ref), NA, round(as.numeric(prop_replicating_cell_ref), 2)),
    prop_same_direction_cell_ref = ifelse(is.na(prop_same_direction_cell_ref), NA, round(as.numeric(prop_same_direction_cell_ref), 2))
  ) |>
  gt() |>
  tab_header(
    title = "pi1 Enrichment Summary: fetal L1 vs Jang et al. 2026 L1"
  ) |>
  cols_label(
    cell_type = "Cell Type",
    ref_cell_type = "Reference Cell Type",
    pi1 = "pi1",
    prop_replicating = "Prop. Replicating",
    prop_same_direction = "Prop. Same Dir.",
    overlap_count = "Overlap Count",
    pi1_cell_ref = "pi1",
    prop_replicating_cell_ref = "Prop. Replicating",
    prop_same_direction_cell_ref = "Prop. Same Dir.",
    overlap_count_cell_ref = "Overlap Count"
  ) |>
  tab_spanner(
    label = "Forward Test",
    columns = c(pi1, prop_replicating, prop_same_direction, overlap_count)
  ) |>
  tab_spanner(
    label = "Reference Test",
    columns = c(pi1_cell_ref, prop_replicating_cell_ref, prop_same_direction_cell_ref, overlap_count_cell_ref)
  ) |>
  fmt_missing(
    columns = everything(),
    missing_text = "NA"
  ) |>
  tab_options(
    table.font.size = 12,
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.border.top.width = px(2),
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black"
  )
```

***

#### Heatmap {.tabset}

```{r pi1-jang-heatmap-prep}
read_pi1_jang_results <- function(ct, jang_ct, pi1_dir) {
  file_path <- paste0(pi1_dir, ct, "_vs_", jang_ct, "_pi1_results_tbl.rds")
  if (file.exists(file_path)) {
    pi1_list <- read_rds(file_path)
    tibble(
      cell_type = ct,
      jang_cell_type = jang_ct,
      pi1_forward = pi1_list$forward$pi1,
      pi1_reverse = pi1_list$reverse$pi1
    )
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

pi1_jang_tbl <- map2_dfr(pi1_combinations$cell_type, pi1_combinations$jang_cell_type,
                          ~ read_pi1_jang_results(.x, .y, pi1_dir))

pi1_jang_long <- pi1_jang_tbl |>
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") |>
  mutate(
    direction = recode(direction, pi1_forward = "Forward", pi1_reverse = "Reverse"),
    cell_type = factor(cell_type, levels = pi1_fetal_l1),
    jang_cell_type = factor(jang_cell_type, levels = pi1_jang_l1)
  )

plot_pi1_jang_heatmap <- function(df) {
  ggplot(df, aes(x = jang_cell_type, y = cell_type, fill = pi1)) +
    geom_tile(color = "black", lwd = 1.1, linetype = 1) +
    geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
              color = "black", size = 3.5) +
    scale_fill_gradientn(
      colours = c("white", "yellow", "red"),
      limits = c(0, 1.0),
      na.value = "grey80",
      name = expression(pi[1])
    ) +
    coord_equal() +
    labs(x = "Cell type (Jang Ref)", y = "Cell type (Fetal Query)") +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank()
    )
}
```

##### Forward

```{r pi1-jang-heatmap-forward, fig.dim=c(6,6)}
plot_pi1_jang_heatmap(pi1_jang_long |> filter(direction == "Forward"))
```

##### Reverse

```{r pi1-jang-heatmap-reverse, fig.dim=c(6,6)}
plot_pi1_jang_heatmap(pi1_jang_long |> filter(direction == "Reverse"))
```

***

#### Summary table

```{r pi1-jang-summary-table}
pi1_jang_tbl |>
  mutate(across(c(pi1_forward, pi1_reverse), ~ round(.x, 2))) |>
  select(Fetal_Cell_Type = cell_type, Jang_Cell_Type = jang_cell_type,
         Pi1_Forward = pi1_forward, Pi1_Reverse = pi1_reverse) |>
  gt() |>
  tab_header(title = "pi1 enrichment: fetal L1 vs Jang et al. 2026 L1")
```

***

### OCR enrichment {.tabset}

- LD-pruned fetal-specific, shared, and specific-heterogeneous eQTL SNPs
tested for enrichment in adult cortical OCRs (Li et al. 2023)
- Cortical adult union peak set manually curated

***

#### Fold enrichment

```{r ocr-enrichment-barplot, fig.height=4, fig.width=7}
ocr_enrich |>
  mutate(
    sig_label = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns"),
    snp_set = factor(
      recode(snp_set,
             fetal_specific = "Fetal-specific",
             shared = "Shared",
             specific_heterogeneous = "Specific + heterogeneous"),
      levels = c("Shared", "Fetal-specific", "Specific + heterogeneous")
    )
  ) |>
  ggplot(aes(x = snp_set, y = fold_enrichment, fill = snp_set)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_text(aes(label = sig_label), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Fetal-specific" = "#E41A1C", "Shared" = "#4DAF4A",
                                "Specific + heterogeneous" = "#984EA3")) +
  labs(x = NULL, y = "Fold enrichment (adult cortical union)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

***

#### Results table

```{r ocr-enrichment-table}
ocr_enrich |>
  mutate(p_value = signif(p_value, 3),
         fold_enrichment = round(fold_enrichment, 2),
         mean_permuted_overlap = round(mean_permuted_overlap, 1)) |>
  select(Peak_Set = peak_set, SNP_Set = snp_set,
         N_SNPs = n_snps, N_Peaks = n_peaks,
         Observed = observed_overlap, Mean_Permuted = mean_permuted_overlap,
         Fold_Enrichment = fold_enrichment, P_Value = p_value) |>
  gt() |>
  tab_header(title = "OCR enrichment results") |>
  tab_options(table.font.size = px(11))
```

***

### Subcluster-specific eQTL {.tabset}

- FDR sig. eGenes in an L2 subcluster but not its parent L1
("subcluster-specific"), and vice versa ("supercluster-specific").

***

#### Counts

```{r subcluster-counts}
subcluster_tbl |>
  count(cell_type_L2, parent_L1, name = "n_specific_eGenes") |>
  gt()
```

***

#### Supercluster-specific counts

```{r supercluster-counts}
supercluster_tbl |>
  count(parent_L1, name = "n_L1_only_eGenes") |>
  gt()
```

***

#### Subcluster-specific gene table

```{r subcluster-gene-table}
subcluster_tbl |>
  select(Gene = gene_symbol, ENSG = phenotype_id, eSNP = variant_id,
         cell_type_L1 = parent_L1, Slope_L1 = slope_parent, SE_L1 = se_parent, Pval_L1 = pval_nom_parent,
         cell_type_L2 = cell_type_L2, Slope_L2 = slope_L2, SE_L2 = se_L2, Pval_L2 = pval_nom_L2) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Sibling comparison table

```{r subcluster-sibling-table-prep}
family_order <- list(
  "Glu-UL" = c("Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2"),
  "Glu-DL" = c("Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2"),
  "GABA"   = c("GABA", "GABA-0", "GABA-1", "GABA-2"),
  "NPC"    = c("NPC", "NPC-0", "NPC-1", "NPC-2")
)

build_sibling_table <- function(subcluster_tbl, effect_long, gene_lookup) {

  fam_lookup <- bind_rows(lapply(names(family_order), function(f) {
    tibble(parent_L1 = f, cell_type_L2 = family_order[[f]][-1])
  }))

  base <- subcluster_tbl |>
    select(phenotype_id, variant_id, gene_symbol, parent_L1,
           sig_cell_type = cell_type_L2,
           Slope_Sig = slope_L2, SE_Sig = se_L2, Pval_Sig = pval_nom_L2,
           Slope_L1 = slope_parent, SE_L1 = se_parent, Pval_L1 = pval_nom_parent)

  sibling_map <- fam_lookup |>
    inner_join(base |> distinct(phenotype_id, variant_id, parent_L1, sig_cell_type),
               by = "parent_L1", relationship = "many-to-many") |>
    filter(cell_type_L2 != sig_cell_type) |>
    group_by(phenotype_id, variant_id) |>
    mutate(sib_slot = paste0("Sib", row_number())) |>
    ungroup()

  sib_stats <- sibling_map |>
    left_join(
      effect_long |> select(phenotype_id, variant_id, cell_type,
                             Slope = slope, SE = slope_se, Pval = pval_nominal),
      by = c("phenotype_id", "variant_id", "cell_type_L2" = "cell_type")
    )

  sib_wide <- sib_stats |>
    select(phenotype_id, variant_id, sib_slot, cell_type_L2, Slope, SE, Pval) |>
    pivot_wider(
      names_from = sib_slot,
      values_from = c(cell_type_L2, Slope, SE, Pval),
      names_glue = "{sib_slot}_{.value}"
    )

  base |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    left_join(sib_wide, by = c("phenotype_id", "variant_id"))
}

sibling_tbl <- build_sibling_table(subcluster_tbl, effect_long, gene_lookup)
```

```{r subcluster-sibling-table}
sibling_tbl |>
  select(
    Gene = external_gene_name, ENSG = phenotype_id, eSNP = variant_id,
    cell_type_L2a = sig_cell_type, Slope_L2a = Slope_Sig, SE_L2a = SE_Sig, Pval_L2a = Pval_Sig,
    cell_type_L1 = parent_L1, Slope_L1, SE_L1, Pval_L1,
    cell_type_L2b = Sib1_cell_type_L2, Slope_L2b = Sib1_Slope, SE_L2b = Sib1_SE, Pval_L2b = Sib1_Pval,
    cell_type_L2c = Sib2_cell_type_L2, Slope_L2c = Sib2_Slope, SE_L2c = Sib2_SE, Pval_L2c = Sib2_Pval
  ) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Family heatmaps {.tabset}

For each L1 family with L2 subtypes, betas across the parent and its 3
children for every subcluster-specific (gene, SNP) pair in that family.
Black border marks the cell where the pair was actually classified
subcluster-specific.

```{r subcluster-family-heatmap-prep}
build_family_matrix <- function(family, subcluster_tbl, effect_long, gene_lookup) {
  members <- family_order[[family]]

  fam_pairs <- subcluster_tbl |>
    filter(parent_L1 == family) |>
    distinct(phenotype_id, variant_id, cell_type_L2) |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    mutate(pair_label = paste0(coalesce(external_gene_name, phenotype_id), " (", variant_id, ")"))

  if (nrow(fam_pairs) == 0) return(NULL)

  slopes <- effect_long |>
    filter(cell_type %in% members,
           phenotype_id %in% fam_pairs$phenotype_id,
           variant_id %in% fam_pairs$variant_id) |>
    inner_join(fam_pairs |> select(phenotype_id, variant_id, pair_label),
               by = c("phenotype_id", "variant_id"))

  slope_wide <- slopes |>
    select(pair_label, cell_type, slope) |>
    distinct(pair_label, cell_type, .keep_all = TRUE) |>
    pivot_wider(names_from = cell_type, values_from = slope)

  for (m in members) if (!(m %in% names(slope_wide))) slope_wide[[m]] <- NA_real_

  mat <- as.matrix(slope_wide[, members])
  rownames(mat) <- slope_wide$pair_label

  sig_mat <- matrix(FALSE, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  for (i in seq_len(nrow(fam_pairs))) {
    pl <- fam_pairs$pair_label[i]
    sig_ct <- fam_pairs$cell_type_L2[i]
    if (pl %in% rownames(sig_mat) && sig_ct %in% colnames(sig_mat)) {
      sig_mat[pl, sig_ct] <- TRUE
    }
  }

  list(mat = mat, sig_mat = sig_mat)
}

col_fun_family <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

draw_family_heatmap <- function(family, row_height_in = NULL) {
  fam_data <- build_family_matrix(family, subcluster_tbl, effect_long, gene_lookup)
  if (is.null(fam_data)) {
    cat("No subcluster-specific genes for", family, ".\n")
    return(invisible(NULL))
  }

  ht_args <- list(
    matrix = fam_data$mat,
    name = "Slope",
    col = col_fun_family,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    na_col = "grey90",
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 10),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (fam_data$sig_mat[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", lwd = 2, fill = NA))
      }
    }
  )

  if (!is.null(row_height_in)) {
    ht_args$height <- unit(row_height_in, "in") * nrow(fam_data$mat)
  }

  do.call(Heatmap, ht_args)

}
```

##### Glu-UL

```{r subcluster-family-glu-ul, fig.height=max(8, 0.18 * nrow(build_family_matrix("Glu-UL", subcluster_tbl, effect_long, gene_lookup)$mat)), fig.width=6, results='asis'}
print(draw_family_heatmap("Glu-UL", row_height_in = 0.18))
```

##### Glu-DL

```{r subcluster-family-glu-dl, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("Glu-DL"))
```

##### GABA

```{r subcluster-family-gaba, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("GABA"))
```

##### NPC

```{r subcluster-family-npc, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("NPC"))
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
files <- list.files(bmark_dir, pattern = "dev_specificity\\..*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]
  # "dev_specificity.extract_universe.txt" -> parts[2] = "extract_universe"
  rule <- parts[2]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

summary <- benchmarks |>
  group_by(rule) |>
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) |>
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  arrange(rule)

summary |>
  gt() |>
  tab_header(
    title = "Benchmarking",
    subtitle = "Means and SD across replicates"
  ) |>
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) |>
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) |>
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

```{r runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```

```{r session-info}
sessionInfo()
```
