---
title: "Developmental Specificity of Fetal Brain eQTL"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
params:
  eqtl_universe: null
  classification: null
  ocr_enrich: null
  eqtl_effects: null
  beta_correlation: null
  perm_dir: null
  gene_lookup: null
  bmark_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(gt)
  library(DT)
  library(data.table)
  library(tidyverse)
})

# --- Params, not snakemake@input: this Rmd is rendered via
# rmarkdown::render(params = list(...)), not Snakemake's script: directive ---
eqtl_universe_file    <- params$eqtl_universe
classification_file   <- params$classification
ocr_enrich_file        <- params$ocr_enrich
eqtl_effects_file      <- params$eqtl_effects
beta_correlation_file  <- params$beta_correlation
perm_dir               <- params$perm_dir
bmark_dir              <- params$bmark_dir

gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)

# Smaller-font DT wrapper for wide tables (Opposite effects, Subcluster-
# specific gene table) so all columns fit on screen without horizontal scroll
small_dt <- function(df, page_length = 20) {
  datatable(
    df, filter = "top", rownames = FALSE, class = "compact stripe hover",
    options = list(
      pageLength = page_length, scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '75%'});",
        "}"
      )
    )
  )
}
```

```{r load-data}
eqtl_universe    <- read_rds(eqtl_universe_file)
classification   <- read_rds(classification_file)
ocr_enrich       <- read_rds(ocr_enrich_file)
eqtl_effects     <- read_rds(eqtl_effects_file)
beta_correlation <- read_rds(beta_correlation_file)

# Unpack classification
any_table    <- classification$any_table
any_summary  <- classification$any_summary
cs_table     <- classification$cell_specific_table
cs_summary   <- classification$cell_specific_summary

# Unpack subcluster/supercluster effect-size object
subcluster_tbl   <- eqtl_effects$subcluster_specific
supercluster_tbl <- eqtl_effects$supercluster_specific
de_markers       <- eqtl_effects$de_markers_raw

# Full L1+L2+adult pairwise slope matrix (reinstated from the prior report
# version). ASSUMED still present under these two names in eqtl_effects --
# please confirm eqtl_effect_sizes.R still writes effect_size_matrix /
# effect_size_long alongside the subcluster/supercluster tables.
effect_matrix <- eqtl_effects$effect_size_matrix
effect_long   <- eqtl_effects$effect_size_long

# Unpack beta correlation object (real structure from devspec_beta_correlation.R)
paired_betas     <- beta_correlation$paired_betas
opposite_effects <- beta_correlation$opposite_effects
celltype_map     <- beta_correlation$celltype_map

# Add gene symbols (paired_betas/opposite_effects only carry ENSG in `gene`)
paired_betas <- paired_betas |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

opposite_effects <- opposite_effects |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

# Per-comparison Pearson r (no cor_summary is written by the beta_correlation
# rule, so compute it here)
beta_cor_summary <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n = n(),
    r = cor(beta_my, beta_fugita, use = "complete.obs"),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type))

comparison_levels <- paste(celltype_map$my_cell, "vs", celltype_map$fugita_cell)

# exp_pc_map for all 19 cell types (L1 + L2), hardcoded per instruction
exp_pc_map <- c(
  "Glu-UL" = 50, "Glu-DL" = 40, "NPC" = 30, "GABA" = 30,
  "Endo-Peri" = 30, "OPC" = 30, "MG" = 30,
  "Glu-UL-0" = 30, "Glu-UL-1" = 50, "Glu-UL-2" = 40,
  "Glu-DL-0" = 40, "Glu-DL-1" = 40, "Glu-DL-2" = 30,
  "GABA-0" = 30, "GABA-1" = 40, "GABA-2" = 40,
  "NPC-0" = 50, "NPC-1" = 30, "NPC-2" = 20
)
geno_pc     <- 4
norm_method <- "quantile"

# GO gene universe = every phenotype_id tested in that cell type's own perm
# file (unfiltered), matching the same path pattern used in
# devspec_beta_correlation.R
read_perm_universe <- function(cell_type, perm_dir) {
  expPC <- exp_pc_map[[cell_type]]
  f <- paste0(perm_dir, cell_type, "_", norm_method, "_genPC_", geno_pc,
              "_expPC_", expPC, "/", cell_type, "_", norm_method, "_perm.cis_qtl.txt.gz")
  if (!file.exists(f)) {
    warning(paste("perm file not found for", cell_type, ":", f))
    return(character(0))
  }
  fread(f, select = "phenotype_id") |> pull(phenotype_id) |> unique()
}
```

## Developmental specificity pipeline steps {.tabset}

***

1. `extract_eqtl_universe`: Extract unique fetal eGene-eSNP universe (L1+L2)
2. `classify_eqtl_specificity`: Classify pairs as testable/untestable and shared/fetal-specific vs Fujita
3. `ocr_enrich`: Test fetal-specific eQTL enrichment in fetal (Ziffra) and adult (Li et al.) OCRs
4. `eqtl_effect_sizes`: Extract subcluster/supercluster-specific eQTL and DE marker crossref
5. `beta_correlation`: Fetal-adult eQTL beta correlations and opposite-effect allele checks
6. `dev_specificity_report`: Summarise developmental specificity results

***

### Classification of fetal eQTL {.tabset}

- 6,844 unique (eGene, eSNP) pairs identified across all 19 fetal cell types
- Test for overlap in 7 Fugita adult cell types
- Categories:
  - `testable_shared`: gene expressed & SNP genotyped in Fujita; the pair is sig. in at least one adult cell type.
  - `testable_specific`: gene expressed & SNP genotyped in Fujita; the pair is NOT sig. in any adult cell type.
  - `untestable_genotyped_not_expressed`: SNP genotyped, gene not expressed in any Fujita cell type
  - `untestable_not_genotyped_expressed`: gene expressed, SNP not genotyped in Fujita data
  - `untestable_not_genotyped_not_expressed`: neither gene nor SNP is present in Fujita.
- Tiers:
  - `Any`: Unique fetal eGene-eSNP pair 'shared' if significant in at least one Fugita cell type.
  - `Cell-specifc`: Unique cell-specific fetal eGene-eSNP pairs tested against only label-matched Fujita cell type

***

#### "Any" tier {.tabset}

##### Summary table

```{r any-tier-summary}
any_summary |>
  mutate(pct = percent(prop, accuracy = 0.1)) |>
  select(Category = final_category, Count = n, Proportion = pct) |>
  gt() |>
  tab_header(title = '"Any" tier classification')
```

##### Barplot

```{r any-tier-barplot, fig.height=5, fig.width=7}
any_summary |>
  mutate(
    final_category = fct_reorder(final_category, n),
    fill_group = case_when(
      str_detect(final_category, "shared")   ~ "Shared",
      str_detect(final_category, "specific") ~ "Fetal-specific",
      TRUE                                   ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = final_category, y = n, fill = fill_group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70")) +
  labs(x = NULL, y = "Number of unique (eGene, eSNP) pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

##### Testable-pairs pie

```{r any-tier-testable-pie, fig.height=4, fig.width=5}
any_table |>
  filter(str_detect(final_category, "^testable")) |>
  count(final_category) |>
  ggplot(aes(x = "", y = n, fill = final_category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("testable_shared" = "#4DAF4A", "testable_specific" = "#E41A1C")) +
  labs(fill = "Classification") +
  theme_void(base_size = 13)
```

***

#### Cell-specific tier {.tabset}

##### Per-cell-type table

```{r cs-tier-summary}
cs_summary |>
  filter(!is.na(ref_cell_type)) |>
  select(Cell_Type = cell_type, Fujita_Match = ref_cell_type,
         Category = final_category, Count = n, Proportion = prop) |>
  datatable(filter = "top", rownames = FALSE,
            options = list(pageLength = 20, scrollX = TRUE))
```

##### Barplot

```{r cs-tier-barplot, fig.height=8, fig.width=10}
cs_summary |>
  filter(!is.na(ref_cell_type)) |>
  mutate(
    fill_group = case_when(
      str_detect(final_category, "shared")    ~ "Shared",
      str_detect(final_category, "specific")  ~ "Fetal-specific",
      final_category == "no_adult_equivalent" ~ "No adult equivalent",
      TRUE                                    ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = cell_type, y = n, fill = fill_group)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70", "No adult equivalent" = "grey40")) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Proportion of eGene-eSNP pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

### OCR enrichment {.tabset}

***

- Fetal-specific and shared eQTL SNPs tested for enrichment in fetal (Ziffra
et al. 2021) and adult (Li et al. 2023) open chromatin regions via block
bootstrap permutation.

For the adult data, the following peaks were mapped to fetal-like cell types:

```{r li-to-l1-mapping, eval=FALSE}
li_to_L1 <- list(
  "Glu-UL" = c("ITL23_1", "ITL23_2", "ITL23_3", "ITL23_4", "ITL23_5", "ITL23_6",
               "ITL34", "ITL4_1", "ITL4_2", "ITL45_1", "ITL45_2",
               "ITV1C_1", "ITV1C_2", "ITV1C_3"),
  "Glu-DL" = c("ITL5_1", "ITL5_2", "ITL5_3", "ITL5_4",
               "ITL6_1_1", "ITL6_1_2", "ITL6_2_1", "ITL6_2_2",
               "CT_1", "CT_2", "L6B_1", "L6B_2", "NP_1", "NP_2", "NP_3", "ET"),
  "GABA"   = c("SST_1", "SST_2", "SST_3", "SST_4", "SST_5", "SST_CHODL",
               "PVALB_1", "PVALB_2", "PVALB_3", "PVALB_4", "PV_ChCs",
               "VIP_1", "VIP_2", "VIP_3", "VIP_4", "VIP_5", "VIP_6", "VIP_7",
               "SNCG_1", "SNCG_2", "SNCG_3", "SNCG_4", "SNCG_5",
               "LAMP5", "LAMP5_LHX6"),
  "MG"     = c("MGC_1", "MGC_2"),
  "OPC"    = c("OPC", "COP"),
  "Endo-Peri" = c("EC", "SMC")
)
```

***

#### Fold enrichment

***

```{r ocr-enrichment-barplot, fig.height=7, fig.width=10}
ocr_enrich |>
  mutate(
    peak_set = str_remove(peak_set, "^(fetal_|adult_)"),
    sig_label = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns")
  ) |>
  ggplot(aes(x = reorder(peak_set, fold_enrichment), y = fold_enrichment, fill = stage)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_text(aes(label = sig_label),
            position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  facet_wrap(~ snp_set, ncol = 1, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = c(fetal = "#984EA3", adult = "#FF7F00")) +
  labs(x = NULL, y = "Fold enrichment (observed / mean permuted)", fill = "Peak stage") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

#### Results table

```{r ocr-enrichment-table}
ocr_enrich |>
  mutate(p_value = signif(p_value, 3),
         fold_enrichment = round(fold_enrichment, 2),
         mean_permuted_overlap = round(mean_permuted_overlap, 1)) |>
  select(Peak_Set = peak_set, Stage = stage, SNP_Set = snp_set,
         N_SNPs = n_snps, N_Peaks = n_peaks,
         Observed = observed_overlap, Mean_Permuted = mean_permuted_overlap,
         Fold_Enrichment = fold_enrichment, P_Value = p_value) |>
  gt() |>
  tab_header(title = "OCR enrichment results") |>
  tab_options(table.font.size = px(11))
```

***

### Effect-size concordance {.tabset}

***

- Pearson correlation of eQTL slopes between each fetal cell type and its
matched Fujita adult cell type
- Opposite-effect pairs - defined as having a discordant sign, highlighted
- Allele-coding status against Fujita's REF/ALT and 1000G EUR annotated.

***

#### Scatter grid

***

```{r beta-scatter-grid, fig.height=8, fig.width=14}
plot_df <- paired_betas |>
  mutate(
    comparison = factor(paste(my_cell_type, "vs", fugita_cell_type), levels = comparison_levels),
    opposite_effect = sign(beta_my) != sign(beta_fugita)
  )

r_labels <- beta_cor_summary |>
  mutate(comparison = factor(comparison, levels = comparison_levels),
         label = paste0("r = ", round(r, 3), "\nn = ", n))

ggplot(plot_df, aes(x = beta_my, y = beta_fugita)) +
  geom_point(aes(colour = opposite_effect), alpha = 0.4, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#377EB8", linewidth = 0.6) +
  geom_text(data = r_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  scale_colour_manual(values = c(`TRUE` = "#E41A1C", `FALSE` = "grey60"),
                       labels = c(`TRUE` = "Opposite effect", `FALSE` = "Concordant")) +
  facet_wrap(~ comparison, ncol = 4, scales = "free") +
  labs(x = "Fetal slope", y = "Adult (Fujita) slope", colour = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

***

#### Correlation heatmap

***

```{r effect-size-heatmap, fig.height=8, fig.width=9}
slope_cols <- grep("^slope_", names(effect_matrix), value = TRUE)
slope_mat <- as.matrix(effect_matrix[, slope_cols])
colnames(slope_mat) <- str_remove(colnames(slope_mat), "^slope_")

fetal_L1_cols <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
fetal_L2_cols <- sort(setdiff(
  colnames(slope_mat)[!str_starts(colnames(slope_mat), "Fujita_")],
  fetal_L1_cols
))
fujita_cols <- sort(colnames(slope_mat)[str_starts(colnames(slope_mat), "Fujita_")])

col_order <- c(fetal_L1_cols, fetal_L2_cols, fujita_cols)
col_order <- col_order[col_order %in% colnames(slope_mat)]
slope_mat <- slope_mat[, col_order]

cor_mat <- cor(slope_mat, use = "pairwise.complete.obs")

stage_anno <- ifelse(str_starts(colnames(cor_mat), "Fujita_"), "Adult", "Fetal")
ha <- HeatmapAnnotation(
  Stage = stage_anno,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

Heatmap(
  cor_mat,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)

# Save to file
slope_full_tbl <- effect_matrix |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(external_gene_name, phenotype_id, variant_id,
         all_of(paste0("slope_", col_order))) |>
  rename(Symbol = external_gene_name, ENSG_ID = phenotype_id, SNP = variant_id) |>
  rename_with(~ str_remove(., "^slope_"), starts_with("slope_")) |>
  distinct(ENSG_ID, SNP, .keep_all = TRUE)

write_tsv(slope_full_tbl, "../reports/19DEV-SPECIFICITY/slope_full_table.tsv")

```

***

#### Pairwise n

***

```{r pairwise-n, fig.height=8, fig.width=9}
n_mat <- matrix(NA, ncol(slope_mat), ncol(slope_mat),
                dimnames = list(colnames(slope_mat), colnames(slope_mat)))
for (i in seq_len(ncol(slope_mat))) {
  for (j in seq_len(ncol(slope_mat))) {
    n_mat[i, j] <- sum(complete.cases(slope_mat[, c(i, j)]))
  }
}

col_fun_n <- colorRamp2(c(0, max(n_mat, na.rm = TRUE)), c("white", "#08519C"))

Heatmap(
  n_mat,
  name = "n pairs",
  col = col_fun_n,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = ha,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(n_mat[i, j], x, y, gp = gpar(fontsize = 5))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

***

#### Opposite effects

***

```{r opposite-effects-table}
opposite_effects |>
  select(Gene = external_gene_name, ENSG = gene, eSNP = snp,
         Fetal_Cell = my_cell_type, Fujita_Cell = fugita_cell_type,
         Beta_Fetal = beta_my, Beta_Fujita = beta_fugita,
         Allele_Status_Fujita = allele_status_fugita,
         Allele_Status_1000G = allele_status_1000G) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Subcluster-specific eQTL {.tabset}

- `subcluster-specific`: eGenes significant at FDR < 0.05 in an L2 subcluster but not its parent L1
- `supercluster-specific`: vice versa
- GO universe (background) is every gene tested in that cell type's own TensorQTL perm file

***

#### Counts

```{r subcluster-counts}
subcluster_tbl |>
  count(cell_type_L2, parent_L1, name = "n_specific_eGenes") |>
  gt()
```

***

#### Parent nominal statistics

```{r subcluster-parent-stats, fig.height=6, fig.width=9}
p1 <- subcluster_tbl |>
  filter(!is.na(pval_nom_parent)) |>
  ggplot(aes(x = pval_nom_parent)) +
  geom_histogram(bins = 50, fill = "#377EB8", colour = "white") +
  facet_wrap(~ cell_type_L2, scales = "free_y") +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "red") +
  labs(x = "Nominal p-value in parent L1", y = "Count") +
  theme_classic(base_size = 11)

p2 <- subcluster_tbl |>
  filter(!is.na(slope_L2), !is.na(slope_parent)) |>
  ggplot(aes(x = slope_L2, y = slope_parent)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linetype = "dashed") +
  facet_wrap(~ cell_type_L2, scales = "free") +
  labs(x = "Slope in L2 subcluster", y = "Slope in parent L1") +
  theme_classic(base_size = 11)

p1 / p2
```

***

#### Summary per L2 type

```{r subcluster-summary-table}
subcluster_tbl |>
  group_by(cell_type_L2, parent_L1) |>
  summarise(
    n_eGenes = n_distinct(phenotype_id),
    median_pval_parent = round(median(pval_nom_parent, na.rm = TRUE), 4),
    prop_same_direction = round(mean(sign(slope_L2) == sign(slope_parent), na.rm = TRUE), 3),
    prop_nom_sig_parent = round(mean(pval_nom_parent < 0.05, na.rm = TRUE), 3),
    n_de_markers = sum(is_de_marker, na.rm = TRUE),
    .groups = "drop"
  ) |>
  gt() |>
  tab_options(table.font.size = px(11))
```

***

#### Supercluster-specific counts

```{r supercluster-counts}
supercluster_tbl |>
  count(parent_L1, name = "n_L1_only_eGenes") |>
  gt()
```

***

#### DE marker overlap

```{r de-overlap, fig.height=5, fig.width=7}
subcluster_tbl |>
  count(cell_type_L2, is_de_marker) |>
  ggplot(aes(x = cell_type_L2, y = n, fill = is_de_marker)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "grey70"),
                     labels = c("TRUE" = "DE marker", "FALSE" = "Not DE")) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Proportion of subcluster-specific eGenes", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

***

#### GO enrichment (subcluster-specific)

```{r go-subcluster, fig.height=10, fig.width=10, results='asis'}
go_results_sub <- list()

for (ct in unique(subcluster_tbl$cell_type_L2)) {
  genes <- subcluster_tbl |> filter(cell_type_L2 == ct) |> pull(phenotype_id) |> unique()
  if (length(genes) < 5) next

  universe_genes <- read_perm_universe(ct, perm_dir)

  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)
  universe_entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = universe_genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)

  if (is.null(entrez) || sum(!is.na(entrez)) < 5) {
    cat("GO universe/entrez mapping failed or insufficient for", ct, "\n\n")
    next
  }

  ego <- tryCatch(
    enrichGO(gene = na.omit(entrez), universe = na.omit(universe_entrez),
             OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH",
             pvalueCutoff = 0.05, readable = TRUE),
    error = function(e) { cat("enrichGO failed for", ct, ":", conditionMessage(e), "\n\n"); NULL })

  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) go_results_sub[[ct]] <- ego
}

for (ct in names(go_results_sub)) {
  cat("\n\n#####", ct, "\n\n")
  print(dotplot(go_results_sub[[ct]], showCategory = 15, title = ct))
}
if (length(go_results_sub) == 0) cat("No cell types produced GO output.\n")
```

***

#### GO enrichment (supercluster-specific)

***

```{r go-supercluster, fig.height=10, fig.width=10, results='asis'}
go_results_super <- list()

for (ct in unique(supercluster_tbl$parent_L1)) {
  genes <- supercluster_tbl |> filter(parent_L1 == ct) |> pull(phenotype_id) |> unique()
  if (length(genes) < 5) next

  universe_genes <- read_perm_universe(ct, perm_dir)

  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)
  universe_entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db, keys = universe_genes,
                          keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first"),
    error = function(e) NULL)

  if (is.null(entrez) || sum(!is.na(entrez)) < 5) {
    cat("GO universe/entrez mapping failed or insufficient for", ct, "\n\n")
    next
  }

  ego <- tryCatch(
    enrichGO(gene = na.omit(entrez), universe = na.omit(universe_entrez),
             OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH",
             pvalueCutoff = 0.05, readable = TRUE),
    error = function(e) { cat("enrichGO failed for", ct, ":", conditionMessage(e), "\n\n"); NULL })

  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) go_results_super[[ct]] <- ego
}

for (ct in names(go_results_super)) {
  cat("\n\n#####", ct, "\n\n")
  print(dotplot(go_results_super[[ct]], showCategory = 15, title = ct))
}
if (length(go_results_super) == 0) cat("No cell types produced GO output.\n")
```

***

#### Subcluster-specific gene table

***

```{r subcluster-gene-table}
subcluster_tbl |>
  select(Gene = gene_symbol, ENSG = phenotype_id, eSNP = variant_id,
         L2_Type = cell_type_L2, Parent_L1 = parent_L1,
         Slope_L2 = slope_L2, QVal_L2 = qval_L2,
         Slope_Parent = slope_parent, Pval_Parent = pval_nom_parent,
         DE_Marker = is_de_marker) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Benchmarking {.tabset}

***

#### Table

***

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
files <- list.files(bmark_dir, pattern = "dev_specificity\\..*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]
  # "dev_specificity.extract_universe.txt" -> parts[2] = "extract_universe"
  rule <- parts[2]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

summary <- benchmarks |>
  group_by(rule) |>
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) |>
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  arrange(rule)

summary |>
  gt() |>
  tab_header(
    title = "Benchmarking",
    subtitle = "Means and SD across replicates"
  ) |>
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) |>
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) |>
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

***

```{r runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```

***

```{r session-info}
sessionInfo()
```

***
