---
title: "Developmental Specificity of Fetal Brain eQTL"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
params:
  eqtl_universe: null
  classification: null
  heterogeneity_test: null
  ocr_enrich: null
  eqtl_effects: null
  beta_correlation: null
  perm_dir: null
  gene_lookup: null
  bmark_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(gt)
  library(DT)
  library(data.table)
  library(tidyverse)
})

# --- Params, not snakemake@input: this Rmd is rendered via
# rmarkdown::render(params = list(...)), not Snakemake's script: directive ---
eqtl_universe_file    <- params$eqtl_universe
classification_file   <- params$classification
heterogeneity_file    <- params$heterogeneity_test
ocr_enrich_file        <- params$ocr_enrich
eqtl_effects_file      <- params$eqtl_effects
beta_correlation_file  <- params$beta_correlation
bmark_dir              <- params$bmark_dir

gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)

# Smaller-font DT wrapper for wide tables (Opposite effects, Heterogeneity
# results, Subcluster-specific gene table) so all columns fit on screen
# without horizontal scroll
small_dt <- function(df, page_length = 20) {
  datatable(
    df, filter = "top", rownames = FALSE, class = "compact stripe hover",
    options = list(
      pageLength = page_length, scrollX = TRUE,
      initComplete = DT::JS(
        "function(settings, json) {",
        "$(this.api().table().container()).css({'font-size': '75%'});",
        "}"
      )
    )
  )
}
```

```{r load-data}
eqtl_universe    <- read_rds(eqtl_universe_file)
classification   <- read_rds(classification_file)
het_test         <- read_rds(heterogeneity_file)
ocr_enrich       <- read_rds(ocr_enrich_file)
eqtl_effects     <- read_rds(eqtl_effects_file)
beta_correlation <- read_rds(beta_correlation_file)

# Unpack classification -- "any" tier only; cell_specific tier retired
# (superseded by the heterogeneity test, below)
any_table    <- classification$any_table
any_summary  <- classification$any_summary

# Collapse the two untestable-not-expressed categories for display only --
# B's underlying categories stay granular in its own TSV output.
any_summary <- any_summary |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable_not_expressed",
    untestable_not_genotyped_not_expressed = "untestable_not_expressed"
  )) |>
  group_by(final_category) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(prop = round(n / sum(n), 3))

any_table <- any_table |>
  mutate(final_category = recode(final_category,
    untestable_genotyped_not_expressed     = "untestable_not_expressed",
    untestable_not_genotyped_not_expressed = "untestable_not_expressed"
  ))

# Unpack heterogeneity test (Cochran's Q on fetal-specific pairs). het_usable
# is the detail level (one row per fetal-adult comparison); het_heterogeneous
# is the aggregated, stringent pair-level call (heterogeneous in ALL tested
# adult comparisons).
het_usable        <- het_test$usable
het_heterogeneous <- het_test$heterogeneous_pairs
het_funnel        <- het_test$funnel

# Unpack subcluster/supercluster effect-size object
subcluster_tbl   <- eqtl_effects$subcluster_specific
supercluster_tbl <- eqtl_effects$supercluster_specific

# Full L1+L2+adult pairwise slope matrix
effect_matrix <- eqtl_effects$effect_size_matrix
effect_long   <- eqtl_effects$effect_size_long

# Unpack beta correlation object (real structure from devspec_beta_correlation.R)
paired_betas     <- beta_correlation$paired_betas
opposite_effects <- beta_correlation$opposite_effects
celltype_map     <- beta_correlation$celltype_map
af_tolerance     <- beta_correlation$af_tolerance

# Add gene symbols (paired_betas/opposite_effects only carry ENSG in `gene`)
paired_betas <- paired_betas |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

opposite_effects <- opposite_effects |>
  left_join(gene_lookup, by = c("gene" = "ensembl_gene_id"))

# Three-way point classification for the scatter grid: grey (not significant in
# both datasets), blue (significant in both, concordant), red (significant in
# both, discordant). Plotted on every overlapping pair regardless of category --
# only the colour/highlight is restricted to the both-significant subset.
paired_betas <- paired_betas |>
  mutate(
    significant_both = significant_fugita == "Yes",
    concordant = sign(beta_my) == sign(beta_fugita),
    point_category = case_when(
      significant_both & concordant  ~ "concordant_both_sig",
      significant_both & !concordant ~ "discordant_both_sig",
      TRUE                            ~ "other"
    )
  )

# Per-comparison Pearson r (no cor_summary is written by the beta_correlation
# rule, so compute it here)
beta_cor_summary <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n = n(),
    r = cor(beta_my, beta_fugita, use = "complete.obs"),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type))

comparison_levels <- paste(celltype_map$my_cell, "vs", celltype_map$fugita_cell)

# Significance-count summary table (one row per comparison): significant in
# fetal & present in adult; also significant in Fujita; of those, concordant;
# of those, discordant.
significance_counts <- paired_betas |>
  group_by(my_cell_type, fugita_cell_type) |>
  summarise(
    n_fetal_sig_present_adult = n(),
    n_also_sig_fujita = sum(significant_both, na.rm = TRUE),
    n_both_sig_concordant = sum(significant_both & concordant, na.rm = TRUE),
    n_both_sig_discordant = sum(significant_both & !concordant, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(comparison = paste(my_cell_type, "vs", fugita_cell_type)) |>
  arrange(factor(comparison, levels = comparison_levels))
```

## Developmental specificity pipeline steps {.tabset}

***

1. `extract_eqtl_universe`: Extract unique fetal eGene-eSNP universe (L1+L2)
2. `classify_eqtl_specificity`: Classify pairs as testable/untestable and shared/fetal-specific vs Fujita ("any" tier)
3. `heterogeneity_test`: Cochran's Q test on fetal-specific pairs vs all tested adult Fujita cell types
4. `prep_ld_clump_input` / `run_ld_clump` / `parse_ld_clump`: LD-independence pruning of the eGene-eSNP universe
5. `ocr_enrich`: Test fetal-specific/shared/specific-heterogeneous eQTL enrichment in adult (Li et al.) cortical union OCRs
6. `eqtl_effect_sizes`: Extract subcluster/supercluster-specific eQTL
7. `beta_correlation`: Fetal-adult eQTL beta correlations and opposite-effect allele checks
8. `dev_specificity_report`: Summarise developmental specificity results

***

### Classification of fetal eQTL {.tabset}

Unique eGene-eSNP pairs sig. in fetal data classified against adult eGene-eSNP pairs:

- Testable shared: if sig. in >= 1 Fugita cell type
  - Feeds: OCR enrichment (as the "shared" SNP set)
- Testable specific: present in Fugita, but not sig. in any Fugita cell type
  - Feeds: Heterogeneity test (next section), OCR enrichment (as the "fetal-specific" SNP set)

Both categories are LD-pruned (see LD-clumping steps) before being used in OCR enrichment.

***

#### Summary table

```{r any-tier-summary}
any_summary |>
  mutate(pct = percent(prop, accuracy = 0.1)) |>
  select(Category = final_category, Count = n, Proportion = pct) |>
  gt() |>
  tab_header(title = '"Any" tier classification')
```

#### Barplot

```{r any-tier-barplot, fig.height=5, fig.width=7}
any_summary |>
  mutate(
    final_category = fct_reorder(final_category, n),
    fill_group = case_when(
      str_detect(final_category, "shared")   ~ "Shared",
      str_detect(final_category, "specific") ~ "Fetal-specific",
      TRUE                                   ~ "Untestable"
    )
  ) |>
  ggplot(aes(x = final_category, y = n, fill = fill_group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Shared" = "#4DAF4A", "Fetal-specific" = "#E41A1C",
                                "Untestable" = "grey70")) +
  labs(x = NULL, y = "Number of unique (eGene, eSNP) pairs", fill = "Classification") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

#### Testable-pairs pie

```{r any-tier-testable-pie, fig.height=4, fig.width=5}
any_table |>
  filter(str_detect(final_category, "^testable")) |>
  count(final_category) |>
  ggplot(aes(x = "", y = n, fill = final_category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("testable_shared" = "#4DAF4A", "testable_specific" = "#E41A1C")) +
  labs(fill = "Classification") +
  theme_void(base_size = 13)
```

***

### Heterogeneity test (fetal-specific eQTL) {.tabset}

- Cochran's Q: Tests whether the fetal and adult effect sizes differ by more
than combined sampling error predicts
- Each `testable_specific` pair is collapsed to a single fetal reference row
(whichever fetal cell type it's most significant in), then tested against
EVERY adult Fujita cell type that pair was actually tested in -- not just its
a priori biologically-matched type
- FDR (BH) applied once, pooled across all usable rows
- STRINGENT final call: a pair only counts as fetal-specific-and-heterogeneous
if heterogeneous in ALL of its tested adult comparisons, not just one
- Notes:
  - Q is not directional -- it tells you the betas differ, not which way
    (see Delta_Beta in the detail table for direction)
  - Pairs tested against more adult cell types face a harder bar under the
    "all" rule -- see the stringency table below

***

#### Summary

```{r het-test-funnel}
het_funnel |>
  gt() |>
  tab_header(title = "Population funnel")
```

```{r het-test-summary-stats, results = "asis"}
n_row_het    <- sum(het_usable$heterogeneous)
n_row_total  <- nrow(het_usable)
n_pair_het   <- sum(het_heterogeneous$heterogeneous_all)
n_pair_total <- nrow(het_heterogeneous)

cat(sprintf(
  paste0(
    "At the individual fetal-adult comparison level, %d of %d tested comparisons (%s) ",
    "were formally heterogeneous (FDR < 0.05). Applying the stringent rule -- heterogeneous ",
    "in **all** tested adult comparisons -- %d of %d testable_specific pairs (%s) pass."
  ),
  n_row_het, n_row_total, percent(n_row_het / n_row_total, accuracy = 0.1),
  n_pair_het, n_pair_total, percent(n_pair_het / n_pair_total, accuracy = 0.1)
))
```

```{r het-test-stringency-effect}
het_heterogeneous |>
  mutate(n_bucket = if_else(n_comparisons_tested >= 4, "4+", as.character(n_comparisons_tested))) |>
  group_by(n_bucket) |>
  summarise(N_Pairs = n(), N_Passed_All = sum(heterogeneous_all), .groups = "drop") |>
  mutate(Pass_Rate = percent(N_Passed_All / N_Pairs, accuracy = 0.1)) |>
  arrange(n_bucket) |>
  gt() |>
  tab_header(title = "Pass rate by number of adult comparisons tested",
             subtitle = "Pairs tested against more adult cell types face a harder bar under the 'all' rule")
```

***

#### Final pairs (stringent)

The headline result: one row per pair, aggregated across all its tested adult
comparisons.

```{r het-test-final-pairs-table}
het_heterogeneous |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(
    ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
    Fetal_Cell_Type = fetal_cell_type,
    N_Adult_Comparisons_Tested = n_comparisons_tested,
    N_Heterogeneous = n_comparisons_heterogeneous,
    Heterogeneous_All = heterogeneous_all,
    Matched_Type_Tested = matched_type_tested,
    Confirmed_By_Matched_Type = confirmed_by_matched_type
  ) |>
  arrange(desc(Heterogeneous_All), desc(N_Adult_Comparisons_Tested)) |>
  small_dt()
```

***

#### Magnitude-consistent pairs (|fetal| > |adult|)

Restricts the 531 stringent (heterogeneous-in-all) pairs further: retains
only those where |fetal beta| > |adult beta| in EVERY tested comparison for
that pair -- not just heterogeneous, but the fetal effect strictly larger in
magnitude, regardless of sign. This sidesteps delta_beta's ambiguity when
fetal and adult betas have opposite signs (delta_beta can look "fetal-bigger"
even when adult's magnitude is actually larger). NOT yet fed to OCR
enrichment -- report-only for now.

```{r het-test-magnitude-prep}
stringent_magnitude <- het_usable |>
  semi_join(het_heterogeneous |> filter(heterogeneous_all),
            by = c("phenotype_id", "variant_id")) |>
  mutate(
    magnitude_diff = abs(beta_my) - abs(beta_fugita_harmonized),
    fetal_bigger = magnitude_diff > 0
  )

pair_magnitude <- stringent_magnitude |>
  group_by(phenotype_id, variant_id) |>
  summarise(
    fetal_cell_type = first(cell_type),
    n_comparisons = n(),
    n_fetal_bigger = sum(fetal_bigger),
    magnitude_consistent = all(fetal_bigger),
    min_magnitude_diff = min(magnitude_diff),
    max_magnitude_diff = max(magnitude_diff),
    .groups = "drop"
  )
```

```{r het-test-magnitude-summary-stats, results = "asis"}
n_pass <- sum(pair_magnitude$magnitude_consistent)
n_total <- nrow(pair_magnitude)

cat(sprintf(
  "%d of %d stringent pairs (%s) also have |fetal beta| > |adult beta| in EVERY tested comparison.",
  n_pass, n_total, percent(n_pass / n_total, accuracy = 0.1)
))
```

```{r het-test-magnitude-summary-table}
pair_magnitude |>
  count(magnitude_consistent, name = "N_Pairs") |>
  mutate(Proportion = percent(N_Pairs / sum(N_Pairs), accuracy = 0.1)) |>
  gt() |>
  tab_header(title = "Magnitude consistency (|fetal beta| > |adult beta|) across a pair's comparisons")
```

```{r het-test-magnitude-histogram, fig.height=4, fig.width=7}
ggplot(stringent_magnitude, aes(x = magnitude_diff)) +
  geom_histogram(bins = 30, fill = "#4DAF4A", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "|fetal beta| - |adult beta| (per comparison)", y = "N comparisons") +
  theme_classic(base_size = 12)
```

```{r het-test-magnitude-final-table}
pair_magnitude |>
  filter(magnitude_consistent) |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
         Fetal_Cell_Type = fetal_cell_type, N_Comparisons = n_comparisons,
         Min_Magnitude_Diff = min_magnitude_diff, Max_Magnitude_Diff = max_magnitude_diff) |>
  arrange(Min_Magnitude_Diff) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  small_dt()
```

***

#### Adult-larger comparisons

Of the individual fetal-adult comparisons underlying the 531 stringent
pairs, those where |adult beta| actually exceeds |fetal beta| -- the mirror
case to the "Magnitude-consistent" tab. Some of these belong to pairs where
adult wins in every comparison; others belong to pairs where fetal wins in
a *different* comparison for the same gene/SNP ("mixed").

```{r het-test-adult-bigger-prep}
adult_bigger_rows <- stringent_magnitude |>
  filter(magnitude_diff < 0) |>
  arrange(magnitude_diff)

target_pairs <- adult_bigger_rows |>
  distinct(phenotype_id, variant_id, cell_type)

target_rows <- stringent_magnitude |>
  semi_join(target_pairs, by = c("phenotype_id", "variant_id", "cell_type"))

side_by_side <- target_rows |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(phenotype_id, external_gene_name, variant_id, cell_type,
         beta_my, fugita_cell_type, beta_fugita_harmonized, magnitude_diff) |>
  group_by(phenotype_id, variant_id, cell_type) |>
  mutate(Fetal_Beta = first(beta_my)) |>
  ungroup() |>
  select(-beta_my) |>
  pivot_wider(
    id_cols = c(phenotype_id, external_gene_name, variant_id, cell_type, Fetal_Beta),
    names_from = fugita_cell_type,
    values_from = c(beta_fugita_harmonized, magnitude_diff),
    names_glue = "{fugita_cell_type}_{.value}"
  ) |>
  rename(Gene = external_gene_name, ENSG = phenotype_id, eSNP = variant_id,
         Fetal_Cell_Type = cell_type)

mixed_flag <- target_rows |>
  distinct(phenotype_id, variant_id) |>
  left_join(
    stringent_magnitude |>
      group_by(phenotype_id, variant_id) |>
      summarise(all_adult_bigger = all(!fetal_bigger), .groups = "drop"),
    by = c("phenotype_id", "variant_id")
  ) |>
  mutate(pair_type = if_else(all_adult_bigger, "adult_bigger_everywhere",
                              "mixed -- fetal bigger in >=1 other comparison"))
```

```{r het-test-adult-bigger-summary-stats, results = "asis"}
cat(sprintf(
  "%d individual comparisons (across %d unique gene/SNP/fetal-cell-type triples) show |adult beta| > |fetal beta|.",
  nrow(adult_bigger_rows), nrow(target_pairs)
))
```

```{r het-test-adult-bigger-breakdown}
mixed_flag |>
  count(pair_type, name = "N_Triples") |>
  mutate(Proportion = percent(N_Triples / sum(N_Triples), accuracy = 0.1)) |>
  gt() |>
  tab_header(title = "Breakdown: adult-bigger-everywhere vs mixed")
```

```{r het-test-adult-bigger-table}
side_by_side |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Per-comparison detail

Every fetal-adult comparison underlying the table above -- for checking why
a specific pair passed or failed the "all" rule.

```{r het-test-detail-table}
het_usable |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(
    Fetal_Cell_Type = cell_type,
    ENSG = phenotype_id, Symbol = external_gene_name, eSNP = variant_id,
    Fetal_Beta = beta_my, Fetal_SE = se_my, Fetal_AF = af_my,
    Adult_Cell_Type = fugita_cell_type, Biologically_Matched = biologically_matched,
    Adult_Beta = beta_fugita_harmonized, Adult_SE = se_fugita, Adult_AF = af_fugita,
    Delta_Beta = delta_beta,
    Q, P = p, Q_value = q, Heterogeneous = heterogeneous
  ) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### Effect-size concordance {.tabset}

Pearson correlation of eQTL betas between each fetal cell type and its
matched Fujita adult cell type. Points are highlighted only where the pair 
is sig. in both datasets  (significant_by_2step_FDR == "Yes"):

- Blue: sig. in both and concordant sign
- Red: sig. in both and and discordant sign (also shown in oppsite effects tab) 
- Grey: sig. in fetal, not sig. in Fugita

***

#### Significance counts

```{r beta-significance-counts}
significance_counts |>
  select(Comparison = comparison,
         `Fetal-sig, present in adult` = n_fetal_sig_present_adult,
         `Also Fujita-sig` = n_also_sig_fujita,
         `Both-sig, concordant` = n_both_sig_concordant,
         `Both-sig, discordant` = n_both_sig_discordant) |>
  gt() |>
  tab_header(title = "Significance counts per fetal-adult comparison")
```

***

#### Scatter grid

```{r beta-scatter-grid, fig.height=8, fig.width=14}
plot_df <- paired_betas |>
  mutate(
    comparison = factor(paste(my_cell_type, "vs", fugita_cell_type), levels = comparison_levels)
  ) |>
  # Plot "other" (grey) points first so blue/red highlighted points draw on top
  arrange(point_category == "other")

r_labels <- beta_cor_summary |>
  mutate(comparison = factor(comparison, levels = comparison_levels),
         label = paste0("r = ", round(r, 3), "\nn = ", n))

ggplot(plot_df, aes(x = beta_my, y = beta_fugita)) +
  geom_point(aes(colour = point_category, alpha = point_category != "other", size = point_category != "other")) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#377EB8", linewidth = 0.6) +
  geom_text(data = r_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  scale_colour_manual(values = c(concordant_both_sig = "#377EB8",
                                 discordant_both_sig = "#E41A1C",
                                 other = "grey70"),
                       labels = c(concordant_both_sig = "Concordant (sig. both)",
                                  discordant_both_sig = "Discordant (sig. both)",
                                  other = "Not sig. in both")) +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.25), guide = "none") +
  scale_size_manual(values = c(`TRUE` = 1.1, `FALSE` = 0.6), guide = "none") +
  facet_wrap(~ comparison, ncol = 4, scales = "free") +
  labs(x = "Fetal slope", y = "Adult (Fujita) slope", colour = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

***

#### Correlation heatmap {.tabset}

Pairwise Pearson correlations of eQTL effect sizes (slope) across all fetal
L1 cell types, fetal L2 subclusters, and Fujita adult cell types, computed
on the subset of (eGene, eSNP) pairs with non-missing slopes in both cell
types of each comparison.

##### Clustered (original)

```{r effect-size-heatmap, fig.height=8, fig.width=9}
slope_cols <- grep("^slope_", names(effect_matrix), value = TRUE)
slope_mat <- as.matrix(effect_matrix[, slope_cols])
colnames(slope_mat) <- str_remove(colnames(slope_mat), "^slope_")

fetal_L1_cols <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
fetal_L2_cols <- sort(setdiff(
  colnames(slope_mat)[!str_starts(colnames(slope_mat), "Fujita_")],
  fetal_L1_cols
))
fujita_cols <- sort(colnames(slope_mat)[str_starts(colnames(slope_mat), "Fujita_")])

col_order <- c(fetal_L1_cols, fetal_L2_cols, fujita_cols)
col_order <- col_order[col_order %in% colnames(slope_mat)]
slope_mat <- slope_mat[, col_order]

cor_mat <- cor(slope_mat, use = "pairwise.complete.obs")

stage_anno <- ifelse(str_starts(colnames(cor_mat), "Fujita_"), "Adult", "Fetal")
ha <- HeatmapAnnotation(
  Stage = stage_anno,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

Heatmap(
  cor_mat,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Cell-type ordered

Dendrogram removed, rows/columns fixed so each L1 cell type sits next to its
own L2 subclusters

```{r effect-size-heatmap-grouped, fig.height=8, fig.width=9}
grouped_order <- c(
  "GABA", "GABA-0", "GABA-1", "GABA-2",
  "Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
  "Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2",
  "NPC", "NPC-0", "NPC-1", "NPC-2",
  "OPC", "MG", "Endo-Peri",
  "Fujita_Exc", "Fujita_Inh", "Fujita_Mic", "Fujita_OPC", "Fujita_End",
  "Fujita_Ast", "Fujita_Oli"
)
grouped_order <- grouped_order[grouped_order %in% colnames(cor_mat)]

cor_mat_grouped <- cor_mat[grouped_order, grouped_order]
stage_anno_grouped <- ifelse(str_starts(colnames(cor_mat_grouped), "Fujita_"), "Adult", "Fetal")
ha_grouped <- HeatmapAnnotation(
  Stage = stage_anno_grouped,
  col = list(Stage = c(Fetal = "#984EA3", Adult = "#FF7F00"))
)

Heatmap(
  cor_mat_grouped,
  name = "Pearson r",
  col = col_fun,
  top_annotation = ha_grouped,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_grouped[i, j]), x, y, gp = gpar(fontsize = 6))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

##### Fetal only

Adult (Fujita) cell types removed entirely; fetal L1/L2 comparisons only,
re-clustered on this fetal-only subset.

```{r effect-size-heatmap-fetal-only, fig.height=8, fig.width=8}
fetal_only_cols <- colnames(cor_mat)[!str_starts(colnames(cor_mat), "Fujita_")]
cor_mat_fetal <- cor_mat[fetal_only_cols, fetal_only_cols]

Heatmap(
  cor_mat_fetal,
  name = "Pearson r",
  col = col_fun,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", cor_mat_fetal[i, j]), x, y, gp = gpar(fontsize = 7))
  },
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9)
)
```

***

#### Pairwise n

```{r pairwise-n, fig.height=8, fig.width=9}
n_mat <- matrix(NA, ncol(slope_mat), ncol(slope_mat),
                dimnames = list(colnames(slope_mat), colnames(slope_mat)))
for (i in seq_len(ncol(slope_mat))) {
  for (j in seq_len(ncol(slope_mat))) {
    n_mat[i, j] <- sum(complete.cases(slope_mat[, c(i, j)]))
  }
}

col_fun_n <- colorRamp2(c(0, max(n_mat, na.rm = TRUE)), c("white", "#08519C"))

Heatmap(
  n_mat,
  name = "n pairs",
  col = col_fun_n,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = ha,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(n_mat[i, j], x, y, gp = gpar(fontsize = 5))
  },
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8)
)
```

***

#### Full slope table

Betas for every (eGene, eSNP) pair across all 19 fetal cell types (L1+L2)
and all 7 Fujita adult cell types

```{r full-slope-table}
slope_full_tbl <- effect_matrix |>
  left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
  select(external_gene_name, phenotype_id, variant_id,
         all_of(paste0("slope_", col_order))) |>
  rename(Symbol = external_gene_name, ENSG_ID = phenotype_id, SNP = variant_id) |>
  rename_with(~ str_remove(., "^slope_"), starts_with("slope_"))

small_dt(slope_full_tbl)
```

***

#### Opposite effects

- Sig. eQTL Pairs in both fetal and Fujita datasets with discordant effect
sign (the red points in the scatter grid). 
- Fetal/Fujita/1000G columns show `REF/ALT; allele frequency` for each dataset 
- AF concordance tolerance: `r af_tolerance`.

```{r opposite-effects-table}
opposite_effects |>
  mutate(
    Fetal  = paste0(ref_my, "/", alt_my, "; ", round(af_my, 4)),
    Fujita = paste0(ref_fugita, "/", alt_fugita, "; ", round(af_fugita, 4)),
    `1000G` = ifelse(is.na(a1_1000G), NA_character_,
                      paste0(a1_1000G, "/", a2_1000G, "; ", round(maf_1000G, 4)))
  ) |>
  select(Gene = external_gene_name, ENSG = gene, eSNP = snp,
         Fetal_Cell = my_cell_type, Fujita_Cell = fugita_cell_type,
         Beta_Fetal = beta_my, Beta_Fujita = beta_fugita,
         Fetal, Fujita, `1000G`,
         Allele_Status = overall_allele_status) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

### OCR enrichment {.tabset}

- LD-pruned fetal-specific, shared, and specific-heterogeneous eQTL SNPs
tested for enrichment in adult cortical OCRs (Li et al. 2023)
- Cortical adult union peak set manually curated
- Specific-heterogeneous is a SUBSET of fetal-specific (the stringent subset
confirmed via the heterogeneity test above), not an independent comparison

***

#### Fold enrichment

```{r ocr-enrichment-barplot, fig.height=4, fig.width=7}
ocr_enrich |>
  mutate(
    sig_label = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns"),
    snp_set = factor(
      recode(snp_set,
             fetal_specific = "Fetal-specific",
             shared = "Shared",
             specific_heterogeneous = "Specific + heterogeneous"),
      levels = c("Shared", "Fetal-specific", "Specific + heterogeneous")
    )
  ) |>
  ggplot(aes(x = snp_set, y = fold_enrichment, fill = snp_set)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_text(aes(label = sig_label), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Fetal-specific" = "#E41A1C", "Shared" = "#4DAF4A",
                                "Specific + heterogeneous" = "#984EA3")) +
  labs(x = NULL, y = "Fold enrichment (adult cortical union)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

***

#### Results table

```{r ocr-enrichment-table}
ocr_enrich |>
  mutate(p_value = signif(p_value, 3),
         fold_enrichment = round(fold_enrichment, 2),
         mean_permuted_overlap = round(mean_permuted_overlap, 1)) |>
  select(Peak_Set = peak_set, SNP_Set = snp_set,
         N_SNPs = n_snps, N_Peaks = n_peaks,
         Observed = observed_overlap, Mean_Permuted = mean_permuted_overlap,
         Fold_Enrichment = fold_enrichment, P_Value = p_value) |>
  gt() |>
  tab_header(title = "OCR enrichment results") |>
  tab_options(table.font.size = px(11))
```

***

### Subcluster-specific eQTL {.tabset}

- FDR sig. eGenes in an L2 subcluster but not its parent L1
("subcluster-specific"), and vice versa ("supercluster-specific").

***

#### Counts

```{r subcluster-counts}
subcluster_tbl |>
  count(cell_type_L2, parent_L1, name = "n_specific_eGenes") |>
  gt()
```

***

#### Supercluster-specific counts

```{r supercluster-counts}
supercluster_tbl |>
  count(parent_L1, name = "n_L1_only_eGenes") |>
  gt()
```

***

#### Subcluster-specific gene table

```{r subcluster-gene-table}
subcluster_tbl |>
  select(Gene = gene_symbol, ENSG = phenotype_id, eSNP = variant_id,
         cell_type_L1 = parent_L1, Slope_L1 = slope_parent, SE_L1 = se_parent, Pval_L1 = pval_nom_parent,
         cell_type_L2 = cell_type_L2, Slope_L2 = slope_L2, SE_L2 = se_L2, Pval_L2 = pval_nom_L2) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Sibling comparison table

For each subcluster-specific (gene, SNP) pair: betas, SEs, and nominal
p-values across the significant L2 subtype, its parent L1, and its two
non-significant L2 siblings, all on one row. Column order: the L2 subtype
the gene is significant in, then the parent L1, then the remaining L2
siblings. Shows whether the "subcluster-specific" call is driven by one
strong signal diluted by null signal in the siblings when pooled at the L1
level.

```{r subcluster-sibling-table-prep}
family_order <- list(
  "Glu-UL" = c("Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2"),
  "Glu-DL" = c("Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2"),
  "GABA"   = c("GABA", "GABA-0", "GABA-1", "GABA-2"),
  "NPC"    = c("NPC", "NPC-0", "NPC-1", "NPC-2")
)

build_sibling_table <- function(subcluster_tbl, effect_long, gene_lookup) {

  fam_lookup <- bind_rows(lapply(names(family_order), function(f) {
    tibble(parent_L1 = f, cell_type_L2 = family_order[[f]][-1])
  }))

  base <- subcluster_tbl |>
    select(phenotype_id, variant_id, gene_symbol, parent_L1,
           sig_cell_type = cell_type_L2,
           Slope_Sig = slope_L2, SE_Sig = se_L2, Pval_Sig = pval_nom_L2,
           Slope_L1 = slope_parent, SE_L1 = se_parent, Pval_L1 = pval_nom_parent)

  sibling_map <- fam_lookup |>
    inner_join(base |> distinct(phenotype_id, variant_id, parent_L1, sig_cell_type),
               by = "parent_L1", relationship = "many-to-many") |>
    filter(cell_type_L2 != sig_cell_type) |>
    group_by(phenotype_id, variant_id) |>
    mutate(sib_slot = paste0("Sib", row_number())) |>
    ungroup()

  sib_stats <- sibling_map |>
    left_join(
      effect_long |> select(phenotype_id, variant_id, cell_type,
                             Slope = slope, SE = slope_se, Pval = pval_nominal),
      by = c("phenotype_id", "variant_id", "cell_type_L2" = "cell_type")
    )

  sib_wide <- sib_stats |>
    select(phenotype_id, variant_id, sib_slot, cell_type_L2, Slope, SE, Pval) |>
    pivot_wider(
      names_from = sib_slot,
      values_from = c(cell_type_L2, Slope, SE, Pval),
      names_glue = "{sib_slot}_{.value}"
    )

  base |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    left_join(sib_wide, by = c("phenotype_id", "variant_id"))
}

sibling_tbl <- build_sibling_table(subcluster_tbl, effect_long, gene_lookup)
```

```{r subcluster-sibling-table}
sibling_tbl |>
  select(
    Gene = external_gene_name, ENSG = phenotype_id, eSNP = variant_id,
    cell_type_L2a = sig_cell_type, Slope_L2a = Slope_Sig, SE_L2a = SE_Sig, Pval_L2a = Pval_Sig,
    cell_type_L1 = parent_L1, Slope_L1, SE_L1, Pval_L1,
    cell_type_L2b = Sib1_cell_type_L2, Slope_L2b = Sib1_Slope, SE_L2b = Sib1_SE, Pval_L2b = Sib1_Pval,
    cell_type_L2c = Sib2_cell_type_L2, Slope_L2c = Sib2_Slope, SE_L2c = Sib2_SE, Pval_L2c = Sib2_Pval
  ) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  small_dt()
```

***

#### Family heatmaps {.tabset}

For each L1 family with L2 subtypes, betas across the parent and its 3
children for every subcluster-specific (gene, SNP) pair in that family.
Black border marks the cell where the pair was actually classified
subcluster-specific -- the pattern to look for is strong colour in one
column, faint/near-white everywhere else in the row.

```{r subcluster-family-heatmap-prep}
build_family_matrix <- function(family, subcluster_tbl, effect_long, gene_lookup) {
  members <- family_order[[family]]

  fam_pairs <- subcluster_tbl |>
    filter(parent_L1 == family) |>
    distinct(phenotype_id, variant_id, cell_type_L2) |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    mutate(pair_label = paste0(coalesce(external_gene_name, phenotype_id), " (", variant_id, ")"))

  if (nrow(fam_pairs) == 0) return(NULL)

  slopes <- effect_long |>
    filter(cell_type %in% members,
           phenotype_id %in% fam_pairs$phenotype_id,
           variant_id %in% fam_pairs$variant_id) |>
    inner_join(fam_pairs |> select(phenotype_id, variant_id, pair_label),
               by = c("phenotype_id", "variant_id"))

  slope_wide <- slopes |>
    select(pair_label, cell_type, slope) |>
    distinct(pair_label, cell_type, .keep_all = TRUE) |>
    pivot_wider(names_from = cell_type, values_from = slope)

  for (m in members) if (!(m %in% names(slope_wide))) slope_wide[[m]] <- NA_real_

  mat <- as.matrix(slope_wide[, members])
  rownames(mat) <- slope_wide$pair_label

  sig_mat <- matrix(FALSE, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  for (i in seq_len(nrow(fam_pairs))) {
    pl <- fam_pairs$pair_label[i]
    sig_ct <- fam_pairs$cell_type_L2[i]
    if (pl %in% rownames(sig_mat) && sig_ct %in% colnames(sig_mat)) {
      sig_mat[pl, sig_ct] <- TRUE
    }
  }

  list(mat = mat, sig_mat = sig_mat)
}

col_fun_family <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

draw_family_heatmap <- function(family, row_height_in = NULL) {
  fam_data <- build_family_matrix(family, subcluster_tbl, effect_long, gene_lookup)
  if (is.null(fam_data)) {
    cat("No subcluster-specific genes for", family, ".\n")
    return(invisible(NULL))
  }

  ht_args <- list(
    matrix = fam_data$mat,
    name = "Slope",
    col = col_fun_family,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    na_col = "grey90",
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 10),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (fam_data$sig_mat[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", lwd = 2, fill = NA))
      }
    }
  )

  if (!is.null(row_height_in)) {
    ht_args$height <- unit(row_height_in, "in") * nrow(fam_data$mat)
  }

  do.call(Heatmap, ht_args)

}
```

##### Glu-UL

```{r subcluster-family-glu-ul, fig.height=max(8, 0.18 * nrow(build_family_matrix("Glu-UL", subcluster_tbl, effect_long, gene_lookup)$mat)), fig.width=6, results='asis'}
print(draw_family_heatmap("Glu-UL", row_height_in = 0.18))
```

##### Glu-DL

```{r subcluster-family-glu-dl, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("Glu-DL"))
```

##### GABA

```{r subcluster-family-gaba, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("GABA"))
```

##### NPC

```{r subcluster-family-npc, fig.height=8, fig.width=6, results='asis'}
print(draw_family_heatmap("NPC"))
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
files <- list.files(bmark_dir, pattern = "dev_specificity\\..*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]
  # "dev_specificity.extract_universe.txt" -> parts[2] = "extract_universe"
  rule <- parts[2]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

summary <- benchmarks |>
  group_by(rule) |>
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) |>
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  arrange(rule)

summary |>
  gt() |>
  tab_header(
    title = "Benchmarking",
    subtitle = "Means and SD across replicates"
  ) |>
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) |>
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) |>
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

```{r runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min),
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)",
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb),
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)",
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```

```{r session-info}
sessionInfo()
```
