---
title: "Plate 1 - First Pass (2)"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: html_document
---

## Experiment Details {.tabset}

***

+ Plate 1: 96 samples, on MEGA kit v2, 60 samples DC, 36 samples CT
+ Sequenced 15 sub-libraries X 3 runs (one sublibrary substandard)
+ ~30 Billion reads
+ Parse workflow has a low [doublet rate](https://support.parsebiosciences.com/hc/en-us/articles/360053107311-What-is-the-expected-doublet-rate#:~:text=Doublet%20rates%20are%20low%2C%20less,through%20the%20Whole%20Transcriptome%20workflow.): ~3% per 100K cells

***

+ FC samples included

+ 2 filtering steps
+ Stage 1
  + Remove cells with < 300 genes OR < 500 reads
  + Remove samples with fewer than 500 cells
+ Stage 2
  + UMI - 3 Median absolute deviations
  + Genes - 3 Median absolute deviations
  + Mito - 5%
  + Ribo - 5%
  + Filtering MT genes and MALAT1 removed
  
+ Harmony run
+ Explore FC distribution in clusters
+ Bio Psych label transfer

### QC {.tabset}

#### Counts by sample {.tabset}

##### Boxplots

```{r tscp_gene_boxplt , results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_boxplt 
```


***

##### Median counts

```{r median_cnt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_tbl
```

***

#### Filter 1 {.tabset}

##### Low counts

```{r samples_low_cnts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
samples_low_cnts_tbl
```

***

##### Counts post-filter 1

```{r cell_cnts_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
cell_cnts_post_filt_tbl
```

***

##### Median counts post-filter 1

```{r median_cnt_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_post_filt_tbl
```

***

##### Boxplots post-filter 1

```{r tscp_gene_post_filt_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_post_filt_boxplt
```





#### Filter 2 {.tabset}

##### All filters

***

```{r filt_all_counts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
filt_all_counts_tbl
```

***

### Seurat {.tabset}

#### Clusters pre-harmony {.tabset}

***

```{r resolution_cnts, results = 'asis', fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("##### Counts ", i, "\n", sep = "")
  get_meta_col_counts(seurat_obj, paste0('RNA_snn_res.', i)) |>
    #print(n = Inf) |>
    arrange(desc(n)) |>
    flextable() |> 
    set_caption(caption = "Seurat clusters: cell counts") |>
    theme_vanilla() |>
    flextable_to_rmd()
  cat('\n\n')
}
```

***

```{r resolution, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("##### Plots ", i , "\n", sep = "")
  plot_subset <- res_plotlist[grep(i, names(res_plotlist), ignore.case = TRUE)]
  #plot_subset <- res_plotlist[names(res_plotlist)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("\n\n")
}
```

***

#### Integration Plots {.tabset}

```{r harmony_cnts, results = 'asis', fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("##### Counts ", i, "\n", sep = "")
  get_meta_col_counts(seurat_obj, paste0('harmony_clusters_', i)) |>
    #print(n = Inf) |>
    arrange(desc(n)) |>
    flextable() |> 
    set_caption(caption = "Seurat clusters: cell counts") |>
    theme_vanilla() |>
    flextable_to_rmd()
  cat('\n\n')
}
```

***

```{r Integration, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("  \n##### Plots **", i , "**\n", sep = "")
  plot_subset <- integration_plotlist[grep(i, names(integration_plotlist), ignore.case = TRUE)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("  \n")
}
```

#### Violin Plots {.tabset}

```{r violin, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in 1:length(violin_plt_list)) {
  cat("##### Violin ", i, "\n", sep = "")
  plot(violin_plt_list[[i]])
  cat("\n\n")
}
```

#### Cluster Assignment {.tabset}

##### Biol Psych vln

***
```{r biol_pysch_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(biol_psych_vln_plot)
```

***

##### Public 2017 vln

***

```{r public_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(public_vln_plot)
```

***

##### Label transfer 1

***

```{r label_transfer_1, fig.dim = c(20, 20), echo = FALSE, results="asis"}
#print(cowplot::plot_grid(harm_umap, cam_umap, poulio_umap, nowakowski_umap))
```

##### Label transfer 2

***

```{r label_transfer_2, fig.dim = c(20, 60), echo = FALSE, results="asis"}
print(cowplot::plot_grid(cam_fcx_plt, cam_ge_plt, cam_hip_plt, cam_tha_plt, cam_cer_plt, ncol = 1))
```


##### FCX clusters {.tabset}

##### Clusters

***

```{r fcx_fact_clusters, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(cowplot::plot_grid(fcx_umap_plt, fcx_feat_plt))
```

***

##### Barplot

***

```{r fcx_fact_barplt, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(fcx_fact_barplt)
```

***