---
title: "Plate 1 - First Pass (3)"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: html_document
---

## Experiment Details {.tabset}

***

+ Plate 1: 96 samples(60/36), on MEGA kit v2, seq 15 sub-lib X 3 runs
+ Plate 2: 96 samples (88/8), on MEGA kit v2, seq 16 sub-lib X 3 runs
+ ~30 Billion reads per plate
+ Parse workflow has a low [doublet rate](https://support.parsebiosciences.com/hc/en-us/articles/360053107311-What-is-the-expected-doublet-rate#:~:text=Doublet%20rates%20are%20low%2C%20less,through%20the%20Whole%20Transcriptome%20workflow.): ~3% per 100K cells

***

- FC samples included
- 2 filtering steps
- Stage 1
  - Remove cells with < 300 genes OR < 500 reads
  - Remove samples with fewer than 500 cells
- Stage 2
  - UMI - 3 Median absolute deviations
  - Genes - 3 Median absolute deviations
  - Mito - 5%
  - Ribo - 5%
  - Filtering MT genes and MALAT1 removed
- Harmony run
- Bio Psych label transfer
- Explore FC distribution in clusters

### QC {.tabset}

#### Pre-filter counts {.tabset}

##### Boxplots

```{r tscp_gene_boxplt , results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_boxplt 
```

***

##### Median counts

```{r median_cnt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_tbl
```

***

##### Low count cells

```{r samples_low_cnts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
samples_low_cnts_tbl
```

***

##### Metadata

```{r seurat_metadata, results='asis', fig.dim = c(10,10), echo=FALSE}
metadata_lst[['seurat_meta']] |>
  flextable() |> 
  set_caption(caption = "Seurat object metadata") |>
  theme_vanilla() |>
  flextable_to_rmd()
```

#### Filter 1 {.tabset}

##### Counts post-filter 1

```{r cell_cnts_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
cell_cnts_post_filt_tbl
```

***

##### Median counts post-filter 1

```{r median_cnt_post_filt_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_post_filt_tbl
```

***

##### Boxplots post-filter 1

```{r tscp_gene_post_filt_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_post_filt_boxplt
```

***

#### Filter 2 {.tabset}

##### Counts post-filter 2

```{r cell_cnts_final_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
cell_cnts_final_tbl
```

***

##### Median counts post-filter 2

```{r median_cnt_final_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
median_cnt_final_tbl
```

***

##### Boxplots post-filter 2

```{r tscp_gene_final_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_final_boxplt
```

***

##### All filters


```{r filt_all_counts_tbl, results='asis', fig.dim = c(10,10), echo=FALSE}
filt_all_counts_tbl
```

***

### Seurat {.tabset}

#### Clusters pre-harmony {.tabset}

***

```{r resolution_cnts, results = 'asis', fig.dim = c(20, 20), echo=FALSE, results="asis", message=FALSE}
for(i in resolution_set) {
  cat("##### Counts ", i, "\n", sep = "")
  get_meta_col_counts(seurat_obj, paste0('RNA_snn_res.', i)) |>
    #print(n = Inf) |>
    arrange(desc(n)) |>
    flextable() |> 
    set_caption(caption = "Seurat clusters: cell counts") |>
    theme_vanilla() |>
    flextable_to_rmd()
  cat('\n\n')
}
```

***

```{r resolution, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("##### Plots ", i , "\n", sep = "")
  plot_subset <- res_plotlist[grep(i, names(res_plotlist), ignore.case = TRUE)]
  #plot_subset <- res_plotlist[names(res_plotlist)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("\n\n")
}
```

***

#### Integration Plots {.tabset}

```{r harmony_cnts, results = 'asis', fig.dim = c(20, 20), echo = FALSE, results="asis",message=FALSE}
for(i in resolution_set) {
  cat("##### Counts ", i, "\n", sep = "")
  get_meta_col_counts(seurat_obj, paste0('harmony_clusters_', i)) |>
    #print(n = Inf) |>
    arrange(desc(n)) |>
    flextable() |> 
    set_caption(caption = "Seurat clusters: cell counts") |>
    theme_vanilla() |>
    flextable_to_rmd()
  cat('\n\n')
}
```

***

```{r Integration, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("  \n##### Plots **", i , "**\n", sep = "")
  plot_subset <- integration_plotlist[grep(i, names(integration_plotlist), ignore.case = TRUE)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("  \n")
}
```

***

#### Cluster Assignment {.tabset}

##### Biol Psych vln

***
```{r biol_pysch_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(vln_plt_lst[['biol_psych_vln_plot']])
```

***

##### Public 2017 vln

***

```{r public_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(vln_plt_lst[['public_vln_plot']])
```

***

```{r violin, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in tail(names(violin_plt_lst), -2)) {
  cat("##### ", i , "\n", sep = "")
  plot_subset <- violin_plt_list[grep(i, names(violin_plt_list), ignore.case = TRUE)]
  #plot_subset <- res_plotlist[names(res_plotlist)]
  plot(cowplot::plot_grid(plotlist = plot_subset))
  cat("\n\n")
}
```

##### Label transfer 1

***

```{r label_transfer_1, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(cowplot::plot_grid(harm_umap, cam_fcx_umap, poulio_umap, nowakowski_umap))
```

##### Label transfer 2

***

```{r label_transfer_2, fig.dim = c(20, 60), echo = FALSE, results="asis"}
print(cowplot::plot_grid(cam_fcx_plt, cam_ge_plt, cam_hip_plt, cam_tha_plt, cam_cer_plt, ncol = 1))
```

***

#### FCX clusters {.tabset}

##### Clusters

***

```{r fcx_fact_clusters, fig.dim = c(20, 10), echo = FALSE, results="asis"}
print(cowplot::plot_grid(fcx_umap_plt, fcx_feat_plt))
```

##### Clusters (split)

***

```{r fcx_fact_clusters_split, fig.dim = c(10, 10), echo = FALSE, results="asis"}
print(fcx_umap_split_plt)
```


***

##### Barplot

***

```{r fcx_fact_barplt, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(fcx_fact_barplt)
```

***

##### Barplot (pcw)

***

```{r fcx_pcw_barplt, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(fcx_pcw_barplt)
```

***

##### Counts

***

```{r fcx_fact_counts, fig.dim = c(20, 20), echo = FALSE, results="asis"}
fcx_in_clust_tbl
```

***
