---
title: "Genotyping QC pre-Imputation"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```

## Genotyping pre-imputation QC steps

***

Steps 3-7 were carried out using the [GenotypeQCtoHRC package](https://github.com/CardiffMRCPathfinder/GenotypeQCtoHRC). 
See the `eqtlgenotypeshg38pretm.qc_report.html` for those metrics. Some additional details for some of those tests are 
below and in the `reports/03GENOTYPES-PRE`directory. The citation for GenotypeQCtoHRC is [Lynham et al. 2023 (DOI:10.1192/bjo.2022.636)](https://www.cambridge.org/core/journals/bjpsych-open/article/dragondata-a-platform-and-protocol-for-integrating-genomic-and-phenotypic-data-across-large-psychiatric-cohorts/2531261CD4300FCF01FA33C6B8693375).

1. Rm duplicate sample (14493 and 144FC same sample. 14493 removed)
2. Check autosomal heterozygosity
3. Missingness:
  - Sample rm if > 5% markers missing, 
  - SNPs excluded if missing in > 5% samples
4. HWE p < 1 x 10-6
5. Samples ambig. sex assignments removed (all passed)
6. Relatedness assessed (duplicate sample identified 14493 and removed)
7. Ancestry assessed
8. Liftover hg19 to hg38
9. Conversion of markers to dbSNP 156 (check this) using 1000G Phase 3
10. [Genotype Harmonizer](https://github.com/molgenis/systemsgenetics/wiki/Genotype-Harmonizer) used to align curated markers to the Trans-Omics for Precision Medicine ([TopMED](https://topmed.nhlbi.nih.gov/)) imputation panel
11. Remove ambiguous A/T, G/C SNP with MAF > 0.4 (none rm here does GenotypeQCtoHRC do this?), 
12. Remove SNPs MAF < 0.01

***

## QC tables and plots {.tabset}

***

### QC 2: Autosomal Heterozygosity {.tabset}

***

#### Tables

```{r, autosomal_het_table, results='asis', echo=FALSE}
het_tbl <- read.table(paste0(in_dir, 'eqtl_genotypes_hg19_nodup.het'), header = T) |>
  tibble() |>
  rename(O_HOM = 'O.HOM.', E_HOM = 'E.HOM.', N_NM = 'N.NM.') |>
  mutate(het_rate = (N_NM - O_HOM) / N_NM) |>
  # ID outliers based on het rate
  mutate(
    mean_het = mean(het_rate),
    sd_het = sd(het_rate),
    outlier_het = abs(het_rate - mean_het) > 3 * sd_het
  ) |>
  # ID outliers based on inbreeding coefficient F
    mutate(
    mean_F = mean(F),
    sd_F = sd(F),
    outlier_F = abs(F - mean_F) > 3 * sd_F | F < -0.15 | F > 0.15
  ) |>
  mutate(across(where(is.numeric), ~round(., 3)))
datatable(het_tbl, caption = 'Table 1. Autosomal heterozygosity table')
```

***

```{r, het_outliers}
outliers_tbl <- het_tbl %>%
  filter(outlier_het | outlier_F) %>%
  select(FID, IID, outlier_het, outlier_F)
datatable(outliers_tbl, caption = 'Table 2. Autosomal heterozygosity oultliers table')
write_tsv(outliers_tbl, paste0(in_dir, "het_outliers.txt"), col_names = FALSE)
```

***

#### Plot

***

```{r, het_plot, message = FALSE, fig.dim=c(10, 5)}
het_plot <- ggplot(het_df, aes(x = het_rate)) +
  geom_histogram(binwidth = 0.005, fill = "#4C78A8", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean_het + 3 * sd_het), color = "#E31A1C", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean_het - 3 * sd_het), color = "#E31A1C", linetype = "dashed", size = 1) +
  labs(
    title = "Heterozygosity Rate Distribution",
    x = "Heterozygosity Rate",
    y = "Count",
    caption = "Red dashed lines indicate ±3 SD thresholds"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.caption = element_text(size = 10, hjust = 0)
  )

# Create publication-ready F distribution plot
f_plot <- ggplot(het_df, aes(x = F)) +
  geom_histogram(binwidth = 0.01, fill = "#F28E2B", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean_F + 3 * sd_F), color = "#E31A1C", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean_F - 3 * sd_F), color = "#E31A1C", linetype = "dashed", size = 1) +
  geom_vline(xintercept = c(-0.15, 0.15), color = "#E31A1C", linetype = "solid", size = 1) +
  labs(
    title = "Inbreeding Coefficient (F) Distribution",
    x = "Inbreeding Coefficient (F)",
    y = "Count",
    caption = "Red lines: solid = [-0.15, 0.15], dashed = ±3 SD"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    plot.caption = element_text(size = 10, hjust = 0)
  )

combined_plot <- plot_grid(het_plot, f_plot, labels = c("A", "B"),
                           label_size = 14, label_fontface = "bold",
                           ncol = 2, align = "h")
combined_plot
#ggsave(paste0(ind_dir, "het_F_distribution_combined.png"), combined_plot,
#       width = 10, height = 5, dpi = 300, bg = "white")
```

### QC 5: Sex ambiguity

***

```{r sex_summary, results='asis', echo=FALSE}
sex_data <- system(
  paste0("cat ", in_dir, "eqtl_genotypes_hg19.qc3.sexcheck"),
  intern = TRUE
)

temp_file <- tempfile(fileext = ".txt")
writeLines(sex_data, temp_file)

sex_tbl <- read.table(
  file = temp_file,
  header = FALSE,  # Header is included in sex_data
  skip = 1,        # Skip the header row
  col.names = c("FID", "IID", "PEDSEX", "SNPSEX", "STATUS", "F", "YCOUNT"),
  colClasses = c("character", "character", "integer", "integer", "character", "numeric", "integer"),
  sep = ""  # Let read.table auto-detect whitespace (spaces or tabs)
) %>%
  as_tibble()  # Convert to tibble for consistency

file.remove(temp_file)

datatable(sex_tbl, caption = 'Table 3. Sex ambiguities')
```

***

### QC 6: Kinship 

+ Note: 

***

```{r kinship_summary, results='asis', echo=FALSE}
kinship_data <- system(
  paste0("cat ", in_dir, "eqtl_genotypes_hg19.qc4.pcrelate.kin"),
  intern = TRUE
)

kinship_tbl <- read_tsv(file = I(kinship_data), col_types = cols())

datatable(kinship_tbl, caption = 'Table 4. Related samples') 
```

***

### QC 7: Ancestry {.tabset}

***

#### All

```{r ancestry_summary, results='asis', echo=FALSE}
ancestry_data <- system(
  paste0("cat ", in_dir, "eqtl_genotypes_hg19.qc5.ancestry_inference.txt | cut -f 1-9"),
  intern = TRUE
)

ancestry_tbl <- read_tsv(file = I(ancestry_data), col_types = cols())

datatable(ancestry_tbl, caption = 'Table 5. Ancestry prediction per sample') 
```

***

#### Non_European

```{r ancestry_nonEur_summary, results='asis', echo=FALSE}
ancestry_nonEur_tbl <- ancestry_tbl |>
  filter(biogeographic_prediction != 'European')

datatable(ancestry_nonEur_tbl, caption = 'Table 6. Ancestry (non European) prediction per sample') 
```

**Non-European sample IDs**: `r ancestry_tbl |> filter(biogeographic_prediction != 'European') |> pull(FID)`

***

### QC 6: Post-GenotypeQCtoHRC checks 

***

+ `r cat(length(readLines(paste0(in_dir, 'ambig_high_maf_snp_lst.tsv'))), 'ambig SNPs with MAF > 0.4 removed')` 
+ `r cat(readLines(paste0(in_dir, 'rm_rare_snps.txt')), ", MAF < 0.01", sep = '')`

***

```{r, snp counts, echo = FALSE, message = FALSE, warning = FALSE}
pre_snp_cnts_tbl <- read_tsv(paste0(in_dir, 'gQC2hrc_vcf_chr_counts.tsv'), col_names = c('chr', 'gQCtoHRC_cnts'))
post_snp_cnts_tbl <- read_tsv(paste0(in_dir, 'final_vcf_chr_counts.tsv'), col_names = c('chr', 'final_cnts'))
final_snp_cnt_tbl <- pre_snp_cnts_tbl |>
  left_join(post_snp_cnts_tbl)
datatable(final_snp_cnt_tbl, caption = 'Table 7. SNP counts after GenotypeQCtoHRC and post-SNP checks') 
```

***

### Benchmarking {.tabset}

***

+ Why is the GenotypeQCtoHRC benchmark file missing?

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files
files <- list.files(bmark_dir, pattern = "geno_pre_impute.*.benchmark.txt", full.names = TRUE)

# Read and combine
benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  # Add rule + sublib from filename
  parts <- strsplit(basename(f), "\\.")[[1]]
  tibble(
    file = f,
    rule = parts[2],           # e.g., "cellsnp_lite", "sort_bam"
    sublib = parts[3],         # e.g., "10_plate1"
    runtime_sec = df$s,        # runtime in seconds
    runtime_hms = df$h.m.s.,   # hh:mm:ss string
    max_rss = df$max_rss       # peak RAM in MB
  )
})

# Summarize across replicates
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) 

datatable(summary, caption = 'Table 8. Rule benchmarking. Sd is null when only on instance of job is run.') 

# Optional save summary tables
#write.csv(benchmarks, paste0(bmark_dir, "benchmark_summary_detailed.csv"), row.names = FALSE)
#write.csv(summary, "benchmark_summary_rules.csv", row.names = FALSE)
```

***

#### Plot

***

```{r, runtime_plt, fig.dim=c(10,10)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule (± SD)") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule (± SD)") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```

***
