---
title: "Genotyping Post-imputation report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

## Genotyping post-imputation QC steps {.tabset}

***

+ From workflow dir run `mkdir -p ../results/04GENOTYPES-POST/TOPMED/`. Copy imputed output file into that dircetory

1. `impute_check`: Evaluate imputation quality for each chromosome
2. `impute_check_cat`: Cat impute check counts for each chr into single file
3. `dwnld_dbsnp_ref`: Download dbSNP reference
4. `add_rsID`: Add dbSNP rsIDs to each imputed chr-specific VCF file
5. `vcf_cat`: Cat imputed chr-specific VCF files into single file
6. `filter_tags`: Compute HWE and filt cated VCF; Keep SNPs pass MAF, R^2 and HWE limits
    + HWE: <0.0001
    + MAF < 0.05
    + R^2 < 0.8
7. `check_VCF`: Validate the filt VCF against ref genome and ID problematic SNPs
8. `exclude_SNPs`: Exclude SNPs IDed in CheckVCF
9. `idx_vcf`: Index the final filtered VCF
10. `create_combined_log`: Summarise logs from previous rules
11. `get_sample_list`: Extract final sample list
12. `vcf_to_plink`: Convert genotypes VCF to plink format
13. `get_ld_pruned_snps`: LD prune SNPs before running PCA on genotypes
    + `--indep-pairwise 250 5 0.2`
14. `prune_genotypes`: Create genotypes plink-format file containing only pruned SNPs
15. `calc_genotype_pcs`: Run PCA on pruned genotypes to get genotype-specific PC covariates

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(gtExtras)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```

****

### Check imputation

```{r, check_imputation_tbl, echo = FALSE, message = FALSE, warning = FALSE}
imp_check_tbl <- read_tsv(paste0(in_dir, 'impute_check/chr_ALL_summary.tsv'))

# Create the GT table
imp_check_tbl %>%
  gt() %>%
  # Add a title and subtitle
  tab_header(
    title = "Imputation Quality Summary by Chromosome",
    subtitle = "Metrics include SNP counts, MAF, and R² thresholds"
  ) %>%
  # Customize column labels for clarity
  cols_label(
    chr = "Chromosome",
    snp_count = "Total SNPs",
    maf_lt_0.05 = "MAF < 0.05",
    maf_gteq_0.05 = "MAF ≥ 0.05",
    rsq_lt_0.08 = "R² < 0.8",
    rsq_gteq_0.08 = "R² ≥ 0.8",
    maf_gteq_0.05_rsq_gteq_0.08 = "MAF ≥ 0.05 & R² ≥ 0.8"
  ) %>%
  # Format numbers with commas for readability
  fmt_number(
    columns = c(snp_count, maf_lt_0.05, maf_gteq_0.05, rsq_lt_0.08, rsq_gteq_0.08, maf_gteq_0.05_rsq_gteq_0.08),
    decimals = 0,
    use_seps = TRUE
  ) %>%
  # Apply a color scale to highlight high/low values in numeric columns
  data_color(
    columns = c(snp_count, maf_lt_0.05, maf_gteq_0.05, rsq_lt_0.08, rsq_gteq_0.08, maf_gteq_0.05_rsq_gteq_0.08),
    palette = "viridis",
    domain = NULL
  ) %>%
  # Emphasize the "Overall" row with bold text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = chr == "Overall")
  ) %>%
  # Adjust column alignment
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  # Apply a clean theme
  gt_theme_538() %>%
  # Add a footnote for context
  tab_footnote(
    footnote = "MAF: Minor Allele Frequency; R²: Imputation quality score. Data includes chromosomes 21, 22, X, and overall totals.",
    locations = cells_title(groups = "subtitle")
  ) %>%
  # Set table width and font size
  tab_options(
    table.width = pct(100),
    table.font.size = 14,
    column_labels.font.weight = "bold"
  )
```

***

### Genotype PCs

```{r, genotype_PCs, echo = FALSE, message = FALSE, warning = FALSE}
eigenval <- read_table("../reports/04GENOTYPES-POST/pca.eigenval", col_names = "eigenval")

# calculate proportion and cumulative variance explained
variance <- eigenval %>%
  mutate(PC = row_number(),
         variance_explained = eigenval / sum(eigenval) * 100,
         cumulative = cumsum(variance_explained))

# elbow plot
p <- ggplot(variance, aes(x = PC, y = variance_explained)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Principal Component",
    y = "Variance Explained (%)",
    title = "Genotype PCA Variance Explained"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
p
```


***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "geno_post_impute.*\\.txt", full.names = TRUE)

# Read and combine benchmark files
benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  # Extract rule and optional chromosome from filename
  parts <- strsplit(basename(f), "\\.")[[1]]
  rule_name <- parts[2]  # e.g., "impute_check_1", "add_rsID_X", "vcf_cat"
  # Check if rule name contains a chromosome identifier (e.g., "_1", "_X")
  chr_match <- str_match(rule_name, "^(.+)_([0-9X]+)$")
  if (!is.na(chr_match[1, 1])) {
    base_rule <- chr_match[1, 2]  # e.g., "impute_check"
    chr_part <- chr_match[1, 3]   # e.g., "1", "X"
  } else {
    base_rule <- rule_name
    chr_part <- "single"
  }
  
  tibble(
    file = f,
    rule = base_rule,
    chromosome = chr_part,     # Chromosome (e.g., "1", "X") or "single" for non-chromosome-specific rules
    runtime_sec = df$s,        # Runtime in seconds
    runtime_hms = df$h.m.s.,   # hh:mm:ss string
    max_rss = df$max_rss       # Peak RAM in MB
  )
})

# Summarize across rules, collapsing chromosome-specific rules
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  # Replace NA in sd columns with 0 for single runs
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  # Round numeric columns
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  # Sort by rule for clarity
  arrange(rule)

# Create a DT table
datatable(
  summary,
  caption = 'Table 1. Rule benchmarking. Sd is 0 when only one instance of a job is run or for collapsed chromosome-specific rules.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)
```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule (± SD)") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule (± SD)") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```


