---
title: "JOBS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(UpSetR)
library(patchwork)
library(ggrepel)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```

## JOBS pipeline steps {.tabset}

***

We used [JOBS](https://github.com/LidaWangPSU/JOBS) to generate integrate our sc-eQTL TensorQTL nominial data with
publically available data [Wen et al 2024]() to boost our eQTL discovery power. 
The citation for JOBS is [Wang et al. (2025), DOI: 10.1016/j.xgen.2025.100820](https://www.cell.com/cell-genomics/fulltext/S2666-979X(25)00076-X).

***

1. `run_jobs`: Run Jobs script on TensorQTL nominial eQTL data and public bulk dataset
2. `jobs_report`: Calculate per gene FDR (per cell type) and report sig eGenes (i.e. top eSNP per eGene)

Settings:

+ Run on L1 cell type eQTL only
+ Wen input file: `devBrain_eQTL_EUR_nominal_50HCP_FDR_0.05.txt.gz`

***

```{r, calculate_FDR, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
in_dir <- '~/Desktop/test/jobs/'
cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")

# Load beta and SE files
ref_beta <- read_tsv(paste0(in_dir, "jobs_ref_beta_genomewide.tsv.gz"))
ref_se <- read_tsv(paste0(in_dir, "jobs_ref_se_genomewide.tsv.gz"))

# Add gene column if missing (extract from ID)
if (!"gene" %in% colnames(ref_beta)) {
  ref_beta <- ref_beta %>%
    mutate(gene = str_remove(ID, "-.*"))
}

# Initialize lists for results
sig_esnps_list <- list()  # For all sig eSNPs (FDR < 0.05) per cell
sig_egenes_list <- list()  # For top eSNP per sig eGene per cell (keep FDR ties)
sig_egenes_list_top <- list()  # For top eSNP per sig eGene per cell (only retains 1 top SNP)

# Loop over cell types
for (cell in cell_types) {
  message(paste("Processing cell:", cell))
  
  if (!cell %in% colnames(ref_beta)) {
    message(paste("Skipping", cell, "- column missing"))
    next
  }
  
  # Compute nominal p-values
  beta_col <- ref_beta[[cell]]
  se_col <- ref_se[[cell]]
  p_nom <- 2 * pnorm(-abs(beta_col / se_col))
  
  # Skip if all NA
  if (all(is.na(p_nom))) {
    message(paste("Skipping", cell, "- all p-values NA"))
    next
  }
  
  # Add p-value column to ref_beta
  ref_beta_temp <- ref_beta %>%
    mutate(!!paste0(cell, "_p") := p_nom)
  
  # Compute per-gene FDR
  fdr_temp <- ref_beta_temp %>%
    group_by(gene) %>%
    mutate(fdr = p.adjust(.data[[paste0(cell, "_p")]], method = "BH")) %>%
    ungroup() %>%
    select(ID, gene, !!paste0(cell, "_p") := .data[[paste0(cell, "_p")]], fdr) %>%
    mutate(cell = cell)
  
  # Sig eSNPs: All FDR < 0.05 for this cell
  sig_esnps_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr))
  sig_esnps_list[[cell]] <- sig_esnps_cell
  
  # Sig eGenes: Top eSNP (min FDR) per gene for this cell, if any FDR < 0.05
  sig_egenes_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr)) %>%
    group_by(gene) %>%
    slice_min(fdr, n = 1) %>%  # Top (lowest FDR) eSNP per gene
    ungroup() |>
    arrange(fdr)
  sig_egenes_list[[cell]] <- sig_egenes_cell
  
  # Sig eGenes: Top eSNP (min FDR) per gene for this cell, if any FDR < 0.05
  # Sort by fdr and take exactly 1 per gene to handle FDR ties
  sig_egenes_cell_top <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr)) %>%
    arrange(fdr) %>%  # Sort ascending FDR
    group_by(gene) %>%
    slice_head(n = 1) %>%  # Take first (lowest FDR) row per gene
    arrange(fdr) %>%  # Sort ascending FDR
    ungroup()
  sig_egenes_top_list[[cell]] <- sig_egenes_cell_top
  
  # Warn on NAs in FDR
  if (any(is.na(fdr_temp$fdr))) {
    message(paste("Warning: NAs in FDR for", cell, "- check small gene groups"))
  }
}

# Combine into data frames
sig_esnps <- bind_rows(sig_esnps_list)
sig_egenes <- bind_rows(sig_egenes_list)
sig_egenes_top <- bind_rows(sig_egenes_top_list)
```

***

### eGenes

```{r, eGenes, echo = FALSE, message = FALSE, warning = FALSE}
sig_egenes_summary <- sig_egenes_top %>%
  group_by(cell) %>%
  summarise(
    n_egenes = n(),
    .groups = "drop"
  )

# Original TensorQTL eGenes (from previous analysis)
tensorqtl_egenes <- tibble(
  cell = c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri"),
  tensorqtl_egenes = c(2027, 2048, 937, 1860, 51, 117, 90)
)

# Join and compute percentage increase
egenes_comparison <- sig_egenes_summary %>%
  left_join(tensorqtl_egenes, by = "cell") %>%
  mutate(
    percentage_increase = ((n_egenes - tensorqtl_egenes) / tensorqtl_egenes) * 100
  ) %>%
  select(cell, tensorqtl_egenes, n_egenes, percentage_increase) %>%
  rename(
    `Cell Type` = cell,
    `TensorQTL eGenes` = tensorqtl_egenes,
    `JOBs eGenes` = n_egenes,
    `Percentage Increase (%)` = percentage_increase
  )

egenes_table <- egenes_comparison %>%
  gt() %>%
  tab_header(
    title = "Comparison of Significant eGenes per Cell Type (FDR < 0.05)"
  ) %>%
  cols_label(
    `Cell Type` = "Cell Type",
    `TensorQTL eGenes` = "TensorQTL eGenes",
    `JOBs eGenes` = "JOBs eGenes",
    `Percentage Increase (%)` = "Percentage Increase (%)"
  ) %>%
  fmt_number(
    columns = c(`TensorQTL eGenes`, `JOBs eGenes`),
    decimals = 0
  ) %>%
  fmt_number(
    columns = `Percentage Increase (%)`,
    decimals = 1,
    scale_by = -1  # For percentages
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
  )

egenes_table
write_tsv(sig_egenes_top, paste0(in_dir, "sig_egenes_top_fdr05.tsv.gz"))
```

***

### eGenes plot

***

```{r, eGenes_plt, echo = FALSE, message = FALSE, warning = FALSE}
plot_data <- egenes_comparison %>%
  pivot_longer(
    cols = c(`TensorQTL eGenes`, `JOBs eGenes`),
    names_to = "method",
    values_to = "count"
  ) %>%
  mutate(method = factor(method, levels = c("TensorQTL eGenes", "JOBs eGenes")))

p <- ggplot(plot_data, aes(x = `Cell Type`, y = count, fill = method)) +
  geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
  scale_fill_manual(values = c("TensorQTL eGenes" = "#1f77b4", "JOBs eGenes" = "#ff7f0e")) +
  labs(
    title = "eGenes per Cell Type: TensorQTL vs. JOBs (FDR < 0.05)",
    x = "Cell Type",
    y = "Number of eGenes",
    fill = "Method"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(expand = c(0, 0), labels = scales::comma)

p
```
***

### eGenes (allow mult SNPs)

***

```{r, eGenes_with_ties, echo = FALSE, message = FALSE, warning = FALSE}
sig_egenes_summary <- sig_egenes %>%
  group_by(cell) %>%
  summarise(
    n_egenes = n(),
    .groups = "drop"
  )

# Original TensorQTL eGenes (from previous analysis)
tensorqtl_egenes <- tibble(
  cell = c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri"),
  tensorqtl_egenes = c(2027, 2048, 937, 1860, 51, 117, 90)
)

# Join and compute percentage increase
egenes_comparison <- sig_egenes_summary %>%
  left_join(tensorqtl_egenes, by = "cell") %>%
  mutate(
    percentage_increase = ((n_egenes - tensorqtl_egenes) / tensorqtl_egenes) * 100
  ) %>%
  select(cell, tensorqtl_egenes, n_egenes, percentage_increase) %>%
  rename(
    `Cell Type` = cell,
    `TensorQTL eGenes` = tensorqtl_egenes,
    `JOBs eGenes` = n_egenes,
    `Percentage Increase (%)` = percentage_increase
  )

egenes_table <- egenes_comparison %>%
  gt() %>%
  tab_header(
    title = "Comparison of Significant eGenes per Cell Type (FDR < 0.05)"
  ) %>%
  cols_label(
    `Cell Type` = "Cell Type",
    `TensorQTL eGenes` = "TensorQTL eGenes",
    `JOBs eGenes` = "JOBs eGenes",
    `Percentage Increase (%)` = "Percentage Increase (%)"
  ) %>%
  fmt_number(
    columns = c(`TensorQTL eGenes`, `JOBs eGenes`),
    decimals = 0
  ) %>%
  fmt_number(
    columns = `Percentage Increase (%)`,
    decimals = 1,
    scale_by = -1  # For percentages
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
  )

egenes_table
write_tsv(sig_egenes, paste0(in_dir, "sig_egenes_top_fdr05_with_ties.tsv.gz"))
```

### eSNPs

***

```{r, eSNPs, echo = FALSE, message = FALSE, warning = FALSE}
sig_esnps_summary <- sig_esnps %>%
  group_by(cell) %>%
  summarise(
    n_esnps = n(),
    .groups = "drop"
  )

esnps_table <- sig_esnps_summary %>%
  gt() %>%
  tab_header(
    title = "Significant eSNPs per Cell Type (FDR < 0.05)"
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_esnps = "Number of eSNPs"
  ) %>%
  fmt_number(
    columns = n_esnps,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
    )

esnps_table 
write_tsv(sig_esnps, paste0(in_dir, "sig_esnps_fdr05.tsv.gz"))
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "12jobs.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
datatable(
  summary,
  caption = 'Table X. Rule benchmarking. Means and SD across replicates.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```