---
title: "JOBS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
  log_dir: null
  smr_dir: null
  p_smr: null
  p_heidi: null
  tensor_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(UpSetR)
library(patchwork)
library(ggrepel)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
tensor_dir <- params$tensor_dir 
log_dir <- params$log_dir
smr_dir <- params$smr_dir
p_smr <- as.numeric(params$p_smr)
p_heidi <- as.numeric(params$p_heidi)

cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")


local <- FALSE
if (local == TRUE) {
  in_dir <- '~/Desktop/test/jobs/'
  tensor_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/05TENSORQTL/logs/'
  bmark_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/benchmarks/'
}

options(encoding = "UTF-8") # For consistent UTF-8 handling remotely
Sys.setenv(R_GSCMD = "gs")
```

## JOBS pipeline steps {.tabset}

***

We used [JOBS](https://github.com/LidaWangPSU/JOBS) to integrate our sc-eQTL TensorQTL nominal data with
publicly available data [Wen et al 2024]() to boost our eQTL discovery power. 
The citation for JOBS is [Wang et al. (2025), DOI: 10.1016/j.xgen.2025.100820](https://www.cell.com/cell-genomics/fulltext/S2666-979X(25)00076-X).

***

1. `run_jobs`: Run Jobs script on TensorQTL nominial eQTL data and public bulk dataset
2. `merge_jobs`: Merge JOBS boosted sc-eQTL signal with original TensorQTL nominal output
  + Addresses JOBS SE = 0 artifact via gene-specific mean SE imputation (excluding zeros in mean calc)
  + Recalculates nominal p-values and per-gene FDR after imputation
3. `jobs_report`: Calculate per gene FDR (per cell type) and report sig eGenes (i.e. top eSNP per eGene)

Settings:

+ Run on L1 cell type eQTL only
+ Wen input file: `devBrain_eQTL_EUR_nominal_50HCP_FDR_0.05.txt.gz`

***

```{r, calculate_FDR, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
hybrid_files <- list.files(
  path = in_dir,
  pattern = "_jobs_mrg_fdr\\.tsv$",  # Fixed pattern: no .gz, no _hybrid
  full.names = TRUE
)

hybrid_files <- hybrid_files[
  str_detect(basename(hybrid_files), str_c(cell_types, collapse = "|"))
]

hybrid <- hybrid_files %>%
  map(~ {
    dt <- read_tsv(.x, col_types = cols(.default = "c")) %>%
      type_convert()
    dt %>%
      mutate(
        cell = str_extract(basename(.x), "^([^-]+)(?=-)"),
        ID = str_c(phenotype_id, variant_id, sep = "-")
      )
  }) %>%
  list_rbind()

## --------------------------------------------------------------
## 1. Significant eSNPs (FDR < 0.05)
## --------------------------------------------------------------
sig_esnps <- hybrid %>%
  filter(fdr < 0.05, !is.na(fdr)) %>%
  transmute(
    ID = str_c(phenotype_id, variant_id, sep = "-"),
    gene,
    snp = variant_id,
    beta = slope,
    se = slope_se,
    pval_nominal,
    fdr,
    cell
  )

sig_esnps_list <- sig_esnps %>% split(.$cell)

## --------------------------------------------------------------
## 2. Significant eGenes – two versions:
##   • sig_egenes:      all significant eSNPs (FDR < 0.05), allows ties
##   • sig_egenes_top:  one eSNP per gene (lowest FDR), no ties
## --------------------------------------------------------------

# All significant eSNPs (FDR < 0.05) — allows multiple per gene
sig_egenes <- hybrid %>%
  filter(fdr < 0.05, !is.na(fdr)) %>%
  transmute(
    ID = str_c(phenotype_id, variant_id, sep = "-"),
    gene,
    snp = variant_id,
    beta = slope,
    se = slope_se,
    pval_nominal,
    fdr,
    cell
  )

# Top eSNP per gene: lowest FDR, no ties
sig_egenes_top <- hybrid %>%
  filter(fdr < 0.05, !is.na(fdr)) %>%
  group_by(gene, cell) %>%
  slice_min(fdr, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    ID = str_c(phenotype_id, variant_id, sep = "-"),
    gene,
    snp = variant_id,
    beta = slope,
    se = slope_se,
    pval_nominal,
    fdr,
    cell
  )

sig_egenes_list     <- sig_egenes %>% split(.$cell)
sig_egenes_top_list <- sig_egenes_top %>% split(.$cell)

## --------------------------------------------------------------
## 3. TensorQTL-only comparison
## --------------------------------------------------------------
pval_comparisons <- list()
egenes_results <- tibble()

for (cell in cell_types) {
  tensor_file <- file.path(
    tensor_dir,
    paste0(cell, "_quantile_genPC_4_expPC_40/"),
    paste0(cell, "_quantile_perm.cis_qtl.txt.gz")
  )
  if (!file.exists(tensor_file)) next

  tensor_data <- read_tsv(tensor_file, col_types = cols(.default = "c")) %>%
    type_convert() %>%
    filter(qval < 0.05) %>%
    transmute(
      gene = phenotype_id,
      snp = variant_id,
      tensor_p = pval_nominal,
      ID = str_c(phenotype_id, variant_id, sep = "-")  # ENSURE ID IN tensor_data
    )

  jobs_p <- hybrid %>%
    filter(cell == !!cell, ID %in% tensor_data$ID) %>%
    transmute(ID, jobs_p = pval_nominal)

  comp <- tensor_data %>%
    left_join(jobs_p, by = "ID") %>%
    select(gene, snp, tensor_p, jobs_p)

  pval_comparisons[[cell]] <- comp

  tensor_genes <- unique(comp$gene)
  jobs_genes   <- unique(sig_egenes_top$gene[sig_egenes_top$cell == cell])

  set_overlap      <- intersect(tensor_genes, jobs_genes)
  set_tensor_only  <- setdiff(tensor_genes, jobs_genes)
  set_jobs_only    <- setdiff(jobs_genes, tensor_genes)

  egenes_results <- egenes_results %>%
    bind_rows(tibble(
      cell                  = cell,
      eqtl_pre_jobs         = length(tensor_genes),
      intersection          = length(set_overlap),
      tensor_only_na_jobs   = length(set_tensor_only),
      jobs_only_not_tensor  = length(set_jobs_only),
      eqtl_post_jobs        = length(set_overlap) + length(set_tensor_only) + length(set_jobs_only),
      percent_increase      = if (length(tensor_genes) > 0)
                                round(100 * (length(set_jobs_only) + length(set_overlap) - length(tensor_genes)) /
                                      length(tensor_genes), 1) else NA_real_
    ))
}

## --------------------------------------------------------------
## Combine for downstream tables/plots
## --------------------------------------------------------------
sig_esnps       <- bind_rows(sig_esnps_list)
sig_egenes_top  <- bind_rows(sig_egenes_top_list)

pval_all <- pval_comparisons %>%
  bind_rows(.id = "cell") %>%
  mutate(
    log_tensor_p = -log10(pmax(tensor_p, 1e-300)),
    log_jobs_p   = -log10(pmax(jobs_p,   1e-300))
  )

## Comparison summary
comparison_summary <- pval_all %>%
  group_by(cell) %>%
  summarise(
    tot_eQTL_pre_jobs = n(),
    non_na_pairs      = sum(!is.na(jobs_p) & !is.na(tensor_p)),
    na_jobs           = sum(is.na(jobs_p)),
    stronger_tensor   = sum(tensor_p < jobs_p & !is.na(jobs_p) & !is.na(tensor_p)),
    equal             = sum(abs(jobs_p - tensor_p) < 1e-10 & !is.na(jobs_p) & !is.na(tensor_p)),
    stronger_jobs     = sum(jobs_p < tensor_p & !is.na(jobs_p) & !is.na(tensor_p)),
    fdr_sig_jobs      = sum(jobs_p < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(cell, tot_eQTL_pre_jobs, non_na_pairs, na_jobs,
         stronger_tensor, equal, stronger_jobs, fdr_sig_jobs)

## NA summary
na_summary <- pval_comparisons %>%
  imap_dfr(~ tibble(cell = .y, na_rate = mean(is.na(.x$jobs_p))))
```

***

### eGenes

***

```{r, eGenes, echo = FALSE, message = FALSE, warning = FALSE}
# Display per-cell eGenes summary as GT table (fixed labels and computations)
egenes_table <- egenes_results %>%
  gt() %>%
  tab_header(
    title = "Table 1. eGene Counts per Cell Type: TensorQTL vs. JOBS Categories"
  ) %>%
  cols_label(
    cell = "Cell Type",
    eqtl_pre_jobs = "eQTLs pre-JOBS (TensorQTL)",
    intersection = "Shared eGenes",
    tensor_only_na_jobs = "TensorQTL-only eGenes",
    jobs_only_not_tensor = "JOBS-only eGenes",
    eqtl_post_jobs = "eQTLs post-JOBS",
    percent_increase = "% Increase"
  ) %>%
  fmt_number(
    columns = -cell,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

egenes_table

# Optional: Save results
# write_csv(egenes_results, paste0(tensor_dir, "../eGene_category_counts_per_cell.csv"))
```

***

### Totals

***

```{r, total_cnts, echo = FALSE, message = FALSE, warning = FALSE}
# Totals counts across all cell types 
grand_egenes <- data.frame(
  category = c("Intersect", "TensorQTL Only (NA in JOBS)", "JOBS Only (Not in TensorQTL)", "Total eGenes"),
  count = c(
    sum(egenes_results$intersection, na.rm = TRUE),
    sum(egenes_results$tensor_only_na_jobs, na.rm = TRUE),
    sum(egenes_results$jobs_only_not_tensor, na.rm = TRUE),
    sum(egenes_results$eqtl_post_jobs, na.rm = TRUE)  # Fixed: use eqtl_post_jobs
  )
)

# Display grand totals
grand_egenes %>%
  gt() %>%
  tab_header(title = "Table 2. Total eGenes Across All Cell Types") %>%
  cols_label(category = "Category", count = "Count") %>%
  fmt_number(columns = count, decimals = 0) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

# write_csv(grand_egenes, paste0(tensor_dir, "../eGene_category_grand_totals.csv"))
```

***

### eGenes plot

***

```{r, eGenes_plt, echo = FALSE, message = FALSE, warning = FALSE}
# Prepare data for stacked bar plot (fixed column references: use intersection for "Both")
plot_data_stacked <- egenes_results %>%
  select(cell, `TensorQTL Only` = tensor_only_na_jobs, `Both` = intersection, `JOBS Only` = jobs_only_not_tensor) %>%
  pivot_longer(
    cols = -cell,
    names_to = "category",
    values_to = "count"
  ) %>%
  mutate(category = factor(category, levels = c("TensorQTL Only", "Both", "JOBS Only")))

# Create stacked bar plot
p_stacked <- ggplot(plot_data_stacked, aes(x = cell, y = count, fill = category)) +
  geom_col(alpha = 0.8, width = 0.7) +
  scale_fill_manual(
    values = c("TensorQTL Only" = "#1f77b4", "Both" = "#2ca02c", "JOBS Only" = "#ff7f0e")
  ) +
  labs(
    title = "eGenes per Cell Type (FDR < 0.05)",
    x = "Cell Type",
    y = "Number of eGenes",
    fill = "Category"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(expand = c(0, 0), labels = scales::comma)

p_stacked
```
***

### eGenes (allow mult SNPs)

***

```{r, eGenes_with_ties, echo = FALSE, message = FALSE, warning = FALSE}
sig_egenes_summary <- sig_egenes %>%
  group_by(cell) %>%
  summarise(
    n_egenes = n(),
    .groups = "drop"
  )

egenes_table <- sig_egenes_summary %>%
  gt() %>%
  tab_header(
    title = "Table 3. Significant eGenes per Cell Type (FDR < 0.05)",
    subtitle = "eGenes with ≥1 sig eSNP. Retains ties for the lowest eSNP/gene FDR."
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_egenes = "Number of eGenes"
  ) %>%
  fmt_number(
    columns = n_egenes,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
  )

egenes_table
# write_tsv(sig_egenes, paste0(in_dir, "sig_egenes_top_fdr05_with_ties.tsv.gz"))
```

***

### eSNPs

***

```{r, eSNPs, echo = FALSE, message = FALSE, warning = FALSE}
sig_esnps_summary <- sig_esnps %>%
  group_by(cell) %>%
  summarise(
    n_esnps = n(),
    .groups = "drop"
  )

esnps_table <- sig_esnps_summary %>%
  gt() %>%
  tab_header(
    title = "Table 4. Significant eSNPs per Cell Type (FDR < 0.05)",
    subtitle = "All eSNPs (variants) per gene that pass FDR < 0.05, across all genes/cell's"
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_esnps = "Number of eSNPs"
  ) %>%
  fmt_number(
    columns = n_esnps,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
    )

esnps_table 
#write_tsv(sig_esnps, paste0(in_dir, "sig_esnps_fdr05.tsv.gz"))
```

***

### P-values {.tabset}

***

#### Table

```{r, p_val_comparision, echo = FALSE, message = FALSE, warning = FALSE}
# GT table
comparison_table <- comparison_summary %>%
  gt() %>%
  tab_header(
    title = "Table 5. P-Value Strength Comparison: TensorQTL vs. JOBS per Cell Type"
  ) %>%
  cols_label(
    cell = "Cell Type",
    tot_eQTL_pre_jobs = "Total eQTL (pre-JOBS)",
    non_na_pairs = "Intersection",
    na_jobs = "NA in JOBS",
    stronger_tensor = "P stronger in TensorQTL",
    equal = "Equal P",
    stronger_jobs = "P Stronger in JOBS",
    fdr_sig_jobs = "FDR sig in JOBS"
  ) %>%
  fmt_number(
    columns = -cell,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

# Display fixed table
comparison_table
```

***

#### Plot

***

```{r, p_val_comparision_plot, echo = FALSE, message = FALSE, warning = FALSE}
# A: Faceted plot of p-value distributions (-log10 for scale)
# p_plot <- pval_all |>
#   filter(!is.na(jobs_p)) |>
#   ggplot(aes(x = log_tensor_p, fill = "TensorQTL")) +
#   geom_density(alpha = 0.5, color = NA) +
#   geom_density(aes(x = log_jobs_p, fill = "JOBs"), alpha = 0.5, color = NA) +
#   facet_wrap(~ cell, scales = "free_y", ncol = 4) +
#   scale_fill_manual(values = c("TensorQTL" = "#1f77b4", "JOBs" = "#ff7f0e")) +
#   labs(
#     title = "Nominal P-Value Distributions: TensorQTL vs. JOBs",
#     x = "-log10(P-value)",
#     y = "Density",
#     fill = "Method"
#   ) +
#   theme_classic(base_size = 12) +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
#     strip.text = element_text(size = 10),
#     legend.position = "bottom"
#   )
# p_plot
```
***

### NAs {.tabset}

***

#### Table

***

```{r, p_val_comparision_NAs, echo = FALSE, message = FALSE, warning = FALSE}
# na_summary <- imap_dfr(pval_comparisons, ~ tibble(cell = .y, na_rate = mean(is.na(.x$jobs_p))))
# na_summary
```

***

#### Plot

```{r, p_val_comparision_NAs_plot, echo = FALSE, message = FALSE, warning = FALSE}
# pval_all %>% 
#     mutate(na_jobs = is.na(jobs_p)) %>% 
#     ggplot(aes(x = -log10(tensor_p), fill = na_jobs)) + 
#     geom_density(alpha = 0.7) + 
#     facet_wrap(~ cell) + 
#     labs(title = "TensorQTL Sig Pairs: NA in JOBs?") 
```
***

### SMR {.tabset}

***

```{r smr_results, echo = FALSE, message = FALSE, warning = FALSE}
# Function to read and summarize SMR results
summarize_smr <- function(cell_type, smr_dir, disorder) {
  file_path <- file.path(smr_dir, cell_type, paste0(cell_type, "_", disorder, ".smr"))
  if (file.exists(file_path)) {
    smr <- suppressMessages(read_tsv(file_path))
    n_probes <- nrow(smr)
    smr_sig <- smr |> filter(p_SMR < (p_smr / n_probes))
    smr_sig_heidi <- smr_sig |> filter(p_HEIDI > p_heidi)
    return(data.frame(
      Cell_Type = cell_type,
      Total_Probes = n_probes,
      Significant_SMR = nrow(smr_sig),
      Significant_SMR_HEIDI = nrow(smr_sig_heidi)
    ))
  } else {
    message("File not found for cell type / disorder: ", cell_type, " / ", disorder)
    return(data.frame(
      Cell_Type = cell_type,
      Total_Probes = 0,
      Significant_SMR = 0,
      Significant_SMR_HEIDI = 0
    ))
  }
}

# Function to read and summarize SMR results
extract_smr_log_params <- function(log_dir, cell_types, disorder) {
  # Initialize parameters to extract
  params <- c(
    "GWAS SNPs Included",
    "eQTL SNPs Included",
    "Individuals in LD Reference",
    "SNPs After Allele Check",
    "SNPs Excluded for Frequency",
    "Probes Analyzed"
  )
  
  # Initialize result data frame
  result <- data.frame(Parameter = params, stringsAsFactors = FALSE)
  
  # Process each cell type
  for (cell_type in cell_types) {
    log_file <- file.path(log_dir, paste0("smr_", cell_type, "_", disorder, ".log"))
    
    # Initialize default values
    values <- rep("Unknown", length(params))
    
    if (file.exists(log_file)) {
      log_content <- read_lines(log_file)
      
      # Extract parameters using regular expressions
      # GWAS SNPs Included
      gwas_snps <- str_extract(log_content, "GWAS summary data of (\\d+) SNPs to be included")
      gwas_snps <- gwas_snps[!is.na(gwas_snps)]
      if (length(gwas_snps) > 0) {
        values[1] <- str_extract(gwas_snps[1], "\\d+")
      }
      
      # eQTL SNPs Included
      eqtl_snps <- str_extract(log_content, "(\\d+) SNPs to be included from \\[.*\\.esi\\]")
      eqtl_snps <- eqtl_snps[!is.na(eqtl_snps)]
      if (length(eqtl_snps) > 0) {
        values[2] <- str_extract(eqtl_snps[1], "\\d+")
      }
      
      # Individuals in LD Reference
      individuals <- str_extract(log_content, "(\\d+) individuals to be included from \\[.*\\.fam\\]")
      individuals <- individuals[!is.na(individuals)]
      if (length(individuals) > 0) {
        values[3] <- str_extract(individuals[1], "\\d+")
      }
      
      # SNPs After Allele Check
      allele_check <- str_extract(log_content, "(\\d+) SNPs are included after allele checking")
      allele_check <- allele_check[!is.na(allele_check)]
      if (length(allele_check) > 0) {
        values[4] <- str_extract(allele_check[1], "\\d+")
      }
      
      # SNPs Excluded for Frequency
      freq_excluded <- str_extract(log_content, "(\\d+) SNPs.*with allele frequency differences > 0.20")
      freq_excluded <- freq_excluded[!is.na(freq_excluded)]
      if (length(freq_excluded) > 0) {
        values[5] <- str_extract(freq_excluded[1], "\\d+")
      }
      
      # Probes Analyzed
      probes <- str_extract(log_content, "SMR analysis \\(SMR and HEIDI tests\\) for (\\d+) probes")
      probes <- probes[!is.na(probes)]
      if (length(probes) > 0) {
        values[6] <- str_extract(probes[1], "\\d+")
      }
      
    } else {
      message("Log file not found for cell type: ", cell_type)
    }
    
    # Add values to result data frame
    result[[cell_type]] <- values
  }
  
  return(result)
}

get_heidi <- function(cell_type, smr_dir, disorder) {
  file_path <- file.path(smr_dir, cell_type, paste0(cell_type, "_", disorder, ".smr"))
  
  if (file.exists(file_path)) {
    smr <- suppressMessages(readr::read_tsv(file_path, show_col_types = FALSE))
    n_probes <- nrow(smr)
    
    smr %>%
      filter(p_SMR < (p_smr / n_probes), p_HEIDI > p_heidi & !is.na(p_HEIDI)) %>%
      mutate(Cell_Type = cell_type)
  } else {
    message("File not found for cell type / disorder: ", cell_type, " / ", disorder)
    return(NULL)
  }
}

# Get gene lookup file
gene_lookup_file <- '../../resources/sheets/gene_lookup_hg38.tsv' # Relative to Rmd file
if (file.exists(gene_lookup_file)) {
  
  gene_lookup_hg38_tbl <- suppressMessages(read_tsv(gene_lookup_file))
  
} else {
  
  library(biomaRt)
  
  ## Create lookup table 
  mart <- useMart('ENSEMBL_MART_ENSEMBL')
  mart <- useDataset('hsapiens_gene_ensembl', mart)
  
  gene_lookup_hg38_tbl <- getBM(
    mart = mart,
    attributes = c("ensembl_gene_id", "chromosome_name", "start_position", "end_position", 
                   "hgnc_symbol", "external_gene_name", "strand", "gene_biotype"),
    uniqueRows = TRUE)
  
  write_tsv(gene_lookup_hg38_tbl, gene_lookup_file)
  
}
```

***

#### SCZ {.tabset}

***

##### SMR summary

***

```{r smr_summary_scz, results='asis', echo=FALSE}
disorder <- 'scz'
summary_data <- bind_rows(
  lapply(cell_types, function(ct) summarize_smr(ct, smr_dir, disorder))
)
datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_scz, results='asis', echo=FALSE}
all_sig_results <- bind_rows(
  lapply(cell_types, function(ct) get_heidi(ct, smr_dir, disorder))
)
datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_scz, echo=FALSE}
gene_lookup_file <- '../../resources/sheets/gene_lookup_hg38.tsv'

sig_tbl <- all_sig_results |>
  left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
  select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)

datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_scz, echo=FALSE}
log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

#### BPD {.tabset}

***

##### SMR summary

***

```{r smr_summary_bpd, results='asis', echo=FALSE}
# disorder <- 'bpd'
# summary_data <- bind_rows(
#   lapply(cell_types, function(ct) summarize_smr(ct, smr_dir, disorder))
# )
# datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_bpd, results='asis', echo=FALSE}
# all_sig_results <- bind_rows(
#   lapply(cell_types, function(ct) get_heidi(ct, smr_dir, disorder))
# )
# datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_bpd, echo=FALSE}
# sig_tbl <- all_sig_results |>
#   left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
#   select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)
# 
# datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_bpd, echo=FALSE}
# log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
# datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

#### MDD{.tabset}

***

##### SMR summary

***

```{r smr_summary_mdd, results='asis', echo=FALSE}
# disorder <- 'mdd'
# summary_data <- bind_rows(
#   lapply(cell_types, function(ct) summarize_smr(ct, smr_dir, disorder))
# )
# datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_mdd, results='asis', echo=FALSE}
# all_sig_results <- bind_rows(
#   lapply(cell_types, function(ct) get_heidi(ct, smr_dir, disorder))
# )
# datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_mdd, echo=FALSE}
# sig_tbl <- all_sig_results |>
#   left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
#   select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)
# 
# datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_mdd, echo=FALSE}
# log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
# datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "12jobs.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
datatable(
  summary,
  caption = 'Table X. Rule benchmarking. Means and SD across replicates.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```