---
title: "JOBS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
  tensor_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(UpSetR)
library(patchwork)
library(ggrepel)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
tensor_dir <- params$tensor_dir 

cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")

local <- FALSE
if (local == TRUE) {
  in_dir <- '~/Desktop/test/jobs/'
  tensor_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/05TENSORQTL/logs/'
  bmark_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/benchmarks/'
}
```

## JOBS pipeline steps {.tabset}

***

We used [JOBS](https://github.com/LidaWangPSU/JOBS) to generate integrate our sc-eQTL TensorQTL nominial data with
publically available data [Wen et al 2024]() to boost our eQTL discovery power. 
The citation for JOBS is [Wang et al. (2025), DOI: 10.1016/j.xgen.2025.100820](https://www.cell.com/cell-genomics/fulltext/S2666-979X(25)00076-X).

***

1. `run_jobs`: Run Jobs script on TensorQTL nominial eQTL data and public bulk dataset
2. `jobs_report`: Calculate per gene FDR (per cell type) and report sig eGenes (i.e. top eSNP per eGene)

Settings:

+ Run on L1 cell type eQTL only
+ Wen input file: `devBrain_eQTL_EUR_nominal_50HCP_FDR_0.05.txt.gz`

***

```{r, calculate_FDR, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
# Load beta and SE files once
ref_beta <- read_tsv(paste0(in_dir, "jobs_ref_beta_genomewide.tsv.gz"))
ref_se <- read_tsv(paste0(in_dir, "jobs_ref_se_genomewide.tsv.gz"))

# Add gene column if missing (extract from ID)
if (!"gene" %in% colnames(ref_beta)) {
  ref_beta <- ref_beta %>%
    mutate(gene = str_remove(ID, "-.*"))
}

# Initialize lists for results
sig_esnps_list <- list()
sig_egenes_list <- list()
sig_egenes_top_list <- list()
pval_comparisons <- list()
egenes_results <- data.frame()

# Loop over cell types: Compute JOBS p-values/FDRs, sig eSNPs/eGenes, and TensorQTL comparisons
for (cell in cell_types) {
  message(paste("Processing cell:", cell))
  
  # Skip if column missing
  if (!cell %in% colnames(ref_beta)) {
    message(paste("Skipping", cell, "- column missing"))
    next
  }
  
  # Compute nominal p-values once per cell
  beta_col <- ref_beta[[cell]]
  se_col <- ref_se[[cell]]
  p_nom <- 2 * pnorm(-abs(beta_col / se_col))
  
  # Add p-value column to ref_beta (temporary for this cell)
  ref_beta_temp <- ref_beta %>%
    mutate(!!paste0(cell, "_p") := p_nom)
  
  # Compute per-gene FDR once per cell
  fdr_temp <- ref_beta_temp %>%
    group_by(gene) %>%
    mutate(fdr = p.adjust(.data[[paste0(cell, "_p")]], method = "BH")) %>%
    ungroup() %>%
    select(ID, gene, !!paste0(cell, "_p") := .data[[paste0(cell, "_p")]], fdr) %>%
    mutate(cell = cell)
  
  # Sig eSNPs: All FDR < 0.05 for this cell
  sig_esnps_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr))
  sig_esnps_list[[cell]] <- sig_esnps_cell
  
  # Sig eGenes: Top eSNP (min FDR) per gene for this cell, if any FDR < 0.05 (retains ties via slice_min)
  sig_egenes_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr)) %>%
    group_by(gene) %>%
    slice_min(fdr, n = 1, with_ties = TRUE) %>%  # Retains ties explicitly
    ungroup() |>
    arrange(fdr)
  sig_egenes_list[[cell]] <- sig_egenes_cell
  
  # Sig eGenes top: Exactly 1 top eSNP per gene (handles ties by sorting and slicing first)
  sig_egenes_cell_top <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr)) %>%
    arrange(fdr) %>%
    group_by(gene) %>%
    slice_head(n = 1) %>%
    ungroup() %>%
    arrange(fdr)
  sig_egenes_top_list[[cell]] <- sig_egenes_cell_top
  
  # TensorQTL comparison 
  tensor_file <- paste0(tensor_dir, cell, "_quantile_genPC_4_expPC_40/", cell, "_quantile_perm.cis_qtl.txt.gz")
  if (file.exists(tensor_file)) {
    tensor_data <- read_tsv(tensor_file, show_col_types = FALSE) %>%
      filter(qval < 0.05) %>%
      select(gene = phenotype_id, snp = variant_id, tensor_p = pval_nominal) %>%
      mutate(ID = paste(gene, snp, sep = "-"))
    
    # Match: Inner join on ID (only pairs in both)
    comparison_cell <- tensor_data %>%
      inner_join(ref_beta_temp %>% select(ID, !!paste0(cell, "_p")), by = "ID") %>%
      rename(jobs_p = !!paste0(cell, "_p")) %>%
      select(gene, snp, tensor_p, jobs_p)
    
    # Save per-cell tibble
    pval_comparisons[[cell]] <- comparison_cell
    
    # TensorQTL sig eGenes: unique genes
    tensor_genes_cell <- unique(comparison_cell$gene)
    
    # JOBS sig eGenes: unique genes with FDR < 0.05
    jobs_genes_cell <- unique(sig_egenes_cell$gene)
    
    # Compute sets
    set_overlap <- intersect(tensor_genes_cell, jobs_genes_cell)
    set_tensor_only <- setdiff(tensor_genes_cell, jobs_genes_cell)
    set_jobs_only <- setdiff(jobs_genes_cell, tensor_genes_cell)
    
    # Counts
    pre_eqtl <- length(tensor_genes_cell)  # Fixed: use unique genes, not nrow (which counts eSNPs)
    post_eqtl <- length(set_overlap) + length(set_tensor_only) + length(set_jobs_only)
    percent_increase <- ifelse(pre_eqtl > 0, round((post_eqtl - pre_eqtl) / pre_eqtl * 100, 1), NA)
    
    # Add to results
    egenes_results <- rbind(egenes_results, data.frame(
      cell = cell,
      eqtl_pre_jobs = pre_eqtl,
      intersection = length(set_overlap),
      tensor_only_na_jobs = length(set_tensor_only),
      jobs_only_not_tensor = length(set_jobs_only),
      eqtl_post_jobs = post_eqtl,  # Added missing column
      percent_increase = percent_increase,
      stringsAsFactors = FALSE
    ))
  } else {
    message(paste("TensorQTL file missing for", cell))
  }
}

# Combine into data frames
sig_esnps <- bind_rows(sig_esnps_list)
sig_egenes <- bind_rows(sig_egenes_list)
sig_egenes_top <- bind_rows(sig_egenes_top_list)

# Combine all p-values for plots/tables
pval_all <- bind_rows(pval_comparisons, .id = "cell") %>%
  mutate(
    log_tensor_p = -log10(pmax(tensor_p, 1e-300)),  # For plot (avoid inf)
    log_jobs_p = -log10(pmax(jobs_p, 1e-300))
  )

# Comparison summary (moved here to avoid re-computation)
comparison_summary <- pval_all %>%
  group_by(cell) %>%
  summarise(
    tot_eQTL_pre_jobs = n(),
    non_na_pairs = sum(!is.na(jobs_p) & !is.na(tensor_p), na.rm = TRUE),
    na_jobs = sum(is.na(jobs_p)),
    stronger_tensor = sum(tensor_p < jobs_p & !is.na(jobs_p) & !is.na(tensor_p), na.rm = TRUE),
    equal = sum(abs(jobs_p - tensor_p) < 1e-10 & !is.na(jobs_p) & !is.na(tensor_p), na.rm = TRUE),
    stronger_jobs = sum(jobs_p < tensor_p & !is.na(jobs_p) & !is.na(tensor_p), na.rm = TRUE),
    fdr_sig_jobs = sum(jobs_p < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(cell, tot_eQTL_pre_jobs, non_na_pairs, na_jobs, stronger_tensor, equal, stronger_jobs, fdr_sig_jobs)

# NA summary (moved here)
na_summary <- imap_dfr(pval_comparisons, ~ tibble(cell = .y, na_rate = mean(is.na(.x$jobs_p))))
```

***

### eGenes

***

```{r, eGenes, echo = FALSE, message = FALSE, warning = FALSE}
# Display per-cell eGenes summary as GT table (fixed labels and computations)
egenes_table <- egenes_results %>%
  gt() %>%
  tab_header(
    title = "eGene Counts per Cell Type: TensorQTL vs. JOBS Categories"
  ) %>%
  cols_label(
    cell = "Cell Type",
    eqtl_pre_jobs = "eQTLs pre-JOBS (TensorQTL)",
    intersection = "Shared eGenes",
    tensor_only_na_jobs = "TensorQTL-only eGenes",
    jobs_only_not_tensor = "JOBS-only eGenes",
    eqtl_post_jobs = "eQTLs post-JOBS",
    percent_increase = "% Increase"
  ) %>%
  fmt_number(
    columns = -cell,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

egenes_table

# Optional: Save results
# write_csv(egenes_results, paste0(tensor_dir, "../eGene_category_counts_per_cell.csv"))
```

***

### Totals

***

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Totals counts across all cell types 
grand_egenes <- data.frame(
  category = c("Intersect", "TensorQTL Only (NA in JOBS)", "JOBS Only (Not in TensorQTL)", "Total eGenes"),
  count = c(
    sum(egenes_results$intersection, na.rm = TRUE),
    sum(egenes_results$tensor_only_na_jobs, na.rm = TRUE),
    sum(egenes_results$jobs_only_not_tensor, na.rm = TRUE),
    sum(egenes_results$eqtl_post_jobs, na.rm = TRUE)  # Fixed: use eqtl_post_jobs
  )
)

# Display grand totals
grand_egenes %>%
  gt() %>%
  tab_header(title = "Grand Total eGenes Across All Cell Types") %>%
  cols_label(category = "Category", count = "Count") %>%
  fmt_number(columns = count, decimals = 0) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

# write_csv(grand_egenes, paste0(tensor_dir, "../eGene_category_grand_totals.csv"))
```

***

### eGenes plot

***

```{r, eGenes_plt, echo = FALSE, message = FALSE, warning = FALSE}
# Prepare data for stacked bar plot (fixed column references: use intersection for "Both")
plot_data_stacked <- egenes_results %>%
  select(cell, `TensorQTL Only` = tensor_only_na_jobs, `Both` = intersection, `JOBS Only` = jobs_only_not_tensor) %>%
  pivot_longer(
    cols = -cell,
    names_to = "category",
    values_to = "count"
  ) %>%
  mutate(category = factor(category, levels = c("TensorQTL Only", "Both", "JOBS Only")))

# Create stacked bar plot
p_stacked <- ggplot(plot_data_stacked, aes(x = cell, y = count, fill = category)) +
  geom_col(alpha = 0.8, width = 0.7) +
  scale_fill_manual(
    values = c("TensorQTL Only" = "#1f77b4", "Both" = "#2ca02c", "JOBS Only" = "#ff7f0e")
  ) +
  labs(
    title = "eGenes per Cell Type: TensorQTL vs. JOBS Categories (FDR < 0.05)",
    x = "Cell Type",
    y = "Number of eGenes",
    fill = "Category"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(expand = c(0, 0), labels = scales::comma)

p_stacked
```
***

### eGenes (allow mult SNPs)

***

```{r, eGenes_with_ties, echo = FALSE, message = FALSE, warning = FALSE}
sig_egenes_summary <- sig_egenes %>%
  group_by(cell) %>%
  summarise(
    n_egenes = n(),
    .groups = "drop"
  )

egenes_table <- sig_egenes_summary %>%
  gt() %>%
  tab_header(
    title = "Table 2. Significant eGenes per Cell Type (FDR < 0.05)",
    subtitle = "eGenes with ≥1 sig eSNP. Retains ties for the lowest eSNP/gene FDR."
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_egenes = "Number of eGenes"
  ) %>%
  fmt_number(
    columns = n_egenes,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
    )

egenes_table
#write_tsv(sig_egenes, paste0(in_dir, "sig_egenes_top_fdr05_with_ties.tsv.gz"))
```

***

### eSNPs

***

```{r, eSNPs, echo = FALSE, message = FALSE, warning = FALSE}
sig_esnps_summary <- sig_esnps %>%
  group_by(cell) %>%
  summarise(
    n_esnps = n(),
    .groups = "drop"
  )

esnps_table <- sig_esnps_summary %>%
  gt() %>%
  tab_header(
    title = "Table 3. Significant eSNPs per Cell Type (FDR < 0.05)",
    subtitle = "All eSNPs (variants) per gene that pass FDR < 0.05, across all genes/cell's"
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_esnps = "Number of eSNPs"
  ) %>%
  fmt_number(
    columns = n_esnps,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
    )

esnps_table 
#write_tsv(sig_esnps, paste0(in_dir, "sig_esnps_fdr05.tsv.gz"))
```

***

### Stats

```{r}
# GT table
comparison_table <- comparison_summary %>%
  gt() %>%
  tab_header(
    title = "P-Value Strength Comparison: TensorQTL vs. JOBS per Cell Type"
  ) %>%
  cols_label(
    cell = "Cell Type",
    tot_eQTL_pre_jobs = "Total eQTL (pre-JOBS)",
    non_na_pairs = "Intersection",
    na_jobs = "NA in JOBS",
    stronger_tensor = "P stronger in TensorQTL",
    equal = "Equal P",
    stronger_jobs = "P Stronger in JOBS",
    fdr_sig_jobs = "FDR sig in JOBS"
  ) %>%
  fmt_number(
    columns = -cell,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = px(12))

# Display fixed table
comparison_table
```

***

### Pval dist {.tabset}

***

#### tQTL vs. Jobs

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# A: Faceted plot of p-value distributions (-log10 for scale)
p_plot <- pval_all |>
  filter(!is.na(jobs_p)) |>
  ggplot(aes(x = log_tensor_p, fill = "TensorQTL")) +
  geom_density(alpha = 0.5, color = NA) +
  geom_density(aes(x = log_jobs_p, fill = "JOBs"), alpha = 0.5, color = NA) +
  facet_wrap(~ cell, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("TensorQTL" = "#1f77b4", "JOBs" = "#ff7f0e")) +
  labs(
    title = "Nominal P-Value Distributions: TensorQTL vs. JOBs",
    x = "-log10(P-value)",
    y = "Density",
    fill = "Method"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    strip.text = element_text(size = 10),
    legend.position = "bottom"
  )
p_plot
```
***

#### NAs {.tabset}

***

##### Summary

***

```{r, }
na_summary <- imap_dfr(pval_comparisons, ~ tibble(cell = .y, na_rate = mean(is.na(.x$jobs_p))))
na_summary
```

***

##### Plot

```{r}
pval_all %>% 
    mutate(na_jobs = is.na(jobs_p)) %>% 
    ggplot(aes(x = -log10(tensor_p), fill = na_jobs)) + 
    geom_density(alpha = 0.7) + 
    facet_wrap(~ cell) + 
    labs(title = "TensorQTL Sig Pairs: NA in JOBs?") 
```

### Benchmarking {.tabset}

***

#### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "12jobs.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
datatable(
  summary,
  caption = 'Table X. Rule benchmarking. Means and SD across replicates.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```