---
title: "JOBS report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(UpSetR)
library(patchwork)
library(ggrepel)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```

## JOBS pipeline steps {.tabset}

***

We used [JOBS](https://github.com/LidaWangPSU/JOBS) to generate integrate our sc-eQTL TensorQTL nominial data with
publically available data [Wen et al 2024]() to boost our eQTL discovery power. 
The citation for JOBS is [Wang et al. (2025), DOI: 10.1016/j.xgen.2025.100820](https://www.cell.com/cell-genomics/fulltext/S2666-979X(25)00076-X).

***

1. `run_jobs`: Run Jobs script on TensorQTL nominial eQTL data and public bulk dataset
2. `jobs_report`: Calculate per gene FDR (per cell type) and report sig eGenes (i.e. top eSNP per eGene)

Settings:

+ Run on L1 cell type eQTL only
+ Wen input file: `devBrain_eQTL_EUR_nominal_50HCP_FDR_0.05.txt.gz`

***

### Level 1 {.tabset}

***

```{r, calculate_FDR, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")

# Load beta and SE files
ref_beta <- read_tsv(paste0(in_dir, "jobs_ref_beta_genomewide.tsv.gz"))
ref_se <- read_tsv(paste0(in_dir, "jobs_ref_se_genomewide.tsv.gz"))

# Add gene column if missing (extract from ID)
if (!"gene" %in% colnames(ref_beta)) {
  ref_beta <- ref_beta %>%
    mutate(gene = str_remove(ID, "-.*"))
}

# Initialize lists for results
sig_esnps_list <- list()  # For all sig eSNPs (FDR < 0.05) per cell
sig_egenes_list <- list()  # For top eSNP per sig eGene per cell

# Loop over cell types
for (cell in cell_types) {
  message(paste("Processing cell:", cell))
  
  if (!cell %in% colnames(ref_beta)) {
    message(paste("Skipping", cell, "- column missing"))
    next
  }
  
  # Compute nominal p-values
  beta_col <- ref_beta[[cell]]
  se_col <- ref_se[[cell]]
  p_nom <- 2 * pnorm(-abs(beta_col / se_col))
  
  # Skip if all NA
  if (all(is.na(p_nom))) {
    message(paste("Skipping", cell, "- all p-values NA"))
    next
  }
  
  # Add p-value column to ref_beta
  ref_beta_temp <- ref_beta %>%
    mutate(!!paste0(cell, "_p") := p_nom)
  
  # Compute per-gene FDR
  fdr_temp <- ref_beta_temp %>%
    group_by(gene) %>%
    mutate(fdr = p.adjust(.data[[paste0(cell, "_p")]], method = "BH")) %>%
    ungroup() %>%
    select(ID, gene, !!paste0(cell, "_p") := .data[[paste0(cell, "_p")]], fdr) %>%
    mutate(cell = cell)
  
  # Sig eSNPs: All FDR < 0.05 for this cell
  sig_esnps_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr))
  sig_esnps_list[[cell]] <- sig_esnps_cell
  
  # Sig eGenes: Top eSNP (min FDR) per gene for this cell, if any FDR < 0.05
  sig_egenes_cell <- fdr_temp %>%
    filter(fdr < 0.05 & !is.na(fdr)) %>%
    group_by(gene) %>%
    slice_min(fdr, n = 1) %>%  # Top (lowest FDR) eSNP per gene
    ungroup()
  sig_egenes_list[[cell]] <- sig_egenes_cell
  
  # Warn on NAs in FDR
  if (any(is.na(fdr_temp$fdr))) {
    message(paste("Warning: NAs in FDR for", cell, "- check small gene groups"))
  }
}

# Combine into data frames
sig_esnps <- bind_rows(sig_esnps_list)
sig_egenes <- bind_rows(sig_egenes_list)
```

***

### eGenes

```{r, eGenes, echo = FALSE, message = FALSE, warning = FALSE}
sig_egenes_summary <- sig_egenes %>%
  group_by(cell) %>%
  summarise(
    n_egenes = n(),
    .groups = "drop"
  )

egenes_table <- sig_egenes_summary %>%
  gt() %>%
  tab_header(
    title = "Table 1. Significant eGenes per Cell Type (Top eSNP, FDR < 0.05)"
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_egenes = "Number of eGenes"
  ) %>%
  fmt_number(
    columns = n_egenes,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
  )

egenes_table
#write_tsv(sig_egenes, paste0(in_dir, "sig_egenes_top_fdr05.tsv.gz"))
```

***

### eSNPs

```{r, eSNPs, echo = FALSE, message = FALSE, warning = FALSE}
sig_esnps_summary <- sig_esnps %>%
  group_by(cell) %>%
  summarise(
    n_esnps = n(),
    .groups = "drop"
  )

esnps_table <- sig_esnps_summary %>%
  gt() %>%
  tab_header(
    title = "Significant eSNPs per Cell Type (FDR < 0.05)"
  ) %>%
  cols_label(
    cell = "Cell Type",
    n_esnps = "Number of eSNPs"
  ) %>%
  fmt_number(
    columns = n_esnps,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8f9fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14)
    )

esnps_table 
#write_tsv(sig_esnps, paste0(in_dir, "sig_esnps_fdr05.tsv.gz"))summary.csv"))
```