---
title: "SLDSR report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:   
  html_document:
    df_print: paged
params:
  in_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(gt)
in_dir  <- params$in_dir
```

## SLDSR pipeline steps 

***

1. `get_hg38_refs`: Download SLDSR hg38 reference files
2. `get_hg19_refs`: Download SLDSR hg38 reference files
3. `make_annot_maxCPP`: Make max CPP reference files from SuSiE output
4. `lift_hapmap3_snps`: Lift Hapmap 3 reference SNPs to hg38
5. `ldsr_ld_scores_hg38`: Generate LD score files for each cell type
6. `ldsr_strat_hg38_bl_v12`: Run stratified LD score regression with baseline version 1.2
7. `ldsr_report`: Summarise SLDSR score results per cell type

***

## Results {.tabset}

### CS95 {.tabset}

***

```{r, cs95_full_table, sldsr_table, echo = FALSE, message=FALSE}
# Generate table and plot
ldsr_tbl <- read_tsv(paste0(in_dir, 'ldsr_strat_hg38_bl_v12.CS95.summary.tsv')) |>
  separate(Category, into = c("annot", "file_name"), sep = "/", remove = FALSE) |>
  separate(file_name, into = c("cell_type", "suffix"), sep = "_hg38", remove = TRUE) |>
  mutate(disorder = toupper(str_extract(suffix, "scz|bpd|mdd"))) |>
  select(-Category, -suffix) |>
  relocate(cell_type, disorder) |>
  mutate(ldsr = if_else(`Coefficient_z-score` > 0, -log10(pnorm(`Coefficient_z-score`, lower.tail = FALSE)), 0)) 
```

***

#### SCZ {.tabset}

***

##### Table

***

```{r, cs95_scz_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'SCZ') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Schizophrenia S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, cs95_scz_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_scz_tbl <- ldsr_tbl |> filter(disorder == 'SCZ') 
  
ldsr_p_plt <- ggplot(ldsr_scz_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("SCZ") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_scz_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_scz_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("SCZ")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

#### BPD {.tabset}

***

##### Table

***

```{r, cs95_bpd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'BPD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Bipolar Disorder S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, cs95_bpd_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_bpd_tbl <- ldsr_tbl |> filter(disorder == 'BPD') 
  
ldsr_p_plt <- ggplot(ldsr_bpd_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("BPD") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_bpd_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_bpd_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("BPD")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

#### MDD {.tabset}

***

##### Table

***

```{r, cs95_mdd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'MDD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Major Depressive Disorder S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, cs95_mdd_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_mdd_tbl <- ldsr_tbl |> filter(disorder == 'MDD') 
  
ldsr_p_plt <- ggplot(ldsr_mdd_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("MDD") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_mdd_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_mdd_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("MDD")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

### MaxCPP {.tabset}

***

```{r, maxCPP_full_table, sldsr_table, echo = FALSE, message=FALSE}
# Generate table and plot
ldsr_tbl <- read_tsv(paste0(in_dir, 'ldsr_strat_hg38_bl_v12.maxCPP.summary.tsv')) |>
  separate(Category, into = c("annot", "file_name"), sep = "/", remove = FALSE) |>
  separate(file_name, into = c("cell_type", "suffix"), sep = "_hg38", remove = TRUE) |>
  mutate(disorder = toupper(str_extract(suffix, "scz|bpd|mdd"))) |>
  select(-Category, -suffix) |>
  relocate(cell_type, disorder) |>
  mutate(ldsr = if_else(`Coefficient_z-score` > 0, -log10(pnorm(`Coefficient_z-score`, lower.tail = FALSE)), 0)) 
```

***

#### SCZ {.tabset}

***

##### Table

***

```{r, maxCPP_scz_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'SCZ') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Schizophrenia S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, maxCPP_scz_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_scz_tbl <- ldsr_tbl |> filter(disorder == 'SCZ') 
  
ldsr_p_plt <- ggplot(ldsr_scz_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("SCZ") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_scz_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_scz_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("SCZ")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

#### BPD {.tabset}

***

##### Table

***

```{r, maxCPP_bpd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'BPD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Bipolar Disorder S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, maxCPP_bpd_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_bpd_tbl <- ldsr_tbl |> filter(disorder == 'BPD') 
  
ldsr_p_plt <- ggplot(ldsr_bpd_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("BPD") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_bpd_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_bpd_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("BPD")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

#### MDD {.tabset}

***

##### Table

***

```{r, maxCPP_mdd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'MDD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 1. Major Depressive Disorder S-LDSC Results",
    subtitle = "eQTL Max CPP Used as Annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***



***

##### Plot

***

```{r, maxCPP_mdd_plot, echo=FALSE, fig.dim=c(10,5)}
options(repr.plot.width = 8, repr.plot.height = 7)
ldsr_mdd_tbl <- ldsr_tbl |> filter(disorder == 'MDD') 
  
ldsr_p_plt <- ggplot(ldsr_mdd_tbl, aes(x = ldsr, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab(expression(-log[10](P))) +
  ggtitle("MDD") +
  geom_vline(xintercept=-log10(0.05 / nrow(ldsr_mdd_tbl)), linetype = "dashed", color = "black") +
  geom_vline(xintercept=-log10(0.05), linetype = "dotted", color = "black") 

ldsr_enrich_plt <- ggplot(ldsr_mdd_tbl, aes(x = Enrichment, y = cell_type)) +
  geom_bar(stat = "identity", width = .9, fill = "#648FFF", colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                width = .2) +
  theme_classic() +
  theme(axis.text.x = element_text(size=16),
        axis.text.y = element_text(size=16),
        axis.title = element_text(size=18),
        axis.title.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size=18, face="bold", hjust = 0.5)) +
  xlab('Enrichment') +
  ggtitle("MDD")

cowplot::plot_grid(ldsr_p_plt, ldsr_enrich_plt)
```

***

### test plot

```{r}
# --- Read and prepare table
ldsr_tbl <- read_tsv(paste0(in_dir, 'ldsr_strat_hg38_bl_v12.maxCPP.summary.tsv')) |>
  separate(Category, into = c("annot", "file_name"), sep = "/", remove = FALSE) |>
  separate(file_name, into = c("cell_type", "suffix"), sep = "_hg38", remove = TRUE) |>
  mutate(disorder = toupper(str_extract(suffix, "scz|bpd|mdd"))) |>
  select(-Category, -suffix) |>
  relocate(cell_type, disorder) |>
  mutate(ldsr = if_else(`Coefficient_z-score` > 0,
                        -log10(pnorm(`Coefficient_z-score`, lower.tail = FALSE)), 0))

# --- Custom colour palette
custom_palette <- c(
  "RG" = "#FF5959",
  "ExN-UL" = "#00B6EB",
  "InN" = "#3CBB75FF",
  "ExN-DL" = "#CEE5FD",
  "Endo-Peri" = "#B200ED",
  "MG" = "#F58231",
  "OPC" = "#FDE725FF"
)

# --- Assign main cluster to subclusters
ldsr_tbl <- ldsr_tbl |>
  mutate(main_type = case_when(
    str_detect(cell_type, "ExN-UL") ~ "ExN-UL",
    str_detect(cell_type, "ExN-DL") ~ "ExN-DL",
    str_detect(cell_type, "InN") ~ "InN",
    str_detect(cell_type, "RG") ~ "RG",
    str_detect(cell_type, "OPC") ~ "OPC",
    str_detect(cell_type, "MG") ~ "MG",
    str_detect(cell_type, "Endo-Peri") ~ "Endo-Peri",
    TRUE ~ cell_type
  ))

# --- Specify y-axis order
cell_order <- c(
  "ExN-UL", grep("^ExN-UL", ldsr_tbl$cell_type, value = TRUE),
  "ExN-DL", grep("^ExN-DL", ldsr_tbl$cell_type, value = TRUE),
  "InN", grep("^InN", ldsr_tbl$cell_type, value = TRUE),
  "RG", grep("^RG", ldsr_tbl$cell_type, value = TRUE),
  "OPC", "MG", "Endo-Peri"
) |> unique()

ldsr_tbl <- ldsr_tbl |>
  mutate(cell_type = factor(cell_type, levels = rev(cell_order)))  # reverse for top-down plotting

# --- Plot
ggplot(ldsr_tbl, aes(x = ldsr, y = cell_type, fill = main_type)) +
  geom_col(width = 0.7) +
  facet_wrap(~ disorder, scales = "free_x") +
  scale_fill_manual(values = custom_palette, name = "Cell Type") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    x = expression(-log[10](p)),
    y = "Cell Type",
    title = "Stratified LDSC Enrichment (maxCPP)",
    fill = "Main Type"
  )
```