---
title: "SLDSR report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:   
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(gt)
library(cowplot)
in_dir  <- params$in_dir
bmark_dir <- params$bmark_dir
```

## SLDSR pipeline steps 

***

1. `get_hg38_refs`: Download SLDSR hg38 reference files
2. `get_hg19_refs`: Download SLDSR hg38 reference files
3. `make_annot_maxCPP`: Make max CPP reference files from SuSiE output
4. `lift_hapmap3_snps`: Lift Hapmap 3 reference SNPs to hg38
5. `ldsr_ld_scores_hg38`: Generate LD score files for each cell type
6. `ldsr_strat_hg38_bl_v12`: Run stratified LD score regression with baseline version 1.2
7. `ldsr_report`: Summarise SLDSR score results per cell type

***

## Results {.tabset}

### MaxCPP {.tabset}

***

#### P-value

***

```{r, maxCPP_p_val_plot, echo = FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
# --- Read and prepare table
ldsr_tbl <- read_tsv(paste0(in_dir, 'ldsr_strat_hg38_bl_v12.maxCPP.summary.tsv')) |>
  separate(Category, into = c("annot", "file_name"), sep = "/", remove = FALSE) |>
  separate(file_name, into = c("cell_type", "suffix"), sep = "_hg38", remove = TRUE) |>
  mutate(disorder = toupper(str_extract(suffix, "scz|bpd|mdd|adhd|ocd"))) |>
  select(-Category, -suffix) |>
  relocate(cell_type, disorder) |>
  mutate(ldsr = if_else(`Coefficient_z-score` > 0,
                        -log10(pnorm(`Coefficient_z-score`, lower.tail = FALSE)), 0)) |>
  mutate(disorder = factor(disorder, levels = c("SCZ", "BPD", "MDD", "ADHD", "OCD")))

# --- Custom colour palette
# Main cluster colours match the UMAP cell-type colours used elsewhere.
# Trajectory colours are deliberately distinct from every main-cluster colour
# so pseudotime-bin bars can never be mistaken for a subcluster shade.
custom_palette <- c(
  'Glu-UL' = '#4363d8',
  'Glu-DL' = '#00B6EB',
  'NPC' = '#FF5959',
  'GABA' = '#3CBB75FF',
  'Endo-Peri' = '#B200ED',
  'MG' = '#F58231',
  'OPC' = '#FDE725FF',
  'Trajectory-to-Glu-UL' = '#E91E8C',
  'Trajectory-to-Glu-DL' = '#2F4F4F'
)

# --- Identify pseudotime trajectory cell types (checked BEFORE the main_type
# cascade below, since names like "NPC-to-Glu-UL-Q4-Bin1" contain "Glu-UL" as
# a substring and would otherwise be mis-classified as a Glu-UL subcluster)
ldsr_tbl <- ldsr_tbl |>
  mutate(
    is_trajectory = str_detect(cell_type, "-to-"),
    main_type = case_when(
      is_trajectory & str_detect(cell_type, "to-Glu-UL") ~ "Trajectory-to-Glu-UL",
      is_trajectory & str_detect(cell_type, "to-Glu-DL") ~ "Trajectory-to-Glu-DL",
      str_detect(cell_type, "Glu-UL") ~ "Glu-UL",
      str_detect(cell_type, "Glu-DL") ~ "Glu-DL",
      str_detect(cell_type, "GABA") ~ "GABA",
      str_detect(cell_type, "NPC") ~ "NPC",
      str_detect(cell_type, "OPC") ~ "OPC",
      str_detect(cell_type, "MG") ~ "MG",
      str_detect(cell_type, "Endo-Peri") ~ "Endo-Peri",
      TRUE ~ cell_type
    )
  )

# --- Specify y-axis order
# Subcluster patterns are anchored ("-[0-9]+$") so they can't accidentally
# re-catch the trajectory names (e.g. a loose "^NPC" grep would otherwise
# also match "NPC-to-Glu-UL-Q4-Bin1").
cell_order <- c(
  "Glu-UL", grep("^Glu-UL-[0-9]+$", ldsr_tbl$cell_type, value = TRUE),
  "Glu-DL", grep("^Glu-DL-[0-9]+$", ldsr_tbl$cell_type, value = TRUE),
  "GABA", grep("^GABA-[0-9]+$", ldsr_tbl$cell_type, value = TRUE),
  "NPC", grep("^NPC-[0-9]+$", ldsr_tbl$cell_type, value = TRUE),
  "OPC", "MG", "Endo-Peri",
  sort(grep("^NPC-to-Glu-UL", ldsr_tbl$cell_type, value = TRUE)),
  sort(grep("^NPC-to-Glu-DL", ldsr_tbl$cell_type, value = TRUE))
) |> unique()

ldsr_tbl <- ldsr_tbl |>
  mutate(cell_type = factor(cell_type, levels = rev(cell_order))) |>
  mutate(cluster_level = case_when(
    is_trajectory ~ "Trajectory",
    cell_type == main_type ~ "Main Cluster",
    TRUE ~ "Subcluster"
  )) |>
  mutate(cluster_level = factor(cluster_level, levels = c("Main Cluster", "Subcluster", "Trajectory")))

# --- Bonferroni correction per facet, derived from the data rather than
# hardcoded, so the divisor always matches however many groups are actually
# present in that facet (this previously drifted out of sync: 13 was used
# for a 12-group subcluster facet after a cluster was dropped upstream).
vline_data <- ldsr_tbl |>
  distinct(cluster_level, cell_type) |>
  count(cluster_level, name = "n_groups") |>
  mutate(intercept = -log10(0.05 / n_groups))

# --- Caption reporting the correction actually applied, built from the same
# vline_data used to draw the lines so the two can never disagree
p_val_caption <- vline_data |>
  mutate(lbl = paste0(cluster_level, ": 0.05/", n_groups)) |>
  pull(lbl) |>
  paste(collapse = "; ") |>
  (\(x) paste0("Dashed lines: Bonferroni threshold per facet (", x, "). Dotted line: nominal P = 0.05."))() |>
  str_wrap(width = 110)

# --- Plot
ggplot(ldsr_tbl, aes(x = ldsr, y = cell_type, fill = main_type)) +
  geom_col(width = 0.7, colour = 'black') +
  facet_grid(cluster_level ~ disorder, scales = "free_y", space = "free_y") +
  scale_fill_manual(values = custom_palette, name = "Cell Type") +
  theme_minimal(base_size = 12) +

  # facet-specific dashed/dotted thresholds
  geom_vline(
    data = vline_data,
    aes(xintercept = intercept, linetype = cluster_level),
    color = "black",
    inherit.aes = FALSE
  ) +
  scale_linetype_manual(
    values = c("Main Cluster" = "dashed", "Subcluster" = "dashed", "Trajectory" = "dashed")
  ) +

  # global nominal threshold
  geom_vline(
    xintercept = -log10(0.05),
    linetype = "dotted",
    color = "black"
  ) +

  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.2, "lines"),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    plot.caption = element_text(hjust = 0, size = 9, margin = margin(t = 10))
  ) +
  labs(
    x = expression(-log[10](P)),
    y = "Cell Type",
    title = "Stratified LDSR P (maxCPP)",
    caption = p_val_caption,
    fill = "Main Type",
    linetype = "Cluster Level"
  ) +
  coord_cartesian(xlim = c(0, 6))
```

***

#### Enrichment

***

```{r, maxCPP_enrich_plot, echo = FALSE, message=FALSE, fig.width=12, fig.height=10}

ggplot(ldsr_tbl, aes(x = Enrichment, y = cell_type, fill = main_type)) +
  geom_col(width = 0.7, colour = 'black') +
  geom_errorbar(aes(xmin = (Enrichment-Enrichment_std_error), 
                    xmax = (Enrichment+Enrichment_std_error)), 
                 width = .2) +
  facet_grid(cluster_level ~ disorder, scales = "free_y", space = "free_y") +
  scale_fill_manual(values = custom_palette, name = "Cell Type") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.2, "lines"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    x = 'Enrichment',
    y = "Cell Type",
    title = "Stratified LDSR Enrichment (maxCPP)",
    fill = "Main Type"
  )
```

***

#### Table {.tabset}

***

##### SCZ

***

```{r, maxCPP_scz_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'SCZ') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  select(-main_type, -cluster_level, -is_trajectory) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 7. Schizophrenia S-LDSC Results",
    subtitle = "Max CPP eQTL used as annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***

##### BPD 

***

```{r, maxCPP_bpd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'BPD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
    select(-main_type, -cluster_level, -is_trajectory) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 8. Bipolar Disorder S-LDSC Results",
    subtitle = "Max CPP eQTL used as annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***

##### MDD {.tabset}

***

```{r, maxCPP_mdd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'MDD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
    select(-main_type, -cluster_level, -is_trajectory) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 9. Major Depressive Disorder S-LDSC Results",
    subtitle = "Max CPP eQTL used as annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***

##### ADHD

***

```{r, maxCPP_adhd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'ADHD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  select(-main_type, -cluster_level, -is_trajectory) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 10. ADHD S-LDSC Results",
    subtitle = "95% CS eQTL used as annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***

##### OCD

***

```{r, maxCPP_ocd_table, sldsr_table, echo = FALSE, message=FALSE}
ldsr_tbl %>%
  filter(disorder == 'OCD') %>%
  # Round numeric columns to avoid excessive precision
  mutate(
    Prop._SNPs = signif(Prop._SNPs, 3),
    Prop._h2 = signif(Prop._h2, 3),
    Prop._h2_std_error = signif(Prop._h2_std_error, 3),
    Enrichment = signif(Enrichment, 3),
    Enrichment_std_error = signif(Enrichment_std_error, 3),
    Enrichment_p = signif(Enrichment_p, 3),
    Coefficient = signif(Coefficient, 3),
    Coefficient_std_error = signif(Coefficient_std_error, 3),
    `Coefficient_z-score` = signif(`Coefficient_z-score`, 3),
    ldsr = signif(ldsr, 3)
  ) %>%
  select(-main_type, -cluster_level, -is_trajectory) %>%
  # Create the gt table
  gt() %>%
  # Add table header and subtitle
  tab_header(
    title = "Table 11. OCD S-LDSC Results",
    subtitle = "95% CS eQTL used as annotation"
  ) %>%
  # Rename columns for display
  cols_label(
    cell_type = "Cell Type",
    disorder = "Disorder",
    Prop._SNPs = "Prop. SNPs",
    Prop._h2 = "Prop. Heritability",
    Prop._h2_std_error = "Prop. Heritability SE",
    Enrichment = "Enrichment",
    Enrichment_std_error = "Enrichment SE",
    Enrichment_p = "Enrichment p-value",
    Coefficient = "Coefficient",
    Coefficient_std_error = "Coefficient SE",
    `Coefficient_z-score` = "Coefficient z-score",
    ldsr = "LDSR"
  ) %>%
  # Format small values in scientific notation with 2 decimal places
  fmt_scientific(
    columns = c(Prop._SNPs, Prop._h2, Prop._h2_std_error, Coefficient, Coefficient_std_error),
    decimals = 2
  ) %>%
  # Format other numeric columns with 3 decimal places
  fmt_number(
    columns = c(Enrichment, Enrichment_std_error, Enrichment_p, `Coefficient_z-score`, ldsr),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  # Customize table appearance
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3),
    heading.align = "center",
    column_labels.font.weight = "bold"
  )
```

***

### Benchmarking {.tabset}

***

#### Table

***

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "09sldsr.*\\.txt", full.names = TRUE)

# Rule names as declared in the Snakefile (see `benchmark:` directives).
# Matched by longest-prefix first, since several rule names share a common
# prefix with another rule (e.g. get_hg38_refs / get_hg19_refs both start
# with "get_hg", and ldsr_ld_scores_hg38 is itself a prefix of what
# ldsr_strat_hg38_bl_v12's benchmark filename would also match) - a naive
# "first token split on _" match, as used in the SMR benchmarking template,
# would collapse distinct rules into one bucket here. lift_hapmap3_snps has
# no benchmark: directive in the Snakefile, so it produces no file to match.
known_rules <- c(
  "ldsr_strat_hg38_bl_v12",
  "ldsr_ld_scores_hg38",
  "get_hg38_refs",
  "get_hg19_refs",
  "make_annot",
  "sldsr_report"
)
known_rules <- known_rules[order(-nchar(known_rules))]

match_rule <- function(rule_full, known_rules) {
  hit <- known_rules[str_starts(rule_full, known_rules)]
  if (length(hit) > 0) hit[1] else rule_full
}

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # parts[2] holds things like:
  # "get_hg38_refs", "make_annot_Glu-UL_1",
  # "ldsr_ld_scores_hg38_Glu-UL_maxCPP_1", "ldsr_strat_hg38_bl_v12_Glu-UL_maxCPP_scz"
  rule_full <- parts[2]
  rule <- match_rule(rule_full, known_rules)

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})

# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
summary %>%
  gt() %>%
  tab_header(
    title = "Table 12. Benchmarking",
    subtitle = "Means and SD across replicates"
  ) %>%
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) %>%
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)
```
