---
title: "Parse Testing"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output: html_document
---

## Experiment Details {.tabset}

***

+ 96 samples, on MEGA kit v1, 8 of which are Claire's FCX samples
+ Sequenced 8 (out of 16 total) sub-libraries on a single lane (only half the data here)
+ Parse workflow has a low [doublet rate](https://support.parsebiosciences.com/hc/en-us/articles/360053107311-What-is-the-expected-doublet-rate#:~:text=Doublet%20rates%20are%20low%2C%20less,through%20the%20Whole%20Transcriptome%20workflow.): ~3% per 100K cells

### pre-Seurat {.tabset}

#### Agg list {.tabset}

***

+ Boxplots generated from: agg_samp_ana_summary.csv

***

```{r agg_list, results='asis', fig.dim = c(10,10), echo=FALSE}
cowplot::plot_grid(plotlist = boxplt_lst)
```

***

#### Rnd1 heatmaps {.tabset}

***

+ Red: Cell count
+ Purple: Transcript count
+ Green: Cell loading (cells/µl)

***

```{r rn1_heatmaps, results='asis', fig.dim = c(10,10), echo=FALSE}
rnd1_hmap
```

***

#### Genes by cells 

***

```{r genes_by_cells, results='asis', fig.dim = c(10,10), echo=FALSE}
gene_by_cells_plot
```

***

#### Median Counts 

***

```{r median_cnts, results='asis', fig.dim = c(10,10), echo=FALSE}
knitr::kable(median_cnt_tbl)
```

***

#### Tscp / gene boxplots

***

```{r tscp_gene_boxplt, results='asis', fig.dim = c(10,10), echo=FALSE}
tscp_gene_boxplt
```

***

### Seurat {.tabset}

#### Outliers {.tabset}

Thresholds:

+ UMI - 3 Median absolute deviations
+ Genes - 3 Median absolute deviations
+ Mito - 5%
+ Ribo - 5%

***

```{r outlier_plot, results='asis', fig.dim = c(10,10), echo=FALSE}
outlier_plt 
```

***

#### Counts

*** 

+ Raw counts (after Parse filtering)

```{r raw_counts, results='asis', fig.dim = c(10,10), echo=FALSE}
knitr::kable(counts_preFilt_tbl)
```

***

+ Outlier counts

```{r outlier_counts, results='asis', fig.dim = c(10,10), echo=FALSE}
#knitr::kable(outlier_cnts_tbl)
```

***

+ Counts (after applying our filtering)

```{r counts_post_filtering, results='asis', fig.dim = c(10,10), echo=FALSE}
knitr::kable(counts_postFilt_tbl)
```

***


#### Resolution pre-hramony {.tabset}

+ Checking res params: 0.3, 0.5, 0.8

***

```{r resolution_cnts, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("  \n##### Counts **", i , "**\n", sep = "")
  clust_cnts <- get_meta_col_counts(seurat_obj, paste0('RNA_snn_res.', i))
  print(knitr::kable(clust_cnts))
  cat("  \n")
}
```

```{r resolution, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("  \n##### Plots **", i , "**\n", sep = "")
  plot_subset <- res_plotlist[grep(i, names(res_plotlist), ignore.case = TRUE)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("  \n")
}
```

#### Integration Plots {.tabset}

+ Integration done on sample column
+ Should there be other considerations for Parse assays (e.g. sublibrary, plate round etc.)

***

```{r Integration, fig.dim = c(20, 20), echo = FALSE, results="asis"}
for(i in resolution_set) {
  cat("  \n##### Plots **", i , "**\n", sep = "")
  plot_subset <- integration_plotlist[grep(i, names(integration_plotlist), ignore.case = TRUE)]
  plot(cowplot::plot_grid(plotlist = plot_subset, ncol = 2))
  cat("  \n")
}
```

#### Cluster Assignment {.tabset}

##### Biol Psych vln

***
```{r biol_pysch_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(biol_psych_plot)
```

***

##### Public 2017 vln

***

```{r public_vln_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(public_vln_plot)
```

***

##### Label transfer 1

***

```{r label_transfer_1, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(cowplot::plot_grid(harm_umap, cam_umap, poulio_umap, nowakowski_umap))
```

##### Label transfer 2

***

```{r label_transfer_2, fig.dim = c(20, 60), echo = FALSE, results="asis"}
print(cowplot::plot_grid(cam_fcx_plt, cam_ge_plt, cam_hip_plt, cam_tha_plt, cam_cer_plt, ncol = 1))
```


##### Claire's FCX clusters

***
```{r fcx_only_plots, fig.dim = c(20, 20), echo = FALSE, results="asis"}
print(cowplot::plot_grid(harm_umap, fcx_only_umap))
```