---
title: "pi1 Enrichment Report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  obrien_dir: null
  ziffra_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(cowplot)
obrien_dir <- params$obrien_dir
ziffra_dir <- params$ziffra_dir
bmark_dir <- params$bmark_dir

custom_palette <- c(
  "RG" = "#FF5959",
  "ExN-UL" = "#00B6EB",
  "InN" = "#3CBB75FF",
  "ExN-DL" = "#CEE5FD",
  "Endo-Peri" = "#B200ED",
  "MG" = "#F58231",
  "OPC" = "#FDE725FF"
)
```

### eQTL replication pipeline steps {.tabset}

***

1. `dwnld_obrien`: Download O'Brien 2018 bulk dev-brain eQTL files  PMID:30419947
2. `dwnld_bryois`: Download Bryois 2022 sc-adult-brain eQTL files   PMID:35915177
3. `dwnld_ziffra`: Download Ziffra 2021 snATAC-seq peaks files     PMID:34616060
4. `dwnld_wen`: Download Wen 2024 bulk dev-brain eQTL files      PMID:38781368
5. `snp_lookup`: Generate snp lookup for eQTL / ATAC-seq peak enrichments
6. `atac_enrich`: Test for eQTL enrichments in Ziffra ATAC-seq peaks
7. `cat_nom_qtl`: Combine per-chr nominal eQTL parquet files into gzip file for pi1 enrichment
8. `pi1_enrich_obrien`: Calc. pi1 enrichments between sn-eQTL and O'Brien 2018 bulk brain eQTL
9. `pi1_enrich_wen`: Calc. pi1 enrichments between sn-eQTL and Wen 2024 bulk brain eQTL
10. `pi1_enrich_bryois`: Calc. pi1 enrichments between sn-eQTL and Bryois 2022 adult brain sn-eQTL

Settings:

+ Expression PC: 40 (see tensorQTL_report.html)
+ Genotype PCs: 4 (see genotyping_post_impute_report.html)
+ Norm methods: quantile

***

### Ziffra {.tabset}

***

#### Table

***

```{r, ziffra_enrichments, echo = FALSE, message = FALSE, warning = FALSE}
ziffra_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/06QTL-REPLICATION/'
cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")

# Create a function to read and add cell_type identifier (optional)
read_rds_file <- function(cell_type) {
  file_path <- paste0(ziffra_dir, cell_type, '_quantile_genPC_4_expPC_40.ziffra_peak_enrich.rds')
  read_rds(file_path) %>%
    mutate(source_cell_type = cell_type) # Optional: add column to track source
}

# Read all files and bind rows
combined_data <- map_dfr(cell_types, read_rds_file)

# View the result
print(combined_data)
```

***

#### Heatmap

***

```{r, ziffra_heatmap, echo = FALSE, message = FALSE, warning = FALSE}
ziffra_plt <- combined_data %>%
  mutate(
    logp = -log10(p_value),
    sig = ifelse(p_value < 0.05, "*", ""),
    peak_cell_type = str_remove(peak_cell_type, "_MACSpeaks")
  ) |>
  complete(eqtl_cell_type, peak_cell_type) |>
  ggplot(aes(x = peak_cell_type, y = eqtl_cell_type, fill = fold_enrichment)) +
  geom_tile(color = "black", linewidth = 0.3, na.rm = FALSE) +
  geom_text(aes(label = ifelse(!is.na(p_value) & p_value < 0.05, "*", "")),
            vjust = 0.5, size = 4, color = "black") +
  scale_fill_viridis_c(
    option = "C",
    direction = 1,
    name = "Fold Enrichment",
    na.value = "white"   # blanks appear white
  )+
  labs(
    x = "ATAC-seq peak cell type",
    y = "eQTL cell type",
    title = "eQTL enrichment in ATAC-seq peaks"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
ziffra_plt
```

***

#### Bar plot

```{r, ziffra_barplot, echo = FALSE, message = FALSE, warning = FALSE}
bar_plt <- combined_data %>%
  mutate(
    logp = -log10(p_value),
    sig = ifelse(p_value < 0.05, "*", ""),
    peak_cell_type = str_remove(peak_cell_type, "_MACSpeaks")
  ) |> ggplot(aes(x = eqtl_cell_type, y = fold_enrichment, fill = peak_cell_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, color = "black") +
  geom_text(aes(label = sig),
            position = position_dodge(width = 0.8), hjust = 0.4,
            size = 5, color = "black") +
  labs(
    title = "eQTL enrichment in ATAC-seq peaks",
    x = "eQTL cell type",
    y = "Fold enrichment",
    fill = "snATAC-seq peak cell type"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) + coord_flip()
bar_plt
```

***

### O'Brien {.tabset}

```{r, obrien_enrichments, echo = FALSE, message = FALSE, warning = FALSE}
obrien_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/06QTL-REPLICATION/'
cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")

# Create a function to read and add cell_type identifier (optional)
read_rds_file <- function(cell_type) {
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_40_pi1_enrich.rds')
  read_rds(file_path) %>%
    mutate(source_cell_type = cell_type) # Optional: add column to track source
}

# Read all files and bind rows
pi1_summary_tbl <- map_dfr(cell_types, read_rds_file)

# View the result
print(pi1_summary_tbl)
```

***

#### Heatmap

***

```{r, obrien_heatmap, echo = FALSE, message = FALSE, warning = FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_40_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}

# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)
pi1_long <- pi1_result_tbl %>%
  dplyr::select(cell_type, pi1_forward, pi1_reverse) %>%
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") %>%
  mutate(direction = recode(direction,
                            pi1_forward = "Forward",
                            pi1_reverse = "Reverse"))

ggplot(pi1_long, aes(x = cell_type, y =  direction, fill = pi1)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
            color = "black", size = 3.5) +
  scale_fill_gradientn(
    colors = c("white", "skyblue", "dodgerblue3", "darkblue"),
    na.value = "grey80",
    limits = c(0, 1),
    name = expression(pi[1])
  ) +
  coord_equal() + 
  labs(x = "Cell type",  y = "Direction") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank()
  ) 
```

***

#### Forward histograms

***

```{r, obrien_hist_forward, echo = FALSE, message = FALSE, warning = FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_40_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}


# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)

plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_forward[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_forward, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_forward, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_forward, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```

***

#### Reverse histograms

***

```{r, obrien_hist_reverse, echo = FALSE, message = FALSE, warning = FALSE}
plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_forward[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_reverse, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_reverse, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_reverse, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```