---
title: "Pseudotime eQTL Report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
params:
  in_dir: null
  cell_type_perm_dir: null
---

***

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(gt)
library(patchwork)
library(ggrepel)
library(UpSetR)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

in_dir              <- params$in_dir
cell_type_perm_dir  <- params$cell_type_perm_dir

trajectories <- c("NPC-to-Glu-DL", "NPC-to-Glu-UL")
q_values     <- c(4, 5, 6)
exp_pcs      <- c(10, 20, 30, 40, 50)
geno_pc      <- 4
norm_method  <- "quantile"

cell_type_map <- c(
  "NPC-to-Glu-DL" = "Glu-DL",
  "NPC-to-Glu-UL" = "Glu-UL"
)

bin_palette <- setNames(
  colorRampPalette(c("#FDE725", "#21908C", "#440154"))(6),
  as.character(1:6)
)

# Load gene symbol lookup
gene_lookup_file <- "../../resources/sheets/gene_lookup_hg38.tsv"
gene_lookup <- read_tsv(gene_lookup_file, show_col_types = FALSE) |>
  select(ensembl_gene_id, symbol = external_gene_name) |>
  filter(!is.na(symbol) & symbol != "") |>
  distinct(ensembl_gene_id, .keep_all = TRUE)
message("Gene lookup loaded: ", nrow(gene_lookup), " entries")

get_symbol <- function(ensembl_ids) {
  syms <- gene_lookup$symbol[match(ensembl_ids, gene_lookup$ensembl_gene_id)]
  ifelse(is.na(syms) | syms == "", ensembl_ids, syms)
}
```

## 1. Pipeline Overview {.tabset}

***

This report summarises the pseudotime eQTL analysis for two neurogenic trajectories
derived from the prenatal human cerebral cortex snRNA-seq atlas (Cameron et al.).

***

### Methods

**Pseudotime analysis**

Cells were subsetted to two neurogenic trajectories from the full 679,738-cell atlas:

- **NPC-to-Glu-DL**: NPC (n=103,598) + Glu-DL (n=136,070) = 239,668 cells
- **NPC-to-Glu-UL**: NPC (n=103,598) + Glu-UL (n=271,180) = 374,778 cells

Pseudotime was computed using [Palantir](https://github.com/dpeerlab/Palantir) with:

- Root cell: cell with highest PAX6 expression in NPC population
- Terminal cell: cell with highest HS3ST4 (Glu-DL) or SATB2 (Glu-UL) expression
- Harmony-corrected PCA as diffusion map input (n_pcs=30, n_neighbors=10)
- 1,000 waypoints

**Pseudobulk generation**

Cells were binned into equal-cell-number quantiles (Q4, Q5, Q6) based on Palantir
pseudotime. Raw counts were summed per sample per bin. Only samples with matching
genotype data were retained.

**eQTL mapping**

cis-eQTL mapping was performed using [TensorQTL](https://github.com/broadinstitute/tensorqtl):

- Normalisation: TMM → log2 → quantile
- cis-window: ±500 kb
- Covariates: PCW, sex, 4 genotype PCs, N expression PCs (sensitivity: 10–50)
- Permutation-based FDR correction (q-value < 0.05)
- Selected expression PCs: bin-specific, chosen per bin by maximum eGene count at FDR < 0.05 (see Section 3)

***

### Parameters

```{r params_table, echo=FALSE, message=FALSE, warning=FALSE}
tibble(
  Parameter = c("Trajectories", "Quantile values", "Expression PCs tested",
                "Selected expression PCs", "Genotype PCs",
                "Normalisation", "cis-window", "FDR threshold"),
  Value     = c(paste(trajectories, collapse=", "),
                paste(q_values, collapse=", "),
                paste(exp_pcs, collapse=", "),
                "Bin-specific (see Section 3)",
                as.character(geno_pc), norm_method, "±500 kb", "q-value < 0.05")
) |>
  gt() |>
  tab_header(title = "Analysis parameters") |>
  cols_label(Parameter = "Parameter", Value = "Value") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels())
```

***

## 2. Pseudotime QC {.tabset}

***

```{r pseudotime_qc, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (traj in trajectories) {
  cat(sprintf("\n\n### %s {.tabset}\n\n***\n\n", traj))

  cat("#### Pseudotime UMAP\n\n***\n\n")
  pt_fig <- file.path(in_dir, traj, "pseudotime_umap.png")
  if (file.exists(pt_fig)) {
    cat(sprintf('![Pseudotime UMAP](%s)\n\n', pt_fig))
  } else {
    cat(sprintf("*Figure not found: %s*\n\n", pt_fig))
  }

  cat("#### Pseudobulk QC\n\n***\n\n")
  pb_fig <- file.path(in_dir, traj, "pseudobulk", "pseudobulk_qc.png")
  if (file.exists(pb_fig)) {
    cat(sprintf('![Pseudobulk QC](%s)\n\n', pb_fig))
  } else {
    cat(sprintf("*Figure not found: %s*\n\n", pb_fig))
  }

  cat("#### Marker genes\n\n***\n\n")
  marker_fig <- file.path(in_dir, traj, "pseudotime_markers.png")
  if (file.exists(marker_fig)) {
    cat(sprintf('![Marker gene expression](%s)\n\n', marker_fig))
  } else {
    cat(sprintf("*Figure not found: %s*\n\n", marker_fig))
  }

  cat("#### Bin Summary\n\n***\n\n")
  pb_csv <- file.path(in_dir, traj, "pseudobulk", "pseudobulk_summary.csv")
  if (file.exists(pb_csv)) {
    read_csv(pb_csv, show_col_types = FALSE) |>
      select(-trajectory) |>
      gt() |>
      tab_header(title = sprintf("Pseudobulk bin summary: %s", traj)) |>
      cols_label(
        quantile_n              = "Q",
        bin                     = "Bin",
        n_cells_total           = "Cells (total)",
        n_samples               = "Samples",
        min_cells_per_sample    = "Min cells/sample",
        median_cells_per_sample = "Median cells/sample",
        max_cells_per_sample    = "Max cells/sample",
        n_samples_below_10      = "Samples < 10 cells"
      ) |>
      tab_style(style = cell_text(weight = "bold"),
                locations = cells_column_labels()) |>
      print()
    cat("\n\n")
  } else {
    cat(sprintf("*Summary CSV not found: %s*\n\n", pb_csv))
  }
  cat("***\n\n")
}
```

***

## 3. Sensitivity Analysis {.tabset}

***

eGene counts at FDR < 0.05 across expression PC values (10–50), stratified by
trajectory, Q value, and bin.

***

```{r sensitivity_functions, echo=FALSE, message=FALSE, warning=FALSE}

read_perm_log <- function(in_dir, trajectory, quantile_n, bin, exp_pc, geno_pc = 4) {
  log_file <- file.path(
    in_dir, trajectory, "tensorqtl", "perm",
    sprintf("Q%d_bin%d_genPC%d_expPC%d", quantile_n, bin, geno_pc, exp_pc),
    sprintf("Q%d_bin%d_perm.tensorQTL.cis.log", quantile_n, bin)
  )
  if (!file.exists(log_file)) { message("Not found: ", log_file); return(NULL) }
  lines <- readLines(log_file, warn = FALSE)

  extract <- function(pattern, lines, type = "integer") {
    line <- str_subset(lines, pattern)
    if (length(line) == 0) return(NA)
    val <- str_extract(tail(line, 1), "[0-9]+")
    if (type == "integer") as.integer(val) else as.numeric(val)
  }
  extract_last_num <- function(pattern, lines, type = "numeric") {
    line <- str_subset(lines, pattern)
    if (length(line) == 0) return(NA)
    val <- str_extract(tail(line, 1), "[0-9\\.eE+-]+$")
    if (type == "integer") as.integer(val) else as.numeric(val)
  }

  tibble(
    trajectory   = trajectory, quantile_n = quantile_n,
    bin          = bin,        exp_pc     = exp_pc,
    geno_pc      = geno_pc,
    genes_tested = extract("phenotypes tested:", lines),
    samples      = extract("\\* [0-9]+ samples", lines),
    covariates   = extract("\\* [0-9]+ covariates", lines),
    variants     = extract("\\* [0-9]+ variants", lines),
    egenes_fdr05 = extract_last_num("QTL phenotypes @ FDR 0.05:", lines, "integer"),
    beta_cor     = extract_last_num("Correlation between Beta-approximated", lines),
    prop_sig     = extract_last_num("Proportion of significant phenotypes", lines),
    min_p        = extract_last_num("min p-value threshold @ FDR 0.05:", lines)
  )
}

build_sensitivity_tbl <- function(in_dir, trajectories, q_values, exp_pcs, geno_pc = 4) {
  rows <- list()
  for (traj in trajectories) for (n in q_values) for (b in seq_len(n)) for (pc in exp_pcs) {
    row <- read_perm_log(in_dir, traj, n, b, pc, geno_pc)
    if (!is.null(row)) rows <- c(rows, list(row))
  }
  bind_rows(rows)
}

sensitivity_tbl <- build_sensitivity_tbl(in_dir, trajectories, q_values, exp_pcs)
if (nrow(sensitivity_tbl) == 0) stop("No tensorQTL log files found. Check in_dir: ", in_dir)
message("Loaded sensitivity data: ", nrow(sensitivity_tbl), " rows")

# ── Single source of truth for "best" expPC per (trajectory, quantile_n, bin) ──
# Selected by maximum eGene count at FDR < 0.05, same criterion as the
# "Best expPC per bin" table below. Every section from here on reads from
# this lookup rather than a separately hardcoded value.

best_exp_pc_tbl <- sensitivity_tbl |>
  group_by(trajectory, quantile_n, bin) |>
  slice_max(egenes_fdr05, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(trajectory, quantile_n, bin, exp_pc)

get_best_exp_pc <- function(traj, n, b) {
  best_exp_pc_tbl |>
    filter(trajectory == traj, quantile_n == n, bin == b) |>
    pull(exp_pc)
}
```

```{r sensitivity_plots, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=14, fig.height=5}
for (traj in trajectories) {
  cat(sprintf("\n\n### %s {.tabset}\n\n***\n\n", traj))
  traj_tbl <- sensitivity_tbl |> filter(trajectory == traj)

  cat("#### Sensitivity plot\n\n***\n\n")
  for (n in q_values) {
    qtbl <- traj_tbl |> filter(quantile_n == n) |> mutate(bin_label = paste0("Bin ", bin))
    if (nrow(qtbl) == 0) next
    p <- ggplot(qtbl, aes(x = exp_pc, y = egenes_fdr05,
                          colour = as.factor(bin), group = as.factor(bin))) +
      geom_line(linewidth = 1) + geom_point(size = 2.5) +
      scale_colour_manual(values = bin_palette, name = "Bin") +
      facet_wrap(~ bin_label, nrow = 1) +
      labs(title = sprintf("%s — Q%d: eGenes (FDR < 0.05) by expression PCs", traj, n),
           x = "Expression PCs", y = "eGene count (FDR < 0.05)") +
      theme_minimal(base_size = 12) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"),
            strip.text = element_text(face = "bold"), legend.position = "none",
            panel.grid.major = element_line(colour = "grey90"),
            panel.grid.minor = element_line(colour = "grey95"))
    print(p); cat("\n\n")
  }

  cat("#### Best expPC per bin\n\n***\n\n")
  traj_tbl |>
    group_by(quantile_n, bin) |>
    slice_max(egenes_fdr05, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(quantile_n, bin, exp_pc, genes_tested, samples,
           covariates, egenes_fdr05, beta_cor, prop_sig) |>
    arrange(quantile_n, bin) |>
    gt() |>
    tab_header(title = sprintf("Best expression PC — %s", traj),
               subtitle = "Selected by maximum eGene count at FDR < 0.05") |>
    cols_label(quantile_n = "Q", bin = "Bin", exp_pc = "Best expPC",
               genes_tested = "Genes tested", samples = "Samples",
               covariates = "Covariates", egenes_fdr05 = "eGenes (FDR<0.05)",
               beta_cor = "Beta correlation", prop_sig = "Prop. significant") |>
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels()) |>
    fmt_number(columns = c(beta_cor, prop_sig), decimals = 3) |>
    print()
  cat("\n\n***\n\n")

  cat("#### Full sensitivity table\n\n***\n\n")
  traj_tbl |>
    select(quantile_n, bin, exp_pc, genes_tested, samples,
           covariates, egenes_fdr05, beta_cor, prop_sig) |>
    arrange(quantile_n, bin, exp_pc) |>
    gt() |>
    tab_header(title = sprintf("Full sensitivity results — %s", traj)) |>
    cols_label(quantile_n = "Q", bin = "Bin", exp_pc = "expPC",
               genes_tested = "Genes tested", samples = "Samples",
               covariates = "Covariates", egenes_fdr05 = "eGenes (FDR<0.05)",
               beta_cor = "Beta cor", prop_sig = "Prop. sig") |>
    fmt_number(columns = c(beta_cor, prop_sig), decimals = 3) |>
    print()
  cat("\n\n***\n\n")
}
```

***

## 4. eQTL Landscape Across Pseudotime {.tabset}

***

eGene counts at FDR < 0.05 across pseudotime bins, using each bin's best
expression PC (selected by maximum eGene count; see Section 3).

***

```{r landscape_data, echo=FALSE, message=FALSE, warning=FALSE}
read_perm_egenes <- function(in_dir, trajectory, quantile_n, bin, exp_pc, geno_pc = 4) {
  perm_file <- file.path(
    in_dir, trajectory, "tensorqtl", "perm",
    sprintf("Q%d_bin%d_genPC%d_expPC%d", quantile_n, bin, geno_pc, exp_pc),
    sprintf("Q%d_bin%d_perm.cis_qtl.txt.gz", quantile_n, bin)
  )
  if (!file.exists(perm_file)) { message("Not found: ", perm_file); return(NULL) }
  read_tsv(perm_file, show_col_types = FALSE) |>
    filter(!is.na(qval) & qval < 0.05) |>
    distinct(phenotype_id) |>
    mutate(trajectory = trajectory, quantile_n = quantile_n, bin = bin, exp_pc = exp_pc)
}

egene_list <- list()
for (traj in trajectories) {
  for (n in q_values) for (b in seq_len(n)) {
    pc <- get_best_exp_pc(traj, n, b)
    key <- sprintf("%s_Q%d_bin%d", traj, n, b)
    egene_list[[key]] <- read_perm_egenes(in_dir, traj, n, b, pc)
  }
}

egene_tbl <- bind_rows(egene_list)
landscape_counts <- egene_tbl |>
  group_by(trajectory, quantile_n, bin) |>
  summarise(egenes = n_distinct(phenotype_id), .groups = "drop")
```

```{r landscape_plots, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=5}
for (traj in trajectories) {
  cat(sprintf("\n\n### %s {.tabset}\n\n***\n\n", traj))
  traj_counts <- landscape_counts |> filter(trajectory == traj)

  cat("#### eGene counts across bins\n\n***\n\n")
  p <- traj_counts |>
    ggplot(aes(x = bin, y = egenes, colour = as.factor(bin), group = as.factor(quantile_n))) +
    geom_line(linewidth = 1, colour = "grey70") + geom_point(size = 3) +
    scale_colour_manual(values = bin_palette, name = "Bin") +
    scale_x_continuous(breaks = seq_len(max(traj_counts$bin))) +
    facet_wrap(~ paste0("Q", quantile_n), scales = "free_x") +
    labs(title    = sprintf("%s — eGenes across pseudotime bins", traj),
         subtitle = "FDR < 0.05, permutation-based q-value (bin-specific best expPC)",
         x = "Pseudotime bin", y = "eGene count (FDR < 0.05)") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, colour = "grey50"),
          strip.text = element_text(face = "bold"), legend.position = "none",
          panel.grid.major = element_line(colour = "grey90"),
          panel.grid.minor = element_line(colour = "grey95"))
  print(p); cat("\n\n")

  cat("#### eGene counts table\n\n***\n\n")
  traj_counts |>
    arrange(quantile_n, bin) |>
    gt() |>
    tab_header(title = sprintf("eGene counts — %s", traj),
               subtitle = "FDR < 0.05 (bin-specific best expPC)") |>
    cols_label(trajectory = "Trajectory", quantile_n = "Q",
               bin = "Bin", egenes = "eGenes (FDR<0.05)") |>
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels()) |>
    print()
  cat("\n\n***\n\n")
}
```

***

## 5. Overlap with Full Paper eQTL Analysis {.tabset}

***

Comparison of pseudotime bin eGenes with all significant eGenes (FDR < 0.05)
from the original paper across all L1 and L2 cell types. eGenes classified as:

- **Shared**: detected in both the pseudotime bin and the full paper eGene set
- **Novel**: detected only in the pseudotime bin — not an eGene in any cell type in the original paper
- **Paper only**: significant in the paper but not detected in this pseudotime bin

**Novelty definition:** An eGene is classified as novel if its Ensembl gene ID
(`phenotype_id`) is not present as a significant eGene (FDR < 0.05) in any L1 or
L2 cell type from the original paper. Novelty is assessed at the gene level — the
same gene detected with a different lead eSNP in the pseudotime analysis is still
classified as novel if the gene itself was not previously an eGene. The same novel
eGene may be detected in multiple pseudotime bins; the novel eGene tables below
report each gene once, retaining the bin and statistics from its most significant
detection (lowest nominal p-value).

***

```{r overlap_data, echo=FALSE, message=FALSE, warning=FALSE}

# ── Cell type lists and best expPC maps — from tensorqtl_report.Rmd ──────────

cell_types_L1 <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")
cell_types_L2 <- c("Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
                   "Glu-DL-0", "Glu-DL-1", "Glu-DL-2",
                   "GABA-0",   "GABA-1",   "GABA-2",
                   "NPC-0",    "NPC-1",    "NPC-2")

exppc_map_L1 <- c(
  "Glu-UL"    = 50, "Glu-DL"    = 40, "GABA"      = 30,
  "NPC"       = 30, "MG"        = 30, "OPC"       = 30,
  "Endo-Peri" = 30
)

exppc_map_L2 <- c(
  "Glu-UL-0" = 30, "Glu-UL-1" = 50, "Glu-UL-2" = 40,
  "Glu-DL-0" = 40, "Glu-DL-1" = 40, "Glu-DL-2" = 30,
  "GABA-0"   = 30, "GABA-1"   = 40, "GABA-2"   = 40,
  "NPC-0"    = 50, "NPC-1"    = 30, "NPC-2"    = 20
)

# ── Function to read eGenes for one cell type ─────────────────────────────────

read_egenes_ct <- function(ct, exppc_map, perm_dir,
                            geno_pc = 4, norm_method = "quantile") {
  pc        <- exppc_map[[ct]]
  perm_file <- file.path(
    perm_dir,
    sprintf("%s_%s_genPC_%d_expPC_%d", ct, norm_method, geno_pc, pc),
    sprintf("%s_%s_perm.cis_qtl.txt.gz", ct, norm_method)
  )
  if (!file.exists(perm_file)) {
    message("Not found: ", perm_file)
    return(NULL)
  }
  read_tsv(perm_file, show_col_types = FALSE) |>
    filter(!is.na(qval) & qval < 0.05) |>
    distinct(phenotype_id) |>
    mutate(cell_type = ct)
}

# ── Build full paper eGene reference set — union of L1 and L2 ────────────────

message("Loading L1 cell type eGenes...")
egenes_L1 <- map_dfr(cell_types_L1, read_egenes_ct,
                     exppc_map  = exppc_map_L1,
                     perm_dir   = cell_type_perm_dir)

message("Loading L2 cell type eGenes...")
egenes_L2 <- map_dfr(cell_types_L2, read_egenes_ct,
                     exppc_map  = exppc_map_L2,
                     perm_dir   = cell_type_perm_dir)

paper_egene_set <- unique(c(egenes_L1$phenotype_id, egenes_L2$phenotype_id))
message("Total unique eGenes in full paper (L1 + L2): ", length(paper_egene_set))

# ── Build overlap table ───────────────────────────────────────────────────────

overlap_tbl <- map_dfr(trajectories, function(traj) {
  map_dfr(q_values, function(n) {
    map_dfr(seq_len(n), function(b) {
      key       <- sprintf("%s_Q%d_bin%d", traj, n, b)
      bin_set   <- egene_list[[key]]
      if (is.null(bin_set)) return(NULL)
      bin_genes <- bin_set$phenotype_id
      tibble(
        trajectory   = traj,
        quantile_n   = n,
        bin          = b,
        n_bin        = length(bin_genes),
        n_paper      = length(paper_egene_set),
        n_shared     = length(intersect(bin_genes, paper_egene_set)),
        n_novel      = length(setdiff(bin_genes, paper_egene_set)),
        n_paper_only = length(setdiff(paper_egene_set, bin_genes)),
        pct_shared   = round(100 * length(intersect(bin_genes, paper_egene_set)) /
                               max(length(bin_genes), 1), 1)
      )
    })
  })
})
```

```{r overlap_and_novel, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=12, fig.height=5}

# ── Build novel eGene tables ──────────────────────────────────────────────────

read_perm_novel <- function(in_dir, trajectory, quantile_n, bin, exp_pc,
                             paper_egenes, geno_pc = 4) {
  perm_file <- file.path(
    in_dir, trajectory, "tensorqtl", "perm",
    sprintf("Q%d_bin%d_genPC%d_expPC%d", quantile_n, bin, geno_pc, exp_pc),
    sprintf("Q%d_bin%d_perm.cis_qtl.txt.gz", quantile_n, bin)
  )
  if (!file.exists(perm_file)) return(NULL)
  read_tsv(perm_file, show_col_types = FALSE) |>
    filter(!is.na(qval) & qval < 0.05) |>
    filter(!phenotype_id %in% paper_egenes) |>
    distinct(phenotype_id, .keep_all = TRUE) |>
    select(phenotype_id, variant_id, slope, pval_nominal, qval) |>
    mutate(trajectory = trajectory, quantile_n = quantile_n, bin = bin)
}

novel_all <- map_dfr(trajectories, function(traj) {
  map_dfr(q_values, function(n) {
    map_dfr(seq_len(n), function(b) {
      pc <- get_best_exp_pc(traj, n, b)
      read_perm_novel(in_dir, traj, n, b, pc, paper_egene_set)
    })
  })
})

novel_union <- novel_all |>
  group_by(trajectory, quantile_n, phenotype_id) |>
  slice_min(pval_nominal, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(symbol = get_symbol(phenotype_id))

novel_summary <- novel_all |>
  group_by(trajectory, quantile_n) |>
  summarise(
    total_novel_detections = n(),
    novel_egene_union      = n_distinct(phenotype_id),
    single_bin             = sum(table(phenotype_id) == 1),
    multi_bin              = sum(table(phenotype_id) > 1),
    .groups = "drop"
  )

# ── Combined per-trajectory output ────────────────────────────────────────────

for (traj in trajectories) {

  cat(sprintf("\n\n### %s {.tabset}\n\n***\n\n", traj))

  traj_overlap <- overlap_tbl |> filter(trajectory == traj)

  overlap_long <- traj_overlap |>
    select(quantile_n, bin, n_shared, n_novel, n_paper_only) |>
    pivot_longer(cols      = c(n_novel, n_shared, n_paper_only),
                 names_to  = "category",
                 values_to = "count") |>
    mutate(category = factor(category,
                             levels = c("n_novel", "n_shared", "n_paper_only"),
                             labels = c("Novel", "Shared", "Paper only")))

  # ── Tab 1: Overlap plot (Q4) ─────────────────────────────────────────────

  cat("#### Overlap plot\n\n***\n\n")

  p <- overlap_long |>
    filter(quantile_n == 4) |>
    mutate(bin_label = paste0("Bin ", bin)) |>
    group_by(bin_label) |>
    mutate(prop = count / sum(count)) |>
    ggplot(aes(x = prop, y = fct_rev(bin_label), fill = category)) +
    geom_col(position = "stack", colour = "black", width = 0.7) +
    geom_text(aes(label = ifelse(count > 0, count, "")),
              position = position_stack(vjust = 0.5),
              size = 3, colour = "black", fontface = "bold") +
    scale_fill_manual(values = c("Novel"      = "grey70",
                                 "Shared"     = "#B2DF8A",
                                 "Paper only" = "#66C2A5"),
                      name = NULL) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(title    = sprintf("%s — Q4 overlap with full paper eGene set", traj),
         subtitle = sprintf("Paper reference: %d unique eGenes across all L1 + L2 cell types",
                            length(paper_egene_set)),
         x = "Proportion of eGenes", y = NULL) +
    theme_classic(base_size = 12) +
    theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, colour = "grey50"),
          legend.position = "top",
          axis.text.y   = element_text(face = "bold"))
  print(p); cat("\n\n")

  # ── Tab 2: Overlap across all Q values ───────────────────────────────────

  cat("#### Overlap across all Q values\n\n***\n\n")

  p2 <- overlap_long |>
    mutate(bin_label = paste0("Bin ", bin)) |>
    group_by(quantile_n, bin_label) |>
    mutate(prop = count / sum(count)) |>
    ggplot(aes(x = prop, y = fct_rev(bin_label), fill = category)) +
    geom_col(position = "stack", colour = "black", width = 0.7) +
    scale_fill_manual(values = c("Novel"      = "grey70",
                                 "Shared"     = "#B2DF8A",
                                 "Paper only" = "#66C2A5"),
                      name = NULL) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ paste0("Q", quantile_n), scales = "free_y") +
    labs(title = sprintf("%s — overlap with full paper eGene set across Q values", traj),
         x = "Proportion of eGenes", y = NULL) +
    theme_classic(base_size = 11) +
    theme(plot.title      = element_text(hjust = 0.5, face = "bold"),
          legend.position = "top",
          axis.text.y     = element_text(face = "bold"),
          strip.text      = element_text(face = "bold"))
  print(p2); cat("\n\n")

  # ── Tab 3: Overlap table ─────────────────────────────────────────────────

  cat("#### Overlap table\n\n***\n\n")

  traj_overlap |>
    select(quantile_n, bin, n_bin, n_shared, n_novel, n_paper_only, pct_shared) |>
    arrange(quantile_n, bin) |>
    gt() |>
    tab_header(
      title    = sprintf("eGene overlap — %s vs full paper eGene set", traj),
      subtitle = sprintf("Pseudotime: bin-specific best expPC | Paper reference: %d eGenes (L1 + L2)",
                         length(paper_egene_set))
    ) |>
    cols_label(quantile_n   = "Q",
               bin          = "Bin",
               n_bin        = "Bin eGenes",
               n_shared     = "Shared",
               n_novel      = "Novel",
               n_paper_only = "Paper only",
               pct_shared   = "% shared") |>
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels()) |>
    tab_style(style     = cell_fill(color = "#B2DF8A"),
              locations = cells_body(columns = n_shared)) |>
    tab_style(style     = cell_fill(color = "grey90"),
              locations = cells_body(columns = n_novel)) |>
    print()
  cat("\n\n")

  # ── Tab 4: Novel eGene summary ───────────────────────────────────────────

  cat("#### Novel eGene summary\n\n***\n\n")

  novel_summary |>
    filter(trajectory == traj) |>
    select(-trajectory) |>
    gt() |>
    tab_header(
      title    = sprintf("Novel eGene summary — %s", traj),
      subtitle = "eGenes not found in any L1 or L2 cell type in the original paper"
    ) |>
    cols_label(
      quantile_n             = "Q",
      total_novel_detections = "Total detections (with duplicates)",
      novel_egene_union      = "Unique novel eGenes (union)",
      single_bin             = "Detected in 1 bin only",
      multi_bin              = "Detected in >1 bin"
    ) |>
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels()) |>
    tab_style(style     = cell_fill(color = "grey90"),
              locations = cells_body(columns = novel_egene_union)) |>
    print()
  cat("\n\n")

  # ── Tabs 5-7: Novel eGene tables per Q value ─────────────────────────────

  for (n in q_values) {
    cat(sprintf("#### Q%d novel eGenes\n\n***\n\n", n))

    novel_tbl <- novel_union |>
      filter(trajectory == traj, quantile_n == n) |>
      arrange(pval_nominal) |>
      select(symbol, phenotype_id, variant_id, bin, slope, pval_nominal, qval)

    if (nrow(novel_tbl) == 0) {
      cat("*No novel eGenes detected.*\n\n***\n\n")
      next
    }

    novel_tbl |>
      gt() |>
      tab_header(
        title    = sprintf("Novel eGenes — %s Q%d", traj, n),
        subtitle = sprintf("%d unique novel eGenes | ranked by lowest nominal p-value",
                           nrow(novel_tbl))
      ) |>
      cols_label(
        symbol       = "Gene",
        phenotype_id = "Ensembl ID",
        variant_id   = "eSNP",
        bin          = "Best bin",
        slope        = "Slope",
        pval_nominal = "Nominal p",
        qval         = "FDR (q-value)"
      ) |>
      fmt_scientific(columns = c(pval_nominal, qval), decimals = 2) |>
      fmt_number(columns = slope, decimals = 4) |>
      tab_style(style = cell_text(weight = "bold"),
                locations = cells_column_labels()) |>
      tab_style(style = cell_text(style = "italic"),
                locations = cells_body(columns = symbol)) |>
      print()
    cat("\n\n***\n\n")
  }
}

# ── Combined (cross-trajectory) Q4 novel eGene summary ────────────────────────

cat("\n\n### Combined — Q4 novel eGenes {.tabset}\n\n***\n\n")

q4_novel_combined <- novel_union |>
  filter(quantile_n == 4)

combined_summary <- tibble(
  metric = c(
    "Paper eGenes (L1 + L2 reference)",
    sprintf("%s — unique Q4 novel eGenes", trajectories[1]),
    sprintf("%s — unique Q4 novel eGenes", trajectories[2]),
    "Combined — unique Q4 novel eGenes (union across trajectories)"
  ),
  n = c(
    length(paper_egene_set),
    novel_summary |> filter(trajectory == trajectories[1], quantile_n == 4) |> pull(novel_egene_union),
    novel_summary |> filter(trajectory == trajectories[2], quantile_n == 4) |> pull(novel_egene_union),
    n_distinct(q4_novel_combined$phenotype_id)
  )
)

cat("#### Summary\n\n***\n\n")

combined_summary |>
  gt() |>
  tab_header(
    title    = "Q4 novel eGenes — combined across trajectories",
    subtitle = "Pseudotime-derived eGenes not detected as significant (FDR < 0.05) in any L1 or L2 cell type in the original paper"
  ) |>
  cols_label(metric = "Metric", n = "N genes") |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(rows = metric == "Combined — unique Q4 novel eGenes (union across trajectories)")) |>
  print()
cat("\n\n")

cat("#### Combined gene list\n\n***\n\n")

q4_novel_combined |>
  group_by(phenotype_id) |>
  slice_min(pval_nominal, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(pval_nominal) |>
  select(symbol, phenotype_id, trajectory, bin, variant_id, slope, pval_nominal, qval) |>
  gt() |>
  tab_header(
    title    = "Q4 novel eGenes — combined unique gene list",
    subtitle = sprintf("%d unique genes across both trajectories | ranked by lowest nominal p-value",
                       n_distinct(q4_novel_combined$phenotype_id))
  ) |>
  cols_label(
    symbol       = "Gene", phenotype_id = "Ensembl ID", trajectory = "Trajectory",
    bin          = "Best bin", variant_id = "eSNP", slope = "Slope",
    pval_nominal = "Nominal p", qval = "FDR (q-value)"
  ) |>
  fmt_scientific(columns = c(pval_nominal, qval), decimals = 2) |>
  fmt_number(columns = slope, decimals = 4) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(style = "italic"), locations = cells_body(columns = symbol)) |>
  print()
cat("\n\n")
```

***

## 6. Cross-Bin eQTL Sharing {.tabset}

***

UpSet plots showing intersections of eGene sets across Q4 pseudotime bins,
using each bin's best expression PC. Each set represents eGenes detected at
FDR < 0.05 in that bin.

***

```{r upset_data, echo=FALSE, message=FALSE, warning=FALSE}
read_perm_full <- function(in_dir, trajectory, quantile_n, bin, exp_pc, geno_pc = 4) {
  perm_file <- file.path(
    in_dir, trajectory, "tensorqtl", "perm",
    sprintf("Q%d_bin%d_genPC%d_expPC%d", quantile_n, bin, geno_pc, exp_pc),
    sprintf("Q%d_bin%d_perm.cis_qtl.txt.gz", quantile_n, bin)
  )
  if (!file.exists(perm_file)) return(NULL)
  read_tsv(perm_file, show_col_types = FALSE) |>
    filter(!is.na(qval) & qval < 0.05) |>
    distinct(phenotype_id, .keep_all = TRUE) |>
    select(phenotype_id, slope, slope_se, pval_nominal, qval) |>
    mutate(trajectory = trajectory, quantile_n = quantile_n, bin = bin, exp_pc = exp_pc)
}

q4_slope <- map_dfr(trajectories, function(traj) {
  map_dfr(seq_len(4), function(b) {
    pc <- get_best_exp_pc(traj, 4, b)
    read_perm_full(in_dir, traj, 4, b, pc)
  })
})
```

### NPC-to-Glu-DL {.tabset}

***

```{r upset_dl_data, echo=FALSE, message=FALSE, warning=FALSE}
traj <- "NPC-to-Glu-DL"
gene_by_bin_dl <- q4_slope |>
  filter(trajectory == traj) |>
  distinct(phenotype_id, bin) |>
  mutate(present = 1, bin_label = paste0("Bin_", bin)) |>
  pivot_wider(names_from = bin_label, values_from = present, values_fill = 0)

dups <- gene_by_bin_dl |> filter(duplicated(phenotype_id))
message("Duplicate phenotype_ids after pivot_wider: ", nrow(dups))

gene_by_bin_dl <- gene_by_bin_dl |>
  group_by(phenotype_id) |>
  summarise(across(everything(), max), .groups = "drop") |>
  column_to_rownames("phenotype_id")

for (b in paste0("Bin_", 1:4)) {
  if (!b %in% colnames(gene_by_bin_dl)) gene_by_bin_dl[[b]] <- 0
}
gene_by_bin_dl <- gene_by_bin_dl[, paste0("Bin_", 1:4)]
```

#### UpSet plot

***

```{r upset_dl_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.dim=c(12, 5)}
upset(gene_by_bin_dl, nsets = 4, order.by = "freq",
      sets = paste0("Bin_", 1:4), keep.order = TRUE,
      mainbar.y.label = "eGene intersection size",
      sets.x.label    = "eGenes per bin",
      text.scale      = 1.3)
```

***

### NPC-to-Glu-UL {.tabset}

***

```{r upset_ul_data, echo=FALSE, message=FALSE, warning=FALSE}
traj <- "NPC-to-Glu-UL"
gene_by_bin_ul <- q4_slope |>
  filter(trajectory == traj) |>
  distinct(phenotype_id, bin) |>
  mutate(present = 1, bin_label = paste0("Bin_", bin)) |>
  pivot_wider(names_from = bin_label, values_from = present, values_fill = 0)

dups <- gene_by_bin_ul |> filter(duplicated(phenotype_id))
message("Duplicate phenotype_ids after pivot_wider: ", nrow(dups))

gene_by_bin_ul <- gene_by_bin_ul |>
  group_by(phenotype_id) |>
  summarise(across(everything(), max), .groups = "drop") |>
  column_to_rownames("phenotype_id")

for (b in paste0("Bin_", 1:4)) {
  if (!b %in% colnames(gene_by_bin_ul)) gene_by_bin_ul[[b]] <- 0
}
gene_by_bin_ul <- gene_by_bin_ul[, paste0("Bin_", 1:4)]
```

#### UpSet plot

***

```{r upset_ul_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.dim=c(12, 5)}
upset(gene_by_bin_ul, nsets = 4, order.by = "freq",
      sets = paste0("Bin_", 1:4), keep.order = TRUE,
      mainbar.y.label = "eGene intersection size",
      sets.x.label    = "eGenes per bin",
      text.scale      = 1.3)
```

***
