---
title: "pi1 Enrichment Report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  obrien_dir: null
  ziffra_dir: null
  ziffra_indep_dir: null
  wen_dir: null
  bryois_dir: null
  fugita_dir: null
  internal_dir: null
  beta_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(cowplot)
library(ggrepel)
obrien_dir <- params$obrien_dir
ziffra_dir <- params$ziffra_dir
ziffra_indep_dir <- params$ziffra_indep_dir
bmark_dir <- params$bmark_dir
wen_dir <- params$wen_dir
bryois_dir <- params$bryois_dir
fugita_dir <- params$fugita_dir
internal_dir <- params$internal_dir
beta_dir <- params$beta_dir

custom_palette <- c(
  "RG" = "#FF5959",
  "ExN-UL" = "#00B6EB",
  "InN" = "#3CBB75FF",
  "ExN-DL" = "#CEE5FD",
  "Endo-Peri" = "#B200ED",
  "MG" = "#F58231",
  "OPC" = "#FDE725FF"
)

cell_types <- c(
  "Glu-UL", "Glu-UL-0", "Glu-UL-1", "Glu-UL-2",
  "Glu-DL", "Glu-DL-0", "Glu-DL-1", "Glu-DL-2", 
  "NPC", "NPC-0", "NPC-1", "NPC-2", 
  "GABA", "GABA-0", "GABA-1", "GABA-2",
  "Endo-Peri",
  "OPC",
  "MG"
)

cell_types_L1 <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")

# Exp PC map
expPC_map_all <- c(
  "Glu-UL"     = 50,
  "Glu-DL"     = 40,
  "GABA"       = 30,
  "NPC"        = 30,
  "MG"         = 30,
  "OPC"        = 30,
  "Endo-Peri"  = 30,
  "Glu-UL-0"     = 30,
  "Glu-UL-1"     = 50,
  "Glu-UL-2"     = 40,
  "Glu-DL-0"     = 40,
  "Glu-DL-1"     = 40,
  "Glu-DL-2"     = 30,
  "GABA-0"       = 30,
  "GABA-1"       = 40,
  "GABA-2"       = 40,
  "NPC-0"        = 50,
  "NPC-1"        = 30,
  "NPC-2"        = 20
)

gene_lookup <- read_tsv('../../resources/sheets/gene_lookup_hg38.tsv') |>
  select(gene = ensembl_gene_id, symbol = external_gene_name)

```

## eQTL replication pipeline steps {.tabset}

***

1. `dwnld_obrien`: Download O'Brien 2018 bulk dev-brain eQTL files  PMID:30419947
2. `dwnld_bryois`: Download Bryois 2022 sc-adult-brain eQTL files   PMID:35915177
3. `dwnld_ziffra`: Download Ziffra 2021 snATAC-seq peaks files     PMID:34616060
4. `dwnld_wen`: Download Wen 2024 bulk dev-brain eQTL files      PMID:38781368
5. `snp_lookup`: Generate snp lookup for eQTL / ATAC-seq peak enrichments
6. `atac_enrich`: Test for eQTL enrichments in Ziffra ATAC-seq peaks
7. `snp_lookup_indep`: Generate snp lookup for eQTL / ATAC-seq peak enrichments (independent)
8. `atac_enrich_indep`: Test for eQTL enrichments in Ziffra ATAC-seq peaks (independent)
9. `cat_nom_qtl`: Combine per-chr nominal eQTL parquet files into gzip file for pi1 enrichment
10. `pi1_enrich_obrien`: Calc. pi1 enrichments between sn-eQTL and O'Brien 2018 bulk brain eQTL
11. `pi1_enrich_wen`: Calc. pi1 enrichments between sn-eQTL and Wen 2024 bulk brain eQTL
12. `pi1_enrich_bryois`: Calc. pi1 enrichments between sn-eQTL and Bryois 2022 adult brain sn-eQTL
13. `pi1_enrich_fugita`: Calc. pi1 enrichments between sn-eQTL and Fugita 2024 adult brain sn-eQTL
14. `beta_corre`
15. `pi1_enrichments_report`: Rule to create this report

Settings:

+ Expression PC: 40 (see tensorQTL_report.html)
+ Genotype PCs: 4 (see genotyping_post_impute_report.html)
+ Norm methods: quantile

Test direction:

+ Forward: Tests pi1 enrichment of our sig. sn-eQTL in public nominal eQTL data
+ Reverse: Tests pi1 of public sig. sn-eQTL in our nominal sn-eQTL data


+ NA in pi1 test means either:
  + too few overlapping eQTLs (<50) for pi1 calculation
  + pi1 smoother (and bootstap backup test) failed (also due tot few overlaps see: )

***

### Ziffra {.tabset}

***

#### Primary

***

```{r, ziffra_enrichments, echo = FALSE, message = FALSE, warning = FALSE}
# Create a function to read and add cell_type identifier (optional)
read_rds_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(ziffra_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '.ziffra_peak_enrich.rds')
  read_rds(file_path) %>%
    mutate(source_cell_type = cell_type) # Optional: add column to track source
}

# Read all files and bind rows
combined_data <- map_dfr(cell_types, read_rds_file)

# Assume your data is in combined_data
combined_data %>%
  # Optionally, round numeric columns for clarity
  mutate(
    p_value = signif(p_value, 3),
    mean_permuted_overlap = round(mean_permuted_overlap, 1),
    fold_enrichment = round(fold_enrichment, 2),
    peak_cell_type = str_remove(peak_cell_type, '_MACSpeaks')
  ) %>%
  dplyr::select(-source_cell_type) %>%
  # Create the gt table
  gt() %>%
  tab_header(
    title = "Table 1. Primary sc-eQTL Enrichment in Ziffra et. al. (2021) snATAC-seq OCRs",
    subtitle = "Observed vs Permuted Overlap"
  ) %>%
  cols_label(
    eqtl_cell_type = "eQTL Cell Type",
    peak_cell_type = "Peak Cell Type",
    p_value = "P Value",
    observed_overlap = "Observed Overlap",
    mean_permuted_overlap = "Mean Permuted Overlap",
    fold_enrichment = "Fold Enrichment"
  ) %>%
  fmt_number(
    columns = vars(p_value, fold_enrichment),
    decimals = 3
  ) %>%
  fmt_number(
    columns = vars(mean_permuted_overlap),
    decimals = 1
  ) %>%
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3)
  ) %>%
  opt_table_font(
    font = list(default_fonts())
  )

```

***

#### Secondary

***

```{r, ziffra_enrichments_secondary, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
read_rds_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(ziffra_indep_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '.ziffra_peak_enrich_indep.rds')
  res <- read_rds(file_path) 
  
  if (nrow(res) == 0) {
    message("Empty stub file (skipping): ", file_path)
    return(tibble())
  }
  
  res |>
    mutate(source_cell_type = cell_type) 
  
}

# Read all files and bind rows
combined_data <- map_dfr(cell_types, read_rds_file)

# Assume your data is in combined_data
combined_data %>%
  # Optionally, round numeric columns for clarity
  mutate(
    p_value = signif(p_value, 3),
    mean_permuted_overlap = round(mean_permuted_overlap, 1),
    fold_enrichment = round(fold_enrichment, 2),
    peak_cell_type = str_remove(peak_cell_type, '_MACSpeaks')
  ) %>%
  dplyr::select(-source_cell_type) %>%
  # Create the gt table
  gt() %>%
  tab_header(
    title = "Table 2. Secondary sc-eQTL Enrichment in Ziffra et. al. (2021) snATAC-seq OCRs",
    subtitle = "Observed vs Permuted Overlap (only cell types with ≥20 secondary eQTLs)"
  ) %>%
  cols_label(
    eqtl_cell_type = "eQTL Cell Type",
    peak_cell_type = "Peak Cell Type",
    p_value = "P Value",
    observed_overlap = "Observed Overlap",
    mean_permuted_overlap = "Mean Permuted Overlap",
    fold_enrichment = "Fold Enrichment"
  ) %>%
  fmt_number(
    columns = vars(p_value, fold_enrichment),
    decimals = 3
  ) %>%
  fmt_number(
    columns = vars(mean_permuted_overlap),
    decimals = 1
  ) %>%
  tab_options(
    table.font.names = "Arial",
    table.font.size = 12,
    data_row.padding = px(3)
  ) %>%
  opt_table_font(
    font = list(default_fonts())
  )

```

***

### O'Brien {.tabset}

***

#### Table

***

```{r, obrien_enrichments, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
cell_types_L1 <- c("Glu-UL", "Glu-DL", "GABA", "NPC", "MG", "OPC", "Endo-Peri")

read_rds_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '_pi1_enrich_tbl.rds')
  read_rds(file_path) %>%
    mutate(source_cell_type = cell_type) # Optional: add column to track source
}

# Read all files and bind rows
pi1_summary_tbl <- map_dfr(cell_types, read_rds_file)

# View the result
pi1_summary_tbl %>%
  mutate(
    pi1_cell_ref = round(pi1_cell_ref, 2),
    prop_replicating_cell_ref = round(prop_replicating_cell_ref, 2),
    prop_same_direction_cell_ref = round(prop_same_direction_cell_ref, 2),
    pi1 = ifelse(is.na(pi1), NA, round(as.numeric(pi1), 2)),
    prop_replicating = ifelse(is.na(prop_replicating), NA, round(as.numeric(prop_replicating), 2)),
    prop_same_direction = ifelse(is.na(prop_same_direction), NA, round(as.numeric(prop_same_direction), 2))
  ) %>%
  select(-source_cell_type) %>%
  gt() %>%
  tab_header(
    title = "Table 3. Pi1 Enrichment: cell-specific eQTL with O'Brien et. al. (2018) bulk eQTL data"
  ) %>%
  cols_label(
    cell_type = "Cell Type",
    pi1 = "pi1",
    prop_replicating = "Prop. Replicating",
    prop_same_direction = "Prop. Same Dir.",
    overlap_count = "Overlap Count",
    pi1_cell_ref = "pi1",
    prop_replicating_cell_ref = "Prop. Replicating",
    prop_same_direction_cell_ref = "Prop. Same Dir.",
    overlap_count_cell_ref = "Overlap Count"
  ) %>%
  # Add higher-level headers (spanners) for forward vs reference
  tab_spanner(
    label = "Forward Test",
    columns = c(pi1, prop_replicating, prop_same_direction, overlap_count)
  ) %>%
  tab_spanner(
    label = "Reference Test",
    columns = c(pi1_cell_ref, prop_replicating_cell_ref, prop_same_direction_cell_ref, overlap_count_cell_ref)
  ) %>%
  fmt_missing(
    columns = everything(),
    missing_text = "NA"
  ) %>%
  tab_options(
    table.font.size = 12,
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.border.top.width = px(2),   # add line above col labels
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black"
  )
```

***

#### Heatmap

***

```{r, obrien_heatmap, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}

# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)
pi1_long <- pi1_result_tbl %>%
  dplyr::select(cell_type, pi1_forward, pi1_reverse) %>%
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") %>%
  mutate(direction = recode(direction,
                            pi1_forward = "Forward",
                            pi1_reverse = "Reverse"))

ggplot(pi1_long, aes(x = cell_type, y =  direction, fill = pi1)) +
  geom_tile(color = "black", lwd = 1.1, linetype = 1) +
  geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
            color = "black", size = 3.5) +
    scale_fill_viridis_c(
    option = "C",
    direction = 1,
    name = expression(pi[1]),
    na.value = "white"   # blanks appear white
  ) +
  coord_equal() + 
  labs(x = "Cell type",  y = "Direction") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank()
  ) 
```

***

#### Forward histograms

***

```{r, obrien_hist_forward, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 10), eval=FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(obrien_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}


# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)

plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_forward[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_forward, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_forward, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_forward, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```

***

#### Reverse histograms

***

```{r, obrien_hist_reverse, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 10), eval=FALSE}
plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_reverse[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_reverse, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_reverse, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_reverse, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```

***

### Wen {.tabset}

***

#### Table

***

```{r, wen_enrichments, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
# Create a function to read and add cell_type identifier (optional)
read_rds_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(wen_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '_pi1_enrich_tbl.rds')
  read_rds(file_path) %>%
    mutate(source_cell_type = cell_type) # Optional: add column to track source
}

# Read all files and bind rows
pi1_summary_tbl <- map_dfr(cell_types, read_rds_file)

# View the result
pi1_summary_tbl %>%
  mutate(
    pi1_cell_ref = round(pi1_cell_ref, 2),
    prop_replicating_cell_ref = round(prop_replicating_cell_ref, 2),
    prop_same_direction_cell_ref = round(prop_same_direction_cell_ref, 2),
    pi1 = ifelse(is.na(pi1), NA, round(as.numeric(pi1), 2)),
    prop_replicating = ifelse(is.na(prop_replicating), NA, round(as.numeric(prop_replicating), 2)),
    prop_same_direction = ifelse(is.na(prop_same_direction), NA, round(as.numeric(prop_same_direction), 2))
  ) %>%
  select(-source_cell_type) %>%
  gt() %>%
  tab_header(
    title = "Table 4. Pi1 Enrichment: cell-specific eQTL with Wen et. al. (2024) bulk eQTL data"
  ) %>%
  cols_label(
    cell_type = "Cell Type",
    pi1 = "pi1",
    prop_replicating = "Prop. Replicating",
    prop_same_direction = "Prop. Same Dir.",
    overlap_count = "Overlap Count",
    pi1_cell_ref = "pi1",
    prop_replicating_cell_ref = "Prop. Replicating",
    prop_same_direction_cell_ref = "Prop. Same Dir.",
    overlap_count_cell_ref = "Overlap Count"
  ) %>%
  # Add higher-level headers (spanners) for forward vs reference
  tab_spanner(
    label = "Forward Test",
    columns = c(pi1, prop_replicating, prop_same_direction, overlap_count)
  ) %>%
  tab_spanner(
    label = "Reference Test",
    columns = c(pi1_cell_ref, prop_replicating_cell_ref, prop_same_direction_cell_ref, overlap_count_cell_ref)
  ) %>%
  fmt_missing(
    columns = everything(),
    missing_text = "NA"
  ) %>%
  tab_options(
    table.font.size = 12,
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.border.top.width = px(2),   # add line above col labels
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black"
  )
```

***

#### Heatmap

***

```{r, wen_heatmap, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(wen_dir, cell_type, '_quantile_genPC_4_expPC_', expPC,'_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}

# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)
pi1_long <- pi1_result_tbl %>%
  dplyr::select(cell_type, pi1_forward, pi1_reverse) %>%
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") %>%
  mutate(direction = recode(direction,
                            pi1_forward = "Forward",
                            pi1_reverse = "Reverse"))

ggplot(pi1_long, aes(x = cell_type, y =  direction, fill = pi1)) +
  geom_tile(color = "black", lwd = 1.1, linetype = 1) +
  geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
            color = "black", size = 3.5) +
    scale_fill_viridis_c(
    option = "C",
    direction = 1,
    name = expression(pi[1]),
    na.value = "white"   # blanks appear white
  ) +
  coord_equal() + 
  labs(x = "Cell type",  y = "Direction") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank()
  ) 
```

***

#### Forward histograms

***

```{r, wen_hist_forward, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 10), eval=FALSE}
# Create a function to read and add cell_type identifier (optional)
read_pi1_file <- function(cell_type) {
  
  expPC <- expPC_map_all[[cell_type]]
  
  file_path <- paste0(wen_dir, cell_type, '_quantile_genPC_4_expPC_', expPC, '_pi1_results_tbl.rds')
  pi1_list <- read_rds(file_path)
  
  tibble(
    cell_type = cell_type,
    pvals_forward = list(pi1_list$forward$pvals),
    pvals_reverse = list(pi1_list$reverse$pvals),
    pi1_forward = pi1_list$forward$pi1,
    pi1_reverse = pi1_list$reverse$pi1,
    prop_replicating_forward = pi1_list$forward$prop_replicating,
    prop_replicating_reverse = pi1_list$reverse$prop_replicating,
    prop_same_direction_forward = pi1_list$forward$prop_same_direction,
    prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
  )
}


# Read all files and bind rows
pi1_result_tbl <- map_dfr(cell_types, read_pi1_file)

plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_forward[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_forward, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_forward, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_forward, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```

***

#### Reverse histograms

***

```{r, wen_hist_reverse, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 10), eval=FALSE}
plot_list <- list()

for (cell_type in cell_types) {
  this_row <- filter(pi1_result_tbl, cell_type == !!cell_type)

  # extract the p-values (from the list column)
  pvals <- this_row$pvals_reverse[[1]]

  # skip if no p-values
  if (is.null(pvals) || length(pvals) == 0) {
    next
  }

  plot_data <- data.frame(pvalue = pvals)
  max_density <- max(ggplot_build(
    ggplot(plot_data, aes(x = pvalue)) +
      geom_histogram(aes(y = ..density..), bins = 30)
  )$data[[1]]$density)

  p <- ggplot(plot_data, aes(x = pvalue)) +
    geom_histogram(aes(y = ..density..), bins = 30,
                   fill = custom_palette[cell_type],
                   color = "black", alpha = 0.8) +
    annotate("text", x = 0.15, y = 0.90 * max_density,
             label = paste(
               "pi1 =", round(this_row$pi1_reverse, 2),
               "\nProp. replicating at 5% FDR:", round(this_row$prop_replicating_reverse, 2),
               "\nProp. same direction:", round(this_row$prop_same_direction_reverse, 2)
             ), hjust = 0) +
    labs(title = cell_type, x = "P-value", y = "Density") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 0.5),
      plot.caption = element_text(size = 10, hjust = 0)
    )

  plot_list[[cell_type]] <- p
}

combined_plot <- cowplot::plot_grid(plotlist = plot_list, ncol = 3)
combined_plot
```

***

### Internal {.tabset}

***

#### Table

***

```{r, Internal_enrichments, echo = FALSE, message = FALSE, warning = FALSE}
read_enrich_tbl <- function(ct, ref_ct) {
  
  expPC <- expPC_map_all[[ct]]
  expPCref <- expPC_map_all[[ref_ct]]
  file_path <- paste0(internal_dir, ct, "_vs_", ref_ct, "_quantile_genPC_4_expPC_", 
                      expPC ,"_expPCref_", expPCref, "_pi1_enrich_tbl.rds")
  if (file.exists(file_path)) {
    read_rds(file_path) %>%
      mutate(cell_type = ct,
             ref_cell_type = ref_ct)
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

# Generate all combinations
combinations <- expand.grid(cell_type = cell_types_L1, ref_cell_type = cell_types_L1, stringsAsFactors = FALSE)

# Read all enrich tables
pi1_enrich_tbl <- map2_dfr(combinations$cell_type, combinations$ref_cell_type, read_enrich_tbl)
pi1_enrich_tbl %>%
  mutate(
    pi1_cell_ref = ifelse(is.na(pi1_cell_ref), NA, round(as.numeric(pi1_cell_ref), 2)),
    prop_replicating_cell_ref = ifelse(is.na(prop_replicating_cell_ref), NA, round(as.numeric(prop_replicating_cell_ref), 2)),
    prop_same_direction_cell_ref = ifelse(is.na(prop_same_direction_cell_ref), NA, round(as.numeric(prop_same_direction_cell_ref), 2)),
    pi1 = ifelse(is.na(pi1), NA, round(as.numeric(pi1), 2)),
    prop_replicating = ifelse(is.na(prop_replicating), NA, round(as.numeric(prop_replicating), 2)),
    prop_same_direction = ifelse(is.na(prop_same_direction), NA, round(as.numeric(prop_same_direction), 2))
  ) %>%
  gt() %>%
  tab_header(
    title = "Table 5. Internal Pi1 Enrichment Summary cell-specific eQTL"
  ) %>%
  cols_label(
    cell_type = "Cell Type",
    ref_cell_type = "Reference Cell Type",
    pi1 = "pi1",
    prop_replicating = "Prop. Replicating",
    prop_same_direction = "Prop. Same Dir.",
    overlap_count = "Overlap Count",
    pi1_cell_ref = "pi1",
    prop_replicating_cell_ref = "Prop. Replicating",
    prop_same_direction_cell_ref = "Prop. Same Dir.",
    overlap_count_cell_ref = "Overlap Count"
  ) %>%
  # Add higher-level headers (spanners) for forward vs reference
  tab_spanner(
    label = "Forward Test",
    columns = c(pi1, prop_replicating, prop_same_direction, overlap_count)
  ) %>%
  tab_spanner(
    label = "Reference Test",
    columns = c(pi1_cell_ref, prop_replicating_cell_ref, prop_same_direction_cell_ref, overlap_count_cell_ref)
  ) %>%
  fmt_missing(
    columns = everything(),
    missing_text = "NA"
  ) %>%
  tab_options(
    table.font.size = 12,
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.border.top.width = px(2),   # add line above col labels
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black"
  )
```

***

#### Heatmap

***

```{r, interal_heatmap, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,5)}
# Function to read pi1 results
read_pi1_results <- function(ct, ref_ct) {
    
  expPC <- expPC_map_all[[ct]]
  expPCref <- expPC_map_all[[ref_ct]]
  
  file_path <- paste0(internal_dir, ct, "_vs_", ref_ct, "_quantile_genPC_4_expPC_", 
                      expPC ,"_expPCref_", expPCref, "_pi1_results_tbl.rds")
  
  if (file.exists(file_path)) {
    pi1_list <- read_rds(file_path)
    
    tibble(
      cell_type = ct,
      ref_cell_type = ref_ct,
      pi1_forward = pi1_list$forward$pi1,
      pi1_reverse = pi1_list$reverse$pi1,
      prop_replicating_forward = pi1_list$forward$prop_replicating,
      prop_replicating_reverse = pi1_list$reverse$prop_replicating,
      prop_same_direction_forward = pi1_list$forward$prop_same_direction,
      prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
    )
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

# Generate all combinations
combinations <- expand.grid(cell_type = cell_types_L1, ref_cell_type = cell_types_L1, stringsAsFactors = FALSE)

# Read all pi1 results
pi1_result_tbl <- map2_dfr(combinations$cell_type, combinations$ref_cell_type, read_pi1_results)

# Pivot for plotting
pi1_long <- pi1_result_tbl %>%
  select(cell_type, ref_cell_type, pi1_forward, pi1_reverse) %>%
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") %>%
  mutate(direction = recode(direction,
                            pi1_forward = "Forward",
                            pi1_reverse = "Reverse"))

# Split the data by direction
pi1_forward_tbl <- pi1_long %>% filter(direction == "Forward")
pi1_reverse_tbl <- pi1_long %>% filter(direction == "Reverse")

# Function to generate a heatmap
plot_pi1_heatmap <- function(df, title) {
  ggplot(df, aes(x = cell_type, y = ref_cell_type, fill = pi1)) +
  geom_tile(color = "black", lwd = 1.1, linetype = 1) +
  geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
            color = "black", size = 3.5) +
    scale_fill_viridis_c(
    option = "C",
    direction = 1,
    name = expression(pi[1]),
    na.value = "white"   # blanks appear white
  ) +
    coord_equal() +
    labs(x = "Cell type", y = "Cell type (Ref)", title = title) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank()
    )
}

# Generate heatmaps
heatmap_forward <- plot_pi1_heatmap(pi1_forward_tbl, "Forward")
heatmap_reverse <- plot_pi1_heatmap(pi1_reverse_tbl, "Reverse")

# Combine side by side
plot_grid(heatmap_forward, heatmap_reverse, ncol = 2)
```

***


### Fugita {.tabset}

***

#### Table

***

```{r, Fugita_enrichments, echo = FALSE, message = FALSE, warning = FALSE}
fugita_cell_types <- c('Ast', 'Exc', 'Inh', 'Oli', 'OPC', 'Mic', 'End')

read_enrich_tbl <- function(ct, ref_ct) {
  
    expPC <- expPC_map_all[[ct]]
  
  file_path <- paste0(fugita_dir, ct, "_vs_", ref_ct, "_quantile_genPC_4_expPC_", expPC, "_pi1_enrich_tbl.rds")
  if (file.exists(file_path)) {
    read_rds(file_path) %>%
      mutate(cell_type = ct,
             ref_cell_type = ref_ct)
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

# Generate all combinations
combinations <- expand.grid(cell_type = cell_types_L1, ref_cell_type = fugita_cell_types, stringsAsFactors = FALSE)

# Read all enrich tables
pi1_enrich_tbl <- map2_dfr(combinations$cell_type, combinations$ref_cell_type, read_enrich_tbl)
pi1_enrich_tbl %>%
  mutate(
    pi1_cell_ref = ifelse(is.na(pi1_cell_ref), NA, round(as.numeric(pi1_cell_ref), 2)),
    prop_replicating_cell_ref = ifelse(is.na(prop_replicating_cell_ref), NA, round(as.numeric(prop_replicating_cell_ref), 2)),
    prop_same_direction_cell_ref = ifelse(is.na(prop_same_direction_cell_ref), NA, round(as.numeric(prop_same_direction_cell_ref), 2)),
    pi1 = ifelse(is.na(pi1), NA, round(as.numeric(pi1), 2)),
    prop_replicating = ifelse(is.na(prop_replicating), NA, round(as.numeric(prop_replicating), 2)),
    prop_same_direction = ifelse(is.na(prop_same_direction), NA, round(as.numeric(prop_same_direction), 2))
  ) %>%
  gt() %>%
  tab_header(
    title = "Table 6. Fugita Pi1 Enrichment Summary cell-specific eQTL"
  ) %>%
  cols_label(
    cell_type = "Cell Type",
    ref_cell_type = "Reference Cell Type",
    pi1 = "pi1",
    prop_replicating = "Prop. Replicating",
    prop_same_direction = "Prop. Same Dir.",
    overlap_count = "Overlap Count",
    pi1_cell_ref = "pi1",
    prop_replicating_cell_ref = "Prop. Replicating",
    prop_same_direction_cell_ref = "Prop. Same Dir.",
    overlap_count_cell_ref = "Overlap Count"
  ) %>%
  # Add higher-level headers (spanners) for forward vs reference
  tab_spanner(
    label = "Forward Test",
    columns = c(pi1, prop_replicating, prop_same_direction, overlap_count)
  ) %>%
  tab_spanner(
    label = "Reference Test",
    columns = c(pi1_cell_ref, prop_replicating_cell_ref, prop_same_direction_cell_ref, overlap_count_cell_ref)
  ) %>%
  fmt_missing(
    columns = everything(),
    missing_text = "NA"
  ) %>%
  tab_options(
    table.font.size = 12,
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    column_labels.border.top.width = px(2),   # add line above col labels
    column_labels.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black"
  )
```

***

#### Heatmap

***

```{r, fugita_heatmap, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,5)}
# Function to read pi1 results
read_pi1_results <- function(ct, ref_ct) {
  
  expPC <- expPC_map_all[[ct]]
  
  file_path <- paste0(fugita_dir, ct, "_vs_", ref_ct, "_quantile_genPC_4_expPC_", expPC,"_pi1_results_tbl.rds")
  
  if (file.exists(file_path)) {
    pi1_list <- read_rds(file_path)
    
    tibble(
      cell_type = ct,
      ref_cell_type = ref_ct,
      pi1_forward = pi1_list$forward$pi1,
      pi1_reverse = pi1_list$reverse$pi1,
      prop_replicating_forward = pi1_list$forward$prop_replicating,
      prop_replicating_reverse = pi1_list$reverse$prop_replicating,
      prop_same_direction_forward = pi1_list$forward$prop_same_direction,
      prop_same_direction_reverse = pi1_list$reverse$prop_same_direction
    )
  } else {
    message("File missing: ", file_path)
    NULL
  }
}

# Generate all combinations
combinations <- expand.grid(cell_type = cell_types, ref_cell_type = fugita_cell_types, stringsAsFactors = FALSE)

# Read all enrich tables
pi1_enrich_tbl <- map2_dfr(combinations$cell_type, combinations$ref_cell_type, read_enrich_tbl)

# Read all pi1 results
pi1_result_tbl <- map2_dfr(combinations$cell_type, combinations$ref_cell_type, read_pi1_results)

# Pivot for plotting
pi1_long <- pi1_result_tbl %>%
  select(cell_type, ref_cell_type, pi1_forward, pi1_reverse) %>%
  pivot_longer(cols = c(pi1_forward, pi1_reverse),
               names_to = "direction", values_to = "pi1") %>%
  mutate(direction = recode(direction,
                            pi1_forward = "Forward",
                            pi1_reverse = "Reverse"))

# Split the data by direction
pi1_forward_tbl <- pi1_long %>% filter(direction == "Forward")
pi1_reverse_tbl <- pi1_long %>% filter(direction == "Reverse")

# Function to generate a heatmap
plot_pi1_heatmap <- function(df, title) {
  ggplot(df, aes(x = cell_type, y = ref_cell_type, fill = pi1)) +
  geom_tile(color = "black", lwd = 1.1, linetype = 1) +
  geom_text(aes(label = ifelse(is.na(pi1), "NA", sprintf("%.2f", pi1))),
            color = "black", size = 3.5) +
    scale_fill_viridis_c(
    option = "C",
    direction = 1,
    name = expression(pi[1]),
    na.value = "white"   # blanks appear white
  ) +
    coord_equal() +
    labs(x = "Cell type", y = "Cell type (Ref)", title = title) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank()
    )
}

# Generate heatmaps
heatmap_forward <- plot_pi1_heatmap(pi1_forward_tbl, "Forward")
heatmap_reverse <- plot_pi1_heatmap(pi1_reverse_tbl, "Reverse")

# Combine side by side
plot_grid(heatmap_forward, heatmap_reverse, ncol = 2)
```

***

### Beta correlation

***

```{r, beta_correlation, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,5)}
paired_betas <- read_tsv(paste0(beta_dir, 'replication_beta_correlation.tsv')) |>
  separate(key, into = c('snp', 'gene')) |>
  left_join(gene_lookup) |>
  group_by(gene) |>
  slice_max(abs(beta_my), n = 1, with_ties = FALSE) |> # Drops from ~3.6K to ~1.7K
  ungroup() |>
  distinct()

# Fit lm
model <- lm(beta_my ~ beta_fugita, data = paired_betas)

# Calculate perpendicular-ish distance from line + absolute residual
paired_betas <- paired_betas %>%
  mutate(.resid = resid(model),.dist  = abs(.resid),
         discordant = sign(beta_my) != sign(beta_fugita) & !is.na(beta_my) & !is.na(beta_fugita),
         label_text = ifelse(is.na(symbol) | symbol == "NA", gene, symbol),
         
         # Decide who deserves a label (adjust thresholds!)
         is_strong_outlier = .dist > quantile(.dist, 0.92, na.rm = TRUE),  # ~top 8%
         should_label = is_strong_outlier | discordant)

# How many labels will we actually try to show?
n_labels <- sum(paired_betas$should_label, na.rm = TRUE)
message("Number of points selected for labelling: ", n_labels)

# Plot
beta_cor_plt <- ggplot(paired_betas, aes(x = beta_fugita, y = beta_my)) +
  geom_point(alpha = 0.2, color = "grey70", size = 1.4) +
  geom_point(data = filter(paired_betas, is_strong_outlier & discordant), color = "#d32f2f", size = 2, alpha = 0.5) +
  geom_point(data = filter(paired_betas, is_strong_outlier & !discordant),
             color = "#1976d2", size = 2, alpha = 0.5) +   # blue for concordant strength) +
  geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", 
              color = "grey50", linewidth = 0.5) +
  geom_text_repel(
    data = filter(paired_betas, should_label),
    aes(label = label_text),
    size          = 3.1,
    color         = "black",
    fontface      = "bold",
    box.padding   = 0.45,
    point.padding = 0.5,
    segment.color = "grey50",
    segment.size  = 0.25,
    min.segment.length = 0,
    max.overlaps  = 25,             # ← increased from default 10
    force         = 1.5,            # pull labels apart more aggressively
    force_pull    = 0.8,
    direction     = "both",
    seed          = 2025            # for reproducibility of label placement
  ) +
  annotate("text", x = Inf, y = Inf, label = cor_label,
           hjust = 2.3, vjust = 3, size = 5, fontface = "bold") +
  labs(x = "Beta (Fujita)", y = "Beta (eGenes)",) +
  theme_minimal(base_size = 13) +
  theme(plot.title   = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.title   = element_text(face = "bold"),
        plot.margin  = margin(t = 20, r = 35, b = 30, l = 25, unit = "pt"))
beta_cor_plt
```

***

### Benchmarking {.tabset}

***

#### Table

***

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
#bmark_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/benchmarks/'
files <- list.files(bmark_dir, pattern = "qtl_replication.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
summary %>%
  gt() %>%
  tab_header(
    title = "Table 5. Benchmarking",
    subtitle = "Means and SD across replicates"
  ) %>%
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) %>%
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

#### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```

