---
title: "Spatial Validation: SingleR Label Transfer"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  singler_csv: null
  umap_csv:    null
---

***

## Changelog

| Version | Date       | Changes |
|---------|------------|---------|
| v1.0    | 2026-05-19 | Initial report. Heatmap used mean SingleR scores row-normalised to fractions. |
| v1.1    | 2026-05-19 | Fixed heatmap metric: switched from mean scores to fraction of cells with winning H2 label. Added aggregated H2 group UMAP tab. Updated MG negative control note to clarify pruning behaviour. |
| v1.2    | 2026-05-19 | Combined heatmap and max fraction bar chart into single figure. Merged all UMAPs into single 2x2 panel. Added H2 group breakdown table per L1 population. Added interpretive note on Glu-UL signal strength. |

***

## Overview {.tabset}

***

This report presents the results of SingleR-based label transfer from the Qian et al. 2025
MERFISH spatial transcriptomics dataset onto our snRNA-seq data. The analysis addresses
Reviewer Comment 2, which requested spatial and laminar validation of our L1 cell type
annotations — specifically the distinction between upper layer (Glu-UL) and deep layer
(Glu-DL) glutamatergic neuron populations.

**Reference:** Qian et al. 2025, *Nature*. Single-cell spatial transcriptomics of the
developing human cortex (MERFISH, 300-gene panel, GW15–GW34).

**Method:** SingleR (Bioconductor) was used to transfer H2-level spatial annotations
from the Qian MERFISH reference onto all 679,738 cells in our snRNA-seq dataset.
The reference was subsampled to 500 cells per H2 cluster (33 clusters, 16,500 cells total)
after excluding V1/V2 cortical areas (transcriptionally discontinuous per Qian et al.).
270 of 300 MERFISH panel genes (90%) were shared with our snRNA-seq data.

***

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
library(DT)
library(gt)
library(cowplot)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)

knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)

# ── Parameters ────────────────────────────────────────────────────────────────────────

singler_csv <- params$singler_csv
umap_csv    <- params$umap_csv

# ── Cell type colour palette (from scanpy_gene_lists.py) ─────────────────────────────

custom_palette <- c(
  'Glu-UL'    = '#4363d8',
  'Glu-DL'    = '#00B6EB',
  'GABA'      = '#3CBB75',
  'NPC'       = '#FF5959',
  'OPC'       = '#FDE725',
  'MG'        = '#F58231',
  'Endo-Peri' = '#B200ED'
)

# ── H2 group colour palette ───────────────────────────────────────────────────────────

group_palette <- c(
  'Upper EN'   = '#4363d8',
  'Deep EN'    = '#00B6EB',
  'Inhibitory' = '#3CBB75',
  'Progenitor' = '#FF5959',
  'OPC'        = '#FDE725',
  'EC'         = '#B200ED',
  'Migrating'  = '#A0A0A0',
  'Astrocyte'  = '#D3D3D3'
)

# ── L1 population order for plots ────────────────────────────────────────────────────

l1_order <- c('Glu-UL', 'Glu-DL', 'GABA', 'NPC', 'OPC', 'MG', 'Endo-Peri')

# ── H2 column order grouped by expected biological correspondence ─────────────────────

h2_order <- c(
  # Glu-UL block
  'EN-L2', 'EN-IT-L2/3', 'EN-IT-L3/4', 'EN-IT-L4', 'EN-IT-L4/5',
  # Glu-DL block
  'EN-IT-L6', 'EN-ET-L5/6', 'EN-ET-SP-early', 'EN-ET-SP', 'EN-ET-SP-P', 'EN-ET-L6-early',
  # GABA block
  'IN-CGE', 'IN-MGE', 'IN-SST', 'IN-VZ/GE', 'INP-VZ/GE',
  # NPC block
  'RG1', 'oRG1', 'tRG', 'vRG-late',
  'IPC-VZ/SVZ', 'IPC-iSVZ', 'IPC-SVZ-1', 'IPC-SVZ-2', 'IPC-oSVZ',
  # OPC
  'OPC',
  # EC (Endo-Peri)
  'EC',
  # Migrating/immature - no clear L1 match
  'EN-IZ-1', 'EN-IZ-2', 'EN-oSVZ-1', 'En-oSVZ-2',
  # Astrocytes - present in Qian but not in our L1
  'Astro-1', 'Astro-late1'
)

# ── H2 to group mapping ───────────────────────────────────────────────────────────────

h2_group_map <- c(
  'EN-L2'          = 'Upper EN',   'EN-IT-L2/3'     = 'Upper EN',
  'EN-IT-L3/4'     = 'Upper EN',   'EN-IT-L4'       = 'Upper EN',
  'EN-IT-L4/5'     = 'Upper EN',
  'EN-IT-L6'       = 'Deep EN',    'EN-ET-L5/6'     = 'Deep EN',
  'EN-ET-SP-early' = 'Deep EN',    'EN-ET-SP'       = 'Deep EN',
  'EN-ET-SP-P'     = 'Deep EN',    'EN-ET-L6-early' = 'Deep EN',
  'IN-CGE'         = 'Inhibitory', 'IN-MGE'         = 'Inhibitory',
  'IN-SST'         = 'Inhibitory', 'IN-VZ/GE'       = 'Inhibitory',
  'INP-VZ/GE'      = 'Inhibitory',
  'RG1'            = 'Progenitor', 'oRG1'           = 'Progenitor',
  'tRG'            = 'Progenitor', 'vRG-late'       = 'Progenitor',
  'IPC-VZ/SVZ'     = 'Progenitor', 'IPC-iSVZ'       = 'Progenitor',
  'IPC-SVZ-1'      = 'Progenitor', 'IPC-SVZ-2'      = 'Progenitor',
  'IPC-oSVZ'       = 'Progenitor',
  'OPC'            = 'OPC',
  'EC'             = 'EC',
  'EN-IZ-1'        = 'Migrating',  'EN-IZ-2'        = 'Migrating',
  'EN-oSVZ-1'      = 'Migrating',  'En-oSVZ-2'      = 'Migrating',
  'Astro-1'        = 'Astrocyte',  'Astro-late1'    = 'Astrocyte'
)

# ── Load data ─────────────────────────────────────────────────────────────────────────

singler <- read_csv(singler_csv, show_col_types = FALSE)
umap    <- read_csv(umap_csv,    show_col_types = FALSE)

# Factor L1 populations in biological order
singler$cell_type <- factor(singler$cell_type, levels = l1_order)
umap$cell_type    <- factor(umap$cell_type,    levels = l1_order)

# Join H2 labels and groups to UMAP at load time
umap <- umap %>%
  left_join(singler %>% select(cell_id, H2_label), by = 'cell_id') %>%
  mutate(H2_group = h2_group_map[H2_label])

```

***

### Classification summary

***

```{r summary_table}

# Cell counts per L1 population
cell_counts <- singler %>%
  count(cell_type, name = 'N cells') %>%
  arrange(match(cell_type, l1_order))

# Pruning rate per L1 population
pruning <- singler %>%
  group_by(cell_type) %>%
  summarise(
    `N pruned` = sum(is.na(H2_pruned)),
    `% pruned` = sprintf('%.2f%%', 100 * mean(is.na(H2_pruned))),
    .groups = 'drop'
  )

# Top H2 winning label per L1 population with percentage
top_h2 <- singler %>%
  group_by(cell_type) %>%
  count(H2_label, name = 'n') %>%
  mutate(pct = sprintf('%.1f%%', 100 * n / sum(n))) %>%
  slice_max(n, n = 1) %>%
  rename(`Top H2 label` = H2_label, `N assigned` = n, `% of population` = pct) %>%
  ungroup()

summary_tbl <- cell_counts %>%
  left_join(pruning, by = 'cell_type') %>%
  left_join(top_h2,  by = 'cell_type')

summary_tbl %>%
  gt() %>%
  tab_header(
    title    = 'Table 1. SingleR classification summary',
    subtitle = 'Per L1 population cell counts, pruning rates, and top H2 winning label assignment'
  ) %>%
  cols_label(cell_type = 'Cell Type') %>%
  tab_style(
    style     = cell_text(weight = 'bold'),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.font.size           = px(14),
    table.border.top.style    = 'none',
    table.border.bottom.style = 'solid',
    table.border.bottom.width = px(2),
    table.border.bottom.color = 'black',
    heading.align             = 'center'
  )

```

***

**Note on Microglia (MG):** Microglia are absent from the Qian MERFISH reference panel,
which was designed for cortical neurons and glia. MG cells will therefore be assigned
to the nearest transcriptional match in the reference, which has no biological
interpretation. The MG row in the heatmap below serves as a **negative control** — we
would not expect enrichment on the diagonal.

Note that SingleR assigned labels to all but 3 MG cells rather than returning NA. This
is expected behaviour — pruning occurs only when the score margin between the top and
second label falls below a statistical threshold. With no microglial genes in the 300-gene
MERFISH panel, MG scores are uniformly low and compressed across all reference clusters,
but the margin between top and second label remains just above the pruning threshold.
The resulting assignments are biologically uninterpretable and confirm the absence of a
microglial reference in the Qian panel.

***

### Correspondence heatmap

***

Replicating the approach of Qian et al. Extended Data Figure 3d, the heatmap below shows
the **fraction of cells from each L1 population whose winning SingleR label was each Qian
H2 cluster**. Each row sums to 1. Columns are ordered by expected biological correspondence
to produce diagonal enrichment. The bar chart on the right shows the maximum fraction
value per row.

H2 clusters are grouped as follows:

- **Upper layer EN** (EN-L2, EN-IT-L2/3, EN-IT-L3/4, EN-IT-L4, EN-IT-L4/5) → expected to enrich for Glu-UL
- **Deep layer EN** (EN-IT-L6, EN-ET-L5/6, EN-ET-SP\*, EN-ET-L6-early) → expected to enrich for Glu-DL
- **Inhibitory** (IN-CGE, IN-MGE, IN-SST, IN-VZ/GE, INP-VZ/GE) → expected to enrich for GABA
- **Progenitors** (RG1, oRG1, tRG, vRG-late, IPC\*) → expected to enrich for NPC
- **OPC, EC** → expected to enrich for OPC and Endo-Peri respectively
- **Migrating/immature EN, Astrocytes** → no expected L1 match (right margin)

**Note on Glu-UL signal strength:** The Glu-UL row shows diffuse enrichment across the
Upper EN block rather than a single dominant hit. This is expected: upper layer neurons
at 12–21 PCW are less transcriptionally mature and more heterogeneous than deep layer
neurons at the same developmental stage. The subplate (EN-ET-SP-P) dominates Glu-DL
with a strong signal (~60%) because subplate neurons are among the earliest and most
transcriptionally distinct excitatory neurons, and are already well-established by
mid-gestation. The weaker Glu-UL signal reflects genuine developmental biology rather
than a classification failure, and is consistent with the known trajectory of upper
layer neuron maturation occurring later in cortical development.

***

```{r heatmap_combined, fig.width=20, fig.height=5}

# ── Build fraction matrix from winning labels ─────────────────────────────────────────

frac_mat <- singler %>%
  group_by(cell_type, H2_label) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(cell_type) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup() %>%
  select(cell_type, H2_label, fraction) %>%
  pivot_wider(
    names_from  = H2_label,
    values_from = fraction,
    values_fill = 0
  ) %>%
  column_to_rownames('cell_type')

frac_mat <- frac_mat[
  l1_order[l1_order %in% rownames(frac_mat)],
  h2_order[h2_order %in% colnames(frac_mat)]
]

# ── Colour scale ──────────────────────────────────────────────────────────────────────

heatmap_colours <- colorRampPalette(
  c('#FFFFFF', '#F2E8F5', '#C9A8D4', '#8B5BA6', '#4B1C6E')
)(100)

# ── Row annotation ────────────────────────────────────────────────────────────────────

row_ann <- data.frame(
  `Cell Type` = rownames(frac_mat),
  row.names   = rownames(frac_mat),
  check.names = FALSE
)
row_ann_colours <- list(`Cell Type` = custom_palette[rownames(frac_mat)])

# ── Column annotation ─────────────────────────────────────────────────────────────────

col_ann <- data.frame(
  Group     = h2_group_map[colnames(frac_mat)],
  row.names = colnames(frac_mat),
  check.names = FALSE
)

group_colours <- c(
  'Upper EN'   = '#4363d8',
  'Deep EN'    = '#00B6EB',
  'Inhibitory' = '#3CBB75',
  'Progenitor' = '#FF5959',
  'OPC'        = '#FDE725',
  'EC'         = '#B200ED',
  'Migrating'  = '#D3D3D3',
  'Astrocyte'  = '#A9A9A9'
)

col_ann_colours <- list(Group = group_colours)

# ── Max fraction bar chart ────────────────────────────────────────────────────────────

max_fractions <- frac_mat %>%
  as.data.frame() %>%
  rownames_to_column('cell_type') %>%
  pivot_longer(-cell_type, names_to = 'H2', values_to = 'fraction') %>%
  group_by(cell_type) %>%
  summarise(max_fraction = max(fraction), .groups = 'drop') %>%
  mutate(cell_type = factor(cell_type, levels = rev(l1_order)))

bar_plt <- ggplot(max_fractions, aes(x = cell_type, y = max_fraction, fill = cell_type)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = custom_palette, guide = 'none') +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = 'Max fraction',
    title = 'Max fraction\nper L1 population'
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title  = element_text(hjust = 0.5, face = 'bold', size = 11),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.title  = element_text(size = 10)
  )

# ── Draw heatmap then combine with bar chart using cowplot ────────────────────────────

heatmap_grob <- pheatmap(
  frac_mat,
  color                = heatmap_colours,
  cluster_rows         = FALSE,
  cluster_cols         = FALSE,
  annotation_row       = row_ann,
  annotation_col       = col_ann,
  annotation_colors    = c(row_ann_colours, col_ann_colours),
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  border_color         = NA,
  fontsize             = 9,
  fontsize_row         = 10,
  fontsize_col         = 8,
  angle_col            = 45,
  legend_breaks        = c(0, 0.25, 0.5, 0.75, 1.0),
  legend_labels        = c('0.0', '0.25', '0.5', '0.75', '1.0'),
  main                 = 'Correspondence with Qian et al. 2025 MERFISH H2 clusters (fraction of cells with winning label)',
  silent               = TRUE
)

plot_grid(
  as.grob(heatmap_grob),
  bar_plt,
  ncol        = 2,
  rel_widths  = c(4, 0.6),
  align       = 'h',
  axis        = 'tb'
)

```

***

### H2 group breakdown

***

The table below summarises the percentage of cells from each L1 population that mapped
to each aggregated H2 group. This provides a cleaner summary statistic than the full
heatmap — the key comparison is the fraction of Glu-UL cells mapping to Upper EN versus
Deep EN, and vice versa for Glu-DL.

***

```{r group_breakdown_table}

# Compute fraction of cells per L1 population mapping to each H2 group
group_breakdown <- singler %>%
  mutate(H2_group = h2_group_map[H2_label]) %>%
  group_by(cell_type, H2_group) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(cell_type) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(cell_type, H2_group, pct) %>%
  pivot_wider(
    names_from  = H2_group,
    values_from = pct,
    values_fill = 0
  )

# Order rows by L1 and columns by biological relevance
col_order <- c('cell_type', 'Upper EN', 'Deep EN', 'Inhibitory',
               'Progenitor', 'OPC', 'EC', 'Migrating', 'Astrocyte')
col_order <- col_order[col_order %in% colnames(group_breakdown)]

group_breakdown <- group_breakdown %>%
  arrange(match(cell_type, l1_order)) %>%
  select(all_of(col_order))

group_breakdown %>%
  gt() %>%
  tab_header(
    title    = 'Table 2. H2 group breakdown per L1 population',
    subtitle = 'Percentage of cells whose winning SingleR label fell in each H2 group'
  ) %>%
  cols_label(cell_type = 'Cell Type') %>%
  # Highlight the diagonal - expected correct group per L1 population
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'Upper EN', rows = cell_type == 'Glu-UL')
  ) %>%
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'Deep EN', rows = cell_type == 'Glu-DL')
  ) %>%
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'Inhibitory', rows = cell_type == 'GABA')
  ) %>%
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'Progenitor', rows = cell_type == 'NPC')
  ) %>%
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'OPC', rows = cell_type == 'OPC')
  ) %>%
  tab_style(
    style     = cell_fill(color = '#EAF4FB'),
    locations = cells_body(columns = 'EC', rows = cell_type == 'Endo-Peri')
  ) %>%
  tab_style(
    style     = cell_text(weight = 'bold'),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.font.size           = px(14),
    table.border.top.style    = 'none',
    table.border.bottom.style = 'solid',
    table.border.bottom.width = px(2),
    table.border.bottom.color = 'black',
    heading.align             = 'center'
  )

```

***

### UMAP

***

```{r umap_combined, fig.width=16, fig.height=12}

umap_theme <- theme_classic(base_size = 11) +
  theme(
    plot.title   = element_text(hjust = 0.5, face = 'bold', size = 11),
    legend.text  = element_text(size = 8),
    axis.text    = element_blank(),
    axis.ticks   = element_blank(),
    axis.line    = element_blank(),
    legend.key.size = unit(0.4, 'cm')
  )

# Panel A: L1 cell type
p_l1 <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
  geom_point(size = 0.1, alpha = 0.5) +
  scale_colour_manual(values = custom_palette, name = 'Cell type') +
  labs(title = 'A: L1 cell type annotations') +
  umap_theme +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

# Panel B: H2 group (aggregated)
group_order <- c('Upper EN', 'Deep EN', 'Inhibitory', 'Progenitor',
                 'OPC', 'EC', 'Migrating', 'Astrocyte')

p_h2_group <- umap %>%
  mutate(H2_group = factor(H2_group, levels = group_order)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = H2_group)) +
  geom_point(size = 0.1, alpha = 0.4) +
  scale_colour_manual(values = group_palette, name = 'H2 group', na.value = 'grey80') +
  labs(title = 'B: SingleR H2 groups (aggregated)') +
  umap_theme +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

# Panel C: H2 label (full 33 clusters)
p_h2 <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = H2_label)) +
  geom_point(size = 0.1, alpha = 0.4) +
  labs(title = 'C: SingleR H2 labels (all clusters)', colour = 'H2 label') +
  umap_theme +
  guides(colour = guide_legend(
    override.aes = list(size = 3, alpha = 1),
    ncol = 2
  ))

# Panel D: PCW
p_pcw <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = as.numeric(PCW))) +
  geom_point(size = 0.1, alpha = 0.5) +
  scale_colour_viridis_c(name = 'PCW', option = 'plasma') +
  labs(title = 'D: Post-conception weeks (PCW)') +
  umap_theme

plot_grid(p_l1, p_h2_group, p_h2, p_pcw,
          ncol   = 2,
          nrow   = 2,
          align  = 'hv',
          labels = NULL)

```

***

### Methods note

***

**Reference dataset:** Qian et al. (2025) *Nature*. MERFISH spatial transcriptomics of
the developing human cortex. 18 million cells profiled across 8 cortical areas and
7 developmental timepoints (GW15–GW34) with a 300-gene panel.

**Reference preparation:** V1/V2 cortical areas were excluded from the reference as these
regions show transcriptional discontinuity from the rest of the cortex (Qian et al.).
All gestational weeks were retained as H2 cluster labels were assigned across all
timepoints jointly. The reference was subsampled to 500 cells per H2 cluster
(33 clusters, 16,500 cells total) — sufficient for stable cluster median estimation
in SingleR.

**Gene overlap:** 270 of 300 MERFISH panel genes (90%) were present in our snRNA-seq
data. The 30 absent genes are likely lowly expressed or absent in dissociated nuclei
relative to intact tissue.

**Classification:** SingleR (v2.14.0) was run with `classifySingleR()` using 8 threads
and logNorm-normalised counts. Only 58 of 679,738 cells (0.009%) were pruned as low
confidence. All classifications used `assay.type = "logcounts"`.

**Heatmap metric (v1.1):** The heatmap shows the fraction of cells in each L1 population
whose winning SingleR label was each H2 cluster. This is computed from hard per-cell
label assignments rather than mean scores, directly replicating the metric used in Qian
et al. Extended Data Figure 3d.

**H2 group breakdown table (v1.2):** Winning labels are aggregated into H2 groups
(Upper EN, Deep EN, Inhibitory, Progenitor, OPC, EC, Migrating, Astrocyte) and
reported as percentages per L1 population. Expected diagonal cells are highlighted.

**Interpretation:** Diagonal enrichment in the heatmap and group breakdown table
confirms that our transcriptionally-defined L1 populations correspond to
spatially-validated Qian H2 clusters. Glu-DL shows strong enrichment in the Deep EN
group (~60% to EN-ET-SP-P alone), consistent with the early transcriptional
specification of subplate neurons. Glu-UL enrichment in the Upper EN group is more
diffuse, reflecting the known later maturation of upper layer neurons at mid-gestation
PCW stages. The Microglia (MG) row serves as a negative control — assignments are
biologically uninterpretable and the low pruning rate (3 cells) reflects score
compression across reference clusters lacking any microglial representation.

***
