---
title: "Spatial Validation: SingleR Label Transfer"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  singler_csv: null
  umap_csv:    null
  pred_rds:    null
  sce_rds:     null
---

***

## Changelog

| Version | Date       | Changes |
|---------|------------|---------|
| v1.0    | 2026-05-19 | Initial report. Heatmap used mean SingleR scores row-normalised to fractions. |
| v1.1    | 2026-05-19 | Fixed heatmap metric: switched from mean scores to fraction of cells with winning H2 label. Added aggregated H2 group UMAP tab. Updated MG negative control note. |
| v1.2    | 2026-05-19 | Combined heatmap and bar chart. Merged UMAPs into single 2x2 panel. Added H2 group breakdown table. Added interpretive note on Glu-UL signal strength. |
| v1.3    | 2026-05-19 | Rewrote heatmaps using ComplexHeatmap with native anno_barplot(). Added mean score heatmap alongside winning label heatmap. Updated colour scheme to blue-yellow-red for clarity. Added full SingleR diagnostics section: delta distribution, score heatmap, leiden concordance. |

***

## Overview

***

This report presents the results of SingleR-based label transfer from the Qian et al. 2025
MERFISH spatial transcriptomics dataset onto our snRNA-seq data. The analysis addresses
Reviewer Comment 2, which requested spatial and laminar validation of our L1 cell type
annotations — specifically the distinction between upper layer (Glu-UL) and deep layer
(Glu-DL) glutamatergic neuron populations.

**Reference:** Qian et al. 2025, *Nature*. Single-cell spatial transcriptomics of the
developing human cortex (MERFISH, 300-gene panel, GW15–GW34).

**Method:** SingleR (Bioconductor) was used to transfer H2-level spatial annotations
from the Qian MERFISH reference onto all 679,738 cells in our snRNA-seq dataset.
The reference was subsampled to 500 cells per H2 cluster (33 clusters, 16,500 cells total)
after excluding V1/V2 cortical areas (transcriptionally discontinuous per Qian et al.).
270 of 300 MERFISH panel genes (90%) were shared with our snRNA-seq data.

***

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
library(DT)
library(gt)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(SingleR)
library(matrixStats)

knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)

# ── Parameters ────────────────────────────────────────────────────────────────────────

singler_csv <- params$singler_csv
umap_csv    <- params$umap_csv
pred_rds    <- params$pred_rds
sce_rds     <- params$sce_rds

# ── Cell type colour palette (from scanpy_gene_lists.py) ─────────────────────────────

custom_palette <- c(
  'Glu-UL'    = '#4363d8',
  'Glu-DL'    = '#00B6EB',
  'GABA'      = '#3CBB75',
  'NPC'       = '#FF5959',
  'OPC'       = '#FDE725',
  'MG'        = '#F58231',
  'Endo-Peri' = '#B200ED'
)

# ── H2 group colour palette ───────────────────────────────────────────────────────────

group_palette <- c(
  'Upper EN'   = '#4363d8',
  'Deep EN'    = '#00B6EB',
  'Inhibitory' = '#3CBB75',
  'Progenitor' = '#FF5959',
  'OPC'        = '#FDE725',
  'EC'         = '#B200ED',
  'Migrating'  = '#A0A0A0',
  'Astrocyte'  = '#D3D3D3'
)

# ── Heatmap colour scale: white -> yellow -> orange -> dark red ───────────────────────

heatmap_col_fun <- colorRamp2(
  c(0, 0.15, 0.35, 0.6),
  c('#FFFFFF', '#FFFFB2', '#FD8D3C', '#800026')
)

# ── L1 population order ───────────────────────────────────────────────────────────────

l1_order <- c('Glu-UL', 'Glu-DL', 'GABA', 'NPC', 'OPC', 'MG', 'Endo-Peri')

# ── H2 column order grouped by biological correspondence ─────────────────────────────

h2_order <- c(
  'EN-L2', 'EN-IT-L2/3', 'EN-IT-L3/4', 'EN-IT-L4', 'EN-IT-L4/5',
  'EN-IT-L6', 'EN-ET-L5/6', 'EN-ET-SP-early', 'EN-ET-SP', 'EN-ET-SP-P', 'EN-ET-L6-early',
  'IN-CGE', 'IN-MGE', 'IN-SST', 'IN-VZ/GE', 'INP-VZ/GE',
  'RG1', 'oRG1', 'tRG', 'vRG-late',
  'IPC-VZ/SVZ', 'IPC-iSVZ', 'IPC-SVZ-1', 'IPC-SVZ-2', 'IPC-oSVZ',
  'OPC', 'EC',
  'EN-IZ-1', 'EN-IZ-2', 'EN-oSVZ-1', 'En-oSVZ-2',
  'Astro-1', 'Astro-late1'
)

# ── H2 to group mapping ───────────────────────────────────────────────────────────────

h2_group_map <- c(
  'EN-L2'          = 'Upper EN',   'EN-IT-L2/3'     = 'Upper EN',
  'EN-IT-L3/4'     = 'Upper EN',   'EN-IT-L4'       = 'Upper EN',
  'EN-IT-L4/5'     = 'Upper EN',
  'EN-IT-L6'       = 'Deep EN',    'EN-ET-L5/6'     = 'Deep EN',
  'EN-ET-SP-early' = 'Deep EN',    'EN-ET-SP'       = 'Deep EN',
  'EN-ET-SP-P'     = 'Deep EN',    'EN-ET-L6-early' = 'Deep EN',
  'IN-CGE'         = 'Inhibitory', 'IN-MGE'         = 'Inhibitory',
  'IN-SST'         = 'Inhibitory', 'IN-VZ/GE'       = 'Inhibitory',
  'INP-VZ/GE'      = 'Inhibitory',
  'RG1'            = 'Progenitor', 'oRG1'           = 'Progenitor',
  'tRG'            = 'Progenitor', 'vRG-late'       = 'Progenitor',
  'IPC-VZ/SVZ'     = 'Progenitor', 'IPC-iSVZ'       = 'Progenitor',
  'IPC-SVZ-1'      = 'Progenitor', 'IPC-SVZ-2'      = 'Progenitor',
  'IPC-oSVZ'       = 'Progenitor',
  'OPC'            = 'OPC',
  'EC'             = 'EC',
  'EN-IZ-1'        = 'Migrating',  'EN-IZ-2'        = 'Migrating',
  'EN-oSVZ-1'      = 'Migrating',  'En-oSVZ-2'      = 'Migrating',
  'Astro-1'        = 'Astrocyte',  'Astro-late1'    = 'Astrocyte'
)

# ── Load data ─────────────────────────────────────────────────────────────────────────

singler <- read_csv(singler_csv, show_col_types = FALSE)
umap    <- read_csv(umap_csv,    show_col_types = FALSE)
pred    <- readRDS(pred_rds)
sce_obj <- readRDS(sce_rds)
query_sub <- sce_obj$query_sub
pred_sub  <- sce_obj$pred_sub

singler$cell_type <- factor(singler$cell_type, levels = l1_order)
umap$cell_type    <- factor(umap$cell_type,    levels = l1_order)

umap <- umap %>%
  left_join(singler %>% select(cell_id, H2_label), by = 'cell_id') %>%
  mutate(H2_group = h2_group_map[H2_label])

# Score columns
score_cols <- setdiff(colnames(singler),
                      c('cell_id', 'cell_type', 'H2_label', 'H2_pruned', 'delta'))

# ── Helper: build fraction matrix from winning labels ─────────────────────────────────

build_frac_mat <- function(df) {
  mat <- df %>%
    group_by(cell_type, H2_label) %>%
    summarise(n = n(), .groups = 'drop') %>%
    group_by(cell_type) %>%
    mutate(fraction = n / sum(n)) %>%
    ungroup() %>%
    select(cell_type, H2_label, fraction) %>%
    pivot_wider(names_from = H2_label, values_from = fraction, values_fill = 0) %>%
    column_to_rownames('cell_type')
  mat[l1_order[l1_order %in% rownames(mat)],
      h2_order[h2_order %in% colnames(mat)]]
}

# ── Helper: build mean score matrix ──────────────────────────────────────────────────

build_score_mat <- function(df) {
  mat <- df %>%
    select(cell_type, all_of(score_cols)) %>%
    group_by(cell_type) %>%
    summarise(across(everything(), mean), .groups = 'drop') %>%
    column_to_rownames('cell_type')
  mat <- mat / rowSums(mat)
  mat[l1_order[l1_order %in% rownames(mat)],
      h2_order[h2_order %in% colnames(mat)]]
}

frac_mat  <- build_frac_mat(singler)
score_mat <- build_score_mat(singler)

# ── Helper: draw ComplexHeatmap ───────────────────────────────────────────────────────

draw_ch <- function(mat, title) {

  # Row annotation: L1 cell type colour bar
  row_ha <- rowAnnotation(
    `Cell type` = rownames(mat),
    col         = list(`Cell type` = custom_palette[rownames(mat)]),
    show_legend = TRUE,
    annotation_width = unit(3, 'mm')
  )

  # Column annotation: H2 group colour bar
  h2_grps <- h2_group_map[colnames(mat)]
  col_ha  <- HeatmapAnnotation(
    Group = h2_grps,
    col   = list(Group = group_palette[unique(h2_grps)]),
    show_legend = TRUE,
    annotation_height = unit(3, 'mm')
  )

  # Max fraction bar chart as right row annotation
  max_vals <- apply(mat, 1, max)
  bar_ha   <- rowAnnotation(
    `Max fraction` = anno_barplot(
      max_vals,
      width       = unit(2, 'cm'),
      gp          = gpar(fill = custom_palette[names(max_vals)], col = NA),
      axis_param  = list(side = 'bottom', labels_rot = 0)
    )
  )

  Heatmap(
    as.matrix(mat),
    name                  = 'Fraction',
    col                   = heatmap_col_fun,
    cluster_rows          = FALSE,
    cluster_columns       = FALSE,
    show_row_names        = TRUE,
    show_column_names     = TRUE,
    row_names_side        = 'right',
    column_names_rot      = 45,
    column_names_gp       = gpar(fontsize = 8),
    row_names_gp          = gpar(fontsize = 10),
    top_annotation        = col_ha,
    left_annotation       = row_ha,
    right_annotation      = bar_ha,
    column_title          = title,
    column_title_gp       = gpar(fontsize = 11, fontface = 'bold'),
    border                = FALSE,
    use_raster            = TRUE,
    raster_quality        = 2
  )
}

```

***

## Classification summary

***

```{r summary_table}

cell_counts <- singler %>%
  count(cell_type, name = 'N cells') %>%
  arrange(match(cell_type, l1_order))

pruning <- singler %>%
  group_by(cell_type) %>%
  summarise(
    `N pruned` = sum(is.na(H2_pruned)),
    `% pruned` = sprintf('%.2f%%', 100 * mean(is.na(H2_pruned))),
    .groups = 'drop'
  )

top_h2 <- singler %>%
  group_by(cell_type) %>%
  count(H2_label, name = 'n') %>%
  mutate(pct = sprintf('%.1f%%', 100 * n / sum(n))) %>%
  slice_max(n, n = 1) %>%
  rename(`Top H2 label` = H2_label, `N assigned` = n, `% of population` = pct) %>%
  ungroup()

delta_summary <- singler %>%
  group_by(cell_type) %>%
  summarise(
    `Mean delta` = round(mean(delta, na.rm = TRUE), 3),
    `Median delta` = round(median(delta, na.rm = TRUE), 3),
    .groups = 'drop'
  )

summary_tbl <- cell_counts %>%
  left_join(pruning,       by = 'cell_type') %>%
  left_join(top_h2,        by = 'cell_type') %>%
  left_join(delta_summary, by = 'cell_type')

summary_tbl %>%
  gt() %>%
  tab_header(
    title    = 'Table 1. SingleR classification summary',
    subtitle = 'Per L1 population counts, pruning rates, top H2 assignment, and delta statistics'
  ) %>%
  cols_label(cell_type = 'Cell Type') %>%
  tab_style(
    style     = cell_text(weight = 'bold'),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.font.size           = px(14),
    table.border.top.style    = 'none',
    table.border.bottom.style = 'solid',
    table.border.bottom.width = px(2),
    table.border.bottom.color = 'black',
    heading.align             = 'center'
  )

```

***

**Note on Microglia (MG):** Microglia are absent from the Qian MERFISH reference panel.
MG cells receive labels rather than NA because SingleR's pruning threshold is based on
outlier detection within each label — with no microglial genes in the 300-gene panel,
MG scores are uniformly compressed and the delta margin just clears the pruning threshold.
MG assignments are biologically uninterpretable and the MG row serves as a **negative control**.

***

## Correspondence heatmap {.tabset}

***

H2 clusters are grouped as follows:

- **Upper layer EN** (EN-L2, EN-IT-L2/3, EN-IT-L3/4, EN-IT-L4, EN-IT-L4/5) → expected to enrich for Glu-UL
- **Deep layer EN** (EN-IT-L6, EN-ET-L5/6, EN-ET-SP\*, EN-ET-L6-early) → expected to enrich for Glu-DL
- **Inhibitory** (IN-CGE, IN-MGE, IN-SST, IN-VZ/GE, INP-VZ/GE) → expected to enrich for GABA
- **Progenitors** (RG1, oRG1, tRG, vRG-late, IPC\*) → expected to enrich for NPC
- **OPC, EC** → expected to enrich for OPC and Endo-Peri
- **Migrating/immature EN, Astrocytes** → no expected L1 match (right margin)

**Note on Glu-UL signal strength:** Glu-UL shows diffuse enrichment across the Upper EN
block rather than a single dominant hit. This is expected — upper layer neurons at 12–21 PCW
are less transcriptionally mature and more heterogeneous than deep layer neurons at the same
stage. Subplate neurons (EN-ET-SP-P) dominate Glu-DL with a strong signal (~60%) because
they are among the earliest and most transcriptionally distinct excitatory neurons, already
well-established by mid-gestation. The weaker Glu-UL signal reflects genuine developmental
biology rather than a classification failure.

***

### Winning label fractions

***

Fraction of cells from each L1 population whose **winning** SingleR label was each Qian H2
cluster. Each row sums to 1. Directly replicates the metric in Qian et al. Extended Data Fig 3d.

***

```{r heatmap_winning, fig.width=16, fig.height=5}

draw_ch(frac_mat, 'Correspondence with Qian et al. 2025 MERFISH H2 clusters\n(fraction of cells with winning label)')

```

***

### Mean scores (row-normalised)

***

Mean SingleR score per L1 population per H2 cluster, row-normalised to fractions. This was
the v1.0 metric — shown here for comparison. It is less sharp than the winning label metric
but captures the continuous score distribution.

***

```{r heatmap_mean, fig.width=16, fig.height=5}

draw_ch(score_mat, 'Correspondence with Qian et al. 2025 MERFISH H2 clusters\n(mean scores, row-normalised)')

```

***

## H2 group breakdown

***

```{r group_breakdown_table}

group_breakdown <- singler %>%
  mutate(H2_group = h2_group_map[H2_label]) %>%
  group_by(cell_type, H2_group) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(cell_type) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(cell_type, H2_group, pct) %>%
  pivot_wider(names_from = H2_group, values_from = pct, values_fill = 0)

col_order <- c('cell_type', 'Upper EN', 'Deep EN', 'Inhibitory',
               'Progenitor', 'OPC', 'EC', 'Migrating', 'Astrocyte')
col_order <- col_order[col_order %in% colnames(group_breakdown)]

group_breakdown <- group_breakdown %>%
  arrange(match(cell_type, l1_order)) %>%
  select(all_of(col_order))

group_breakdown %>%
  gt() %>%
  tab_header(
    title    = 'Table 2. H2 group breakdown per L1 population',
    subtitle = 'Percentage of cells whose winning SingleR label fell in each H2 group. Expected diagonal cells highlighted.'
  ) %>%
  cols_label(cell_type = 'Cell Type') %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'Upper EN',   rows = cell_type == 'Glu-UL')) %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'Deep EN',    rows = cell_type == 'Glu-DL')) %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'Inhibitory', rows = cell_type == 'GABA')) %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'Progenitor', rows = cell_type == 'NPC')) %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'OPC',        rows = cell_type == 'OPC')) %>%
  tab_style(style = cell_fill(color = '#EAF4FB'),
            locations = cells_body(columns = 'EC',         rows = cell_type == 'Endo-Peri')) %>%
  tab_style(style = cell_text(weight = 'bold'),
            locations = cells_column_labels(everything())) %>%
  tab_options(
    table.font.size           = px(14),
    table.border.top.style    = 'none',
    table.border.bottom.style = 'solid',
    table.border.bottom.width = px(2),
    table.border.bottom.color = 'black',
    heading.align             = 'center'
  )

```

***

## SingleR diagnostics {.tabset}

***

### Delta distribution

***

The per-cell delta is the difference between the score for the assigned label and the median
score across all labels. High deltas indicate confident, unambiguous assignments. Low deltas
indicate the cell's true type may be absent from the reference or that scores are uniformly
compressed. Points are coloured by whether the cell was pruned (NA) by SingleR's outlier-based
threshold. Note that with only 58 cells pruned overall, the pruning threshold was very
permissive — the delta distributions themselves are more informative than the pruning counts.

***

```{r delta_distribution, fig.width=14, fig.height=7}

plotDeltaDistribution(pred, ncol = 6)

```

***

### Score heatmap (subsampled)

***

Per-cell score heatmap for a random subsample of 200 cells per L1 population (1,400 cells
total). Each column is a cell; each row is a Qian H2 reference label. Cells are grouped by
their L1 annotation. Unambiguous assignments show one dark row per cell; flat profiles
indicate uncertain assignments. Note that scores shown are pre-fine-tuning and the visually
top-scoring label may differ from the final assigned label.

***

```{r score_heatmap, fig.width=14, fig.height=8}

plotScoreHeatmap(
  pred_sub,
  annotation_col = as.data.frame(
    colData(query_sub)[, c('cell_type'), drop = FALSE]
  ),
  show.pruned = TRUE
)

```

***

### Leiden concordance

***

Concordance between your original unsupervised leiden_harmony_0.2 clusters and the SingleR
H2 winning labels. Each row is a leiden cluster and each column is a Qian H2 label. The
fraction of cells in each leiden cluster assigned to each H2 label is shown. Strong diagonal
enrichment here — where a leiden cluster maps predominantly to a single H2 label — provides
additional validation that the label transfer is capturing real transcriptional structure
rather than noise.

***

```{r leiden_concordance, fig.width=14, fig.height=5}

# Leiden cluster is stored in the umap CSV
leiden_concordance <- umap %>%
  left_join(singler %>% select(cell_id, H2_label), by = 'cell_id') %>%
  group_by(leiden_harmony_0.2, H2_label) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(leiden_harmony_0.2) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup() %>%
  select(leiden_harmony_0.2, H2_label, fraction) %>%
  pivot_wider(names_from = H2_label, values_from = fraction, values_fill = 0) %>%
  column_to_rownames('leiden_harmony_0.2')

# Order columns by h2_order
leiden_concordance <- leiden_concordance[,
  h2_order[h2_order %in% colnames(leiden_concordance)]]

# Column annotation: H2 group
h2_grps <- h2_group_map[colnames(leiden_concordance)]
col_ha  <- HeatmapAnnotation(
  Group = h2_grps,
  col   = list(Group = group_palette),
  show_legend = TRUE,
  annotation_height = unit(3, 'mm')
)

Heatmap(
  as.matrix(leiden_concordance),
  name              = 'Fraction',
  col               = heatmap_col_fun,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_side    = 'left',
  column_names_rot  = 45,
  column_names_gp   = gpar(fontsize = 8),
  row_names_gp      = gpar(fontsize = 10),
  top_annotation    = col_ha,
  column_title      = 'Leiden cluster vs H2 label concordance',
  column_title_gp   = gpar(fontsize = 11, fontface = 'bold'),
  border            = FALSE,
  use_raster        = TRUE
)

```

***

## UMAP

***

```{r umap_combined, fig.width=16, fig.height=12}

umap_theme <- theme_classic(base_size = 11) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = 'bold', size = 11),
    legend.text     = element_text(size = 8),
    axis.text       = element_blank(),
    axis.ticks      = element_blank(),
    axis.line       = element_blank(),
    legend.key.size = unit(0.4, 'cm')
  )

group_order <- c('Upper EN', 'Deep EN', 'Inhibitory', 'Progenitor',
                 'OPC', 'EC', 'Migrating', 'Astrocyte')

p_l1 <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
  geom_point(size = 0.1, alpha = 0.5) +
  scale_colour_manual(values = custom_palette, name = 'Cell type') +
  labs(title = 'A: L1 cell type annotations') +
  umap_theme +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

p_h2_group <- umap %>%
  mutate(H2_group = factor(H2_group, levels = group_order)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = H2_group)) +
  geom_point(size = 0.1, alpha = 0.4) +
  scale_colour_manual(values = group_palette, name = 'H2 group', na.value = 'grey80') +
  labs(title = 'B: SingleR H2 groups (aggregated)') +
  umap_theme +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

p_h2 <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = H2_label)) +
  geom_point(size = 0.1, alpha = 0.4) +
  labs(title = 'C: SingleR H2 labels (all clusters)', colour = 'H2 label') +
  umap_theme +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 2))

p_pcw <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = as.numeric(PCW))) +
  geom_point(size = 0.1, alpha = 0.5) +
  scale_colour_viridis_c(name = 'PCW', option = 'plasma') +
  labs(title = 'D: Post-conception weeks (PCW)') +
  umap_theme

plot_grid(p_l1, p_h2_group, p_h2, p_pcw, ncol = 2, nrow = 2, align = 'hv')

```

***

## Methods note

***

**Reference dataset:** Qian et al. (2025) *Nature*. MERFISH spatial transcriptomics of
the developing human cortex. 18 million cells across 8 cortical areas and 7 developmental
timepoints (GW15–GW34), 300-gene panel.

**Reference preparation:** V1/V2 excluded (transcriptionally discontinuous per Qian et al.).
All gestational weeks retained. Subsampled to 500 cells per H2 cluster (33 clusters,
16,500 cells total).

**Gene overlap:** 270 of 300 MERFISH panel genes (90%) present in our snRNA-seq data.

**Classification:** SingleR v2.14.0, `classifySingleR()`, 8 threads, logNorm counts,
`assay.type = "logcounts"`. 58 of 679,738 cells pruned (0.009%).

**Heatmap metrics:** Winning label fractions (v1.1+) use hard per-cell label assignments,
directly replicating Qian et al. Extended Data Fig 3d. Mean score heatmap (v1.0 metric)
shown for comparison.

**Colour scale (v1.3):** White–yellow–orange–dark red replaces purple gradient for
improved perceptual contrast of enrichment blocks.

**Diagnostics (v1.3):** Delta distribution and score heatmap use the built-in SingleR
`plotDeltaDistribution()` and `plotScoreHeatmap()` functions on a subsampled query
(200 cells per L1 population). Leiden concordance uses leiden_harmony_0.2 cluster
assignments from the original snRNA-seq analysis.

**Interpretation:** Diagonal enrichment confirms transcriptionally-defined L1 populations
correspond to spatially-validated Qian H2 clusters. Glu-DL shows strong Deep EN enrichment
(~60% to EN-ET-SP-P). Glu-UL enrichment in the Upper EN group is more diffuse, consistent
with the later maturation of upper layer neurons at mid-gestation stages. MG row is a
negative control — assignments are biologically uninterpretable.

***
