---
title: "SingleR - Label transfer"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  cameron_csv:     null
  polioudakis_csv: null
  nowakowski_csv:  null
  umap_csv:        null
---

***

## Overview

***

SingleR label transfer from three published fetal cortex snRNA-seq reference datasets
onto our 679,738-cell prenatal cortex snRNA-seq query. For each reference, classification
was run at two levels: fine-grained (reference-specific cluster labels) and L1 group
(mapped to our seven L1 cell type categories). Queries were subsetted to genes shared
with each reference prior to classification.

| Reference | Cells | Shared genes | Fine labels |
|-----------|-------|-------------|-------------|
| Cameron et al. 2023 | 10,083 | 14,840 | 17 |
| Polioudakis et al. 2019 | 33,986 | 15,561 | 16 |
| Nowakowski et al. 2017 | 4,261 | 13,308 | 48 |

***

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(knitr)
library(DT)
library(gt)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggplot2)
library(dplyr)
library(tidyr)

# Fix namespace conflicts from Bioconductor packages if loaded
count  <- dplyr::count
rename <- dplyr::rename
select <- dplyr::select
filter <- dplyr::filter

knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)

# ── Parameters ────────────────────────────────────────────────────────────────────────

cameron_csv     <- params$cameron_csv
polioudakis_csv <- params$polioudakis_csv
nowakowski_csv  <- params$nowakowski_csv
umap_csv        <- params$umap_csv

# ── Colour palette ─────────────────────────────────────────────────────────────────

custom_palette <- c(
  'Glu-UL'    = '#4363d8',
  'Glu-DL'    = '#00B6EB',
  'GABA'      = '#3CBB75',
  'NPC'       = '#FF5959',
  'OPC'       = '#FDE725',
  'MG'        = '#F58231',
  'Endo-Peri' = '#B200ED'
)

l1_order <- c('Glu-UL', 'Glu-DL', 'GABA', 'NPC', 'OPC', 'MG', 'Endo-Peri')

# Polychrome 36 palette - fixed set, no package needed
polychrome_36 <- c(
  "#5A5156", "#E4E1E3", "#F6222E", "#FE00FA", "#16FF32", "#3283FE",
  "#FEAF16", "#B00068", "#1CFFCE", "#90AD1C", "#2ED9FF", "#DEA0FD",
  "#AA0DFE", "#F8A19F", "#325A9B", "#C4451C", "#1C8356", "#85660D",
  "#B10DA1", "#FBE426", "#1CBE4F", "#FA0087", "#FC1CBF", "#F7E1A0",
  "#C075A6", "#782AB6", "#AAF400", "#BDCDFF", "#822E1C", "#B5EFB5",
  "#7ED7D1", "#1C7F93", "#D85FF7", "#683B79", "#66B0FF", "#3B00FB"
)

# ── Heatmap colour scale ──────────────────────────────────────────────────────────────

heatmap_col_fun <- colorRamp2(
  c(0, 0.15, 0.35, 0.6),
  c('#FFFFFF', '#FFFFB2', '#FD8D3C', '#800026')
)

# ── Reference fine label -> L1 group mappings ─────────────────────────────────────────

cameron_map <- c(
  'FC-ExN-1'   = 'Glu-UL',  'FC-ExN-2'   = 'Glu-UL',  'FC-ExN-3'   = 'Glu-UL',
  'FC-ExN-4'   = 'Glu-DL',  'FC-ExN-5'   = 'Glu-DL',
  'FC-InN-1'   = 'GABA',    'FC-InN-2'   = 'GABA',
  'FC-InN-3'   = 'GABA',    'FC-InN-4'   = 'GABA',
  'FC-RG-1'    = 'NPC',     'FC-RG-2'    = 'NPC',
  'FC-CycPro'  = 'NPC',     'FC-IP'      = 'NPC',
  'FC-OPC'     = 'OPC',     'FC-MG'      = 'MG',
  'FC-Endo'    = 'Endo-Peri',
  'FC-N-undef' = 'unassigned'
)

polioudakis_map <- c(
  'ExN'    = 'Glu-UL',   'ExM'    = 'Glu-UL',    'ExM-U'  = 'Glu-UL',
  'ExDp1'  = 'Glu-DL',   'ExDp2'  = 'Glu-DL',
  'InCGE'  = 'GABA',     'InMGE'  = 'GABA',
  'oRG'    = 'NPC',      'vRG'    = 'NPC',
  'PgS'    = 'NPC',      'PgG2M'  = 'NPC',        'IP'     = 'NPC',
  'OPC'    = 'OPC',      'Mic'    = 'MG',
  'End'    = 'Endo-Peri', 'Per'   = 'Endo-Peri'
)

nowakowski_map <- c(
  'EN-PFC1' = 'Glu-UL',   'EN-PFC2' = 'Glu-UL',   'EN-PFC3' = 'Glu-UL',
  'EN-V1-1' = 'Glu-UL',   'EN-V1-2' = 'Glu-UL',   'EN-V1-3' = 'Glu-UL',
  'nEN-early1' = 'Glu-UL', 'nEN-early2' = 'Glu-UL', 'nEN-late' = 'Glu-UL',
  'IN-CTX-MGE1' = 'GABA', 'IN-CTX-MGE2' = 'GABA',
  'IN-STR'  = 'GABA',
  'nIN1'    = 'GABA',    'nIN2'    = 'GABA',    'nIN3'    = 'GABA',
  'nIN4'    = 'GABA',    'nIN5'    = 'GABA',
  'oRG'     = 'NPC',     'vRG'     = 'NPC',     'tRG'     = 'NPC',
  'RG-early' = 'NPC',    'RG-div1' = 'NPC',     'RG-div2' = 'NPC',
  'IPC-div1' = 'NPC',    'IPC-div2' = 'NPC',
  'IPC-nEN1' = 'NPC',    'IPC-nEN2' = 'NPC',    'IPC-nEN3' = 'NPC',
  'MGE-div'  = 'NPC',    'MGE-IPC1' = 'NPC',    'MGE-IPC2' = 'NPC',
  'MGE-IPC3' = 'NPC',    'MGE-RG1'  = 'NPC',    'MGE-RG2'  = 'NPC',
  'OPC'        = 'OPC',
  'Microglia'  = 'MG',
  'Endothelial' = 'Endo-Peri', 'Mural' = 'Endo-Peri',
  'Astrocyte'  = 'unassigned', 'Choroid' = 'unassigned',
  'Glyc'       = 'unassigned',
  'U1' = 'unassigned',   'U2' = 'unassigned',
  'U3' = 'unassigned',   'U4' = 'unassigned'
)

# Group colour palette - includes ExN and nEN for Nowakowski
group_palette <- c(
  'Glu-UL'     = '#4363d8',
  'Glu-DL'     = '#00B6EB',
  'GABA'       = '#3CBB75',
  'NPC'        = '#FF5959',
  'OPC'        = '#FDE725',
  'MG'         = '#F58231',
  'Endo-Peri'  = '#B200ED',
  'unassigned' = '#D3D3D3'
)

# ── Load data ─────────────────────────────────────────────────────────────────────────

cameron     <- read_csv(cameron_csv,     show_col_types = FALSE)
polioudakis <- read_csv(polioudakis_csv, show_col_types = FALSE)
nowakowski  <- read_csv(nowakowski_csv,  show_col_types = FALSE)
umap        <- read_csv(umap_csv,        show_col_types = FALSE)

# Factor cell_type in L1 order
for (df_name in c('cameron', 'polioudakis', 'nowakowski')) {
  df <- get(df_name)
  df$cell_type <- factor(df$cell_type, levels = l1_order)
  assign(df_name, df)
}
umap$cell_type <- factor(umap$cell_type, levels = l1_order)

# ── Helper: build winning label fraction matrix ───────────────────────────────────────

build_frac_mat <- function(df, col = 'fine_label') {
  mat <- df %>%
    dplyr::group_by(cell_type, .data[[col]]) %>%
    dplyr::summarise(n = n(), .groups = 'drop') %>%
    dplyr::group_by(cell_type) %>%
    dplyr::mutate(fraction = n / sum(n)) %>%
    dplyr::ungroup() %>%
    dplyr::select(cell_type, label = .data[[col]], fraction) %>%
    pivot_wider(names_from = label, values_from = fraction, values_fill = 0) %>%
    column_to_rownames('cell_type')
  mat[l1_order[l1_order %in% rownames(mat)], , drop = FALSE]
}

# ── Helper: build mean score matrix ──────────────────────────────────────────────────

build_score_mat <- function(df, score_prefix = 'fine_score_') {
  score_cols <- grep(paste0('^', score_prefix), colnames(df), value = TRUE)
  clean_names <- sub(score_prefix, '', score_cols)
  mat <- df %>%
    dplyr::select(cell_type, all_of(score_cols)) %>%
    dplyr::group_by(cell_type) %>%
    dplyr::summarise(across(everything(), mean), .groups = 'drop') %>%
    column_to_rownames('cell_type')
  colnames(mat) <- clean_names
  mat <- mat / rowSums(mat)
  mat[l1_order[l1_order %in% rownames(mat)], , drop = FALSE]
}

# ── Helper: draw ComplexHeatmap with group annotation bar ────────────────────────────

draw_heatmap <- function(mat, group_map, title) {

  # Order columns by group to produce diagonal enrichment
  group_order <- c('Glu-UL', 'Glu-DL', 'GABA', 'unassigned',
                   'NPC', 'OPC', 'MG', 'Endo-Peri')

  col_groups_all <- group_map[colnames(mat)]
  col_groups_all[is.na(col_groups_all)] <- 'unassigned'

  # Sort columns by group order, then alphabetically within group
  col_order <- order(
    match(col_groups_all, group_order),
    colnames(mat)
  )
  mat <- mat[, col_order, drop = FALSE]

  # Row annotation: L1 cell type
  row_ha <- rowAnnotation(
    `Cell type` = rownames(mat),
    col         = list(`Cell type` = custom_palette[rownames(mat)]),
    show_legend = TRUE,
    annotation_width = unit(3, 'mm')
  )

  # Column annotation: reference group
  col_groups <- group_map[colnames(mat)]
  col_groups[is.na(col_groups)] <- 'unassigned'

  col_ha <- HeatmapAnnotation(
    Group = col_groups,
    col   = list(Group = group_palette[unique(col_groups)]),
    show_legend = TRUE,
    annotation_height = unit(3, 'mm')
  )

  # Max fraction bar on right
  max_vals <- apply(mat, 1, max)
  bar_ha   <- rowAnnotation(
    `Max fraction` = anno_barplot(
      max_vals,
      width      = unit(2, 'cm'),
      gp         = gpar(fill = custom_palette[names(max_vals)], col = NA),
      axis_param = list(side = 'bottom', labels_rot = 0)
    )
  )

  Heatmap(
    as.matrix(mat),
    name              = 'Fraction',
    col               = heatmap_col_fun,
    cluster_rows      = FALSE,
    cluster_columns   = FALSE,
    show_row_names    = TRUE,
    show_column_names = TRUE,
    row_names_side    = 'right',
    column_names_rot  = 45,
    column_names_gp   = gpar(fontsize = 8),
    row_names_gp      = gpar(fontsize = 10),
    top_annotation    = col_ha,
    left_annotation   = row_ha,
    right_annotation  = bar_ha,
    column_title      = title,
    column_title_gp   = gpar(fontsize = 11, fontface = 'bold'),
    border            = FALSE,
    use_raster        = TRUE,
    raster_quality    = 2
  )
}

# ── Helper: summary table ────────────────────────────────────────────────────────────

make_summary_table <- function(df, ref_name) {

  cell_counts <- df %>%
    dplyr::count(cell_type, name = 'N cells') %>%
    dplyr::arrange(match(cell_type, l1_order))

  pruning <- df %>%
    dplyr::group_by(cell_type) %>%
    dplyr::summarise(
      `% pruned (fine)`  = sprintf('%.2f%%', 100 * mean(is.na(fine_pruned))),
      `% pruned (group)` = sprintf('%.2f%%', 100 * mean(is.na(group_pruned))),
      `Mean delta (fine)`  = round(mean(delta_fine,  na.rm = TRUE), 3),
      `Mean delta (group)` = round(mean(delta_group, na.rm = TRUE), 3),
      .groups = 'drop'
    )

  top_fine <- df %>%
    dplyr::group_by(cell_type) %>%
    dplyr::count(fine_label, name = 'n') %>%
    dplyr::mutate(pct = sprintf('%.1f%%', 100 * n / sum(n))) %>%
    dplyr::slice_max(n, n = 1) %>%
    dplyr::rename(`Top fine label` = fine_label, `% of population` = pct) %>%
    dplyr::select(cell_type, `Top fine label`, `% of population`) %>%
    dplyr::ungroup()

  top_group <- df %>%
    dplyr::group_by(cell_type) %>%
    dplyr::count(group_label, name = 'n') %>%
    dplyr::mutate(pct = sprintf('%.1f%%', 100 * n / sum(n))) %>%
    dplyr::slice_max(n, n = 1) %>%
    dplyr::rename(`Top group label` = group_label, `% (group)` = pct) %>%
    dplyr::select(cell_type, `Top group label`, `% (group)`) %>%
    dplyr::ungroup()

  cell_counts %>%
    dplyr::left_join(top_fine,  by = 'cell_type') %>%
    dplyr::left_join(top_group, by = 'cell_type') %>%
    dplyr::left_join(pruning,   by = 'cell_type') %>%
    gt() %>%
    tab_header(
      title    = paste('Table:', ref_name),
      subtitle = 'L1 cell counts, top assignments and classification confidence'
    ) %>%
    cols_label(cell_type = 'Cell Type') %>%
    tab_style(
      style     = cell_text(weight = 'bold'),
      locations = cells_column_labels(everything())
    ) %>%
    tab_options(
      table.font.size           = px(13),
      table.border.top.style    = 'none',
      table.border.bottom.style = 'solid',
      table.border.bottom.width = px(2),
      table.border.bottom.color = 'black',
      heading.align             = 'center'
    )
}

# ── Helper: discrete palette for fine labels ──────────────────────────────────────────

make_fine_palette <- function(labels) {
  unique_labels <- sort(unique(na.omit(labels)))
  n             <- length(unique_labels)
  pal           <- rep(polychrome_36, length.out = n)
  names(pal)    <- unique_labels
  pal
}

# ── Helper: 2x2 UMAP panel ───────────────────────────────────────────────────────────

make_umap_panel <- function(df_singler, ref_name) {

  # Join both group and fine labels to UMAP
  umap_joined <- umap %>%
    dplyr::left_join(
      df_singler %>% dplyr::select(cell_id, group_label, fine_label),
      by = 'cell_id'
    ) %>%
    dplyr::mutate(group_label = factor(group_label,
                                       levels = names(group_palette)))

  # Generate discrete palette sized to fine labels in this reference
  fine_pal <- make_fine_palette(umap_joined$fine_label)

  # Panel A: original L1 annotations
  p_l1 <- ggplot(umap, aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
    geom_point(size = 0.1, alpha = 0.4) +
    scale_colour_manual(values = custom_palette, name = 'L1 cell type') +
    labs(title = 'A: L1 annotations (this study)') +
    umap_theme +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

  # Panel B: SingleR group label transfer
  p_group <- ggplot(umap_joined, aes(x = UMAP1, y = UMAP2, colour = group_label)) +
    geom_point(size = 0.1, alpha = 0.4) +
    scale_colour_manual(values = group_palette, name = 'SingleR group',
                        na.value = 'grey80') +
    labs(title = paste('B: SingleR group labels\n(', ref_name, ')')) +
    umap_theme +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

  # Panel C: SingleR fine labels with discrete palette
  p_fine <- ggplot(umap_joined, aes(x = UMAP1, y = UMAP2, colour = fine_label)) +
    geom_point(size = 0.1, alpha = 0.4) +
    scale_colour_manual(values = fine_pal, name = 'Fine label',
                        na.value = 'grey80') +
    labs(title = paste('C: SingleR fine labels\n(', ref_name, ')')) +
    umap_theme +
    guides(colour = guide_legend(
      override.aes = list(size = 3, alpha = 1),
      ncol = 2
    ))

  # 2x2 grid with blank 4th panel
  plot_grid(p_l1, p_group, p_fine, NULL,
            ncol  = 2,
            nrow  = 2,
            align = 'hv')
}
```

***

## Cameron et al. 2023 {.tabset}

***

**Reference:** 10,083 cells from frontal cortex (PFC), 30,290 genes, 17 cell types.
Shared genes with query: **14,840 (49.0% of reference)**.

***

### Summary table

***

```{r cameron_table}
make_summary_table(cameron, "Cameron et al. 2023")
```

***

### Heatmap {.tabset}

***

#### Winning label fractions

***

```{r cameron_heatmap_frac, fig.width=14, fig.height=5}
frac_mat <- build_frac_mat(cameron, 'fine_label')
draw_heatmap(frac_mat, cameron_map,
             'Cameron et al. 2023 - fraction of cells with winning fine label')
```

***

#### Mean scores (row-normalised)

***

```{r cameron_heatmap_scores, fig.width=14, fig.height=5}
score_mat <- build_score_mat(cameron, 'fine_score_')
draw_heatmap(score_mat, cameron_map,
             'Cameron et al. 2023 - mean scores (row-normalised)')
```

***

### UMAP

***

```{r cameron_umap, fig.width=14, fig.height=12}
make_umap_panel(cameron, "Cameron et al. 2023")
```

***

## Polioudakis et al. 2019 {.tabset}

***

**Reference:** 33,986 cells from frontal cortex, 35,543 genes, 16 cell types.
Shared genes with query: **15,561 (43.8% of reference)**.

***

### Summary table

***

```{r pol_table}
make_summary_table(polioudakis, "Polioudakis et al. 2019")
```

***

### Heatmap {.tabset}

***

#### Winning label fractions

***

```{r pol_heatmap_frac, fig.width=14, fig.height=5}
frac_mat <- build_frac_mat(polioudakis, 'fine_label')
draw_heatmap(frac_mat, polioudakis_map,
             'Polioudakis et al. 2019 - fraction of cells with winning fine label')
```

***

#### Mean scores (row-normalised)

***

```{r pol_heatmap_scores, fig.width=14, fig.height=5}
score_mat <- build_score_mat(polioudakis, 'fine_score_')
draw_heatmap(score_mat, polioudakis_map,
             'Polioudakis et al. 2019 - mean scores (row-normalised)')
```

***

### UMAP

***

```{r pol_umap, fig.width=14, fig.height=12}
make_umap_panel(polioudakis, "Polioudakis et al. 2019")
```

***

## Nowakowski et al. 2017 {.tabset}

***

**Reference:** 4,261 cells from developing cortex and GE, 56,864 genes, 48 cell types.
Shared genes with query: **13,308 (23.4% of reference)**.

**Note on EN-PFC and nEN clusters:** The Nowakowski 2017 cluster labels EN-PFC1/2/3, EN-V1-1/2/3
and nEN-early1/2/nEN-late do not carry explicit upper/deep layer designations in the original publication.
These are annotated here as ExN (area-specific excitatory neurons) and nEN (newborn excitatory neurons)
respectively, without layer assignment.

***

### Summary table

***

```{r now_table}
make_summary_table(nowakowski, "Nowakowski et al. 2017")
```

***

### Heatmap {.tabset}

***

#### Winning label fractions

***

```{r now_heatmap_frac, fig.width=18, fig.height=5}
frac_mat <- build_frac_mat(nowakowski, 'fine_label')
draw_heatmap(frac_mat, nowakowski_map,
             'Nowakowski et al. 2017 - fraction of cells with winning fine label')
```

***

#### Mean scores (row-normalised)

***

```{r now_heatmap_scores, fig.width=18, fig.height=5}
score_mat <- build_score_mat(nowakowski, 'fine_score_')
draw_heatmap(score_mat, nowakowski_map,
             'Nowakowski et al. 2017 - mean scores (row-normalised)')
```

***

### UMAP

***

```{r now_umap, fig.width=14, fig.height=12}
make_umap_panel(nowakowski, "Nowakowski et al. 2017")
```

***

## Methods note

***

**References:** Cameron et al. 2023 *Biological Psychiatry*; Polioudakis et al. 2019 *Cell*;
Nowakowski et al. 2017 *Science*.

**Classification:** SingleR v2.14.0 run with `classifySingleR()` using 8 threads.
Each reference was run twice — once with fine-grained cluster labels and once with L1 group labels.

**Memory strategy:** Query h5ad files store counts as sparse CSR in `layers/counts`. The R
classification script reads these sparse components directly via `rhdf5` without densification,
normalises with `scuttle::logNormCounts()` maintaining sparsity throughout, avoiding the
~200GB memory requirement of `anndataR::as_SingleCellExperiment()`.

**Gene overlap:** Queries were subsetted to genes shared with each reference prior to
classification. Shared gene counts: Cameron 14,840 (49.0%), Polioudakis 15,561 (43.8%),
Nowakowski 13,308 (23.4%).

**Heatmap colour scale:** White → pale yellow → orange → dark red (0 to 0.6+ fraction).
Row annotation bar shows L1 cell type colours. Column annotation bar shows the L1 group
each reference cluster maps to. Max fraction bar chart shown on right of each heatmap.

***
