---
title: "SMR summary"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  cell_types: null
  in_dir: null
  p_smr: null
  p_heidi: null
  bmark_dir: null
---

## SMR pipeline steps {.tabset}

***

We used the the Yang Lab toll for [SMR](https://yanglab.westlake.edu.cn/software/smr/#Overview).The citation for SMR is [Nature Genetics, 48:481-487](https://www.nature.com/articles/ng.3538).

***

```{r set_up, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(DT)
library(gt)
library(cowplot)

# Set variables from params
cell_types <- params$cell_types
in_dir <- params$in_dir
p_smr <- as.numeric(params$p_smr)
p_heidi <- as.numeric(params$p_heidi)
bmark_dir <- params$bmark_dir
```

***

1. `get_smr_binary`: Download the SMR binary file
2. `cat_refs`: Cat hg38 refs into a single file
3. `cat_tensorqtl_nom_snps`: Cat TensorQTL parquet files into single file for SMR
4. `get_snp_positions`: Get SNP positions from bim file
5. `create_query`: Create query file for SMR
6. `create_besd`: Create besd file for SMR
7. `format_gwas`: Convert GWAS into .ma format for SMR
8. `smr`: Run SMR
9. `smr_report`: Generate SMR report

***

Params:

- SMR p: `r p_smr`
  - Applying a BF correction P / Total_probes per cell type
- HEIDI p: `r p_heidi`

***

### Results {.tabset}

***

```{r functions, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
# Function to read and summarize SMR results
summarize_smr <- function(cell_type, in_dir, disorder) {
  file_path <- file.path(in_dir, cell_type, paste0(cell_type, "_", disorder, ".smr"))
  if (file.exists(file_path)) {
    smr <- suppressMessages(read_tsv(file_path))
    n_probes <- nrow(smr)
    smr_sig <- smr |> filter(p_SMR < (p_smr / n_probes))
    smr_sig_heidi <- smr_sig |> filter(p_HEIDI > p_heidi)
    return(data.frame(
      Cell_Type = cell_type,
      Total_Probes = n_probes,
      Significant_SMR = nrow(smr_sig),
      Significant_SMR_HEIDI = nrow(smr_sig_heidi)
    ))
  } else {
    message("File not found for cell type / disorder: ", cell_type, " / ", disorder)
    return(data.frame(
      Cell_Type = cell_type,
      Total_Probes = 0,
      Significant_SMR = 0,
      Significant_SMR_HEIDI = 0
    ))
  }
}

# Function to read and summarize SMR results
log_dir <- "../../results/00LOG/10SMR/"
extract_smr_log_params <- function(log_dir, cell_types, disorder) {
  # Initialize parameters to extract
  params <- c(
    "GWAS SNPs Included",
    "eQTL SNPs Included",
    "Individuals in LD Reference",
    "SNPs After Allele Check",
    "SNPs Excluded for Frequency",
    "Probes Analyzed"
  )
  
  # Initialize result data frame
  result <- data.frame(Parameter = params, stringsAsFactors = FALSE)
  
  # Process each cell type
  for (cell_type in cell_types) {
    log_file <- file.path(log_dir, paste0("smr_", cell_type, "_", disorder, ".log"))
    
    # Initialize default values
    values <- rep("Unknown", length(params))
    
    if (file.exists(log_file)) {
      log_content <- read_lines(log_file)
      
      # Extract parameters using regular expressions
      # GWAS SNPs Included
      gwas_snps <- str_extract(log_content, "GWAS summary data of (\\d+) SNPs to be included")
      gwas_snps <- gwas_snps[!is.na(gwas_snps)]
      if (length(gwas_snps) > 0) {
        values[1] <- str_extract(gwas_snps[1], "\\d+")
      }
      
      # eQTL SNPs Included
      eqtl_snps <- str_extract(log_content, "(\\d+) SNPs to be included from \\[.*\\.esi\\]")
      eqtl_snps <- eqtl_snps[!is.na(eqtl_snps)]
      if (length(eqtl_snps) > 0) {
        values[2] <- str_extract(eqtl_snps[1], "\\d+")
      }
      
      # Individuals in LD Reference
      individuals <- str_extract(log_content, "(\\d+) individuals to be included from \\[.*\\.fam\\]")
      individuals <- individuals[!is.na(individuals)]
      if (length(individuals) > 0) {
        values[3] <- str_extract(individuals[1], "\\d+")
      }
      
      # SNPs After Allele Check
      allele_check <- str_extract(log_content, "(\\d+) SNPs are included after allele checking")
      allele_check <- allele_check[!is.na(allele_check)]
      if (length(allele_check) > 0) {
        values[4] <- str_extract(allele_check[1], "\\d+")
      }
      
      # SNPs Excluded for Frequency
      freq_excluded <- str_extract(log_content, "(\\d+) SNPs.*with allele frequency differences > 0.20")
      freq_excluded <- freq_excluded[!is.na(freq_excluded)]
      if (length(freq_excluded) > 0) {
        values[5] <- str_extract(freq_excluded[1], "\\d+")
      }
      
      # Probes Analyzed
      probes <- str_extract(log_content, "SMR analysis \\(SMR and HEIDI tests\\) for (\\d+) probes")
      probes <- probes[!is.na(probes)]
      if (length(probes) > 0) {
        values[6] <- str_extract(probes[1], "\\d+")
      }
      
    } else {
      message("Log file not found for cell type: ", cell_type)
    }
    
    # Add values to result data frame
    result[[cell_type]] <- values
  }
  
  return(result)
}

get_heidi <- function(cell_type, in_dir, disorder) {
  file_path <- file.path(in_dir, cell_type, paste0(cell_type, "_", disorder, ".smr"))
  
  if (file.exists(file_path)) {
    smr <- suppressMessages(readr::read_tsv(file_path, show_col_types = FALSE))
    n_probes <- nrow(smr)
    
    smr %>%
      filter(p_SMR < (p_smr / n_probes), p_HEIDI > p_heidi & !is.na(p_HEIDI)) %>%
      mutate(Cell_Type = cell_type)
  } else {
    message("File not found for cell type / disorder: ", cell_type, " / ", disorder)
    return(NULL)
  }
}

# Get gene lookup file
gene_lookup_file <- '../../resources/sheets/gene_lookup_hg38.tsv' # Relative to Rmd file
if (file.exists(gene_lookup_file)) {
  
  gene_lookup_hg38_tbl <- suppressMessages(read_tsv(gene_lookup_file))
  
} else {
  
  library(biomaRt)
  
  ## Create lookup table 
  mart <- useMart('ENSEMBL_MART_ENSEMBL')
  mart <- useDataset('hsapiens_gene_ensembl', mart)
  
  gene_lookup_hg38_tbl <- getBM(
    mart = mart,
    attributes = c("ensembl_gene_id", "chromosome_name", "start_position", "end_position", 
                   "hgnc_symbol", "external_gene_name", "strand", "gene_biotype"),
    uniqueRows = TRUE)
  
  write_tsv(gene_lookup_hg38_tbl, gene_lookup_file)
  
}
```

***

#### SCZ {.tabset}

***

##### SMR summary

***

```{r smr_summary_scz, results='asis', echo=FALSE}
disorder <- 'scz'
summary_data <- bind_rows(
  lapply(cell_types, function(ct) summarize_smr(ct, in_dir, disorder))
)
datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_scz, results='asis', echo=FALSE}
all_sig_results <- bind_rows(
  lapply(cell_types, function(ct) get_heidi(ct, in_dir, disorder))
)
datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_scz, echo=FALSE}
gene_lookup_file <- '../../resources/sheets/gene_lookup_hg38.tsv'

sig_tbl <- all_sig_results |>
  left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
  select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)

datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_scz, echo=FALSE}
log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

#### BPD {.tabset}

***

##### SMR summary

***

```{r smr_summary_bpd, results='asis', echo=FALSE}
disorder <- 'bpd'
summary_data <- bind_rows(
  lapply(cell_types, function(ct) summarize_smr(ct, in_dir, disorder))
)
datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_bpd, results='asis', echo=FALSE}
all_sig_results <- bind_rows(
  lapply(cell_types, function(ct) get_heidi(ct, in_dir, disorder))
)
datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_bpd, echo=FALSE}
sig_tbl <- all_sig_results |>
  left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
  select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)

datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_bpd, echo=FALSE}
log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

#### MDD{.tabset}

***

##### SMR summary

***

```{r smr_summary_mdd, results='asis', echo=FALSE}
disorder <- 'mdd'
summary_data <- bind_rows(
  lapply(cell_types, function(ct) summarize_smr(ct, in_dir, disorder))
)
datatable(as.data.frame(summary_data), caption = 'Summary of SMR and HEIDI Results by Cell Type')
```

***

##### SMR and HEIDI sig (full)

```{r smr_heidi_sig_mdd, results='asis', echo=FALSE}
all_sig_results <- bind_rows(
  lapply(cell_types, function(ct) get_heidi(ct, in_dir, disorder))
)
datatable(as.data.frame(all_sig_results), caption = 'Significant SMR and HEIDI results by Cell Type (full)')
```

***

##### SMR and HEIDI sig (abbr. w. symbols)

```{r smr_heidi_symbol_mdd, echo=FALSE}
sig_tbl <- all_sig_results |>
  left_join(gene_lookup_hg38_tbl, by = join_by(probeID == ensembl_gene_id)) |>
  select(Cell_Type, ensembl_id = probeID, symbol = external_gene_name, gene_biotype, topSNP, topSNP_chr, p_SMR,	p_HEIDI,	nsnp_HEIDI)

datatable(as.data.frame(sig_tbl), caption = 'Significant SMR and HEIDI results by Cell Type (abbr. with gene symbols)') 
```


***

##### SMR logs

```{r logs_mdd, echo=FALSE}
log_tbl <- extract_smr_log_params(log_dir, cell_types, disorder)
datatable(as.data.frame(log_tbl), caption = 'SMR log file summary') 
```

***

#### Benchmarking {.tabset}

***

##### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
#bmark_dir <- '~/Desktop/eQTL_study_2025/workflow/reports/benchmarks/'
files <- list.files(bmark_dir, pattern = "10smr.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
summary %>%
  gt() %>%
  tab_header(
    title = "Table 5. Benchmarking",
    subtitle = "Means and SD across replicates"
  ) %>%
  cols_label(
    rule = "Rule",
    n = "N",
    mean_runtime_min = "Mean Runtime (min)",
    sd_runtime_min = "SD Runtime (min)",
    mean_ram_gb = "Mean RAM (GB)",
    sd_ram_gb = "SD RAM (GB)"
  ) %>%
  fmt_number(
    columns = c(mean_runtime_min, sd_runtime_min, mean_ram_gb, sd_ram_gb),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = px(14),
    table.border.top.style = "none",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
    table.border.bottom.color = "black",
    heading.align = "center"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

***

##### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min), 0),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = pmax(mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb), 0),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```