---
title: "SuSiE report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:   
  html_document:
    df_print: paged
params:
  cell_types: null
  in_dir: null
---

```{r setup, include=FALSE}
library(tidyverse)
library(rmarkdown)
library(gt)
#cell_types <- c("Endo-Peri", "ExN-2", "InN", "OPC", "ExN-1", "InN", "MG", "RG")
#in_dir <- '~/Desktop/test/susie_test/'
in_dir <- params$in_dir
cell_types <- params$cell_types
```

## SuSiE finemapping pipeline steps {.tabset}

***

1. `get_sig_eGenes`: Get sig. eGenes per cell type from TensorQTL perm output
2. `prep_susie_gene_meta`: Create gene metadata file per cell type for SuSiE
3. `vcf_to_dosage`: Generate a allele dosage file from genotypes VCF for SuSiE
4. `prep_susie_input`: Extract genotype, expression, and covariate data per cell type for SuSiE fine-mapping
5. `run_susie`: Run SuSiE fine-mapping per cell type
6. `merge_susie`: Merge SuSiE fine-mapping output into a single file per cell type
7. `sort_susie`: Sort SuSiE fine-mapping output
8. `susie_report`: Summarise SuSiE results

***


### Summary

***

```{r susie_summary_tbl, echo = FALSE, message=FALSE}
susie_summary_tbl <- map_dfr(cell_types, function(cell_type) {
  
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)
  
  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    arrange(desc(cs_size))
  
  tibble(
    cell_type = cell_type,
    uniq_cs = n_distinct(susie_tbl$cs_id),
    uniq_genes = n_distinct(susie_tbl$phenotype_id),
    uniq_snps = n_distinct(susie_tbl$variant_id),
    one_snp_cs = sum(cs_tbl$cs_size == 1),
    two_snp_cs = sum(cs_tbl$cs_size == 2)
  )
})

gt_tbl <- gt(susie_summary_tbl) |>
    tab_header(title = "Susie summary table") |>
    tab_options(
      table.width = pct(80),
      table.font.size = 12,
      data_row.padding = px(5)
    ) |>
    cols_label(
      uniq_cs = "Unique Credible Sets",
      cell_type = "Cell Type",
      uniq_genes = "Unique Genes",
      uniq_snps = "Unique Variants",
      one_snp_cs = "Single SNP CS",
      two_snp_cs = "Two SNP CS"
    )
  
# Render gt table
gt_tbl
```

***

### Boxplots

***

```{r cred_set_boxplots, echo = FALSE, message=FALSE, results='asis', fig.dim=c(10,8)}
cs_size_plts <- list()

for (cell_type in cell_types) {
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)

  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    arrange(desc(cs_size)) |>
    mutate(cell_type = cell_type) 

  p <- ggplot(cs_tbl, aes(x = cell_type, y = cs_size)) +
    geom_boxplot(outlier.shape = NA, fill = "#69b3a2") +
    geom_jitter(width = 0.2, alpha = 0.1) +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
          axis.text.y = element_text(size = 14),
          axis.title.x = element_blank(),
          axis.title = element_text(size = 16),
          legend.position = "none",
          plot.title = element_text(size = 18, face="bold", hjust = 0.5)) +
    ylab('CS Size') +
    ggtitle(cell_type)

  # Store in list
  cs_size_plts[[cell_type]] <- p
}
cowplot::plot_grid(plotlist = cs_size_plts)
```

***

### Credible sets {.tabset}

```{r cred_set_tbl, echo = FALSE, message=FALSE, results='asis'}
for (cell_type in cell_types) {
  cat('####', cell_type, '\n\n')
  
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)
  
  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    arrange(desc(cs_size))
  

  gt_tbl <- gt(cs_tbl) |>
      tab_header(title = paste("Credible Sets for", cell_type)) |>
      tab_options(
        table.width = pct(80),
        table.font.size = 12,
        data_row.padding = px(5)
      ) |>
      cols_label(
        cs_id = "CS ID",
        cs_size = "CS Size",
        finemapped_region = "Finemapped Region",
        phenotype_id = "Phenotype ID"
      )
    
    # Render gt table
    print(gt_tbl)
  cat('\n\n\n')
}

```

