---
title: "SuSiE report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:   
  html_document:
    df_print: paged
params:
  cell_types: null
  in_dir: null
  gene_meta_dir: null
  gene_lookup: null
---

```{r setup, include=FALSE}
library(tidyverse)
library(rmarkdown)
library(gt)
library(scales)
#cell_types <- c("Endo-Peri", "ExN-2", "InN", "OPC", "ExN-1", "InN", "MG", "RG")
#in_dir <- '~/Desktop/test/susie_test/'
#gene_meta_dir <- '~/Desktop/test/susie_test/gene_meta/'
#gene_lookup <- '~/Desktop/test/susie_test/gene_lookup_hg38.tsv'
in_dir <- params$in_dir
cell_types <- params$cell_types
gene_meta_dir <- params$gene_meta_dir

# Gene symbol lookup (used to add a Gene Symbol column to credible set tables)
gene_lookup <- read_tsv(params$gene_lookup, show_col_types = FALSE) |>
  select(ensembl_gene_id, external_gene_name) |>
  distinct(ensembl_gene_id, .keep_all = TRUE)
```

## SuSiE finemapping pipeline steps {.tabset}

***

We used [susieR](https://stephenslab.github.io/susieR/index.html) for finemapping. The citation for SuSiE is [https://doi.org/10.1111/rssb.12388](https://academic.oup.com/jrsssb/article/82/5/1273/7056114).

***

1. `get_sig_eGenes`: Get sig. eGenes per cell type from TensorQTL perm output
2. `prep_susie_gene_meta`: Create gene metadata file per cell type for SuSiE
3. `vcf_to_dosage`: Generate a allele dosage file from genotypes VCF for SuSiE
4. `prep_susie_input`: Extract genotype, expression, and covariate data per cell type for SuSiE fine-mapping
5. `run_susie`: Run SuSiE fine-mapping per cell type
6. `merge_susie`: Merge SuSiE fine-mapping output into a single file per cell type
7. `sort_susie`: Sort SuSiE fine-mapping output
8. `susie_report`: Summarise SuSiE results

***


### Summary

***

```{r susie_summary_tbl, echo = FALSE, message=FALSE}
susie_summary_tbl <- map_dfr(cell_types, function(cell_type) {
  
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)
  
  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    arrange(desc(cs_size))
  
  tibble(
    cell_type = cell_type,
    uniq_cs = n_distinct(susie_tbl$cs_id),
    uniq_genes = n_distinct(susie_tbl$phenotype_id),
    uniq_snps = n_distinct(susie_tbl$variant_id),
    one_snp_cs = sum(cs_tbl$cs_size == 1),
    two_snp_cs = sum(cs_tbl$cs_size == 2)
  )
})

gt_tbl <- gt(susie_summary_tbl) |>
    tab_header(title = "Susie summary table") |>
    tab_options(
      table.width = pct(80),
      table.font.size = 12,
      data_row.padding = px(5)
    ) |>
    cols_label(
      uniq_cs = "Unique Credible Sets",
      cell_type = "Cell Type",
      uniq_genes = "Unique Genes",
      uniq_snps = "Unique Variants",
      one_snp_cs = "Single SNP CS",
      two_snp_cs = "Two SNP CS"
    )
  
# Render gt table
gt_tbl
```

***

### Boxplots

***

```{r cred_set_boxplots, echo = FALSE, message=FALSE, results='asis', fig.dim=c(10,8)}
cs_size_plts <- list()

for (cell_type in cell_types) {
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)

  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    arrange(desc(cs_size)) |>
    mutate(cell_type = cell_type) 

  p <- ggplot(cs_tbl, aes(x = cell_type, y = cs_size)) +
    geom_boxplot(outlier.shape = NA, fill = "#69b3a2") +
    geom_jitter(width = 0.2, alpha = 0.1) +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
          axis.text.y = element_text(size = 14),
          axis.title.x = element_blank(),
          axis.title = element_text(size = 16),
          legend.position = "none",
          plot.title = element_text(size = 18, face="bold", hjust = 0.5)) +
    ylab('CS Size') +
    ggtitle(cell_type)

  # Store in list
  cs_size_plts[[cell_type]] <- p
}
cowplot::plot_grid(plotlist = cs_size_plts)
```

***

### Credible sets {.tabset}

```{r cred_set_tbl, echo = FALSE, message=FALSE, results='asis'}
for (cell_type in cell_types) {
  cat('####', cell_type, '\n\n')
  
  file_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  susie_tbl <- read_tsv(file_path, show_col_types = FALSE)
  
  cs_tbl <- susie_tbl[!duplicated(susie_tbl$cs_id), ] |>
    select(cs_id, cs_size, finemapped_region, phenotype_id) |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    arrange(cs_size)
  

  gt_tbl <- gt(cs_tbl) |>
      tab_header(title = paste("Credible Sets for", cell_type)) |>
      tab_options(
        table.width = pct(80),
        table.font.size = 12,
        data_row.padding = px(5)
      ) |>
      cols_label(
        cs_id = "CS ID",
        cs_size = "CS Size",
        finemapped_region = "Finemapped Region",
        phenotype_id = "Phenotype ID",
        external_gene_name = "Gene Symbol"
      )
    
    # Render gt table
    print(gt_tbl)
  cat('\n\n\n')
}

```

***

### Fine-mapping diagnostics (reviewer response) {.tabset}

***

The sections below address reviewer 2's questions about fine-mapping performance:

- Table of credible set LBF estimates per credible set per gene
- How many genes had at least one credible set per cell type
- How many credible sets were identified per successfully finemapped gene
- Average size in terms of the number of variants of the credible sets per cell type
- Average distance to the TSS of the credible set index variants
  - is this dependent on the effect size of the credible set index variant


- A credible set's "index variant" is defined here as the SNP in the set with the highest PIP.
- This is distinct from SuSiE's own `cs_index` (`L1`, `L2`, etc.), the number of independent 
credible sets SuSiE found at that locus. Note that these labels don't guarantee that 
L1 is "the strongest" signal.

```{r compute_reviewer_response_summary, echo=FALSE, message=FALSE, warning=FALSE}
# Index variant = highest-PIP SNP within each cs_id (not the same as SuSiE's
# internal cs_index L1/L2/... label, which is just a model slot and is not
# guaranteed to be ordered by effect size or evidence strength).
get_index_variants <- function(cell_type) {

  susie_path <- paste0(in_dir, cell_type, '/', cell_type, '.mrgd.srtd.susie.cred.hp.txt.gz')
  meta_path  <- paste0(gene_meta_dir, cell_type, '_gene_meta.tsv')  # adjust suffix if your snakemake output differs

  if (!file.exists(susie_path) || !file.exists(meta_path)) {
    message("Skipping ", cell_type, " - missing susie or gene meta file")
    return(NULL)
  }

  susie_tbl <- read_tsv(susie_path, show_col_types = FALSE)
  gene_meta <- read_tsv(meta_path, show_col_types = FALSE) |>
    select(phenotype_id, phenotype_pos)

  susie_tbl |>
    group_by(cs_id) |>
    slice_max(pip, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(phenotype_id, cs_id, cs_size, cs_min_r2, cs_avg_r2, cs_log10bf,
           variant_id, chr, pos, pip, z) |>
    rename(index_variant_id = variant_id, index_pip = pip, index_z = z) |>
    left_join(gene_meta, by = "phenotype_id") |>
    left_join(gene_lookup, by = c("phenotype_id" = "ensembl_gene_id")) |>
    rename(gene_symbol = external_gene_name) |>
    mutate(dist_to_tss = abs(pos - phenotype_pos)) |>
    group_by(phenotype_id) |>
    mutate(
      n_cs_for_gene = n(),
      cs_rank = rank(-abs(index_z), ties.method = "first")
    ) |>
    ungroup() |>
    mutate(
      cs_rank_label = case_when(
        cs_rank == 1 ~ "primary",
        cs_rank == 2 ~ "secondary",
        TRUE         ~ "tertiary_or_lower"
      ),
      cell_type = cell_type
    )
}

cs_summary_all <- map_dfr(cell_types, get_index_variants)

# Write the CS-level supplementary table (in_dir already has a trailing slash)
if (nrow(cs_summary_all) > 0) {
  cs_summary_all |>
    select(cell_type, phenotype_id, gene_symbol, cs_id, n_cs_for_gene, cs_rank, cs_rank_label,
           index_variant_id, chr, pos, phenotype_pos, dist_to_tss,
           index_pip, index_z, cs_size, cs_min_r2, cs_avg_r2, cs_log10bf) |>
    arrange(cell_type, phenotype_id, cs_rank) |>
    write_tsv(paste0(in_dir, "susie_credible_set_summary.tsv"))
}
```

#### Credible sets per gene

```{r cs_per_gene_tbl, echo=FALSE, message=FALSE}
cs_per_gene_summary <- cs_summary_all |>
  distinct(cell_type, phenotype_id, n_cs_for_gene) |>
  group_by(cell_type) |>
  summarise(
    n_genes_with_cs     = n_distinct(phenotype_id),
    mean_cs_per_gene    = mean(n_cs_for_gene),
    pct_genes_1_cs      = mean(n_cs_for_gene == 1) * 100,
    pct_genes_2_cs      = mean(n_cs_for_gene == 2) * 100,
    pct_genes_3plus_cs  = mean(n_cs_for_gene >= 3) * 100,
    .groups = "drop"
  )

gt(cs_per_gene_summary) |>
  tab_header(title = "Credible sets per fine-mapped gene") |>
  tab_options(
    table.width = pct(80),
    table.font.size = 12,
    data_row.padding = px(5)
  ) |>
  fmt_number(columns = c(mean_cs_per_gene, pct_genes_1_cs, pct_genes_2_cs, pct_genes_3plus_cs),
             decimals = 1) |>
  cols_label(
    cell_type           = "Cell Type",
    n_genes_with_cs     = "Genes with \u22651 CS",
    mean_cs_per_gene    = "Mean CS / Gene",
    pct_genes_1_cs      = "% Genes, 1 CS",
    pct_genes_2_cs      = "% Genes, 2 CS",
    pct_genes_3plus_cs  = "% Genes, \u22653 CS"
  )
```

```{r cs_per_gene_plot, echo=FALSE, message=FALSE, fig.dim=c(10,6)}
cs_summary_all |>
  distinct(cell_type, phenotype_id, n_cs_for_gene) |>
  mutate(n_cs_capped = if_else(n_cs_for_gene >= 4, "4+", as.character(n_cs_for_gene))) |>
  count(cell_type, n_cs_capped) |>
  ggplot(aes(x = n_cs_capped, y = n, fill = cell_type)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "Credible sets per gene", y = "Number of genes", fill = "Cell Type",
       title = "Distribution of credible-set counts per fine-mapped gene")
```

#### Credible set size & LBF

```{r cs_size_lbf_tbl, echo=FALSE, message=FALSE}
cs_size_summary <- cs_summary_all |>
  group_by(cell_type) |>
  summarise(
    n_credible_sets = n_distinct(cs_id),
    mean_cs_size    = mean(cs_size),
    median_cs_size  = median(cs_size),
    mean_log10bf    = mean(cs_log10bf),
    .groups = "drop"
  )

gt(cs_size_summary) |>
  tab_header(title = "Credible set size and Bayes factor summary") |>
  tab_options(
    table.width = pct(80),
    table.font.size = 12,
    data_row.padding = px(5)
  ) |>
  fmt_number(columns = c(mean_cs_size, median_cs_size, mean_log10bf), decimals = 2) |>
  cols_label(
    cell_type       = "Cell Type",
    n_credible_sets = "Credible Sets",
    mean_cs_size    = "Mean CS Size",
    median_cs_size  = "Median CS Size",
    mean_log10bf    = "Mean log10(BF)"
  )
```

```{r cs_lbf_boxplot, echo=FALSE, message=FALSE, fig.dim=c(10,6)}
ggplot(cs_summary_all, aes(x = cell_type, y = cs_log10bf)) +
  geom_boxplot(outlier.shape = NA, fill = "#69b3a2") +
  geom_jitter(width = 0.2, alpha = 0.1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "log10(Bayes Factor)",
       title = "Credible set log10(BF) distribution by cell type")
```

#### Distance to TSS

- Table: Average / Median distance to TSS per cell type across all credible sets
- Boxplot: 
- Wilcoxon test: rank-sum test comparing dist_to_tss between primary CS 
  and all lower-ranked CS pooled, restricted to genes with >1 credible set. 
  Tests whether primary CS index variants are categorically closer to the 
  TSS than lower-ranked CS index variants. 
- Rank definition: Within each gene with >1 credible set, credible sets are 
  ranked by |z| of their index variant (highest-PIP SNP per CS), descending. 
  Rank is independent of SuSiE's cs_index (L1/L2/...).

```{r dist_to_tss_summary_tbl, echo=FALSE, message=FALSE}
dist_to_tss_summary <- cs_summary_all |>
  group_by(cell_type) |>
  summarise(
    n_credible_sets    = n(),
    mean_dist_to_tss   = mean(dist_to_tss, na.rm = TRUE),
    median_dist_to_tss = median(dist_to_tss, na.rm = TRUE),
    .groups = "drop"
  )

gt(dist_to_tss_summary) |>
  tab_header(title = "Index-variant distance to TSS, all credible sets") |>
  tab_options(
    table.width = pct(80),
    table.font.size = 12,
    data_row.padding = px(5)
  ) |>
  fmt_number(columns = c(mean_dist_to_tss, median_dist_to_tss), decimals = 0) |>
  cols_label(
    cell_type          = "Cell Type",
    n_credible_sets    = "Credible Sets",
    mean_dist_to_tss   = "Mean Distance to TSS (bp)",
    median_dist_to_tss = "Median Distance to TSS (bp)"
  )
```

```{r dist_to_tss_plot, echo=FALSE, message=FALSE, fig.dim=c(8,6)}
cs_summary_all |>
  filter(n_cs_for_gene > 1) |>
  ggplot(aes(x = cs_rank_label, y = dist_to_tss)) +
  geom_boxplot(outlier.shape = NA, fill = "#69b3a2") +
  geom_jitter(width = 0.2, alpha = 0.1) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(x = NULL, y = "Distance from index variant to TSS (bp)",
       title = "CS index-variant distance to TSS, genes with >1 credible set")
```

```{r dist_to_tss_test, echo=FALSE, message=FALSE, results='asis'}
multi_cs <- cs_summary_all |> filter(n_cs_for_gene > 1)

if (nrow(multi_cs) > 0 && n_distinct(multi_cs$cs_rank_label) > 1) {
  wt <- wilcox.test(dist_to_tss ~ I(cs_rank_label == "primary"), data = multi_cs)
  cat(sprintf(
    "**Wilcoxon rank-sum test** (genes with >1 credible set only): primary vs lower-ranked CS distance to TSS, W = %.0f, p = %.3g\n\n",
    wt$statistic, wt$p.value
  ))
} else {
  cat("Not enough genes with multiple credible sets to test primary vs lower-ranked CS distance to TSS.\n\n")
}
```

#### Effect size vs distance to TSS

- Spearman's rho (rank-based, not loess) is the test statistic; loess curves are descriptive only.
- Negative correlation: higher |z| associates with smaller distance to TSS.

```{r effect_size_dist_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.dim=c(11,5)}
ggplot(cs_summary_all, aes(x = abs(index_z), y = dist_to_tss)) +
  geom_point(alpha = 0.15, colour = "#2c3e50") +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  facet_wrap(~ cs_rank_label, nrow = 1) +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt", labels = scales::comma) +
  theme_minimal() +
  labs(x = "|z| of index variant (sqrt scale)", y = "Distance to TSS, bp (sqrt scale)",
       title = "Index-variant effect size vs distance to TSS, by credible set rank")
```

```{r effect_size_dist_test, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
cor_res_all <- cor.test(abs(cs_summary_all$index_z), cs_summary_all$dist_to_tss, method = "spearman")

cor_by_rank <- cs_summary_all |>
  group_by(cs_rank_label) |>
  summarise(
    rho = cor(abs(index_z), dist_to_tss, method = "spearman"),
    p   = suppressWarnings(cor.test(abs(index_z), dist_to_tss, method = "spearman")$p.value),
    n   = n(),
    .groups = "drop"
  )

cat(sprintf(
  "**Spearman correlation, pooled across all credible sets and cell types**: |z| of index variant vs distance to TSS, rho = %.3f, p = %.3g (n = %d)\n\n",
  cor_res_all$estimate, cor_res_all$p.value, nrow(cs_summary_all)
))

cat("**Spearman correlation within each credible set rank** (controls for the fact that primary CS mechanically occupy a higher `|z|` range than secondary/tertiary CS, isolating the within-rank trend):\n\n")

cor_by_rank |>
  mutate(rho = sprintf("%.3f", rho), p = sprintf("%.3g", p)) |>
  knitr::kable(col.names = c("CS rank", "Spearman rho", "p-value", "n"))
```

#### MaxCPP source credible set rank

- Assess rank of credible set that each MaxCPP variant is assigned to.

```{r compute_maxcpp_source_rank, echo=FALSE, message=FALSE, warning=FALSE}
compute_maxcpp_source <- function(ct) {

  susie_path <- paste0(in_dir, ct, '/', ct, '.mrgd.srtd.susie.cred.hp.txt.gz')
  if (!file.exists(susie_path)) {
    message("Skipping ", ct, " - susie file not found")
    return(NULL)
  }
  susie_tbl <- read_tsv(susie_path, show_col_types = FALSE)

  # For each variant, find the credible set that gives it its maxCPP value
  # (mirrors ldsr_make_annot.R's group_by(variant_id) |> mutate(maxCPP = max(pip)),
  # but additionally keeps track of *which* (phenotype_id, cs_id) achieved it)
  winning_cs <- susie_tbl |>
    group_by(variant_id) |>
    slice_max(pip, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(variant_id, phenotype_id, cs_id, maxCPP = pip) |>
    mutate(cell_type = ct)

  cs_ranks_ct <- cs_summary_all |>
    filter(cell_type == ct) |>
    select(phenotype_id, cs_id, cs_rank_label)

  winning_cs |>
    left_join(cs_ranks_ct, by = c("phenotype_id", "cs_id"))
}

maxcpp_source_rank <- map_dfr(cell_types, compute_maxcpp_source)

maxcpp_rank_pct <- maxcpp_source_rank |>
  count(cell_type, cs_rank_label) |>
  group_by(cell_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(names_from = cs_rank_label, values_from = pct, values_fill = 0)

gt(maxcpp_rank_pct) |>
  tab_header(title = "Source credible-set rank of MaxCPP-contributing variants") |>
  tab_options(
    table.width = pct(80),
    table.font.size = 12,
    data_row.padding = px(5)
  ) |>
  fmt_number(columns = -cell_type, decimals = 1) |>
  cols_label(cell_type = "Cell Type")
```

```{r maxcpp_source_rank_plot, echo=FALSE, message=FALSE, fig.dim=c(10,6)}
maxcpp_source_rank |>
  count(cell_type, cs_rank_label) |>
  group_by(cell_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  ggplot(aes(x = cell_type, y = pct, fill = cs_rank_label)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
  labs(x = NULL, y = "% of MaxCPP-contributing variants", fill = "Source CS rank",
       title = "Which credible set rank contributes the MaxCPP value, per cell type")
```

```{r maxcpp_source_rank_overall, echo=FALSE, message=FALSE, results='asis'}
overall_pct <- maxcpp_source_rank |>
  count(cs_rank_label) |>
  mutate(pct = n / sum(n) * 100)

cat("**Overall, pooled across all cell types:**\n\n")
overall_pct |>
  mutate(pct = sprintf("%.1f%%", pct)) |>
  select(cs_rank_label, n, pct) |>
  knitr::kable(col.names = c("Source CS rank", "n variants", "% of total"))
```

***

A full credible-set-level table (cell type, gene, gene symbol, CS rank, index variant, 
distance to TSS, PIP, z, CS size, purity, log10 Bayes factor) for all cell types has 
been written to `r paste0(in_dir, "susie_credible_set_summary.tsv")`.

***
