---
title: "TensorQTL report"
author: "Darren Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  in_dir: null
  bmark_dir: null
---

***

```{r setup, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(cowplot)
library(DT)
library(glue)
library(gt)
library(tidyverse)
library(UpSetR)
library(ggplot2)
library(patchwork)
library(ggrepel)
library(cowplot)
in_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```

### TensorQTL pipeline steps {.tabset}

***

We used [tensorQTL](https://github.com/broadinstitute/tensorqtl) to generate eQTL. 
The citation for tensorQTL is [Taylor-Weiner et al. 2019 (DOI:10.1186/s13059-019-1836-7)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1836-7).

***

1. `prep_tensorQTL_input`: Munge cell-specific pseudobulk gExp and covaraiate data into TensorQTL-ready format
2. `zip_pblk_cnts`: bgzip and index pseudobulk gExp data
3. `convert_genotypes`: Convert genotypes into into TensorQTL-ready plink 2 format 
4. `split_covariates`: Generate covaraiate files for gene_pc and exp_pc combinations
5. `tensorqtl_nom`: Run TensorQTL nominal
6. `tensorqtl_perm`: Run TensorQTL perm

Settings:

+ cis-window: $\pm$ 500kb
+ Expression PC covariates tested: 10, 20, 30, 40, 50
+ Meta data covariates: PCW, sex
+ Genotype PCs: 4 (see genotyping_post_impute_report.html)
+ Norm methods tested: bryois, fugita, quantile, xue

***

```{r, summary_table, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
cell_types <- c("ExN-UL", "ExN-DL", "InN", "RG", "MG", "OPC", "Endo-Peri")
#in_dir <- "~/Desktop/eQTL_study_2025/workflow/reports/05TENSORQTL/logs/"
#bmark_dir <- "~/Desktop/eQTL_study_2025/workflow/reports/benchmarks/"
exp_PCs <- c(10, 20, 30, 40, 50)
gen_PCs <- 4
norm_methods <- c('bryois', 'fujita', 'quantile', 'xue')

summarise_logs <- function(cell_types, in_dir, expPCs = exp_PCs, 
                           genPC = gen_PCs, norm_method = norm_methods) {
  # Generate all combinations of cell types and expPC values
  params_grid <- expand.grid(cell_type = cell_types, expPC = expPCs, norm_method = norm_methods,
                             stringsAsFactors = FALSE)
  
  # Function to process a single log file
  process_log <- function(cell_type, expPC, norm_method) {
    message('Reading eQTL log file for: ', cell_type, ', ', norm_method , 
            ', with genPC_', genPC, ' expPC_', expPC)
    
    # Construct the log file path
    log_dir <- paste0(in_dir, cell_type, '_', norm_method, '_genPC_', genPC, '_expPC_', expPC, '/')
    log_file <- paste0(log_dir, cell_type, '_', norm_method, '_perm.tensorQTL.cis.log')

    # Check if file exists
    if (!file.exists(log_file)) {
      message('Log file not found: ', log_file)
      return(NULL)
    }
    
    # Read and process the log file
    lines <- readLines(log_file, warn = FALSE)
    
    gene_line <- str_subset(lines, 'phenotypes tested:')
    sample_line <- str_subset(lines, 'samples')
    covariate_line <- tail(str_subset(lines, 'covariates'), 1) # need to fix this
    variants_line <- str_subset(lines, 'variants')
    qtl_line <- str_subset(lines, 'QTL phenotypes @ FDR 0.05:')
    window_line <- str_subset(lines, 'cis-window:')
    run_line <- str_subset(lines, 'Time elapsed:')
    beta_line <- str_subset(lines, 'Correlation between Beta-approximated and empirical p-values:')
    min_p_line <- str_subset(lines, 'min p-value threshold @ FDR 0.05:')
    prop_sig_line <- str_subset(lines, 'Proportion of significant phenotypes')
    
    # Extract numbers
    num_genes <- as.integer(str_extract(gene_line, '\\d+$'))
    num_samples <- as.integer(str_extract(sample_line, '\\d+'))
    covariates <- as.integer(str_extract(covariate_line, '\\d+$'))
    variants <- as.integer(str_extract(variants_line, '\\d+'))
    num_qtls <- as.integer(str_extract(qtl_line, '\\d+$'))
    cis_window <- as.integer(str_extract(window_line, '\\d+'))
    run_time <- as.integer(str_extract(run_line, '\\d+'))
    beta_cor <- as.numeric(str_extract(beta_line, '[0-9\\.eE-]+$'))
    min_p <- as.numeric(str_extract(min_p_line, '[0-9\\.eE-]+$'))
    prop_sig <- as.integer(str_extract(prop_sig_line, '\\d+$'))
    
    message('Collating sumstats for: ', cell_type, ' with genPC_', genPC, ' expPC_', expPC, ' ...')
    
    # Create tibble with all relevant fields
    tibble(
      cell_type = cell_type,
      genPC = genPC,
      expPC = expPC,
      norm_methods = norm_method,
      genes_tested = num_genes,
      samples = num_samples,
      covariates = covariates,
      variants = variants,
      qtl_fdr_0.05 = num_qtls,
      cis_window_kb = cis_window,
      beta_cor = beta_cor,
      min_p = formatC(min_p, format = "e", digits = 2),
      prop_sig = prop_sig,
      run_time_min = run_time
    )
  }
  
  # Apply the processing function to all combinations
  summary_list <- pmap(params_grid, function(cell_type, expPC, norm_method) {
    process_log(cell_type, expPC, norm_method)
  })
  
  # Identify which elements are non-null
  valid_idx <- !sapply(summary_list, is.null)
  table(valid_idx)  # how many TRUE vs FALSE
  length(summary_list)
  sum(valid_idx)

  # Keep only valid entries
  summary_list <- summary_list[valid_idx]

  # Name the list elements for clarity
  names(summary_list) <- paste0(
    params_grid$cell_type, '_', 
    params_grid$norm_method, '_genPC_', genPC, 
    '_expPC_', params_grid$expPC
    )[valid_idx]
  
  # Return the list of summary tibbles
  return(summary_list)
}

summary_list <- summarise_logs(cell_types, in_dir)
```

***

#### eQTL plot

***

```{r, eQTL_by_exp_PC_plot, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,7.5)}
# Define custom color palette
custom_palette <- c(
  "RG" = "#FF5959",
  "ExN-UL" = "#00B6EB",
  "InN" = "#3CBB75FF",
  "ExN-DL" = "#CEE5FD",
  "Endo-Peri" = "#B200ED",
  "MG" = "#F58231",
  "OPC" = "#FDE725FF"
)

summary_tbl <- bind_rows(summary_list)

# Function to make one plot per norm_method
make_plot <- function(df, norm) {
  ggplot(df %>% filter(norm_methods == norm),
         aes(x = expPC, y = qtl_fdr_0.05, color = cell_type, group = cell_type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_manual(values = custom_palette) +
    labs(
      x = "Expression PCs",
      y = "eGene count (FDR < 0.05)",
      color = "Cell Type",
      title = paste("Number of Significant QTL by Expression PCs\nNorm method:", norm)
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      legend.text = element_text(size = 10),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_line(color = "grey95"),
      plot.margin = margin(10, 10, 10, 10)
    ) + ylim(0, 2100)
}

# Generate one plot per norm method
plots <- lapply(unique(summary_tbl$norm_methods), function(norm) {
  make_plot(summary_tbl, norm)
})

# Arrange in 2x2 grid
combined_plot <- plot_grid(plotlist = plots, ncol = 2)

# Show the combined plot
print(combined_plot)

# Optionally save
# ggsave("qtl_fdr_by_expPC_all_norms.png", combined_plot, width = 12, height = 10, dpi = 300)
```

***

#### Density plot {.tabset}

##### Individual

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

***

```{r, eQTL_density_plt, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,7.5)}
plt_list <- list()
genPC <- 4
expPC <- 40
norm_method <- 'quantile'
for (cell_type in cell_types) {
  
  message('Reading eQTL data for: ', cell_type)
  log_dir <- paste0(in_dir, cell_type, '_', norm_method, '_genPC_', genPC, '_expPC_', expPC, '/')
  log_file <- paste0(log_dir, cell_type, '_', norm_method, '_perm.cis_qtl.txt.gz')
  eqtl_tbl <- read_tsv(log_file) %>%
    dplyr::filter(!is.na(start_distance)) %>%
    mutate(distance_kb = start_distance / 1000) %>%
    dplyr::filter(qval < 0.05)
  
  message('eQTL detected at FDR < 0.05: ', nrow(eqtl_tbl))
  
  if (nrow(eqtl_tbl) > 0) {
    
    message('Plotting ...')
    
    eqtl_count <- nrow(eqtl_tbl)
    
    eqtl_density_plt <- eqtl_tbl  %>%
      ggplot(aes(x = distance_kb)) +
      geom_density() +
      geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
      annotate("text", x = 100, y = Inf, 
               label = paste("eQTLs:", eqtl_count), 
               hjust = 0, vjust = 1.5, size = 4, fontface = "bold") +
      scale_x_continuous(limits = c(-600, 600), 
                         breaks = seq(-600, 600, by = 200),
                         labels = seq(-600, 600, by = 200)) +
      labs(x = "Distance to TSS (kb)",
           y = "Density",
           title = cell_type) +
      theme_minimal() +
      theme(
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text = element_text(color = "black"),
        axis.title = element_text(face = "bold")
      )
    
    plt_list[[cell_type]] <- eqtl_density_plt
    
  } else {
    
    message('Skipped plotting.')
    
  }
  
}

combined_plot <- cowplot::plot_grid(plotlist = plt_list)
combined_plot
```

***

##### Combined

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

```{r, denisty_combined, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(5,3) }
eqtl_list <- list()
genPC <- 4
expPC <- 40
norm_method <- 'quantile'

for (cell_type in cell_types) {
  f <- file.path(
    in_dir,
    paste0(cell_type, "_", norm_method, "_genPC_", genPC, "_expPC_", expPC),
    paste0(cell_type, "_", norm_method, "_perm.cis_qtl.txt.gz")  # adjust filename
  )
  
  if (file.exists(f)) {
    dat <- read_tsv(f, show_col_types = FALSE) %>%
      filter(!is.na(start_distance)) %>%
      mutate(distance_kb = start_distance / 1000) %>%
      filter(qval < 0.05)   # significance cutoff
    
    if (nrow(dat) > 0) {
      eqtl_list[[cell_type]] <- dat %>% mutate(cell_type = cell_type)
    }
  }
}

eqtls_all <- bind_rows(eqtl_list)

ggplot(eqtls_all, aes(x=distance_kb, color=cell_type)) +
  geom_density() +
  geom_vline(xintercept = 0, linetype="dashed", color = "black") +
  scale_color_manual(values=custom_palette) +
  #scale_fill_manual(values=custom_palette) +
  scale_x_continuous(limits=c(-600,600), breaks=seq(-600,600,200)) +
  labs(x="Distance to TSS (kb)", y="Density") +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(face = "bold")
  )
```

***

#### Pie Chart

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

```{r, pie_chart, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(7.5,7.5) }
gene_cell <- eqtls_all %>%
  distinct(phenotype_id, cell_type)

gene_counts <- gene_cell %>%
  count(phenotype_id) %>%
  count(n)  # number of cell types per gene

pie_dat <- gene_counts %>%
  mutate(
    label = paste0(n, " cell type", ifelse(n > 1, "s", "")),
    count = nn
  ) %>%
  arrange(desc(n)) %>%
  mutate(
    # Calculate cumulative fraction for positioning labels
    fraction = count / sum(count),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    midpoint = (ymax + ymin) / 2
  )

ggplot(pie_dat, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
  geom_rect(color = "white") +
  coord_polar(theta = "y") +
  geom_text_repel(aes(
    x = 1.1, 
    y = midpoint, 
    label = count
  ),
  nudge_x = 0.05,
  segment.color = "grey50",
  segment.size = 0.5,
  show.legend = FALSE
  ) +
  labs(
    fill = "Cell type sharing",
    title = "Shared eGenes across cell types"
  ) +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  )

```

***

#### Counts

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

***

```{r, counts, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(5,5) }
cell_counts <- tibble(
  cell_type = c("ExN-UL","ExN-DL","RG","InN","Endo-Peri","OPC","MG"),
  n_cells   = c(359823, 200545, 178642, 154155, 20160, 13797, 7406)
)
egene_per_cell <- gene_cell %>% count(cell_type)

ggplot(egene_per_cell, aes(x=reorder(cell_type,n), y=n, fill=cell_type)) +
  geom_col() +
  scale_fill_manual(values=custom_palette) +
  labs(x="Cell type", y="Number of eGenes") +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(face = "bold")
  ) +
  theme(legend.position = "none")
```

***

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

#### Upset

```{r, upset, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10,5) }
gene_by_cell <- gene_cell %>%
  pivot_wider(names_from = cell_type, values_from = cell_type,
              values_fill = 0, values_fn = function(x) 1) %>%
  column_to_rownames("phenotype_id")

upset(gene_by_cell, nsets = length(cell_types), order.by = "freq")
```

***

#### Scatter

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

***

```{r, scatter, results='asis', echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(5,5) }
scatter_dat <- egene_per_cell %>%
  left_join(cell_counts, by="cell_type")

ggplot(scatter_dat, aes(x=log10(n_cells), y=n, label=cell_type, color=cell_type)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label=cell_type, color=cell_type), 
                  size=4, 
                  show.legend=FALSE,
                  max.overlaps=20, 
                  force=2,      # pushes labels apart
                  nudge_y=100,  # pushes labels up a bit
  ) +
  scale_color_manual(values=custom_palette)+
  labs(x="log10(Cell count)", y="cis-eQTL (FDR < 0.05)") +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(color = "black", size = 12),
    axis.title = element_text(face = "bold", size =  14)
  ) +
  theme(legend.position = "none")
```

#### Settings

+ Showing plots for geno PCs = 4, Exp PCs = 40, norm method = 'quantile'

***

```{r, variables, results='asis', echo = FALSE, message = FALSE, warning = FALSE}
# Select one row per cell type, keeping only cell_type, genes_tested, samples, and variants
summary_table <- summary_tbl %>%
  dplyr::select(cell_type, genes_tested, samples, variants) %>%
  distinct(cell_type, .keep_all = TRUE) %>%
  arrange(cell_type) # Sort alphabetically for consistency

# Create the gt table
table <- summary_table %>%
  gt() %>%
  tab_header(
    title = "Table 1. TensorQTL, inputs per cell type"
  ) %>%
  cols_label(
    cell_type = "Cell Type",
    genes_tested = "Genes Tested",
    samples = "Samples",
    variants = "Variants"
  ) %>%
  fmt_number(
    columns = c(genes_tested, samples, variants),
    decimals = 0,
    use_seps = TRUE
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.title.font.weight = "bold",
    heading.subtitle.font.size = 14,
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table.border.top.color = "black",
    table.border.bottom.color = "transparent",
    table_body.hlines.color = "grey80",
    table_body.border.bottom.color = "black"
  )

# Save the table as a LaTeX file for publication
#gtsave(table, filename = "summary_table.tex")

# Optionally, display the table
print(table)
```

***

#### Benchmarking {.tabset}

***

##### Table

```{r benchmarking, echo = FALSE, message = FALSE, warning = FALSE}
# List all benchmark files for the pipeline
files <- list.files(bmark_dir, pattern = "05tensorQTL.*\\.txt", full.names = TRUE)

benchmarks <- map_dfr(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  fname <- basename(f)
  parts <- strsplit(fname, "\\.")[[1]]

  # this field holds things like:
  # "prep_tensorQTL_input_Endo-Peri", "split_covariates_Endo-Peri_genPC_4_expPC_10",
  # "perm_ExN-2_genPC_4_expPC_50", "nom_ExN-2_genPC_4_expPC_50", "convert_genotypes"
  rule_full <- parts[2]

  # strip everything after the first "_" → leaves only the base rule
  rule <- str_split(rule_full, "_")[[1]][1]

  tibble(
    file = f,
    rule = rule,
    runtime_sec = df$s,
    runtime_hms = df$h.m.s.,
    max_rss = df$max_rss
  )
})



# Summarise by rule
summary <- benchmarks %>%
  group_by(rule) %>%
  summarise(
    n = n(),
    mean_runtime_min = mean(runtime_sec) / 60,
    sd_runtime_min = sd(runtime_sec) / 60,
    mean_ram_gb = mean(max_rss) / 1024,
    sd_ram_gb = sd(max_rss) / 1024,
    .groups = "drop"
  ) %>%
  mutate(
    sd_runtime_min = if_else(is.na(sd_runtime_min), 0, sd_runtime_min),
    sd_ram_gb = if_else(is.na(sd_ram_gb), 0, sd_ram_gb)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(rule)

# Table
datatable(
  summary,
  caption = 'Table X. Rule benchmarking. Means and SD across replicates.',
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

***

##### Plot

***

```{r, runtime_plt, echo = FALSE, message = FALSE, warning = FALSE, fig.dim=c(10, 5)}
runtime_plt <- ggplot(summary, aes(x = reorder(rule, mean_runtime_min), 
                                 y = mean_runtime_min)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_runtime_min - ifelse(is.na(sd_runtime_min), 0, sd_runtime_min),
                    ymax = mean_runtime_min + ifelse(is.na(sd_runtime_min), 0, sd_runtime_min)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "Runtime (minutes)", 
       title = "Mean runtime per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )



mem_plt <- ggplot(summary, aes(x = reorder(rule, mean_ram_gb), 
                                y = mean_ram_gb)) +
  geom_col(fill = "darkorange") +
  geom_errorbar(aes(ymin = mean_ram_gb - ifelse(is.na(sd_ram_gb), 0, sd_ram_gb),
                    ymax = mean_ram_gb + ifelse(is.na(sd_ram_gb), 0, sd_ram_gb)),
                width = 0.3, colour = "black", linewidth = 0.6) +
  coord_flip() +
  labs(x = "Rule", y = "RAM (GB)", 
       title = "Mean RAM usage per rule") +
  theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_grid(runtime_plt, mem_plt)

# ggsave(paste0(bmark_dir, "genotype_genotype_pre_benchmark.png"), width = 7, height = 5)
```



