---
title: "TWAS Weight Computation Summary"
date: "`r Sys.Date()`"
output: html_document
params:
  cell_types: null
  weights_dir: null
  bmark_dir: null
---

## TWAS weight computation pipeline steps {.tabset}

***

We used [FUSION](http://gusevlab.org/projects/fusion/) to generate gene expression prediction weights. 
The citation for FUSION is [Gusev et al. 2016 (DOI:10.1038/ng.3506)](https://www.nature.com/articles/ng.3506).

***

```{r set_up, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(stringr)
library(DT)
library(gt)
library(cowplot)

# Set variables from params
cell_types <- params$cell_types
weights_dir <- params$in_dir
bmark_dir <- params$bmark_dir
```


1. `get_gemma`: Download the GEMMA binary executable (version 0.98.5) required for FUSION's genetic modeling (BSLMM and BLUP); 
2. `prep_exp_data`: Process gene expression data into PLINK-compatible phenotype files and generates a gene coordinate TSV file for FUSION input. Serves as a foundational checkpoint for gene-specific extractions with outputs aggregated across all genes for the cell type.
3. `convert_vcf`: Convert VCF genotype files to PLINK binary format 
4. `restrict_geno_to_ldref`: Restricts the converted genotypes to SNPs present in the LD reference panel (e.g., `1000G.EUR.hg38`)
5. `get_gene_info`: Checkpoint that dynamically splits the gene coordinate file into individual per-gene info files (one TSV per gene) stored in a cell-type-specific directory; enables parallel processing of genes by providing dynamic inputs to downstream rules. Data aggregation is handled in later rules
6. `extract_cis_snps`: Rule that extracts cis-regulatory SNPs for each gene within a specific window. Creates empty stub files if no SNPs are found.
7. `prepare_gene_pheno`: Extracts and formats expression data for a single gene from the full expression matrix into a FUSION-compatible phenotype file 
8. `compute_weights`: Runs FUSION's `compute_weights.R` script to estimate predictive up to 5 models and generate TWAS weights (.wgt.RDat) for a specific gene. Handles low-heritability skips by creating stub files or log notes
9. `aggregate_per_cell_type`: Aggregation rule triggered post-checkpoint via the `aggregate_input` function, which dynamically lists all per-gene weight files from the checkpoint-expanded gene IDs; outputs a summary `.txt` file per cell type listing processed genes, ensuring completion tracking before full pipeline aggregation.
10. `aggregate_all`: Final aggregation rule that collects completion files from all cell types using expand over config-defined cell types; concatenates summaries into a single master `.txt` file, confirming all gene weights across all cell types are processed for downstream reporting.
11. `twas_weights_report`: Creates twas weights report


Settings:

+ cis-window: $\pm$ 500kb (this is applied in `prep_exp_data` and `extract_cis_snps`; check this)
+ Heritability p-val thresh: `--hsq_p 0.01`
+ Models run: ` --models top1,lasso,enet` (cTWAS uses LASSO)

***

### Summary Table

***

```{r summary_table, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
# Function to process one context
process_context <- function(ct) {
  ct_dir <- file.path(weights_dir, ct)
  wgt_files <- list.files(ct_dir, pattern = "\\.wgt\\.RDat$", full.names = TRUE)
  
  process_file <- function(file) {
    gene_id <- gsub("\\.wgt\\.RDat$", "", basename(file))
    
    # Try to load safely
    env <- new.env()
    load_success <- tryCatch({
      load(file, envir = env)
      TRUE
    }, error = function(e) FALSE, warning = function(w) FALSE)
    
    if (!load_success) {
      return(tibble(
        context = ct,
        gene = gene_id,
        n_snps_lasso = NA_real_,
        reason = "load_failed"
      ))
    }
    
    # Check if wgt.matrix exists and has content
    if (!exists("wgt.matrix", envir = env)) {
      return(tibble(
        context = ct,
        gene = gene_id,
        n_snps_lasso = NA_real_,
        reason = "no_wgt_matrix"
      ))
    }
    
    wgt_mat <- get("wgt.matrix", envir = env)
    
    # Check if lasso column exists and is not empty/all-zero
    if (!"lasso" %in% colnames(wgt_mat) || all(wgt_mat[, "lasso"] == 0) || nrow(wgt_mat) == 0) {
      return(tibble(
        context = ct,
        gene = gene_id,
        n_snps_lasso = NA_real_,
        reason = "stub_or_empty_lasso"
      ))
    }
    
    # Valid lasso model
    lasso_wgts <- wgt_mat[, "lasso"]
    n_snps <- sum(lasso_wgts != 0)
    
    return(tibble(
      context = ct,
      gene = gene_id,
      n_snps_lasso = n_snps,
      reason = "valid"
    ))
  }
  
  map_dfr(wgt_files, process_file)
}

# Full table across all contexts
full_table <- map_dfr(cell_types, process_context)

# Summary table
summary_table <- full_table %>%
  group_by(context) %>%
  summarise(
    total_genes = n(),
    n_valid_lasso   = sum(reason == "valid"),
    n_stub          = sum(reason %in% c("stub_or_empty_lasso", "no_wgt_matrix", "load_failed")),
    n_NA_other      = sum(reason == "valid" & is.na(n_snps_lasso)),  # should be 0
    mean_snps       = mean(n_snps_lasso[reason == "valid"], na.rm = TRUE),
    median_snps     = median(n_snps_lasso[reason == "valid"], na.rm = TRUE),
    min_snps        = min(n_snps_lasso[reason == "valid"], na.rm = TRUE),
    max_snps        = max(n_snps_lasso[reason == "valid"], na.rm = TRUE),
    pct_single_snp  = round(100 * sum(n_snps_lasso == 1 & reason == "valid", na.rm = TRUE) / n_valid_lasso, 1),
    pct_multi_snp   = round(100 * sum(n_snps_lasso > 1 & reason == "valid", na.rm = TRUE) / n_valid_lasso, 1),
    pct_stub        = round(100 * n_stub / total_genes, 1)
  ) %>%
  arrange(desc(total_genes))

print("Summary Table (with stub/empty file counts):")
print(summary_table, n = Inf)
```

***

### NA genes

***

```{r NA genes, include=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
no_model_table <- full_table %>%
  filter(reason != "valid") %>%
  select(context, gene, reason) %>%
  arrange(context, gene)

print("Genes without valid LASSO model (stubs, empty, load issues):")
print(no_model_table, n = Inf)

# Optional: Summary of reasons
reason_counts <- no_model_table %>%
  count(context, reason, name = "count") %>%
  pivot_wider(names_from = reason, values_from = count, values_fill = 0)

print("Breakdown of no-model reasons per context:")
print(reason_counts)
```

***